---
title: "`concept-map.R`: many-to-many column diagrams"
subtitle: "Walkthrough: how the script works, and how to use it in Quarto"
format:
  html:
    code-fold: false
    toc: true
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(fig.align = "center")
```

## Aims

`concept-map.R` draws a general "N columns of ellipses, many-to-many
curved links between adjacent columns" diagram, in the style of
McElreath's *Statistical Rethinking* Figure 1.2 (theories → process
models → statistical models). It's used in `doing-being-science.qmd` to
make the point that the theory is not the hypothesis is not the
statistical test result: several different theories can imply the same
process model, and several different process models can imply the same
statistical test result, so a single statistically significant result
never uniquely confirms one theory.

Unlike `person-grid.R`/`testing-sequence.R`, this script isn't about
pictograms at all — it's a general-purpose node-and-edge diagram builder
that happens to use ellipses and curves rather than a graph library, so it
renders identically wherever Quarto builds the document.

After working through it, you should be able to:

- explain how node layout and edge computation are kept independent until
  the final plot;
- build your own N-column, many-to-many concept map for a different
  topic, including per-column sizing for uneven label lengths;
- diagnose the validation errors/warnings the script raises for bad edge
  indices or too-tight column spacing.

## Prerequisites / assumptions

- **Packages**: `ggplot2` only.
- **Sourcing**: standalone — no other script in this set needs to be
  sourced first.
- **Reader background**: no special background needed beyond basic
  `ggplot2` (`geom_polygon()`, `geom_curve()`).

```{r}
#| label: source-script
source("concept-map.R")
```

## How it works: the process

```{mermaid}
flowchart LR
    A["columns: list of\ncharacter vectors\n(node labels per column)"] --> D["validate edges\n(catch typo'd indices early)"]
    B["edges: list of\nfrom/to index data frames,\none per column gap"] --> D
    C["rx, ry, wrap_width, dx, dy:\nsizing + spacing\n(global or per-column)"] --> E["lay out nodes:\none column at a time,\ncentred vertically"]
    D --> F["compute edge endpoints\nat each node's own\nellipse boundary"]
    E --> F
    F --> G["ggplot():\ngeom_curve() edges +\ngeom_polygon() ellipses +\ngeom_text() labels"]
```

The structure mirrors a typical graph-drawing pipeline: nodes are laid out
first (purely from `columns` + sizing parameters, independent of edges),
then edges are computed as line segments between specific node
boundaries, then both are drawn in one plot. Keeping layout and edges
separate is what makes "many-to-many" straightforward — an edge is just
`(from index in column k, to index in column k+1)`, resolved to actual
coordinates only at the end.

## Element-by-element walkthrough

### Geometry helpers: `ellipse_pts()` and `wrap_label()`

```{r}
#| label: show-ellipse-pts
#| eval: false
ellipse_pts <- function(cx, cy, rx, ry, n = 72) {
  t <- seq(0, 2 * pi, length.out = n)
  data.frame(x = cx + rx * cos(t), y = cy + ry * sin(t))
}

wrap_label <- function(label, width = 16) {
  paste(strwrap(label, width = width), collapse = "\n")
}
```

`ellipse_pts()` is the ellipse analogue of `person-grid.R`'s
`circle_pts()` — a centre, two radii (`rx`, `ry`), and a point count.
`wrap_label()` wraps long node text (e.g. full theory sentences) onto
multiple lines at a given character width, using base R's `strwrap()`,
since node labels here are often full sentences rather than short tags.

```{r}
#| label: demo-wrap-label
cat(wrap_label("Words we experience more often are easier to recall", width = 16))
```

### Per-column parameters: `recycle_per_column()`

A design choice specific to this script: `rx`, `ry`, and `wrap_width` can
each be given as **either** a single value (applied to every column) or a
vector with one entry per column:

```{r}
#| label: show-recycle
#| eval: false
recycle_per_column <- function(val, name) {
  if (length(val) == 1) return(rep(val, k))
  if (length(val) == k) return(val)
  stop(sprintf("`%s` must have length 1 or length(columns) = %d, not %d.",
               name, k, length(val)))
}
```

This matters in practice because columns rarely hold equally long labels —
e.g. a "theories" column of full sentences next to a "statistical tests"
column of short phrases needs a bigger ellipse (and wider wrap width) on
the first column, not a one-size-fits-all setting.

### Validation before layout: edge-index checking

```{r}
#| label: show-edge-validation
#| eval: false
for (ci in seq_len(k - 1)) {
  e <- edges[[ci]]
  n_from <- length(columns[[ci]])
  n_to <- length(columns[[ci + 1]])
  bad_from <- e$from[!(e$from %in% seq_len(n_from))]
  bad_to <- e$to[!(e$to %in% seq_len(n_to))]
  if (length(bad_from) > 0) {
    stop(sprintf(
      "edges[[%d]]$from has out-of-range value(s) %s: column %d ('%s') only has %d item(s), so valid indices are 1..%d.",
      ci, paste(unique(bad_from), collapse = ", "), ci,
      if (!is.null(titles)) titles[ci] else paste("column", ci), n_from, n_from
    ))
  }
  # ... same check for `to` ...
}
```

::: {.callout-note title="Why validate indices before plotting?"}
Every `from`/`to` index in `edges` is checked against the actual length
of the relevant column *before* any plotting is attempted. The failure
mode this avoids is subtle: an out-of-range index used directly as a
lookup would otherwise silently produce `NA` coordinates, and the edge
would simply vanish from the rendered plot with only a generic,
hard-to-trace `ggplot2` warning — rather than an error that names the
exact column and bad index.
:::

A second check guards column spacing rather than edge indices:

```{r}
#| label: show-dx-check
#| eval: false
for (ci in seq_len(k - 1)) {
  min_dx <- rx[ci] + rx[ci + 1]
  if (dx <= min_dx) {
    warning(sprintf(
      paste("dx = %.2f may be too small for columns %d and %d",
            "(rx = %.2f and %.2f): their ellipses may overlap.",
            "Consider increasing dx."),
      dx, ci, ci + 1, rx[ci], rx[ci + 1]
    ))
  }
}
```

This warns if `dx` is too small for two neighbouring columns' `rx`
values, which would otherwise manifest as visibly overlapping ellipses
with no explicit warning pointing at the cause — easy to miss if you've
just widened one column's `rx` for longer labels without revisiting `dx`.

### Laying out nodes and edges: the body of `concept_map()`

```{r}
#| label: show-node-layout
#| eval: false
nodes <- do.call(rbind, lapply(seq_len(k), function(ci) {
  labels <- columns[[ci]]
  m <- length(labels)
  data.frame(
    col = ci, idx = seq_len(m),
    x = (ci - 1) * dx,
    y = (m - 1) / 2 * dy - (seq_len(m) - 1) * dy,
    label = vapply(labels, wrap_label, character(1), width = wrap_width[ci]),
    fill = fill_colors[ci],
    rx = rx[ci], ry = ry[ci],
    stringsAsFactors = FALSE
  )
}))
```

Each column is laid out **independently**: column `ci` sits at
`x = (ci - 1) * dx`, and its own `m` items are centred vertically around
`y = 0` (the formula places item 1 at `+((m-1)/2)*dy` and the last item at
the mirror-image negative value).

::: {.callout-note title="Rows don't align across columns"}
Columns with different numbers of items are *each* centred on `y = 0`
independently, so a specific row position doesn't correspond to the same
item across columns — a 3-item column and a 2-item column don't have
their first items aligned. This matches the McElreath-style figure (where
correspondence is carried by the edges, not by row alignment), but is
worth checking visually for your own content.
:::

```{r}
#| label: show-edge-endpoints
#| eval: false
edge_df <- do.call(rbind, lapply(seq_len(k - 1), function(ci) {
  e <- edges[[ci]]
  do.call(rbind, lapply(seq_len(nrow(e)), function(j) {
    p1 <- node_xy(ci, e$from[j])
    p2 <- node_xy(ci + 1, e$to[j])
    data.frame(x = p1["x"] + p1["rx"], y = p1["y"],
               xend = p2["x"] - p2["rx"], yend = p2["y"])
  }))
}))
```

Edges run from the **right edge** of the source ellipse to the **left
edge** of the target ellipse (`+rx`/`-rx` from each node's own centre, not
a shared/average `rx`), so differently-sized ellipses across columns still
meet their edges at their actual boundaries rather than at a
one-size-fits-all offset.

### Drawing: arrows vs. plain correspondence lines

```{r}
#| label: show-arrows-param
#| eval: false
geom_curve(data = edge_df, aes(x = x, y = y, xend = xend, yend = yend),
           curvature = curvature, color = edge_color, alpha = edge_alpha,
           linewidth = 0.6,
           arrow = if (arrows) arrow(length = unit(0.12, "cm"), type = "closed") else NULL)
```

`arrows = TRUE` (the default) draws a small arrowhead at each curve's
target end, implying directionality. For a diagram about *correspondence*
between theories/processes/statistical tests — not a causal chain — the
deck sets `arrows = FALSE`, giving plain connecting curves closer to
McElreath's original figure.

```{r}
#| label: demo-concept-map
#| fig-cap: "A small concept map, built directly from the script"
#| fig-width: 8
#| fig-height: 4
demo_columns <- list(
  c("Theory A", "Theory B"),
  c("Process 1", "Process 2"),
  c("Test result X")
)
demo_edges <- list(
  data.frame(from = c(1, 1, 2), to = c(1, 2, 2)),
  data.frame(from = c(1, 2), to = c(1, 1))
)
concept_map(demo_columns, demo_edges, titles = c("Theories", "Processes", "Tests"),
            arrows = FALSE, curvature = 0)$plot
```

## Quarto usage / integration

Setup-chunk line:

```r
source("concept-map.R")
```

Real usage from `doing-being-science.qmd` — the theory/process/statistical
test figure, with per-column structure built directly in the chunk
(`theory_columns`, `theory_edges`, `theory_titles`) and plain (arrowless)
curves:

````{verbatim}
```{r}
#| label: fig-theory-process-stats
#| fig-cap: "Different theories can imply the same process, and different processes can imply the same statistical test result"
#| fig-width: 9
#| fig-height: 5.0625

theory_columns <- list(
  c("Words we experience more often are easier to recall",
    "Frequency or age of experience does not matter to recall",
    "Words we experience early in life are easier to recall"),
  c("Neural network connection strength changes",
    "Neural word representation quality changes",
    "Word recognition decision-making strategy changes"),
  c("Experience correlates with word recognition speed",
    "Experience does not correlate with word recognition speed")
)
theory_titles <- c("Theories", "Process models", "Statistical tests")

# theories -> processes: many-to-many
# processes -> statistical models: every process links to both tests
theory_edges <- list(
  data.frame(from = c(1, 1, 2, 3, 3), to = c(1, 2, 3, 1, 2)),
  data.frame(from = rep(1:3, each = 2), to = rep(1:2, times = 3))
)

concept_map(theory_columns, theory_edges, theory_titles, arrows = FALSE, curvature = 0,
            edge_color = "#B0B0B0", edge_alpha = 0.45)$plot
```
````

Notice the edges are **generated** rather than hand-enumerated where a
pattern exists (`rep(1:3, each = 2)`/`rep(1:2, times = 3)` for "every
process links to both tests") — a useful pattern when a column's edges
are systematic rather than ad hoc.

### Parameter quick reference

| Parameter | Default | Meaning |
|---|---|---|
| `columns` | — | List of character vectors; node labels per column. |
| `edges` | — | List (length `length(columns)-1`); each a data frame of 1-based `from`/`to` indices. |
| `titles` | `NULL` | Optional column headings. |
| `rx`, `ry` | `1.15, 0.62` | Ellipse half-width/height; single value or one per column. |
| `dx`, `dy` | `3.3, 1.55` | Column spacing / within-column item spacing. |
| `wrap_width` | `16` | Label wrap width (characters); single value or one per column. |
| `fill_colors` | 5 pastel hexes | One per column; extended via `colorRampPalette()` if too few given. |
| `edge_color`, `edge_alpha` | `"#888888"`, `0.7` | Curve styling. |
| `curvature` | `0.2` | Passed to `geom_curve()`; `0` = straight lines. |
| `arrows` | `TRUE` | `FALSE` for plain correspondence lines (no arrowheads). |

### Common errors

| Message | Likely cause |
|---|---|
| `edges[[k]]$from has out-of-range value(s)...` | A `from`/`to` index in `edges` exceeds the number of items in that column. |
| ``rx` must have length 1 or length(columns) = k` | A per-column argument (`rx`, `ry`, `wrap_width`) was given the wrong-length vector. |
| Warning: `dx = ... may be too small for columns...` | Neighbouring columns' `rx` values are too large for the current `dx`; ellipses may overlap. |

## Exercises

1. Add a fourth theory sentence to `theory_columns[[1]]` and a matching
   row (or rows) to `theory_edges[[1]]`, pointing it at one or more
   existing process models. What happens if you forget to add the edge?
2. Deliberately set one `edges[[...]]$from` value to `99` (out of range)
   and re-render. Read the resulting error message — does it tell you
   exactly which column and index is wrong?
3. Flip `arrows` to `TRUE` and `curvature` to `0.3` on the
   theory/process/stats figure. Discuss: does the figure now seem to
   claim something it shouldn't (causality, rather than
   correspondence)?
4. Give the "Theories" column a larger `rx`/`ry` and a wider
   `wrap_width` (e.g. `rx = 1.6, ry = 0.9, wrap_width = 22`) while
   leaving the other columns at their defaults, using the per-column
   vector form described above. Confirm the long sentences fit inside
   their ellipses afterward.
5. Build a two-column (not three) concept map for a different topic you
   teach, with at least one many-to-many edge, from scratch.

## See also

This script has no dependencies on, and no dependents among, the other
four scripts in this set — it's used independently for the
theory/process/statistical-test slide, thematically preceding the
sampling/measurement-error sequence built from
[testing-sequence-walkthrough.qmd](testing-sequence-walkthrough.qmd) and
[sampling-distributions-walkthrough.qmd](sampling-distributions-walkthrough.qmd).
