---
title: "`error-rates-process-control.R`: false positives and false negatives as process-control error rates"
subtitle: "Walkthrough: how the script works, and how to use it in Quarto"
format:
  html:
    code-fold: false
    toc: true
bibliography: references.bib
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(fig.align = "center")
```

## Aims

`error-rates-process-control.R` reframes false-positive and false-negative
rates — concepts usually introduced as abstract properties of a hypothesis
test — as **properties of a decision procedure applied over and over**, in
the sense Neyman and Pearson built their theory of hypothesis testing around
[@neyman1933]. It does this in two steps, deliberately concrete before
abstract:

- **Part A** starts from something that needs no statistical machinery at
  all: ART3, the author-recognition test from `psych-battery-dgp.R`'s
  battery [@vermeiren2023], already produces real false positives (saying
  "yes" to a name that isn't an author) and real false negatives (saying
  "no" to a name that is) for every simulated participant. This script
  turns that into a **decision rule** — "classify someone as a frequent
  reader if they recognize at least `k` of the 60 authors" — and asks the
  question a rule like that actually needs answering: not "was this one
  classification right?", but "if this exact rule were applied over and
  over, how often would it be wrong, and in which direction?"
- **Part B** points out that a significance test at a fixed alpha is
  exactly this kind of rule — "reject H0 if p < alpha" — and reuses
  `psych-battery-dgp.R`'s sweep engine to show power, the empirical Type I
  rate, sign consistency (Type S), and CI coverage behaving exactly like
  ART3's rule's error rates did: stable long-run properties of a fixed
  procedure, recoverable by repetition, not properties of any one study's
  result. A fourth property gets its own figure: the Type M "exaggeration
  ratio" [@gelman2014] — not whether the procedure's *decision* is right,
  but how distorted the *number* it reports tends to be, conditional on
  having rejected.

@perezgonzalez2015 is a useful accessible companion here: it lays out, side
by side, how Fisher's evidential reading of a single p-value differs from
Neyman and Pearson's decision-procedure reading — and how the hybrid most
textbooks actually teach quietly borrows language from both without always
saying so.

After working through it, you should be able to:

- explain, using ART3's own classification rule as a worked example, why a
  false-positive/false-negative rate is a property of a *rule*, estimable
  only by repetition, not a property of any single person's test result;
- generate a reliability-by-sample-size or a true-correlation-by-sample-size
  error-rate heatmap for a battery pair of your choosing, and read off
  where a fixed testing procedure is well- vs. poorly-behaved;
- explain why a significance filter inflates reported effect sizes at low
  power, and why that inflation is a property of *selecting on
  significance*, not a flaw in any individual study;
- state, in your own words, the difference between Neyman and Pearson's
  long-run error-rate framing of a significance test and Fisher's
  evidential reading of a p-value.

## Prerequisites / assumptions

- **Packages**: `ggplot2` only.
- **Sourcing**: `psych-battery-dgp.R` must be sourced first — this script
  uses `default_test_specs()`, `default_reliability_vec()`,
  `default_r_obs_matrix()`, `get_calibration()`, `simulate_battery()`,
  `simulate_pair()`, `fit_focal_cor()`, `diagnose_cor_result()`,
  `disattenuate()`, `run_sweep()`, `summarize_sweep()`,
  `plot_metrics_vs_x()`, and `plot_power_heatmap()` directly.
- **Reader background**: the measurement-error/sampling-error distinction
  from `sampling-measurement-error-walkthrough.qmd` and the NHST/power
  framing from `nhst-power-walkthrough.qmd` are both drawn on directly here
  and are worth reading first if unfamiliar; some exposure to the terms
  "sensitivity"/"specificity" or "true positive rate"/"false positive rate"
  helps for Part A but is defined as it comes up.

```{r}
#| label: source-script
source("psych-battery-dgp.R")
source("error-rates-process-control.R")
```

## How it works: the process

```{mermaid}
flowchart LR
    A["ART3 item-level responses\nsimulate_art3_battery()"] --> B["evaluate_art3_rule()\none application of a\n'>= k authors' rule"]
    B --> C["sweep_art3_rule()\nrepeat over k x reps,\na fresh sample each time"]
    C --> D["plot_art3_roc()\ncloud = single applications,\nred line = the rule's long-run rates"]
    E["psych-battery-dgp.R\nrun_sweep() / run_reliability_sweep()"] --> F["summarize_sweep()\npower/Type I, sign consistency,\ntype_m_ratio, CI coverage, per grid cell"]
    F --> G["plot_reliability_heatmap() /\nplot_true_corr_heatmap() /\nplot_process_control_overview()"]
    F --> H["plot_type_m_curve()\nexaggeration ratio,\nits own (non 0-1) y-axis"]
```

The two halves share a structure even though they look different: both
repeat a fixed procedure many times (Part A: classify `n` people, `reps`
times, at each threshold `k`; Part B: run a significance test on `n`
participants, `reps` times, at each grid cell) and summarize what fraction
of those repetitions landed in each outcome category. Nothing in Part B is
new simulation machinery — it is entirely orchestration over
`psych-battery-dgp.R`'s existing `run_sweep()`/`summarize_sweep()`, the same
engine `nhst-power.R` uses, now read through a different lens.

## Element-by-element walkthrough

### Part A: ART3 as a concrete anchor

#### `simulate_art3_battery()` and `plot_art3_rates()`

```{r}
#| label: show-simulate-art3-battery
#| eval: false
d <- simulate_battery(n, reliability = list(ART3 = rel_val), seed = seed, keep_latent = TRUE)
hits    <- round((1 - d$ART3_fn_rate) * sp$split[["authors"]])
rejects <- round((1 - d$ART3_fp_rate) * sp$split[["foils"]])
true_ability <- (d$ART3_theta - cal$theta_mu) / cal$theta_sigma
```

Rather than re-implementing ART3's item-level draw, this reuses
`simulate_battery()`'s existing `ART3_fn_rate`/`ART3_fp_rate` columns and
backs out the raw `hits`/`rejects` counts from them — one less place for
the two codebases to drift apart. `true_ability` re-derives the
standardized ability `simulate_battery()` doesn't return directly (only the
logit-scale `theta`), by inverting the calibration it was built from.

```{r}
#| label: demo-art3-rates
#| fig-width: 6
#| fig-height: 5
plot_art3_rates(n = 500, seed = 1)
```

::: {.callout-note title="Why these two rates move together"}
`cor(fp_rate, fn_rate)` across a large simulated sample comes out strongly
**positive** (around .8), and the figure above shows it: people with a low
false-negative rate also tend to have a low false-positive rate. This is a
direct consequence of a modeling choice documented in
`psych-battery-dgp.R` — ART3 uses a single probability `p` per person for
*both* correctly recognizing an author and correctly rejecting a foil
(`hits ~ Binomial(60, p)`, `rejects ~ Binomial(30, p)`), rather than the two
separate parameters (sensitivity and response criterion) a full
signal-detection-theory model would use. That's a deliberate simplification
for this teaching battery, not an attempt at a psychometrically faithful
recognition-memory model — worth knowing before reading too much into the
exact *shape* of this scatter, even though the false-positive/false-negative
rates themselves are genuine.
:::

#### `evaluate_art3_rule()`: one application of a decision rule

```{r}
#| label: show-evaluate-art3-rule-core
#| eval: false
truth_positive     <- d$ART3_theta > cal$theta_mu   # true ability above the population average
predicted_positive <- hits >= k                      # the rule's call

tp <- sum(truth_positive & predicted_positive)
fn <- sum(truth_positive & !predicted_positive)
fp <- sum(!truth_positive & predicted_positive)
tn <- sum(!truth_positive & !predicted_positive)
```

The rule being evaluated — "classify as a frequent reader if `hits >= k`"
— needs a ground truth to be right or wrong *against*, and `theta_mu` (the
population's average true recognition ability, on the calibrated logit
scale) supplies one: being truly above average is independent of how the
noisy classification turns out, which is exactly the property a ground
truth needs.

```{r}
#| label: demo-evaluate-art3-rule
evaluate_art3_rule(n = 300, k = 40, seed = 2)
```

One call, one sample, one set of rates — and, by itself, not yet
informative about the *rule*: a different seed at the same `n` and `k` will
give a visibly different `fpr`/`tpr` pair. That instability is the reason
Part A doesn't stop here.

#### `sweep_art3_rule()` and `plot_art3_roc()`: the rule's long-run operating curve

```{r}
#| label: demo-sweep-art3-rule
sweep_small <- sweep_art3_rule(n = 300, k_grid = c(30, 40, 50), reps = 50, seed_base = 3)
aggregate(cbind(fpr, tpr) ~ k, sweep_small, mean)
```

`sweep_art3_rule()` repeats `evaluate_art3_rule()` at each threshold in
`k_grid`, drawing a **fresh** sample of `n` people each time — the single
mechanism that turns "one classification's noisy outcome" into "the rule's
actual operating characteristics," by averaging over many independent
applications rather than trusting any one of them.

```{r}
#| label: demo-art3-roc
#| fig-width: 6.5
#| fig-height: 5.5
plot_art3_roc(n = 300, k_grid = seq(25, 50, by = 5), reps = 200, seed_base = 2)
```

Read the pale cloud and the red line as two different claims. The cloud —
one dot per replicate — is how noisy a *single* application of the rule's
reported false-positive/true-positive rate can be, even at a respectable
`n = 300`. The red line connects each threshold's *average* over 200
replicates, and it traces a textbook ROC-style trade-off: tightening `k`
(demanding more recognized authors before calling someone a frequent
reader) buys a lower false-positive rate at the cost of a lower true-positive
rate, and no threshold minimizes both. That trade-off — not any particular
point on it — is the property of the *rule*, in the Neyman-Pearson sense:
knowable in advance, by simulation or by theory, before the rule is ever
applied to a real person.

### Part B: the same logic, abstracted to a fixed-alpha NHST procedure

#### `run_reliability_sweep()` and `plot_reliability_heatmap()`: holding the right thing fixed

This is the script's most distinctive piece of code, and it exists because
of a real bug caught while building this script — worth reading in full:

```{r}
#| label: show-run-reliability-sweep-full
run_reliability_sweep
```

::: {.callout-note title="A real bug this function fixes"}
The first version of `plot_reliability_heatmap()` called
`psych-battery-dgp.R`'s `run_sweep()` directly, sweeping `reliability_grid`
while holding `true_corr` fixed on the **observed** scale (`run_sweep()`'s
default interpretation) — e.g. "hold StuVoc1-GK's correlation at .67 while
reliability drops from .9 to .3." That request is not always satisfiable:
disattenuating an observed r = .67 back through a shared reliability of .3
requires a *latent* correlation above 1, which `disattenuate()` can only
handle by capping it at .999 and issuing a warning. Run across a whole
reliability grid, that produced a wall of repeated capping warnings and,
worse, a heatmap whose low-reliability cells were all silently simulating
the *same* near-1.0 latent correlation rather than the requested .67
observed correlation — a wrong figure that happened not to crash.

The fix is the same device `sampling-measurement-error.R`'s
`simulate_attenuation_sweep()` already uses: hold the **latent**
(construct-level) correlation fixed instead, and let the **observed**
correlation — and therefore its significance, sign, and coverage — respond
to reliability, which is the actual point of this figure. `true_corr_latent
* reliability` (both tests share one reliability value here) gives the
*implied* observed correlation at each cell, which is what actually gets
tested for significance/sign/coverage, and which is written into the
output's `true_corr` column so `summarize_sweep()` can be reused with no
changes at all.
:::

```{r}
#| label: demo-reliability-heatmap
#| fig-width: 7.5
#| fig-height: 5
plot_reliability_heatmap("StuVoc1", "GK", n_grid = c(20, 50, 100, 200, 400),
                          reliability_grid = seq(0.3, 0.9, by = 0.1), reps = 150, seed_base = 3)
```

Power climbs with both axes, as it should: more participants and a more
reliable measure both make a real effect easier to detect, and the figure
makes visible just how much reliability alone costs — at `reliability =
.3`, even `n = 20` only reaches 19% power, where the same `n` at
`reliability = .9` already reaches 97%.

#### `plot_true_corr_heatmap()` and `plot_process_control_overview()`

```{r}
#| label: demo-true-corr-heatmap
#| fig-width: 7
#| fig-height: 4
plot_true_corr_heatmap("StuVoc1", "GK", n_grid = c(20, 50, 100, 200, 400),
                        true_corr_grid = c(0, 0.2, 0.67), reps = 200, seed_base = 4)
```

Unlike `plot_reliability_heatmap()`, this one can safely reuse `run_sweep()`
directly, because reliability is held fixed at a sensible default rather
than swept down toward the disattenuation ceiling — the `true_corr = 0` row
is the empirical false-positive rate, and it should (and does) stay low and
flat across every `n`, unlike the other two rows.

```{r}
#| label: demo-process-control-overview
#| fig-width: 7.5
#| fig-height: 6.5
plot_process_control_overview("StuVoc1", "GK", n_grid = c(20, 50, 100, 200, 400), reps = 200, seed_base = 5)
```

All three panels answer the same underlying question Part A's ROC figure
did — "what happens if this exact procedure is applied over and over?" —
just for a significance test rather than a classification rule. The
`true_corr = 0` curves are the figure's real payoff: a flat, low power/Type
I-rate line and a (largely undefined, since there's no "correct sign" under
a true null) blank sign-consistency panel are what a well-behaved testing
procedure looks like when there's nothing there to find, in exactly the
sense Neyman and Pearson meant the "size" of a test's rejection region to
be fixed and known in advance [@neyman1933] — not inferred after the fact
from how convincing any one result felt.

#### A fourth error rate, on its own axis: `plot_type_m_curve()`

`summarize_sweep()` computes one more quantity than the three panels above
show: `type_m_ratio`, the median `|estimate| / |true_corr|` among
*significant* results only — the Type M "exaggeration ratio" [@gelman2014].
It answers a different question from power/Type I/sign-consistency/
coverage: not "does this procedure reach the right decision?", but "when it
rejects, how distorted is the number it hands back?"

::: {.callout-note title="Why this isn't a fourth panel in `plot_process_control_overview()`"}
`type_m_ratio` is deliberately kept out of the three-panel overview above,
not folded in as a fourth facet. Power, sign consistency, and coverage are
all proportions bounded to `[0, 1]`, which is what lets
`plot_metrics_vs_x()` share one y-axis across them. An exaggeration ratio
is `1` at no inflation and can comfortably exceed `2`–`3` at low power —
sharing that `[0, 1]` axis would silently clip the part of the story that
matters most. `plot_type_m_curve()` is a separate figure built around its
own appropriately scaled axis instead.
:::

```{r}
#| label: demo-type-m-curve
#| fig-width: 7
#| fig-height: 4.5
plot_type_m_curve("StuVoc1", "GK", n_grid = c(10, 15, 20, 30, 50, 100, 200),
                   true_corr_grid = 0.2, reps = 500, seed_base = 6)
```

A true correlation of .2 (rather than this pair's own ≈.67) is used here
deliberately, for the same reason `nhst-power-walkthrough.qmd` sweeps a
narrower `n` range for its corridor plot: a strong effect like StuVoc1–GK's
actual correlation reaches high power — and so, low exaggeration — almost
immediately, leaving nothing dramatic to see. At a true correlation of .2
(the same weak-effect benchmark @vermeiren2023's own power illustration uses), a
significant result at `n = 10` overstates the true correlation by a median
factor of roughly 3.5×; by `n = 200` that's fallen to close to 1×. Nothing
is wrong with any individual significant study along the way — the
inflation is a consequence of *selecting on significance*, which forces
every significant result at small `n` to clear a high bar (`|r|` has to
exceed roughly .44 to be significant at `n = 20`, for instance), and the
smaller the true effect relative to that bar, the more a result has to be
pulled upward by sampling variation to clear it at all.

## Quarto usage / integration

This script is standalone, like the other three in this set — not sourced
into `doing-being-science.qmd`. A typical setup chunk:

```r
source("psych-battery-dgp.R")
source("error-rates-process-control.R")
```

### Parameter quick reference

| Function | Parameter | Default | Meaning |
|---|---|---|---|
| `evaluate_art3_rule()` / `sweep_art3_rule()` / `plot_art3_roc()` | `k` / `k_grid` | — | Classification threshold(s): "frequent reader" if `hits >= k`. |
| | `n` | — | Participants per single application of the rule. |
| | `reps` | `200` | Independent applications of the rule per threshold (sweep/plot only). |
| | `reliability` | `NULL` | ART3's reliability; `NULL` uses its Vermeiren-default value. |
| `run_reliability_sweep()` / `plot_reliability_heatmap()` | `reliability_grid` | — | Reliability values to sweep (applied to both tests equally). |
| | `true_corr_latent` | `NULL` | The *true* (disattenuated) correlation held fixed; `NULL` backs it out from the pair's Vermeiren-default observed correlation. |
| `plot_true_corr_heatmap()` / `plot_process_control_overview()` | `true_corr_grid` | `NULL` (overview) / — (heatmap) | Observed-scale correlations to sweep; the overview's `NULL` default uses `c(0, <pair's Vermeiren-default r>)`. |
| | `reliability` | `NULL` | Reliability applied to both tests; `NULL` uses `var_x`'s Vermeiren-default value. |
| `plot_reliability_heatmap()` / `plot_true_corr_heatmap()` | `metric` | `"power_or_type1"` | Which `summarize_sweep()` column to plot: `"power_or_type1"`, `"sign_consistency"`, or `"coverage"` (not `"type_m_ratio"` — see `plot_type_m_curve()`). |
| `plot_type_m_curve()` | `true_corr_grid` | `NULL` | Observed-scale correlation(s) to compute the exaggeration ratio for; `NULL` uses the pair's Vermeiren-default value only (unlike the other Part B functions, **not** `c(0, ...)` — `type_m_ratio` is `NA` under the null). |
| | `n_grid` | — | Sample sizes to sweep. |
| *(sweep-based functions)* | `seed_base` | `1234` | Base seed for deterministic per-cell seeding. |

### Common errors

| Message | Likely cause |
|---|---|
| `` `n` must be a positive integer `` | `n <= 0` passed to `simulate_art3_battery()`/`evaluate_art3_rule()`. |
| `` `k` must be between 0 and 60 (the number of author items in ART3) `` | `k` outside ART3's valid range of recognizable authors. |
| `` `reps` must be at least 1 `` | `sweep_art3_rule()` called with `reps = 0`. |
| `` `metric` must be one of power_or_type1, sign_consistency, coverage `` | A misspelled or unsupported `metric` passed to either heatmap function. |
| Warning: `achieved reliability (...) differs from the requested target (...)` | A `reliability`/`reliability_grid` value pushed calibration outside tolerance — see `psych-battery-dgp.R`'s `calibrate_ability_general()`. |
| Warning: `Target observed correlation r = ... is not achievable at ...` | `plot_true_corr_heatmap()`'s `true_corr_grid` includes a value too large for the `reliability` in effect; **this is the exact warning `run_reliability_sweep()` was built to avoid** when sweeping reliability itself (see the callout above). |

## Exercises

1. Call `evaluate_art3_rule(n = 300, k = 40, seed = s)` for `s` in `1:10`.
   How much does `fpr` vary across the ten single applications? Compare
   that spread to the tight red trajectory in `plot_art3_roc()`'s `k = 40`
   point, built from 200 applications rather than one.
2. Lower ART3's `reliability` to `0.5` in `plot_art3_roc()` (via the
   `reliability` argument) and compare the resulting curve to the default
   (`reliability = .9`) version above. Does the trade-off get better or
   worse, and does that match what you'd predict from more measurement
   error?
3. Run `plot_reliability_heatmap()` for StuVoc1-ART3 (a weaker pair, r
   ≈ .52) instead of StuVoc1-GK. At `n = 100`, how much more does power
   suffer from the same reliability drop, compared to the StuVoc1-GK
   heatmap above?
4. Request `plot_true_corr_heatmap("StuVoc1", "GK", n_grid = c(50, 200),
   true_corr_grid = c(0.67), reliability = 0.2)` — deliberately low
   reliability with a high requested observed correlation. Confirm you get
   the "not achievable" warning from the Common Errors table, and explain
   in your own words why `plot_reliability_heatmap()`'s
   `run_reliability_sweep()` approach doesn't have the same problem.
5. Pick a `metric` other than the default (`"sign_consistency"` or
   `"coverage"`) for `plot_process_control_overview()`'s underlying
   `run_sweep()` call (via `summarize_sweep()`'s output directly, since
   `plot_process_control_overview()` always shows all three). Which of the
   three looks most stable across `n`, and which looks most sensitive to
   it?
6. Re-run `plot_type_m_curve("StuVoc1", "StuVoc2", ...)` (r ≈ .82, a much
   stronger effect than the true correlation of .2 used above) across the
   same `n_grid = c(10, 15, 20, 30, 50, 100, 200)`. How much smaller is the
   exaggeration ratio at `n = 10`, and does it still fall toward 1 by
   `n = 200`, or has it already arrived there well before that?

## See also

- [psych-battery-dgp-walkthrough.qmd](psych-battery-dgp-walkthrough.qmd) —
  the shared simulation/sweep engine (`run_sweep()`, `summarize_sweep()`,
  `disattenuate()`) this script's Part B builds on directly.
- [sampling-measurement-error-walkthrough.qmd](sampling-measurement-error-walkthrough.qmd)
  — `simulate_attenuation_sweep()`'s latent-vs-observed correlation device,
  reused (and the bug it would have prevented, explained) in
  `run_reliability_sweep()` above.
- [nhst-power-walkthrough.qmd](nhst-power-walkthrough.qmd) — the single-study
  and power-curve framing this script's Part B reframes as long-run
  process-control error rates rather than individual-study outcomes.

## References
