---
title: "`nhst-power.R`: null hypothesis significance testing and power"
subtitle: "Walkthrough: how the script works, and how to use it in Quarto"
format:
  html:
    code-fold: false
    toc: true
bibliography: references.bib
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(fig.align = "center")
```

## Aims

`nhst-power.R` carries two correlation pairs from the battery simulated by
`psych-battery-dgp.R` through the same null-hypothesis-significance-testing
machinery, deliberately chosen to have different effect sizes:

- **StuVoc1 – StuVoc2** — observed r ≈ .82 (two related vocabulary
  subtests)
- **StuVoc1 – GK** — observed r ≈ .67 (vocabulary and general
  knowledge)

The point of carrying *two* pairs through the same code, rather than one, is
to make a fact about power concrete rather than asserted: the same test, the
same decision rule, applied to two real correlations from the same battery,
needs very different sample sizes to reach the same power, purely because
their effect sizes differ. StuVoc1–StuVoc2 is reliably detectable at
n = 10–20; StuVoc1–GK needs several times that. Nothing about the
*test* changed between the two — only the thing being tested.

This script adds **no new simulation machinery of its own**. Every function
here is a thin orchestration layer over `psych-battery-dgp.R`'s
`simulate_pair()`, `fit_focal_cor()`, `diagnose_cor_result()`, `run_sweep()`,
`summarize_sweep()`, and `plot_metrics_vs_x()` — see
[psych-battery-dgp-walkthrough.qmd](psych-battery-dgp-walkthrough.qmd) for
how those are built. What's new here is purely the *framing*: one simulated
study and its decision, many studies at one sample size (a direct replica of
@vermeiren2023's own power illustration, their "Power considerations"
figure), and many studies across a sample-size grid (power curves).

After working through it, you should be able to:

- explain, in your own words, why "not significant" and "the null is true"
  are different claims, using this script's own true_corr = 0 curves as the
  concrete counter-example;
- generate a corridor plot and a power curve for a battery pair of your
  choosing, and read off roughly what sample size reaches 80% power;
- explain why StuVoc1–StuVoc2 and StuVoc1–GK need such different
  sample sizes to reach the same power, despite using an identical testing
  procedure.

## Prerequisites / assumptions

- **Packages**: `ggplot2` only — no new dependency beyond what
  `psych-battery-dgp.R` already requires.
- **Sourcing**: `psych-battery-dgp.R` must be sourced first — this
  script uses `default_r_obs_matrix()`, `simulate_pair()`, `fit_focal_cor()`,
  `diagnose_cor_result()`, `run_sweep()`, `summarize_sweep()`, and
  `plot_metrics_vs_x()` directly.
- **Reader background**: familiarity with p-values, confidence intervals,
  and the idea of statistical power is assumed (not re-derived here); see
  [psych-battery-dgp-walkthrough.qmd](psych-battery-dgp-walkthrough.qmd) for
  the underlying simulation model if that's the unfamiliar part.

```{r}
#| label: source-script
source("psych-battery-dgp.R")
source("nhst-power.R")
```

## How it works: the process

```{mermaid}
flowchart LR
    A["psych-battery-dgp.R\nsimulate_pair(), fit_focal_cor(),\ndiagnose_cor_result(), run_sweep(),\nsummarize_sweep(), plot_metrics_vs_x()"] --> B["summarize_single_study() /\nplot_single_study()\none simulated dataset,\none cor.test(), one decision"]
    A --> C["plot_corridor()\nrun_sweep() at one true_corr,\nmany n -- a Vermeiren-Fig.-1-style\nscatter coloured by significance"]
    A --> D["plot_power_curves()\nrun_sweep() + summarize_sweep()\nacross a full n grid,\ntrue_corr = 0 and the pair's real r"]
```

Three views of the same underlying machinery, increasing in scope: one
study (`B`) shows what a single p-value is actually attached to; the
corridor (`C`) shows what happens if that same study were run many times at
one sample size; the power curves (`D`) show what happens across a whole
range of sample sizes, with the `true_corr = 0` curve kept visible
throughout as the benchmark for "how often would this procedure falsely
claim an effect that isn't there."

## Element-by-element walkthrough

### Plain-language decisions: `format_pvalue()` and `interpret_decision()`

```{r}
#| label: show-interpret-decision
#| eval: false
interpret_decision <- function(p_value, alpha = 0.05) {
  p_label <- format_pvalue(p_value)
  if (p_value < alpha) {
    sprintf("reject H0 (%s, threshold %.2f) -- call this result \"significant\"", p_label, alpha)
  } else {
    sprintf("fail to reject H0 (%s, threshold %.2f) -- call this result \"not significant\"", p_label, alpha)
  }
}
```

A small thing worth calling out: both `summarize_single_study()` and
`plot_single_study()` call this one function rather than each writing their
own "significant"/"not significant" string, so the wording a reader sees on
a figure matches the wording in the accompanying table exactly —
avoiding a common source of confusion where a table says "significant" and
a figure caption says something subtly different (or vice versa).

```{r}
#| label: demo-interpret-decision
interpret_decision(0.002)
interpret_decision(0.31)
```

### One study: `summarize_single_study()` and `plot_single_study()`

```{r}
#| label: demo-single-study
summarize_single_study("StuVoc1", "GK", n = 200, seed = 1)
```

```{r}
#| label: demo-plot-single-study
#| fig-width: 6
#| fig-height: 5
plot_single_study("StuVoc1", "GK", n = 200, seed = 1)
```

This is deliberately the simplest possible figure in the whole project: one
scatterplot, one fitted line, one estimate/CI/p-value annotated directly on
it. Everything downstream — the corridor plot, the power curves —
is just this exact same procedure, repeated many times and summarized.
Nothing new is introduced at the level of *what a single study's result is*;
only *how many times it's repeated* changes.

::: {.callout-note title="Same n, different decision, just from a different seed"}
Re-run the chunk above with `seed = 6` instead of `seed = 1`. At `n = 200`
and the pair's true correlation of .67, both calls will almost certainly
still be significant (this pair is easy to detect) — but try the same
comparison at `n = 15` and the decision can flip between seeds. That
flip is not a sign anything went wrong; it's sampling error, made visible at
the level of a single reported decision rather than averaged away across
many studies, which is exactly what the corridor plot below is built to
show systematically.
:::

### Many studies, one N — the corridor plot: `plot_corridor()`

This is the script's most distinctive piece of code — a direct replica
of @vermeiren2023's own "Power considerations" figure, which plotted 2000
simulated studies' correlations against sample size, coloured by
significance, and reported that at N=40 only about a quarter of studies
testing a true r = .2 came back significant. Worth reading in full:

```{r}
#| label: show-plot-corridor-full
plot_corridor
```

It doesn't re-simulate anything new; it just calls `run_sweep()` (from
`psych-battery-dgp.R`) at a single `true_corr` value across a grid of sample
sizes, and plots the *raw*, per-replicate output directly — one point
per simulated study — rather than summarizing it first.

```{r}
#| label: demo-plot-corridor
#| fig-width: 7
#| fig-height: 5
plot_corridor("StuVoc1", "GK", n_grid = c(10, 15, 20, 30, 40), reps = 200, seed_base = 2)
```

::: {.callout-note title="Why n = 10–40 here, not Vermeiren's 40–400"}
@vermeiren2023's own figure swept N from 40 to 400 because they were
illustrating power for a true correlation of r = .2 — a much weaker
effect than any pair in this battery. StuVoc1–GK's true correlation
(≈ .67) is strong enough that by N=40 almost every simulated study is
already significant — try `plot_corridor("StuVoc1", "GK", n_grid = c(40,
100, 200, 400), reps = 200)` yourself and compare: it comes out solid red
throughout, with no mix left to see. Sweeping a *smaller* n range here is
what's needed to see the mix of significant and non-significant outcomes
that makes a corridor plot informative — the right N range to sweep
depends on the effect size being investigated, not on any fixed
convention.
:::

### Many studies, many N — power curves: `plot_power_curves()`

```{r}
#| label: show-plot-power-curves-core
#| eval: false
if (is.null(true_corr_grid)) {
  true_corr_grid <- c(0, unname(default_r_obs_matrix()[var_x, var_y]))
}
sweep <- run_sweep(n_grid = n_grid, true_corr_grid = true_corr_grid, ...)
summ  <- summarize_sweep(sweep)
plot_metrics_vs_x(summ, x = "n", metrics = "power_or_type1", color_by = "true_corr") +
  geom_hline(yintercept = alpha, linetype = "dotted", color = "gray40")
```

The default `true_corr_grid` is worth noticing: it's always `c(0, <this
pair's real correlation>)`, never just the real correlation on its own. That
default isn't a convenience shortcut — it's the whole point of this
figure. A power curve shown on its own answers "how often would we detect
this real effect?" A power curve with the null included alongside it also
answers the question a reader is more likely to actually need: "how do I
tell a detected real effect apart from a false alarm, at this same sample
size?" The dotted horizontal line at `alpha` gives the second curve a
reference point — a well-behaved false-positive rate should hover near
it regardless of `n`, which is a different (and much flatter) shape than
power ever takes.

```{r}
#| label: demo-power-curves-stuvoc1-stuvoc2
#| fig-width: 7
#| fig-height: 4.2
plot_power_curves("StuVoc1", "StuVoc2", n_grid = c(10, 20, 40, 80), reps = 300, seed_base = 3)
```

```{r}
#| label: demo-power-curves-stuvoc1-gk
#| fig-width: 7
#| fig-height: 4.2
plot_power_curves("StuVoc1", "GK", n_grid = c(20, 50, 100, 200, 400), reps = 300, seed_base = 4)
```

Read the two figures side by side: StuVoc1–StuVoc2's power curve is
already near 1 by `n = 20`; StuVoc1–GK needs `n` in the low hundreds to
get there. Both true_corr = 0 curves, meanwhile, sit near the dotted .05
line throughout — flat, low, and (within simulation noise) unaffected
by `n`, which is exactly what a false-positive rate should look like when
nothing about the testing procedure has changed.

## Quarto usage / integration

This script is standalone — like `sampling-measurement-error.R`, it
isn't currently sourced into `doing-being-science.qmd`. A typical setup
chunk:

```r
source("psych-battery-dgp.R")
source("nhst-power.R")
```

and then any of `summarize_single_study()`, `plot_single_study()`,
`plot_corridor()`, or `plot_power_curves()`, each returning a plain data
frame or `ggplot` object respectively.

### Parameter quick reference

| Function | Parameter | Default | Meaning |
|---|---|---|---|
| `summarize_single_study()` / `plot_single_study()` | `var_x`, `var_y` | — | The focal test pair. |
| | `n` | — | Sample size for this one study. |
| | `true_corr` | `NULL` | Population *observed*-score correlation to simulate from; `NULL` uses the pair's Vermeiren-default value. |
| | `alpha` | `0.05` | Significance threshold. |
| `plot_corridor()` | `n_grid` | — | Sample sizes to compare (one row of points each). |
| | `reps` | `150` | Independent simulated studies per sample size. |
| | `true_corr` | `NULL` | A single population correlation (not a grid — this figure fixes the effect size and varies `n`). |
| `plot_power_curves()` | `n_grid` | — | Sample sizes to sweep. |
| | `true_corr_grid` | `NULL` | Correlations to compare; `NULL` uses `c(0, <pair's Vermeiren-default r>)`, always including the null. |
| | `reliability_grid` | `NULL` | Reliability override for both tests; `NULL` uses Vermeiren defaults. |
| | `reps` | `300` | Replicate studies per grid cell. |
| *(all four)* | `seed` / `seed_base` | `NULL` / `1234` | Reproducibility (single-study functions take `seed`; sweep-based functions take `seed_base`, passed through to `run_sweep()`). |

### Common errors

This script introduces no new validation of its own — every
`stopifnot()`/`warning()` a user might see originates in `psych-battery-dgp.R`
(see its own [Common errors
table](psych-battery-dgp-walkthrough.qmd#common-errors)), most commonly:

| Message | Likely cause |
|---|---|
| `` `var_x`/`var_y` must each be one of StuVoc1, StuVoc2, GK, ART3 `` | Misspelled or unsupported test name passed to any function here. |
| `` `var_x` and `var_y` must be different tests `` | The same test name passed twice as both `var_x` and `var_y`. |
| Warning: `achieved reliability (...) differs from the requested target (...)` | A `reliability`/`reliability_grid` value pushed calibration outside its tolerance — see `psych-battery-dgp.R`'s `calibrate_ability_general()`. |

## Exercises

1. Call `summarize_single_study("StuVoc1", "StuVoc2", n = 15, seed = 1)` for
   several different seeds (1 through 10). How many of the ten come back
   significant? Does that roughly match what `plot_power_curves()` predicts
   for `n = 15` on this pair?
2. Build a corridor plot for StuVoc1–ART3 (r ≈ .52) using
   `n_grid = c(10, 20, 40, 80, 150)`. Compare the N at which it turns mostly
   red (significant) to the StuVoc1–GK corridor above — does the
   weaker correlation need a larger N to reach the same mix, as you'd
   expect?
3. Run `plot_power_curves("StuVoc1", "GK", n_grid = c(...), reliability_grid = 0.5)`
   (a reliability well below the Vermeiren default for both tests) and
   compare to the default-reliability power curve above. Which error
   source — sampling or measurement — is responsible for the
   difference, and does that match what
   [sampling-measurement-error-walkthrough.qmd](sampling-measurement-error-walkthrough.qmd)
   would predict?
4. At what sample size does StuVoc1–GK's power first exceed 80%? Use
   `plot_power_curves()`'s underlying `run_sweep()`/`summarize_sweep()` calls
   directly (rather than reading it off the plot) to find a more precise
   answer.
5. Pick an `alpha` other than .05 (e.g. .01 or .10) and re-run
   `plot_power_curves("StuVoc1", "StuVoc2", ...)` with it. What happens to
   both the true_corr = 0 curve and the true_corr = .82 curve, and why do
   they move in opposite directions?

## See also

- [psych-battery-dgp-walkthrough.qmd](psych-battery-dgp-walkthrough.qmd) —
  the shared simulation/sweep engine this script orchestrates
  (`simulate_pair()`, `run_sweep()`, `summarize_sweep()`,
  `plot_metrics_vs_x()`), including its own validation against
  @vermeiren2023's reported power benchmarks.
- [sampling-measurement-error-walkthrough.qmd](sampling-measurement-error-walkthrough.qmd)
  — the measurement-error/sampling-error distinction this script's
  `reliability_grid` argument connects to (Exercise 3 above).
- [error-rates-process-control-walkthrough.qmd](error-rates-process-control-walkthrough.qmd)
  — takes the same `run_sweep()`/`summarize_sweep()` engine used here and
  reframes power, Type I rate, sign consistency, and coverage as long-run
  Neyman–Pearson process-control error rates, varying all three dials
  together rather than one pair at a time.

## References
