---
title: "`peer-review-workflow.R`: the peer-review cycle as a flow diagram"
subtitle: "Walkthrough: how the script works, and how to use it in Quarto"
format:
  html:
    code-fold: false
    toc: true
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(fig.align = "center")
```

## Aims

`peer-review-workflow.R` draws a single, fixed diagram of the peer-reviewed
publication cycle — author submits to editor, editor sends out to two
reviewers (fan-out), reviewers report back (fan-in), editor decides, with a
dashed loop-back arrow for "reject with revise & resubmit → repeat". It
reuses the "person at a computer" icon from `testing-sequence.R` across
four roles (author, editor, reviewer ×2) rather than drawing a new icon
style, so the peer-review figure visually matches the testing-sequence
figures elsewhere in the deck.

Unlike the other four scripts in this set, this one is **not
parametrized** for general reuse — it's a single bespoke diagram, built
with hand-placed coordinates for this specific six-icon layout. This
document explains that layout logic so it can be adapted (e.g. a
different number of reviewers) rather than treated as a black box.

After working through it, you should be able to:

- explain how the diagram's icons, arrows, and labels are each built from
  the same hand-placed `pos` coordinates;
- trace how arrow endpoints are anchored just outside icons, not at their
  centres, and why that matters;
- adapt the diagram for a different number of roles or reviewers.

## Prerequisites / assumptions

- **Packages**: `ggplot2` only.
- **Sourcing**: `person-grid.R` **and** `testing-sequence.R` must both be
  sourced first, in that order — this script calls
  `person_at_computer()` directly, which itself depends on
  `person_parts()`/`monitor_parts()`.
- **Reader background**: read
  [person-grid-walkthrough.qmd](person-grid-walkthrough.qmd) and
  [testing-sequence-walkthrough.qmd](testing-sequence-walkthrough.qmd)
  first — this script is a thin, specific application of both.

```{r}
#| label: source-scripts
source("person-grid.R")
source("testing-sequence.R")
source("peer-review-workflow.R")
```

## How it works: the process

```{mermaid}
flowchart TB
    A["pos: named list of\nhand-placed (x, y) coordinates\nfor each of the 6 icons"] --> B["icon_df() / icon helper\nbuilds each icon via\nperson_at_computer(), tags it,\ncolours it by role"]
    A --> C["right_of() / left_of()\ncompute arrow endpoints\njust outside each icon,\nnot at icon centres"]
    C --> D["straight_arrows, curve_arrows,\nloop_arrow: data frames of\narrow endpoints"]
    B --> E["single ggplot():\nicons + segments + curves\n+ role/step text labels"]
    D --> E
```

The diagram is built from three largely independent ingredients that only
come together in the final `ggplot()` call: **where** each icon sits
(`pos`), **what** each icon looks like (role-specific colour via
`person_at_computer()`), and **how** the icons connect (arrows computed
from positions, not drawn freehand).

## Element-by-element walkthrough

### Layout: the `pos` list and role colours

```{r}
#| label: show-pos
#| eval: false
researcher_col <- list(person = "#3B6EA5", screen = "#CFE8FF")
editor_col     <- list(person = "#B0562D", screen = "#FFDCC5")
reviewer_col   <- list(person = "#3B9E5F", screen = "#D3F5DF")

pos <- list(
  researcher1 = c(0,    0),
  editor1     = c(5.0,  0),
  reviewerA   = c(10.5, 3.4),
  reviewerB   = c(10.5, -3.4),
  editor2     = c(16.0, 0),
  researcher2 = c(21.0, 0)
)
```

Every coordinate here is hand-chosen, not computed from a formula (unlike
`person-grid.R`'s automatic grid spacing). The author and editor appear
**twice** (`researcher1`/`researcher2`, `editor1`/`editor2`) at different
x-positions, because the same two people appear at both the start and end
of the cycle — this is a flow diagram of a *process*, not a single
snapshot of who's present. The two reviewers are offset vertically
(`+3.4`/`-3.4`) at the same x-position, to draw the fan-out/fan-in
visually as a split-and-rejoin. Each role gets its own `person`/`screen`
colour pair, reused across every icon that plays that role.

### Building each icon: the `icon_df()` helper

```{r}
#| label: show-icon-df
#| eval: false
icon_df <- function(name, kind, cols) {
  p <- pos[[name]]
  d <- person_at_computer(kind = kind, ox = p[1], oy = p[2],
                           person_color = cols$person, screen_color = cols$screen)
  d$icon <- name
  d
}

icons <- rbind(
  icon_df("researcher1", "A", researcher_col),
  icon_df("editor1",     "B", editor_col),
  icon_df("reviewerA",   "A", reviewer_col),
  icon_df("reviewerB",   "B", reviewer_col),
  icon_df("editor2",     "B", editor_col),
  icon_df("researcher2", "A", researcher_col)
)
icons$poly_id <- paste(icons$icon, icons$part, icons$role)
```

This is a direct application of `person_at_computer()` (from
`testing-sequence.R`) at each named position, looked up from `pos`. The
`kind` argument (`"A"`/`"B"`) here is just varying the leg silhouette for
visual variety across the six icons — it carries no role meaning (unlike
`person_color`/`screen_color`, which *does* encode role). As with every
other script in this set, `poly_id` is rebuilt to be unique per icon
*and* per sub-part, so `ggplot`'s `group` aesthetic doesn't connect
polygons that happen to share a part name (e.g. every icon's `"head"`)
across different icons.

```{r}
#| label: demo-icon-df
#| fig-cap: "The six positioned icons, before any arrows are added"
researcher_col <- list(person = "#3B6EA5", screen = "#CFE8FF")
editor_col     <- list(person = "#B0562D", screen = "#FFDCC5")
reviewer_col   <- list(person = "#3B9E5F", screen = "#D3F5DF")
pos <- list(
  researcher1 = c(0, 0), editor1 = c(5.0, 0),
  reviewerA = c(10.5, 3.4), reviewerB = c(10.5, -3.4),
  editor2 = c(16.0, 0), researcher2 = c(21.0, 0)
)
icon_df <- function(name, kind, cols) {
  p <- pos[[name]]
  d <- person_at_computer(kind = kind, ox = p[1], oy = p[2],
                           person_color = cols$person, screen_color = cols$screen)
  d$icon <- name
  d
}
demo_icons <- rbind(
  icon_df("researcher1", "A", researcher_col),
  icon_df("editor1", "B", editor_col),
  icon_df("reviewerA", "A", reviewer_col),
  icon_df("reviewerB", "B", reviewer_col),
  icon_df("editor2", "B", editor_col),
  icon_df("researcher2", "A", researcher_col)
)
demo_icons$poly_id <- paste(demo_icons$icon, demo_icons$part, demo_icons$role)
ggplot(demo_icons, aes(x, y, group = poly_id, fill = fill)) +
  geom_polygon() +
  scale_fill_identity() +
  coord_equal() +
  theme_void()
```

### Arrows: `right_of()`/`left_of()` and three arrow data frames

```{r}
#| label: show-arrow-helpers
#| eval: false
right_of <- function(name, dy = 0.9) c(pos[[name]][1] + 1.25, pos[[name]][2] + dy)
left_of  <- function(name, dy = 0.9) c(pos[[name]][1] - 0.55, pos[[name]][2] + dy)
```

::: {.callout-note title="Why not anchor arrows at icon centres?"}
Arrows are anchored just outside each icon (`+1.25`/`-0.55` from the
icon's x-position, with a small `+0.9` vertical nudge) rather than at icon
centres — centring the arrows would draw them straight through the icon
polygons instead of visibly connecting two separate shapes.
:::

Three kinds of connections are built from these anchors — the
fan-out/fan-in pair is the most distinctive, since its `curvature` sign is
chosen deliberately rather than left at a single default:

```{r}
#| label: show-curve-arrows
#| eval: false
curve_arrows <- data.frame(rbind(
  c(right_of("editor1")[1],   right_of("editor1")[2],   left_of("reviewerA")[1], left_of("reviewerA")[2],  0.30),
  c(right_of("editor1")[1],   right_of("editor1")[2],   left_of("reviewerB")[1], left_of("reviewerB")[2], -0.30),
  c(right_of("reviewerA")[1], right_of("reviewerA")[2], left_of("editor2")[1],   left_of("editor2")[2],   -0.30),
  c(right_of("reviewerB")[1], right_of("reviewerB")[2], left_of("editor2")[1],   left_of("editor2")[2],    0.30)
))
names(curve_arrows) <- c("x", "y", "xend", "yend", "curvature")
```

- `straight_arrows` — plain `geom_segment()` arrows: author → editor, and
  editor → author (submission and final decision).
- `curve_arrows` (shown above) — `geom_curve()` arrows with an explicit
  `curvature` value per row (`+0.30` or `-0.30`) for the fan-out (editor →
  two reviewers) and fan-in (two reviewers → editor) — the sign of
  `curvature` is mirrored to match each reviewer's vertical offset, so the
  four curves bow outward symmetrically rather than crossing.
- `loop_arrow` — a single dashed `geom_curve()` well below the main row
  (`y = -7.0`), representing "reject, revise & resubmit" looping back from
  the end to the start.

### Text labels and final assembly

Two more data frames (`step_captions`, `role_labels`) place bold numbered
step captions (e.g. "1. Submit manuscript") and small role labels
("Author", "Editor", "Reviewer 1"/"2") at further hand-offset positions.

::: {.callout-note title="Hand-tuned, not computed"}
A code comment in the script flags that *Reviewer 1*'s label needed a
different vertical offset than the generic pattern, to clear its icon
without crowding the step caption beneath it. That's a reminder that this
diagram's positions were tuned by eye, so changing icon spacing or icon
size will likely require re-tuning label offsets too, rather than
cascading automatically.
:::

Everything (polygons, segments, curves, loop arrow, annotation, two sets
of text labels) is layered onto **one** `ggplot()` object and returned as
`peer_review_workflow()$plot`.

```{r}
#| label: demo-peer-review-workflow
#| fig-cap: "The full peer-review workflow diagram"
#| fig-width: 9
#| fig-height: 6
peer_review_workflow()$plot
```

## Quarto usage / integration

Setup-chunk lines (order matters: both dependencies before this script):

```r
source("person-grid.R")
source("testing-sequence.R")
source("peer-review-workflow.R")
```

Real usage from `doing-being-science.qmd`:

````{verbatim}
```{r}
#| label: fig-peer-review-workflow
#| fig-cap: "The peer-reviewed publication cycle: submission, review, and decision, repeating until accept or reject"
#| fig-width: 9
#| fig-height: 6

peer_review_workflow()$plot
```
````

`peer_review_workflow()` takes **no arguments** — there's nothing to
configure via chunk options beyond `fig-width`/`fig-height` (the plot uses
`coord_equal(clip = "off", ylim = c(-8.8, 5.8))` internally, so it has a
fixed aspect ratio; picking a `fig-width`/`fig-height` far from roughly
9:6 will add letterboxing rather than reflowing the diagram).

### If you need to adapt this diagram

Because nothing here is parametrized, adapting it (e.g. three reviewers
instead of two, or a different number of decision rounds) means editing
the script directly rather than passing new arguments:

1. Add/remove entries in `pos` (and give any new role a colour list like
   `reviewer_col`).
2. Add/remove the corresponding `icon_df()` calls.
3. Add/remove rows in `curve_arrows` for the new fan-out/fan-in
   connections, with signed `curvature` chosen to avoid crossing.
4. Re-check `role_labels`/`step_captions` y-offsets by eye — as noted
   above, these were tuned per-icon, not computed generically.

### Common symptoms

| Symptom | Likely cause |
|---|---|
| Diagram looks squeezed or letterboxed | `fig-width`/`fig-height` far from the ~9:6 ratio the fixed `coord_equal(clip = "off", ylim = ...)` was designed for. |
| New icon added but no arrow reaches it | Forgot to add `right_of()`/`left_of()`-based rows for the new position in `curve_arrows`/`straight_arrows`. |
| New curves cross awkwardly | `curvature` signs not mirrored to match the new icon's vertical offset. |
| A label overlaps its icon or a step caption | Offsets are hand-tuned per icon (see callout above) — a new/moved icon needs its own offset check, not just reuse of an existing one. |

## Exercises

1. Re-render `peer_review_workflow()$plot` at `fig-width = 6, fig-height
   = 6` (narrower than the 9:6 the diagram was designed for). What
   happens, and why (hint: look at the internal `coord_equal(clip =
   "off", ylim = ...)` call)?
2. Add a third reviewer: duplicate `reviewerB`'s entry in `pos` as
   `reviewerC` at a third vertical offset, add a matching `icon_df()`
   call, and add the corresponding fan-out/fan-in rows to
   `curve_arrows`. What else needs adjusting before it looks right?
3. Change `reviewer_col`'s `person` colour and re-render. Which parts of
   the diagram change colour, and which don't? Explain by tracing how
   `person_at_computer()` uses `person_color` vs. `screen_color`.
4. The script notes Reviewer 1's label offset needed manual tuning.
   Deliberately break it (set its `y` back to the generic `+1.4` the
   code comment mentions) and re-render — confirm you can see why the
   fix was needed.
5. This script takes no arguments at all. Sketch (in words, or in code if
   you want to go further) what a parametrized version might look like —
   e.g. `peer_review_workflow(n_reviewers = 3)` — and what would have to
   become programmatic (positions? arrows? labels?) rather than
   hand-placed.

## See also

- [testing-sequence-walkthrough.qmd](testing-sequence-walkthrough.qmd) —
  the source of the `person_at_computer()` icon reused four times here.
- [person-grid-walkthrough.qmd](person-grid-walkthrough.qmd) — the
  underlying geometry/colour primitives.
