---
title: "`person-grid.R`: a parametrized pictogram generator"
subtitle: "Walkthrough: how the script works, and how to use it in Quarto"
format:
  html:
    code-fold: false
    toc: true
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(fig.align = "center")
```

## Aims

`person-grid.R` draws "restroom-sign style" person icons — plain polygons,
not an icon font or image file — and arranges any number of them into a
grid. It's the base building block behind several other figures in this
book: a single icon reused by `testing-sequence.R` (a person at a computer)
and by `peer-review-workflow.R` (author/editor/reviewer roles), plus two
figures used directly:

- a **diversity sample** figure — a grid of icons varying in silhouette,
  age, disability, and skin tone, to make "any group of people will vary"
  concrete rather than abstract;
- an **effect grid** figure — the same grid mechanism, but with each icon's
  fill colour driven by a numeric value (e.g. an effect size or score) on a
  diverging blue–white–orange scale, to show "people vary in how they
  respond" to some intervention or condition.

This document explains how the script is built, then how to call it from a
Quarto chunk.

After working through it, you should be able to:

- explain how one icon's shape, its demographic composition, and its fill
  colour are each decided independently, then combined only at the end;
- generate a diversity-sample grid and an effect-size (value-mode) grid of
  your own, with your own proportions/values;
- read the parameter quick-reference table well enough to adapt a call
  without re-reading the source.

## Prerequisites / assumptions

- **Packages**: `ggplot2` only (loaded by the script itself). No icon
  fonts or external image assets — every icon is drawn from coordinates,
  so it renders identically wherever Quarto builds the document.
- **Sourcing**: standalone — nothing else needs to be sourced first. (It
  is, however, a dependency *of* `testing-sequence.R` and
  `peer-review-workflow.R` — see their own walkthroughs.)
- **Reader background**: basic familiarity with `ggplot2` (`geom_polygon()`,
  `aes()`, `theme_void()`) is assumed; no statistics background needed.

```{r}
#| label: source-script
source("person-grid.R")
```

## How it works: the process

Each icon is just a handful of flat shapes (head, torso, legs or
wheelchair, optional cane) assembled as polygons in a local coordinate
frame, then offset into a grid position. Colour and demographic
composition are decided by sampling from user-specified proportions. The
pipeline for building the whole grid looks like this:

```{mermaid}
flowchart LR
    A["person_parts()\n(one icon's polygons)"] --> E["assemble + offset\ninto grid positions"]
    B["sample_counts()\n(silhouette / age / disability mix)"] --> E
    C["diverging_colors() or\nsampled skin tones\n(fill colour)"] --> E
    D["choose_ncol()\n(grid shape)"] --> E
    E --> F["ggplot() + geom_polygon()\n+ scale_fill_identity()"]
```

In short: **geometry** (what shape is each icon) is decided independently
from **composition** (how many of each kind) and **colour** (what fill
each gets), and all three are combined only at the very end when the
polygon data frame is built and handed to `ggplot()`.

## Element-by-element walkthrough

### Low-level geometry: `circle_pts()` and `person_parts()`

```{r}
#| label: show-circle-pts
#| eval: false
circle_pts <- function(cx, cy, r, n = 24) {
  t <- seq(0, 2 * pi, length.out = n + 1)[-(n + 1)]
  data.frame(x = cx + r * cos(t), y = cy + r * sin(t))
}
```

`circle_pts()` is a tiny helper that turns a centre + radius into `n`
points around a circle (used for the head, and for the wheelchair wheel).
It's deliberately generic — nothing about it is specific to drawing people.

`person_parts(kind, age, disabled, scale, ox, oy)` is where one icon's
shape is actually defined. It builds a **named list** of polygons (head,
torso, legs, etc.) in a fixed local coordinate frame (head centred at
`x = 0, y = 1.62`), then at the very end:

```{r}
#| label: show-offset-logic
#| eval: false
out <- lapply(names(parts), function(nm) {
  p <- parts[[nm]]
  data.frame(part = nm, x = ox + scale * p$x, y = oy + scale * p$y)
})
do.call(rbind, out)
```

scales and offsets every part by `(ox, oy)` — this is *the* mechanism that
lets the same local shape be placed anywhere in a grid (or, in
`testing-sequence.R`, next to a monitor icon) just by changing `ox`/`oy`.

Three design choices worth knowing about, because they explain shapes you
might otherwise find surprising in the rendered figure. The most
distinctive is the leg silhouette, controlled by `kind`:

```{r}
#| label: show-kind-branching
#| eval: false
if (kind == "A") {           # trouser legs: two bars with a slim gap
  parts[["leg_l"]] <- data.frame(x = c(-0.20, -0.03, -0.06, -0.20),
                                  y = c(0.72, 0.72, 0.00, 0.00))
  parts[["leg_r"]] <- data.frame(x = c(0.03, 0.20, 0.20, 0.06),
                                  y = c(0.72, 0.72, 0.00, 0.00))
} else if (kind == "B") {    # skirt
  parts[["legs"]] <- data.frame(x = c(-0.20, 0.20, 0.46, -0.46),
                                 y = c(0.72, 0.72, 0.00, 0.00))
} else if (kind == "N") {    # half trouser (left) / half skirt (right)
  parts[["leg_l"]] <- data.frame(x = c(-0.20, -0.02, -0.05, -0.20),
                                  y = c(0.72, 0.72, 0.00, 0.00))
  parts[["legs_skirt_half"]] <- data.frame(x = c(0.00, 0.20, 0.46, 0.00),
                                            y = c(0.72, 0.72, 0.00, 0.00))
}
```

::: {.callout-note title="Why a third silhouette, instead of just A/B?"}
`"A"` is trouser-coded (two separate leg bars with a slim gap) and `"B"` is
skirt-coded (one flared polygon). Rather than picking one of those two as
a generic "default" person, `"N"` is literally half of each — an
inclusive, non-binary-coded silhouette, not a third stereotyped shape
layered on top of the first two.
:::

The other two choices change less structurally:

- `disabled = TRUE` swaps the torso/legs for a single wheelchair wheel plus
  a footrest peg, redesigned (per a code comment in the script) as one big
  unbroken shape rather than several small ones, because small shapes stop
  reading clearly once icons are shrunk down in a large grid.
- `age = "elder"` adds a cane polygon; `age = "child"` doesn't change the
  shape at all — it's handled later, as an overall `scale` multiplier in
  `person_grid()`, not inside `person_parts()`.

```{r}
#| label: demo-person-parts
#| fig-cap: "A single adult icon, drawn directly from person_parts()"
one_person <- person_parts("A", "adult", disabled = FALSE)
ggplot(one_person, aes(x, y, group = part)) +
  geom_polygon(fill = "#3B6EA5") +
  coord_equal() +
  theme_void()
```

### Composition sampling: `sample_counts()`

```{r}
#| label: show-sample-counts
#| eval: false
sample_counts <- function(n, props, seed) {
  stopifnot(
    "`props` must have non-negative values, with at least one positive" =
      all(props >= 0) && sum(props) > 0
  )
  props <- props / sum(props)
  set.seed(seed)
  raw <- props * n
  counts <- floor(raw)
  remainder <- n - sum(counts)
  if (remainder > 0) {
    order <- order(-(raw - counts))
    for (i in seq_len(remainder)) {
      idx <- order[((i - 1) %% length(order)) + 1]
      counts[idx] <- counts[idx] + 1
    }
  }
  sample(rep(names(props), counts))
}
```

Given target proportions (e.g. `c(adult = 0.70, child = 0.15, elder =
0.15)`) and a grid size `n`, this turns proportions into **exact integer
counts** using the largest-remainder method (the same allocation method
used for apportioning seats to parties in some electoral systems): take the
floor of each `proportion * n`, then hand out the few leftover icons to
whichever categories had the largest fractional remainder. This hits the
requested mix as closely as integers allow, rather than independently
sampling each icon (which could, by chance, give you 0 children out of 60
even at `child = 0.15`). The final `sample()` call shuffles the result so
icons of the same kind aren't clumped together in the grid.

```{r}
#| label: demo-sample-counts
sample_counts(10, c(adult = 0.7, child = 0.15, elder = 0.15), seed = 1)
```

::: {.callout-note title="disabled_prop = 0 is fine"}
The `stopifnot` requires *at least one* positive proportion overall, but
zero is allowed for individual categories — so `disabled_prop = 0` (the
default in some contexts) works fine: that category just gets 0 of the
`n` icons, not an error.
:::

### Colour: `diverging_colors()`

```{r}
#| label: show-diverging-colors
#| eval: false
diverging_colors <- function(value, neg = "#0072B2", mid = "#F2F1EC", pos = "#E69F00") {
  span <- max(abs(value), 1e-9)
  tt <- value / span                     # -1 .. 1, 0 = null
  ramp_lo <- grDevices::colorRamp(c(neg, mid))
  ramp_hi <- grDevices::colorRamp(c(mid, pos))
  to_hex <- function(m) grDevices::rgb(m[, 1], m[, 2], m[, 3], maxColorValue = 255)
  vapply(tt, function(t) {
    if (t <= 0) to_hex(ramp_lo(t + 1))
    else        to_hex(ramp_hi(t))
  }, character(1))
}
```

This is the colour engine behind `person_grid()`'s `mode = "value"` —
e.g. colouring icons by an effect size or score. It's a **symmetric,
diverging** scale: `0` is always treated as the reference/"null" point
(light, near-white), regardless of where your data happen to sit, and the
two endpoints (default blue/orange — the Okabe–Ito colour-blind-safe
palette, not a red/green or red/blue scale) are reached only at
`max(abs(value))`.

::: {.callout-note title="The scale is set by whichever value is furthest from zero"}
A `value` vector of `c(-1, 3)` will *not* make `-1` look strongly
negative, because `3` is furthest from zero and so sets the scale — `-1`
ends up only a third of the way to the "strongly negative" endpoint.
Worth checking if your values aren't already roughly symmetric around
zero.
:::

```{r}
#| label: demo-diverging-colors
#| fig-cap: "The diverging scale at a few sample values"
vals <- c(-2, -1, 0, 1, 2)
data.frame(value = vals, fill = diverging_colors(vals)) |>
  ggplot(aes(x = value, y = 1, fill = fill)) +
  geom_tile(width = 0.9, height = 0.9) +
  scale_fill_identity() +
  theme_void()
```

### Automatic grid shape: `choose_ncol()`

```{r}
#| label: show-choose-ncol
#| eval: false
choose_ncol <- function(n, dx, dy, aspect) {
  ncols <- seq_len(n)
  nrows <- ceiling(n / ncols)
  grid_ratio <- (ncols * dx) / (nrows * dy)
  ncols[which.min(abs(log(grid_ratio / aspect)))]
}
```

Given `n` icons and a target width:height `aspect` ratio (default 16:9),
this brute-forces every possible number of columns from 1 to `n` and picks
whichever makes the resulting grid's bounding box closest to that aspect
ratio — so the figure fills a slide with minimal leftover margin, without
you having to hand-pick `ncol` for every `n`. It compares ratios on a *log*
scale, so "twice too wide" and "twice too tall" count as equally bad,
rather than the search being biased toward wide grids.

### Putting it together: `person_grid()`

This is the function you actually call. It:

1. Resolves `ncol` (via `choose_ncol()` if left as `"auto"`).
2. Samples silhouette/age/disability composition via `sample_counts()`.
3. Resolves fill colour — either sampled skin tones (`mode = "diversity"`)
   or `diverging_colors(value)` (`mode = "value"`).
4. Loops over `1:n`, computing each icon's grid row/column (and hence
   `ox`/`oy`), calling `person_parts()` for its shape (children are drawn
   at `scale = 0.65`), and stamping on that icon's fill colour.
5. Row-binds every icon's polygons into one data frame and builds a single
   `ggplot()` call — one `geom_polygon()` for the whole grid, not `n`
   separate layers, grouped by a per-icon `poly_id`.
6. Also returns `meta` (a tidy one-row-per-icon summary) and `alt_text` (a
   plain-language description of the composition) — see the integration
   section below for how `alt_text` feeds `fig-alt`.

```{r}
#| label: demo-person-grid-diversity
#| fig-cap: "person_grid() in diversity mode"
#| fig-width: 7
#| fig-height: 3
diversity_demo <- person_grid(n = 24, seed = 11)
diversity_demo$plot
```

```{r}
#| label: demo-person-grid-value
#| fig-cap: "person_grid() in value mode, coloured by a simulated effect size"
#| fig-width: 7
#| fig-height: 3
set.seed(2716)
value_demo <- person_grid(n = 24, mode = "value", value = rnorm(24, mean = 0, sd = 1.2))
value_demo$plot
```

## Quarto usage / integration

Minimal setup-chunk line:

```r
source("person-grid.R")
```

Real usage from `doing-being-science.qmd` (the diversity-sample slide):

````{verbatim}
```{r}
#| label: fig-diversity-sample
#| fig-cap: "The people in any group will vary in many ways: just look around you"
#| fig-alt: !expr diversity_grid$alt_text
#| fig-width: 8
#| fig-height: 4.5

diversity_grid <- person_grid(n = 125, seed = 11)
diversity_grid$plot
```
````

And the value-mode / effect-size variant:

````{verbatim}
```{r}
#| label: fig-effect-grid-demo
#| fig-cap: "People vary and how conditions or interventions affect us will vary: from strongly negative (dark blue) to null (near-white) to strongly positive (dark red) impacts"
#| fig-alt: !expr effect_grid$alt_text
#| fig-width: 8
#| fig-height: 4.5

set.seed(2716)
effect_grid <- person_grid(n = 60, mode = "value", value = rnorm(60, mean = 0, sd = 1.2))
effect_grid$plot
```
````

Two things worth keeping in mind when reusing this pattern:

- `fig-alt` is set via `!expr` to pull the *already-computed*
  `alt_text` field out of the object returned by `person_grid()` — so the
  alt text always matches what was actually sampled, rather than being
  hand-written and liable to drift out of sync.
- Because `coord_equal(expand = FALSE)` is used internally, the figure's
  panel is exactly the icons' bounding box with no extra margin — set
  `fig-width`/`fig-height` to roughly match the grid's aspect ratio so
  Reveal/HTML doesn't stretch it oddly.

### Parameter quick reference

| Parameter | Default | Meaning |
|---|---|---|
| `n` | `30` | Number of icons. |
| `ncol` | `"auto"` | Icons per row; `"auto"` picks the best fit to `aspect`. |
| `aspect` | `16/9` | Target width:height ratio used only when `ncol = "auto"`. |
| `comp_kind` | `c(A=.45, B=.45, N=.10)` | Silhouette mix (trouser / skirt / half-half). |
| `comp_age` | `c(adult=.70, child=.15, elder=.15)` | Age-presentation mix. |
| `disabled_prop` | `0.12` | Proportion drawn with the wheelchair silhouette. |
| `mode` | `"diversity"` | `"diversity"` (sampled skin tone) or `"value"` (diverging fill from `value`). |
| `skin_tones` | 6 hex colours | Palette sampled from when `mode = "diversity"`. |
| `value` | `NULL` | Numeric vector, length `n`; required when `mode = "value"`. |
| `seed` | `1` | Reproducibility seed for every sampling step. |
| `dx`, `dy` | `1.15, 2.15` | Icon spacing (horizontal, vertical). |

### Common errors

| Message | Likely cause |
|---|---|
| ``props` must have non-negative values, with at least one positive` | A composition argument (`comp_kind`, `comp_age`, or the derived `disabled_prop` pair) summed to zero — every category was set to 0. At least one must be positive. |
| ``value` must be supplied and length n when mode = 'value'` | `mode = "value"` was set but `value` was left `NULL`, or its length didn't match `n`. |
| Icons overlapping or clipped at the edges | `dx`/`dy` too small for the chosen `ncol`, or `aspect` far from the actual figure's `fig-width`/`fig-height` ratio. |

## Exercises

1. Change `comp_kind` in `person_grid(n = 24, seed = 11)` to
   `c(A = 0.1, B = 0.1, N = 0.8)` and re-render. What changes, and what
   stays the same?
2. Set `disabled_prop = 0` and re-render. Confirm it no longer errors
   (per the `sample_counts()` discussion above) and check the `alt_text`
   updates to match.
3. Re-run the value-mode demo with `value = rnorm(24, mean = 2, sd = 1.2)`
   (note the shifted mean) instead of `mean = 0`. Does `-2` still look as
   "strongly negative" as it did before? Explain why, referring to how
   `diverging_colors()` picks its scale.
4. Try `ncol = 4` instead of `"auto"` on a grid of `n = 24` and compare
   the result to the auto-chosen layout. When might you want to override
   the automatic choice?
5. Write a one-paragraph `fig-alt` description by hand for a `mode =
   "value"` grid, then compare it to the `alt_text` the function
   generates automatically. What would you add or change?

## See also

- [testing-sequence-walkthrough.qmd](testing-sequence-walkthrough.qmd) —
  reuses `person_parts()`, `sample_counts()`, and `diverging_colors()` to
  build a "person at a computer" icon.
- [peer-review-workflow-walkthrough.qmd](peer-review-workflow-walkthrough.qmd) —
  reuses the composite icon from `testing-sequence.R` (and so, transitively,
  this script) for the author/editor/reviewer roles.
