---
title: "`psych-battery-dgp.R`: a four-test psychological battery simulator"
subtitle: "Walkthrough: how the script works, and how to use it in Quarto"
format:
  html:
    code-fold: false
    toc: true
bibliography: references.bib
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(fig.align = "center")
```

## Aims

`psych-battery-dgp.R` is the shared infrastructure every other script in
this set is built on. It simulates a battery of four correlated
psychological tests — StuVoc1, StuVoc2, GK, ART3, grounded in @vermeiren2023
— and exposes exactly three independently controllable "dials", because
those three dials are what the surrounding teaching scripts each take turns
holding fixed and varying:

- **(i) participant sample size** (`n`) — how many people are in a study;
- **(ii) measurement error**, via each test's **reliability** — how noisy a
  single person's score is;
- **(iii) the *true* (construct-level) correlation** between two tests'
  underlying abilities — what the correlation *would* be if both tests were
  measured without error; together with (ii), this determines how strong a
  correlation the *observed* scores actually show.

On top of the simulator sits a generic diagnostic/sweep engine
(`run_sweep()`, `summarize_sweep()`) that repeats a correlation test many
times over a grid of these dials and computes **power, the empirical false
positive (Type I) rate, sign consistency (Type S), the Type M "exaggeration
ratio" [@gelman2014], and confidence-interval coverage** — the same
diagnostic quartet used in this project's heavier Aims 1–2 `glmer`
simulation work, here scaled down to run inline rather than as a background
job.

The simulator isn't just plausible — it's checked against real numbers.
Simulated at its default settings, it reproduces @vermeiren2023's own
reported test statistics (means, reliabilities, inter-test correlations)
closely, and a quick power sweep reproduces the power values the paper
itself reports for a true r = .2 correlation at N = 40 and N = 200 (see the
[Validation](#validation-against-vermeiren-et-al-2023) section).

After working through it, you should be able to:

- explain why reliability, true correlation, and observed correlation
  can't be set independently of one another, and what `build_sigma_true()`
  / `disattenuate()` do about that;
- generate a single study (`simulate_battery()`), a two-test focal pair
  (`simulate_pair()`), or a sweep over sample size/reliability/true
  correlation (`run_sweep()`), and read off power, Type I rate, sign
  consistency, the Type M exaggeration ratio, and coverage from the result;
- explain why `simulate_pair()` exists as a separate function rather than
  always drawing from the full four-test correlation matrix.

## Prerequisites / assumptions

- **Packages**: `ggplot2`; `MASS`, called via `::` only (so it never masks
  `dplyr::select()` if `dplyr` happens to be attached) — same convention as
  this project's Aims 1–2 code; base R `stats` for `uniroot()`,
  `integrate()`, and `cor.test()`.
- **Sourcing**: standalone — no other script needs to be sourced first.
  This script is, in turn, a prerequisite for `sampling-measurement-error.R`,
  `nhst-power.R`, and `error-rates-process-control.R`.
- **Reader background**: classical test theory terms (reliability, true
  score, attenuation) and basic `ggplot2`/`cor.test()` familiarity are
  assumed.

```{r}
#| label: source-script
source("psych-battery-dgp.R")
```

## How it works: the process

```{r}
#| label: model-equation
#| echo: false
#| results: asis
cat(
  "$$",
  "\\text{Standardized true ability (correlated across tests):} \\quad z_i \\sim \\text{MVN}(0, \\Sigma_{\\text{true}})",
  "$$",
  "$$",
  "\\text{Per test } t \\text{, calibrated ability:} \\quad \\theta_{ti} = \\theta_{\\mu,t} + \\theta_{\\sigma,t} \\, z_{it}",
  "$$",
  "$$",
  "\\text{Per-item probability:} \\quad p_{ti} = \\text{guessing}_t + (1 - \\text{guessing}_t) \\cdot \\text{plogis}(\\theta_{ti})",
  "$$",
  "$$",
  "\\text{4AFC tests (StuVoc1, StuVoc2, GK):} \\quad \\text{score}_{ti} \\sim \\text{Binomial}(n_{\\text{items},t}, p_{ti})",
  "$$",
  "$$",
  "\\text{ART3:} \\quad \\text{hits}_i \\sim \\text{Binomial}(60, p_{\\text{ART3},i}), \\quad \\text{rejects}_i \\sim \\text{Binomial}(30, p_{\\text{ART3},i})",
  "$$",
  sep = "\n"
)
```

`reliability` and `Σ_true` are never set on the observed-score scale
directly — the whole script exists to make that translation automatic:

```{mermaid}
flowchart LR
    A["default_test_specs()\ndefault_reliability_vec()\ndefault_r_obs_matrix()\n(Vermeiren-grounded defaults)"] --> B["build_sigma_true()\ndisattenuate observed r\n-> true-score correlation matrix"]
    B --> C["simulate_battery()\nz ~ MVN(0, Sigma_true)\nper-test calibrate + draw scores"]
    B --> D["simulate_pair()\ntwo-test focal-pair version,\navoids 4x4 PD issues in a sweep"]
    D --> E["fit_focal_cor()\ncor.test(var_x, var_y)"]
    E --> F["diagnose_cor_result()\nsignificant? sign-correct? covers?"]
    F --> G["run_sweep()\nrepeat over n x true_corr x reliability x reps"]
    G --> H["summarize_sweep()\npower/Type I, sign consistency, coverage"]
    H --> I["plot_metrics_vs_x() /\nplot_power_heatmap()"]
```

The central design idea, same philosophy as `sampling-distributions.R`'s
`calibrate_ability()`, generalized to four tests and two item-response
mechanisms: **you set the things that matter pedagogically — a test's mean
score, its reliability, and the *observed* correlation you want two tests
to show — and the script solves numerically for the underlying ability
parameters that reproduce them**, rather than requiring you to work out the
ability-scale numbers yourself.

## Element-by-element walkthrough

### Vermeiren-grounded defaults: `default_test_specs()`, `default_reliability_vec()`, `default_r_obs_matrix()`

```{r}
#| label: demo-defaults
default_test_specs()
default_reliability_vec()
default_r_obs_matrix()
```

Every number here is traceable to a specific table or figure in
@vermeiren2023's Study 5: item counts and means from their Table 6
(StuVoc1/StuVoc2/GK) and their ART3 figure, reliabilities (coefficient
alpha) from the same table, and the observed-score correlation matrix from
their Tables 7–8. `guessing = 0.25` on the three 4AFC tests reflects a
one-in-four chance of a correct guess; `guessing = 0` on ART3 reflects that
it's a yes/no recognition task, not a forced choice among four options.

### Calibration machinery: `expect_theta()`, `implied_moments_general()`, `calibrate_ability_general()`, `get_calibration()`

```{r}
#| label: show-implied-moments-general
#| eval: false
implied_moments_general <- function(theta_mu, theta_sigma, n_items, guessing = 0) {
  p_fun  <- function(th) guessing + (1 - guessing) * plogis(th)
  p_bar  <- expect_theta(p_fun, theta_mu, theta_sigma)
  p2_bar <- expect_theta(function(th) p_fun(th)^2, theta_mu, theta_sigma)
  pq_bar <- expect_theta(function(th) { p <- p_fun(th); p * (1 - p) }, theta_mu, theta_sigma)

  mean_score  <- n_items * p_bar
  var_true    <- n_items^2 * (p2_bar - p_bar^2)
  var_error   <- n_items * pq_bar
  reliability <- var_true / (var_true + var_error)
  # ...
}
```

This generalizes `sampling-distributions.R`'s `implied_moments()` with one
change: a `guessing` floor in `p_fun()`, which collapses back to the
original model exactly when `guessing = 0` (ART3's case). Given a candidate
`(theta_mu, theta_sigma)`, it works out the mean score, the between-person
("true") variance, the within-person (measurement-error) variance, and
their ratio — reliability — that choice implies.

`calibrate_ability_general()` runs this backward: a nested `uniroot()`
search (outer over `theta_sigma`, inner over `theta_mu`) for the parameters
whose `implied_moments_general()` output matches a target `mu` and
`reliability`, with the outer search bracket doubled automatically if the
initial range doesn't contain a root.

::: {.callout-note title="A `stop()` and a `warning()`, not just one or the other"}
Two different failure modes are handled two different ways. If no bracket
up to a generous ceiling contains a root at all (e.g. `reliability = 0.999`
requested for a 50-item test), that's a genuinely unachievable target and
`calibrate_ability_general()` `stop()`s with a message explaining why. But
`uniroot()`/`integrate()` are only numerically exact up to their own
tolerance, so even a *successful* search can land slightly off target —
that's checked separately against a `tolerance` argument (default `0.01`)
and only `warning()`s, since the calibration did still return a usable
result.
:::

```{r}
#| label: demo-calibrate-ability-general
cal <- calibrate_ability_general(mu = 32.3, reliability = 0.88, n_items = 50, guessing = 0.25)
cal$moments
```

`get_calibration()` wraps this in a simple cache, keyed on the rounded
`(mu, reliability, n_items, guessing)` arguments — because a sweep calls it
with the *same* arguments many times over (once per replicate within a
grid cell), and re-running `uniroot()`/`integrate()` every time would be
needless, measurable overhead at sweep scale.

### Disattenuation: `disattenuate()` and `build_sigma_true()`

```{r}
#| label: show-disattenuate
disattenuate
```

This is the single formula the whole "three dials" design rests on:
classical test theory says $r_{\text{true}} = r_{\text{observed}} /
\sqrt{\text{reliability}_x \cdot \text{reliability}_y}$ — usually used to
*estimate* a disattenuated correlation from data. Here it's run in the
*opposite* direction: given a target *observed*-score correlation (one of
the Vermeiren-grounded defaults, or an override), solve for the
*true*-score correlation that, once measurement error is added back in via
the per-test calibration above, will reproduce it.

`build_sigma_true()` applies this to every off-diagonal cell of a target
observed-score correlation matrix to build the full 4×4 matrix that's
actually fed to the multivariate draw:

```{r}
#| label: show-build-sigma-true-guard
#| eval: false
ev <- eigen(Sigma, symmetric = TRUE)
if (any(ev$values < 1e-8)) {
  warning("Sigma_true was not positive-definite after disattenuation; nudging to the nearest valid correlation matrix (eigenvalue clipping).")
  vals <- pmax(ev$values, 1e-6)
  adj <- ev$vectors %*% diag(vals) %*% t(ev$vectors)
  d <- sqrt(diag(adj))
  Sigma <- adj / outer(d, d)
  dimnames(Sigma) <- list(nm, nm)
}
```

::: {.callout-note title="Why a 4x4 matrix can stop being positive-definite"}
Disattenuating each pairwise correlation independently doesn't guarantee
the resulting matrix is still a valid correlation matrix (positive
semi-definite) — pushing even one cell to an extreme value (as a sweep
deliberately does) can make the implied matrix mathematically
inconsistent as a whole. Rather than erroring out, `build_sigma_true()`
nudges it to the nearest valid matrix by clipping negative eigenvalues to
a small positive floor and rescaling back to a correlation matrix, with a
`warning()` so this correction isn't silent.
:::

```{r}
#| label: demo-build-sigma-true
S <- build_sigma_true(default_r_obs_matrix(), default_reliability_vec())
round(S, 2)
```

### Simulating one study: `simulate_battery()`

```{r}
#| label: show-simulate-battery-core
#| eval: false
z <- MASS::mvrnorm(n, mu = rep(0, 4), Sigma = Sigma_true)
colnames(z) <- names(specs)

for (t in names(specs)) {
  sp  <- specs[[t]]
  cal <- get_calibration(sp$mu, rel[[t]], sp$n_items, sp$guessing)
  theta <- cal$theta_mu + cal$theta_sigma * z[, t]
  p <- sp$guessing + (1 - sp$guessing) * plogis(theta)
  # 4AFC: out[[t]] <- rbinom(n, sp$n_items, p)
  # ART3: hits/rejects drawn separately, plus fp/fn rate columns
}
```

One `MASS::mvrnorm()` draw gives every simulated participant's correlated
true abilities across all four tests at once; the loop then applies each
test's own calibrated item-response mechanism. This is the function to
reach for when you want **one realistic dataset** — every test, Vermeiren-
default correlations and reliabilities unless overridden:

```{r}
#| label: demo-simulate-battery
d <- simulate_battery(n = 2000, seed = 1)
str(d)
round(cor(d[c("StuVoc1", "StuVoc2", "GK", "ART3")]), 2)
```

### Simulating a focal pair: `simulate_pair()`

`run_sweep()` doesn't use `simulate_battery()` — it uses a leaner sibling,
`simulate_pair()`, that draws only the two tests under test:

```{r}
#| label: demo-simulate-pair
d2 <- simulate_pair(n = 300, var_x = "StuVoc1", var_y = "GK",
                     true_corr = 0.2, corr_scale = "observed", seed = 2)
cor(d2$StuVoc1, d2$GK)
```

::: {.callout-tip title="Why not just reuse the full 4x4 matrix for a sweep?"}
A sweep deliberately pushes one cell of the correlation matrix to extreme
values (including 0, to measure the false-positive rate). Doing that
inside the full 4×4 `Sigma_true` risks exactly the positive-definiteness
problem `build_sigma_true()`'s eigenvalue-clipping guard exists to patch —
and patching it silently changes *other* tests' correlations too, which
would be a confusing side effect in a sweep meant to vary only one pair.
`simulate_pair()` sidesteps this by only ever building a 2×2 correlation
matrix for the two tests actually being tested, which stays
positive-definite for any correlation in `(-1, 1)`.
:::

`corr_scale` controls what `true_corr` means: `"observed"` (the default)
treats it as the *observed*-score correlation and disattenuates
internally — the right choice for NHST/power work, where the population
parameter being tested lives on the observed scale (used throughout
`run_sweep()`, below). `"latent"` treats it as the *true*-score correlation
directly, with no compensation — used by `sampling-measurement-error.R`'s
attenuation figure, where the point is to hold the construct-level
relationship fixed and watch the *observed* correlation respond to
reliability.

### Repeated studies: `simulate_studies_battery()`

```{r}
#| label: demo-simulate-studies-battery
studies <- simulate_studies_battery(n = 40, n_studies = 4, seed = 3)
tapply(studies$StuVoc1, studies$study, mean)
```

Thin wrapper: call `simulate_battery()` once per study (without resetting
the seed between calls, so each study differs but the whole sequence is
reproducible from one top-level `seed`), row-binding the results with a
`study` label — same pattern as `sampling-distributions.R`'s
`simulate_studies()`.

### NHST on one dataset: `fit_focal_cor()` and `diagnose_cor_result()`

```{r}
#| label: demo-fit-diagnose
fit <- fit_focal_cor(d2, "StuVoc1", "GK")
fit
diagnose_cor_result(fit, true_value = 0.2)
```

`fit_focal_cor()` is a thin wrapper around `cor.test()`, returning just the
estimate, CI, and p-value as a one-row data frame. `diagnose_cor_result()`
is where the *teaching* content lives: given the **true** value used to
simulate the data (something a real researcher never gets to know),
classify the fitted result as significant (CI/p excludes zero),
sign-correct (only meaningful when the true value isn't zero — there's no
"correct" nonzero sign to match under the null, hence `NA`), whether the CI
covers the truth, and `inflation_ratio` — `|estimate| / |true_value|`, the
Type M "how exaggerated is this estimate" ratio [@gelman2014], also `NA`
under the null for the same reason `sign_correct` is (dividing by a true
value of zero is undefined, and "exaggeration relative to zero" isn't a
meaningful question anyway).

### The sweep engine: `run_sweep()` and `summarize_sweep()`

This is the script's most distinctive piece of code — the one place the
"three dials, four diagnostics" design is all visible at once — so it's
worth reading in full rather than trimmed:

```{r}
#| label: show-run-sweep-full
run_sweep
```

::: {.callout-note title="Deterministic per-cell seeding"}
`seed_cell <- seed_base + rep_i * 1e6 + n * 1e3 + round(reliability * 100) + round((true_corr + 1) * 1000)`
gives every `(n, true_corr, reliability, rep)` combination its own fixed
seed, derived arithmetically rather than drawn once and reused — the same
scheme used by this project's Aims 1–2 background-job simulations. The
practical payoff: any single cell of a sweep can be reproduced in
isolation (useful for debugging one surprising result) without re-running
the whole sweep, and the whole sweep is reproducible end to end from one
`seed_base`.
:::

```{r}
#| label: demo-run-sweep
sweep <- run_sweep(n_grid = c(40, 200), true_corr_grid = c(0, 0.2, 0.67),
                    reps = 200, var_x = "StuVoc1", var_y = "GK", seed_base = 777)
summ <- summarize_sweep(sweep)
summ
```

Reading this table: at `true_corr = 0` (the null), `power_or_type1` is the
empirical **false positive rate** — it should sit near `alpha` (.05).
Everywhere else, it's **power**. `sign_consistency` and `type_m_ratio` are
both `NA` at `true_corr = 0` (no correct sign to match, and no meaningful
"exaggeration relative to zero") and otherwise computed only among
*significant* results: `sign_consistency` is the proportion that pointed
the right direction, `type_m_ratio` is the **median** `inflation_ratio`
among them (median rather than mean, since significance filtering
disproportionately selects the occasional large outlier estimate,
especially at small `n`). `coverage` is the proportion of 95% CIs that
actually contained the true value — should sit near .95 regardless of
`true_corr`.

### Figures: `plot_metrics_vs_x()` and `plot_power_heatmap()`

```{r}
#| label: demo-plot-metrics-vs-x
#| fig-width: 7
#| fig-height: 4.5
plot_metrics_vs_x(summ, x = "n", metrics = "power_or_type1", color_by = "true_corr")
```

```{r}
#| label: demo-plot-power-heatmap
#| fig-width: 6
#| fig-height: 3.5
plot_power_heatmap(summ, x = "n", y = "true_corr")
```

Both take `summarize_sweep()`'s output directly and are deliberately
generic — any 0–1-bounded metric column (`power_or_type1`,
`sign_consistency`, `coverage`) can be plotted against any swept dial,
which is what lets the same two functions serve `nhst-power.R` and
`error-rates-process-control.R` without modification.

::: {.callout-note title="Why `type_m_ratio` gets its own figure, not these two"}
`type_m_ratio` is deliberately left out of that list. Both
`plot_metrics_vs_x()` (a shared `c(0, 1)` y-axis across metrics) and
`plot_power_heatmap()` (a `c(0, 1)` fill scale, with cell labels formatted
as percentages) assume a 0–1 proportion — true of power, sign consistency,
and coverage, but not of an exaggeration ratio, which is 1 at no inflation
and can comfortably exceed 2–3 at low power. Plotting it through either
function would silently clip or flatten the part of the story that matters
most. `error-rates-process-control.R` instead gives it a dedicated figure,
`plot_type_m_curve()`, with its own appropriately scaled axis.
:::

## Quarto usage / integration

This script is standalone infrastructure — not itself wired into
`doing-being-science.qmd`, and not meant to be used alone for a finished
figure; `sampling-measurement-error.R`, `nhst-power.R`, and
`error-rates-process-control.R` each source it and build their own
figures/narrative on top. A typical setup chunk:

```r
source("psych-battery-dgp.R")
```

### Validation against Vermeiren et al. (2023) {#validation-against-vermeiren-et-al-2023}

```{r}
#| label: validation-moments
specs <- default_test_specs()
rel   <- default_reliability_vec()
validation <- do.call(rbind, lapply(names(specs), function(t) {
  sp  <- specs[[t]]
  cal <- get_calibration(sp$mu, rel[[t]], sp$n_items, sp$guessing)
  data.frame(test = t, target_mean = sp$mu, implied_mean = round(cal$moments$mean_score, 2),
             target_reliability = rel[[t]], implied_reliability = round(cal$moments$reliability, 4))
}))
validation
```

```{r}
#| label: validation-correlations
d_val <- simulate_battery(n = 5000, seed = 101)
cat("Simulated observed correlations (n = 5000):\n")
print(round(cor(d_val[c("StuVoc1", "StuVoc2", "GK", "ART3")]), 2))
cat("\nVermeiren et al. (2023) Study 5 targets:\n")
print(default_r_obs_matrix())
```

```{r}
#| label: validation-power-benchmark
bench <- run_sweep(n_grid = c(40, 200), true_corr_grid = c(0, 0.2),
                    reps = 300, var_x = "StuVoc1", var_y = "GK", seed_base = 777)
summarize_sweep(bench)[c("n", "true_corr", "power_or_type1", "coverage")]
```

@vermeiren2023's own power illustration (their "Power considerations"
section) reports power ≈ .23 at N=40 and ≈ .81 at N=200 for a true
correlation of r = .2 — the `true_corr = 0.2` rows above should land close
to those values (simulation noise from only 300 replicates per cell means
exact agreement isn't expected). The `true_corr = 0` rows give the
empirical false-positive rate, which should sit near the nominal .05.

### Parameter quick reference

| Function | Parameter | Default | Meaning |
|---|---|---|---|
| `simulate_battery()` | `n` | — | Sample size (dial i). |
| | `reliability` | `NULL` | Named override for one or more tests (dial ii); `NULL` uses Vermeiren defaults. |
| | `Sigma_true` | `NULL` | A 4×4 true-score correlation matrix (dial iii); `NULL` builds one from `default_r_obs_matrix()` + whichever reliability is in effect. |
| | `keep_latent` | `FALSE` | If `TRUE`, also return each test's `theta`/`p` columns. |
| `simulate_pair()` | `var_x`, `var_y` | — | The two tests to simulate. |
| | `true_corr` | `0` | Target correlation, on whichever scale `corr_scale` selects. |
| | `corr_scale` | `"observed"` | `"observed"`: disattenuate internally so `true_corr` is the *observed*-score correlation. `"latent"`: use `true_corr` directly as the true-score correlation. |
| `run_sweep()` | `n_grid`, `true_corr_grid`, `reliability_grid` | — / — / `NULL` | The three dials to sweep; `reliability_grid = NULL` holds reliability fixed at `var_x`'s default. |
| | `reps` | `200` | Replicate studies per grid cell. |
| | `alpha` | `0.05` | Significance threshold. |
| | `seed_base` | `1234` | Base seed for deterministic per-cell seeding. |
| *(calibration)* | `tolerance` | `0.01` | Max acceptable gap between requested and achieved reliability before `calibrate_ability_general()` warns. |

### Common errors

| Message | Likely cause |
|---|---|
| `reliability = ... is not achievable at mu = ...` | Requested reliability too close to 1 (or `mu` too close to the guessing floor/ceiling) for the given `n_items`/`guessing`. |
| Warning: `achieved reliability (...) differs from the requested target (...)` | Calibration succeeded but landed outside `tolerance` of the target — usually near an achievability extreme; treat downstream diagnostics with caution. |
| Warning: `Target observed correlation r = ... is not achievable at ...` | Requested observed correlation exceeds what's reachable given the two tests' reliabilities (`\|r_true\| > 1`); capped. |
| Warning: `Sigma_true was not positive-definite after disattenuation...` | A disattenuated correlation matrix wasn't a valid correlation matrix; nudged via eigenvalue clipping. |
| `` `r_obs_matrix` must be named on both dimensions `` | `build_sigma_true()` called with an unnamed matrix. |
| `` `Sigma_true` must be a 4x4 matrix named StuVoc1/StuVoc2/GK/ART3 `` | `simulate_battery()` called with a malformed/mislabeled `Sigma_true`. |
| `` `var_x`/`var_y` must each be one of StuVoc1, StuVoc2, GK, ART3 `` | Misspelled or unsupported test name passed to `simulate_pair()`. |
| `` `var_x` and `var_y` must be different tests `` | Same test name passed twice. |

## Exercises

1. Call `calibrate_ability_general(mu = 42.2, reliability = 0.995, n_items = 65, guessing = 0.25)`
   (GK, pushed to a near-unachievable reliability). Does it warn, error, or
   succeed silently? Compare `$moments$reliability` to the requested value.
2. Build `Sigma_true` from `default_r_obs_matrix()` but with the
   StuVoc1–ART3 cell manually set to `0.95` before calling
   `build_sigma_true()`. Does it warn? Inspect the returned matrix — did
   any *other* pair's correlation change as a side effect?
3. Use `simulate_pair()` with `corr_scale = "latent"` and `corr_scale =
   "observed"` for the same `true_corr = 0.5` on StuVoc1/GK. Why do the two
   calls produce different simulated correlations in `cor(d$StuVoc1,
   d$GK)`?
4. Run `run_sweep()` for the StuVoc1–ART3 pair (r ≈ .52) instead of
   StuVoc1–GK, at the same `n_grid`. At what `n` does power cross 80%, and
   how does that compare to the StuVoc1–GK benchmark above?
5. Re-run the validation power-benchmark chunk with `reps = 1000` instead
   of `300`. Does the gap between the simulated and Vermeiren-reported
   power values shrink, as you'd expect from reducing simulation noise?
6. Re-run `run_sweep()`/`summarize_sweep()` for StuVoc1–GK at
   `true_corr_grid = 0.2` (a weaker effect than the pair's actual ≈.67)
   across `n_grid = c(10, 20, 50, 100, 200)`. How large is `type_m_ratio`
   at `n = 10`, and how far has it fallen by `n = 200`? Compare to how
   `power_or_type1` moves over the same range.

## See also

- [sampling-distributions-walkthrough.qmd](sampling-distributions-walkthrough.qmd)
  — the single-test ancestor of this script's calibration approach
  (`calibrate_ability()`/`implied_moments()`, generalized here to
  `calibrate_ability_general()`/`implied_moments_general()` with a
  guessing floor).
- [sampling-measurement-error-walkthrough.qmd](sampling-measurement-error-walkthrough.qmd)
  — uses `get_calibration()` and `simulate_pair()` from this script to
  separate measurement error (dial ii) from sampling error (dial i).
- [nhst-power-walkthrough.qmd](nhst-power-walkthrough.qmd) — uses
  `run_sweep()` and `summarize_sweep()` from this script to walk through
  NHST and power for the StuVoc1–StuVoc2 and StuVoc1–GK pairs.
- [error-rates-process-control-walkthrough.qmd](error-rates-process-control-walkthrough.qmd)
  — uses the same sweep engine, now varying all three dials together, to
  frame power/Type I/sign-consistency/coverage as long-run Neyman–Pearson
  process-control error rates.

## References
