---
title: "`sampling-measurement-error.R`: measurement error vs. sampling error"
subtitle: "Walkthrough: how the script works, and how to use it in Quarto"
format:
  html:
    code-fold: false
    toc: true
bibliography: references.bib
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(fig.align = "center")
```

## Aims

`sampling-measurement-error.R` uses the four-test psychological battery
simulated by `psych-battery-dgp.R` (StuVoc1, StuVoc2, GK, ART3, grounded in
@vermeiren2023) to draw out a distinction that's easy to state and easy to
blur in practice: **measurement error** and **sampling error** are different
things, caused by different mechanisms, fixed by different means, and this
script gives each one its own pair of figures so they can't be confused.

- **Measurement error** is noise in *one person's* score, caused by a test
  having only finitely many items (or author/foil names) to sample responses
  from. It's controlled by a test's **reliability**. Lower reliability widens
  the scatter of a person's observed score around their true ability, and —
  the consequence that matters most for research — it **attenuates** any
  correlation computed from that noisy score, with no change in sample size
  at all.
- **Sampling error** is noise in *a study's estimate* (a sample mean, a
  sample correlation), caused by observing only `n` participants rather than
  the whole population. It's controlled by **sample size**. A smaller `n`
  widens the *spread* of a statistic's sampling distribution across repeated
  studies, without shifting where that distribution is *centred*.

Put the two together and you get the script's organizing claim, demonstrated
rather than just asserted: **changing N changes the spread of a sampling
distribution; changing reliability changes its centre (via attenuation), not
its spread.** `plot_sampling_distribution()` and
`plot_attenuation_vs_reliability()` are built as a deliberate pair so this
contrast is visible side by side rather than left implicit.

After working through it, you should be able to:

- explain, using this script's own figures, why "the correlation was weaker
  than expected" and "the study was underpowered" are not the same diagnosis,
  and which error source each one points to;
- generate a measurement-error figure (one test, one person at a time) and a
  sampling-error figure (one statistic, repeated studies) for a battery test
  pair of your choosing;
- predict, roughly, how much an observed correlation will be attenuated at a
  given reliability, using the classical test theory formula this script
  simulates directly.

## Prerequisites / assumptions

- **Packages**: `ggplot2`; `ggdist`, for `plot_sampling_distribution()` only
  (guarded the same way as `sampling-distributions.R`'s `plot_dotpile()` —
  loading it isn't required just to source the script).
- **Sourcing**: `psych-battery-dgp.R` must be sourced first — this script
  uses its `default_test_specs()`, `default_reliability_vec()`,
  `default_r_obs_matrix()`, `get_calibration()`, `simulate_pair()`, and
  `disattenuate()` directly rather than re-implementing any calibration
  machinery.
- **Reader background**: classical test theory terms (reliability, true
  score, attenuation) help but aren't required — defined as they come up.
  Basic `ggplot2` familiarity assumed. Prior reading of
  `sampling-distributions-walkthrough.qmd` is useful context (same
  trial-level-up simulation philosophy, single test instead of a battery)
  but not required.

```{r}
#| label: source-script
library(ggdist)
source("psych-battery-dgp.R")
source("sampling-measurement-error.R")
```

## How it works: the process

```{r}
#| label: model-equation
#| echo: false
#| results: asis
cat(
  "$$",
  "\\text{True ability (standardized):} \\quad z_i \\sim N(0, 1)",
  "$$",
  "$$",
  "\\text{Calibrated ability (logit scale):} \\quad \\theta_i = \\theta_\\mu + \\theta_\\sigma z_i",
  "$$",
  "$$",
  "\\text{Per-item probability:} \\quad p_i = \\text{guessing} + (1 - \\text{guessing}) \\cdot \\text{plogis}(\\theta_i)",
  "$$",
  "$$",
  "\\text{4AFC tests (StuVoc1, StuVoc2, GK):} \\quad \\text{score}_i \\sim \\text{Binomial}(n_{\\text{items}}, p_i)",
  "$$",
  sep = "\n"
)
```

Every figure in this script is built from the same single-test generator,
`simulate_single_test()` (or its two-draws-per-person sibling,
`simulate_parallel_forms_battery()`), which is itself a thin wrapper around
`psych-battery-dgp.R`'s `get_calibration()` — the numerical solver that
finds the `(theta_mu, theta_sigma)` pair reproducing a target mean score and
reliability, same device as `sampling-distributions.R`'s
`calibrate_ability()`, generalized to handle a 4AFC guessing floor and
ART3's author/foil split.

```{mermaid}
flowchart LR
    A["psych-battery-dgp.R\nget_calibration()\nsolves theta_mu, theta_sigma\nfor target mu + reliability"] --> B["simulate_single_test()\nz ~ N(0,1)  (true ability)\np = guessing + (1-guessing)*plogis(theta)\nscore ~ Binomial(n_items, p)"]
    B --> C["plot_true_vs_observed()\nscore vs true ability --\nvertical scatter = measurement error"]
    A --> D["simulate_parallel_forms_battery()\ntwo independent score draws\nfrom the same theta_i"]
    D --> E["plot_parallel_forms()\nform1 vs form2 --\nr approx reliability"]
    A --> F["simulate_sampling_distribution()\nrepeat across n_studies,\none N at a time"]
    F --> G["plot_sampling_distribution()\nfacet by N --\nspread shrinks as N grows"]
    A --> H["simulate_attenuation_sweep()\nsimulate_pair(corr_scale='latent'),\ntrue corr fixed, reliability varied"]
    H --> I["plot_attenuation_vs_reliability()\nobserved r vs reliability --\ncentre shifts (attenuation)"]
```

The two right-hand branches are the ones to pay attention to: `F`/`G` holds
reliability fixed and varies `n`, `H`/`I` holds `n` fixed and varies
reliability. Same shared calibration engine, two different "dials" turned
one at a time.

## Element-by-element walkthrough

### Measurement error, one test at a time: `simulate_single_test()` and `plot_true_vs_observed()`

```{r}
#| label: show-simulate-single-test
#| eval: false
sp  <- specs[[test]]
cal <- get_calibration(sp$mu, rel, sp$n_items, sp$guessing)
z     <- rnorm(n)
theta <- cal$theta_mu + cal$theta_sigma * z
p     <- sp$guessing + (1 - sp$guessing) * plogis(theta)
score <- rbinom(n, sp$n_items, p)   # ART3 branches into hits + rejects instead
```

`z` is each simulated person's **true ability**, standardized to mean 0 / SD
1 — something a real researcher never observes directly. `score` is what a
real researcher *does* observe, and the gap between the two is measurement
error. Because `score` is drawn from a `Binomial`, not computed as a smooth
function of `z`, that gap is baked into the simulation structurally, not
added afterwards as noise.

```{r}
#| label: demo-true-vs-observed
#| fig-width: 7
#| fig-height: 4.2
plot_true_vs_observed("StuVoc1", n = 300, seed = 1)
```

The red curve is the *expected* score at each ability level (`n_items *
p`); every point's vertical distance from it is that person's measurement
error for this one administration. Lower the `reliability` argument and the
same plot shows a visibly fatter scatter around an unchanged curve — spread
increases, but the curve (and hence the relationship between true ability
and expected score) doesn't move.

### Parallel forms: `simulate_parallel_forms_battery()` and `plot_parallel_forms()`

```{r}
#| label: show-parallel-forms
#| eval: false
draw_once <- function() rbinom(n, sp$n_items, p)   # same p both times
data.frame(id = seq_len(n), test = test, true_ability = z,
           form1 = draw_once(), form2 = draw_once(), reliability = rel)
```

Two independent score draws from the *same* `theta_i` (and hence the same
`p`) is the textbook definition of reliability: it's the correlation you'd
see if you could test the same person twice under identical conditions.
That's exactly what `cor(form1, form2)` recovers:

```{r}
#| label: demo-parallel-forms
#| fig-width: 6
#| fig-height: 5
plot_parallel_forms("StuVoc1", n = 500, seed = 2)
d_pf <- simulate_parallel_forms_battery("StuVoc1", n = 20000, seed = 22)
cor(d_pf$form1, d_pf$form2)
```

::: {.callout-note title="Why this, and not StuVoc1 vs. StuVoc2?"}
`simulate_parallel_forms_battery()` re-administers the *same* test, not two
different tests. StuVoc1 and StuVoc2 are highly correlated (r ≈ .82 per
@vermeiren2023) but are not strictly parallel forms of one underlying test
— they're two related-but-distinct vocabulary subtests. For a clean
demonstration that reliability *is* parallel-forms correlation, this script
simulates two independent administrations of one test rather than reusing
StuVoc1/StuVoc2 for that purpose.
:::

### Sampling error across repeated studies: `simulate_sampling_distribution()` and `plot_sampling_distribution()`

```{r}
#| label: show-sampling-distribution
#| eval: false
values <- vapply(seq_len(n_studies), function(s) {
  d <- simulate_pair(n, var_x, var_y, true_corr = tc, corr_scale = "observed",
                      reliability_x = rel_x, reliability_y = rel_y, seed = NULL)
  cor(d[[var_x]], d[[var_y]])
}, numeric(1))
```

This is the sampling-distribution idea made concrete: simulate an entire
study — not one person — `n_studies` times, each time computing the same
statistic (a sample mean or a sample correlation), and look at how that
collection of per-study estimates is spread out. Reliability stays fixed
throughout; only `n` (passed in from `plot_sampling_distribution()`'s
`n_grid`) changes across panels.

```{r}
#| label: demo-sampling-distribution
#| fig-width: 7
#| fig-height: 6
plot_sampling_distribution(n_grid = c(30, 100, 400), n_studies = 300, statistic = "cor",
                            var_x = "StuVoc1", var_y = "GK", seed = 4)
```

All three panels are centred at roughly the same value (the population
StuVoc1–GK correlation, ≈ .67) because reliability — and therefore
attenuation — hasn't changed. Only the *spread* shrinks as `n` climbs from
30 to 400, which is sampling error shrinking, not measurement error.

### Attenuation: `simulate_attenuation_sweep()` and `plot_attenuation_vs_reliability()`

This is the script's most distinctive piece of code, and the one most worth
reading in full rather than trimmed, because the single detail that makes it
work — `corr_scale = "latent"` — is easy to get backwards:

```{r}
#| label: show-attenuation-sweep-full
simulate_attenuation_sweep
```

::: {.callout-note title="`corr_scale = \"latent\"`, not `\"observed\"`"}
`simulate_pair()` (from `psych-battery-dgp.R`) supports two ways of
interpreting `true_corr`: `"observed"` treats it as the target *observed*-score
correlation and disattenuates internally to find the latent correlation that
will produce it — the right choice for the NHST/power work, where the
population parameter being tested lives on the observed scale. Here, the
opposite is wanted: a *fixed construct-level relationship*, with reliability
left free to vary and the *observed* correlation left free to respond to it.
That's `corr_scale = "latent"` — `true_corr_latent` is handed to
`simulate_pair()` untouched, and the whole point of the figure is to watch
the *observed* correlation move as reliability does.
:::

```{r}
#| label: demo-attenuation
#| fig-width: 7
#| fig-height: 4.5
plot_attenuation_vs_reliability(reliability_grid = seq(0.2, 0.9, by = 0.1),
                                 n = 300, reps = 300,
                                 var_x = "StuVoc1", var_y = "GK", seed = 5)
```

The blue line (simulated mean observed correlation, with a ribbon showing
the spread across the 300 replicate studies run at each reliability value)
tracks the dashed red theoretical line almost exactly: $r_{\text{obs}} =
r_{\text{true}} \times \text{reliability}$, the classical test theory
attenuation formula, here simulated rather than just asserted. The dotted
line marks the ceiling — the correlation you'd see if both tests were
measured with no error at all.

## Quarto usage / integration

This script is standalone — it isn't currently sourced into
`doing-being-science.qmd` (that deck's figure sequence is built from the
icon/number pairing in `person-grid.R` / `testing-sequence.R` /
`sampling-distributions.R`; this script is a separate, inferential-statistics
demonstration). A typical setup chunk:

```r
library(ggdist)
source("psych-battery-dgp.R")
source("sampling-measurement-error.R")
```

and then any of the four demo calls shown above. All four `plot_*()`
functions return a plain `ggplot` object, so they combine with `patchwork`
or further `ggplot2` layers like any other plot.

### Parameter quick reference

| Function | Parameter | Default | Meaning |
|---|---|---|---|
| `simulate_single_test()` / `plot_true_vs_observed()` | `test` | `"StuVoc1"` | Which battery test: `StuVoc1`, `StuVoc2`, `GK`, or `ART3`. |
| | `n` | — | Number of simulated participants. |
| | `reliability` | `NULL` | Dial (ii); `NULL` uses the test's Vermeiren-default value. |
| `simulate_parallel_forms_battery()` / `plot_parallel_forms()` | *(same as above)* | | Two independent score draws per person from the same true ability. |
| `simulate_sampling_distribution()` / `plot_sampling_distribution()` | `n` / `n_grid` | — | Dial (i); `plot_*()` takes a vector (`n_grid`) for a faceted comparison. |
| | `n_studies` | — | Independent studies simulated per `n`/panel. |
| | `statistic` | — | `"mean"` (one test) or `"cor"` (a pair, needs `var_y`). |
| | `true_corr` | `NULL` | Population *observed*-score correlation to simulate from; `NULL` uses the Vermeiren-default for the pair. |
| `simulate_attenuation_sweep()` / `plot_attenuation_vs_reliability()` | `reliability_grid` | — | Dial (ii), swept; applied equally to `var_x` and `var_y`. |
| | `n` | `300` | Participants per replicate study (held fixed across the sweep). |
| | `reps` | `200` | Replicate studies simulated per reliability value. |
| | `true_corr_latent` | `NULL` | The fixed construct-level correlation; `NULL` backs it out from the Vermeiren-default observed correlation and reliabilities via `disattenuate()`. |
| *(all functions)* | `seed` | `NULL` | Reproducibility. |

### Common errors

| Message | Likely cause |
|---|---|
| `` `test` must be one of StuVoc1, StuVoc2, GK, ART3 `` | Misspelled or unsupported test name passed to `simulate_single_test()` / `simulate_parallel_forms_battery()`. |
| `` `reliability` must be strictly between 0 and 1 `` | A reliability of 0, 1, or outside that range was supplied directly. |
| `` `var_y` must be supplied when statistic = 'cor' `` | `simulate_sampling_distribution(statistic = "cor", ...)` called without `var_y`. |
| `` `n_studies` must be at least 2 to show a distribution `` | `n_studies` too small to compute a spread. |
| `` `var_x` and `var_y` must be different tests `` | Same test name passed twice to `simulate_attenuation_sweep()`. |
| `` `reliability_grid` values must all be strictly between 0 and 1 `` | A sweep value of 0 or 1 (or outside that range) was included. |
| `` `reps` must be at least 2 `` | `reps` too small to compute a spread at each reliability value. |
| `Package 'ggdist' is required for plot_sampling_distribution()` | `ggdist` not installed; install with `install.packages("ggdist")`. |

## Exercises

1. Call `plot_true_vs_observed("StuVoc1", n = 300, reliability = 0.95, seed = 1)`
   and then the same call with `reliability = 0.5`. How does the scatter
   around the red curve change, and does the curve itself move?
2. Run `plot_parallel_forms("ART3", n = 500, seed = 2)` and confirm the
   observed `r(form1, form2)` lands close to ART3's default reliability
   (.90). Then try `reliability = 0.6` — does the point cloud still look
   like a straight line, just noisier, or does its shape change?
3. Use `simulate_sampling_distribution()` with `statistic = "mean"`,
   `var_x = "GK"`, at `n = 20` and `n = 500` (300 studies each). Compare
   `sd()` of the two `value` columns — is the ratio close to what you'd
   predict from $\sqrt{500/20}$?
4. Using `simulate_attenuation_sweep()` for StuVoc1 and ART3 (default
   `true_corr_latent`, backed out from their Vermeiren-default observed
   correlation of ≈ .52), find roughly what reliability would be needed
   for the *observed* correlation to drop below .20.
5. Predict, then check: if reliability is fixed at .95 but `n` shrinks to
   10, does `plot_sampling_distribution()` for the StuVoc1–GK
   correlation look wide or narrow? Explain in your own words why that's a
   sampling-error effect, not a measurement-error one, even though
   reliability is high.

## See also

- [sampling-distributions-walkthrough.qmd](sampling-distributions-walkthrough.qmd)
  — the single-test prior art this script generalizes: same trial-level-up
  simulation philosophy and the same `calibrate_ability()`/parallel-forms
  pattern, applied to one Shipley-style test rather than a four-test
  battery.
- [psych-battery-dgp-walkthrough.qmd](psych-battery-dgp-walkthrough.qmd) —
  the shared calibration/simulation engine this script is built on
  (`get_calibration()`, `simulate_pair()`, `disattenuate()`,
  `default_test_specs()`), including its own validation check against
  @vermeiren2023's reported test statistics.
- [nhst-power-walkthrough.qmd](nhst-power-walkthrough.qmd) — uses the same
  battery and the same `simulate_pair()`/disattenuation machinery to ask a
  different question: not "how spread out is this estimate" but "how often
  would a fixed significance-testing procedure detect a real effect, or be
  fooled by none."
- [error-rates-process-control-walkthrough.qmd](error-rates-process-control-walkthrough.qmd)
  — reuses this script's `simulate_attenuation_sweep()` device (hold the
  *latent* correlation fixed, not the observed one, when sweeping
  reliability) in its own `run_reliability_sweep()`, and explains a real
  bug that device was built to avoid.

## References
