---
title: "`sampling-distributions.R`: simulating and plotting test scores"
subtitle: "Walkthrough: how the script works, and how to use it in Quarto"
format:
  html:
    code-fold: false
    toc: true
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(fig.align = "center")
```

## Aims

`sampling-distributions.R` simulates scores on a hypothetical 0–40 item
vocabulary test, from the trial (item) level up, and draws the synthetic
data half of the sampling/measurement-error narrative: one dot → two dots
→ a "pile of dots" showing a sample's distribution → a histogram → a grid
of histograms (repeated studies) → a scatterplot against a continuous
covariate. It's the data-generating and plotting counterpart to
`testing-sequence.R`'s icon sequence (same narrative, numbers instead of
pictograms).

The simulation is deliberately built **trial-level up** rather than
generating a continuous score and rounding/clipping it: every simulated
person's score is the sum of 40 independent correct/incorrect item
responses, so it is *structurally* impossible for a simulated score to
fall outside `[0, 40]` — there's no clipping step anywhere in the script,
because none is needed.

After working through it, you should be able to:

- explain why `mu`, `reliability`/`sd_true`, and `n_items` jointly
  determine the simulation rather than being three free knobs;
- generate a sample, a repeated-studies grid, and a covariate-correlated
  sample of your own;
- predict, roughly, how a requested `r_with_x` will be attenuated at a
  given `reliability`.

## Prerequisites / assumptions

- **Packages**: `ggplot2`, and `ggdist` for the dot-pile figures
  (`plot_dotpile()`, `plot_dotpile_grid()`).
- **Sourcing**: standalone — no other script in this set needs to be
  sourced first.
- **Reader background**: some familiarity with classical test theory
  terms (reliability, true score) is helpful but not required — the key
  ideas are explained below. Basic `ggplot2` familiarity assumed.

```{r}
#| label: source-script
library(ggdist)
source("sampling-distributions.R")
```

## How it works: the process

```{r}
#| label: model-equation
#| echo: false
#| results: asis
cat(
  "$$",
  "\\text{Ability (logit scale):} \\quad \\theta_i \\sim N(\\theta_\\mu, \\theta_\\sigma)",
  "$$",
  "$$",
  "\\text{Per-item probability:} \\quad p_i = \\text{plogis}(\\theta_i)",
  "$$",
  "$$",
  "\\text{Observed score:} \\quad Y_i \\sim \\text{Binomial}(40, p_i) \\quad \\in \\{0, \\dots, 40\\}",
  "$$",
  sep = "\n"
)
```

A `Binomial(40, p_i)` draw is the vectorized equivalent of summing 40
independent Bernoulli (correct/incorrect) trials, so the model above
*is* the trial-level process described in the aims, just computed
efficiently rather than looped item-by-item.

```{mermaid}
flowchart LR
    A["calibrate_ability()\nsolve theta_mu, theta_sigma\nto hit target mu + reliability\n(or mu + sd_true)"] --> B["simulate_vocab_data()\ndraw theta_i per person,\np_i = plogis(theta_i),\nscore_i ~ Binomial(40, p_i)"]
    B --> C["plot_dots() / plot_dotpile() /\nplot_histogram() / plot_scatter()"]
    B --> D["simulate_studies()\nrepeat simulate_vocab_data()\nper independent 'study'"]
    D --> E["plot_histogram_grid() /\nplot_dotpile_grid()\n(facet_wrap(~study))"]
```

The central design idea: **you don't set the ability distribution's mean
and spread directly.** You set the things that matter pedagogically — the
test's mean score (`mu`) and its reliability (`reliability`, or
alternatively a target spread `sd_true`) — and `calibrate_ability()`
numerically solves for the underlying `theta_mu`/`theta_sigma` that
reproduce them, because for a fixed test length, mean/reliability/spread
are mathematically linked, not three independently free knobs.

## Element-by-element walkthrough

### Calibration machinery: `expect_theta()` and `implied_moments()`

```{r}
#| label: show-expect-theta
#| eval: false
expect_theta <- function(f, theta_mu, theta_sigma) {
  integrate(function(theta) f(theta) * dnorm(theta, theta_mu, theta_sigma),
            lower = theta_mu - 10 * theta_sigma,
            upper = theta_mu + 10 * theta_sigma)$value
}
```

`expect_theta()` computes $E[f(\theta)]$ for $\theta \sim N(\theta_\mu,
\theta_\sigma)$ by plain numerical integration (base R `integrate()`, no
extra dependency) — used to get exact expectations like $E[p_i]$ or
$E[p_i^2]$ without resorting to Monte Carlo.

```{r}
#| label: show-implied-moments
#| eval: false
implied_moments <- function(theta_mu, theta_sigma, n_items) {
  p_bar  <- expect_theta(plogis, theta_mu, theta_sigma)
  p2_bar <- expect_theta(function(th) plogis(th)^2, theta_mu, theta_sigma)
  pq_bar <- expect_theta(function(th) plogis(th) * (1 - plogis(th)), theta_mu, theta_sigma)

  mean_score  <- n_items * p_bar
  var_true    <- n_items^2 * (p2_bar - p_bar^2)
  var_error   <- n_items * pq_bar
  reliability <- var_true / (var_true + var_error)
  # ...
}
```

Given a candidate `(theta_mu, theta_sigma)`, this works out what mean
score, "true"-score variance (between-person spread), measurement-error
variance (within-person, item-sampling noise), and reliability
(`var_true / (var_true + var_error)`) that choice *implies* — i.e. it runs
the model forward analytically. `calibrate_ability()` then runs this
backward: searching for the `(theta_mu, theta_sigma)` whose
`implied_moments()` output matches your targets.

```{r}
#| label: demo-implied-moments
implied_moments(theta_mu = 0, theta_sigma = 1.3, n_items = 40)
```

### Solving for ability parameters: `calibrate_ability()`

```{r}
#| label: show-calibrate-ability-signature
#| eval: false
calibrate_ability <- function(mu, reliability = 0.80, sd_true = NULL, n_items = 40) {
  # ...
}
```

Two calibration targets are supported, and they're **mutually
exclusive** — for a fixed `mu` and `n_items`, fixing one determines the
other:

- `reliability` (the default path) — "how repeatable is this test"; use
  this when the figure is about test-retest/parallel-forms reliability.
- `sd_true` — "how spread out is this population"; use this when a figure
  needs independent control of between-person spread (e.g. two panels
  with the same mean but visibly different spread), supplied instead of
  `reliability`.

Internally, both paths use the same robust pattern: a nested `uniroot()`
search (an outer search over `theta_sigma`, and for each candidate, an
inner search over `theta_mu` so the mean score matches `mu` exactly), with
the search bracket doubled automatically if the initial range doesn't
contain a root, and an informative error (rather than a cryptic numerical
failure) if the target isn't achievable — e.g. `reliability` above about
0.99 would require an implausibly large, numerically unstable ability
spread for a 40-item test. `sd_true` additionally has a closed-form
feasibility check: the theoretical ceiling is `sqrt(mu * (n_items - mu))`,
approached only as the ability spread goes to infinity.

```{r}
#| label: demo-calibrate-ability
cal <- calibrate_ability(mu = 30, reliability = 0.80)
cal$moments
```

### Simulating one sample: `simulate_vocab_data()`

```{r}
#| label: show-simulate-vocab-data-core
#| eval: false
cal <- calibrate_ability(mu, reliability = reliability, sd_true = sd_true, n_items = n_items)
# ... correlation setup (see below) ...
theta <- cal$theta_mu + cal$theta_sigma * theta_z
p_correct <- plogis(theta)
observed_score <- rbinom(n, size = n_items, prob = p_correct)
```

After calibration, simulating is simple: draw standardized ability
`theta_z` per person (optionally correlated with a covariate — see next
section), scale/shift it to `theta`, convert to `p_correct` via
`plogis()`, and draw `observed_score` as a single vectorized
`rbinom(n, 40, p_correct)` call — bounded to `[0, 40]` by construction.

```{r}
#| label: demo-simulate-vocab-data
d <- simulate_vocab_data(n = 10, mu = 30, reliability = 0.80, seed = 1)
d
range(simulate_vocab_data(n = 10000, seed = 2)$observed_score)
```

### Inducing a covariate correlation: `r_with_x`

This is the part of the script most worth reading carefully if you're
adapting it, because of **attenuation**: any correlation involving a
noisy *observed* variable is weaker than the correlation involving the
underlying *true* ability, because measurement error adds noise that
dilutes the relationship. Classical test theory gives
$\text{corr}(\text{observed}, \theta) = \sqrt{\text{reliability}}$, so to
hit a target **observed**-score correlation `r_with_x`, the script works
backward through that relationship:

```{r}
#| label: show-attenuation-logic
#| eval: false
max_r <- sqrt(reliability)
if (abs(r_with_x) > max_r) {
  warning(...)  # cap r_with_x at the achievable ceiling
  r_with_x <- sign(r_with_x) * max_r
}
r_theta_x <- if (r_with_x == 0) 0 else r_with_x / max_r
# theta_z is then a mix of standardized x and independent noise,
# correlated with x at r_theta_x so that, after measurement error,
# the *observed* score ends up correlated with x at (approximately) r_with_x
```

::: {.callout-note title="r_with_x is on the observed scale, and has a ceiling"}
Two consequences follow from the attenuation logic above: (1) `r_with_x`
is specified on the *observed*-score scale — i.e. it's (approximately)
the correlation you'll actually see in `plot_scatter()`, not some larger
hidden value; (2) at low `reliability`, only a limited range of `r_with_x`
is achievable at all (capped at `sqrt(reliability)`), and the function
warns rather than silently failing if you ask for more than that.
:::

```{r}
#| label: demo-r-with-x
d_cor <- simulate_vocab_data(n = 2000, r_with_x = 0.5, seed = 3)
cor(d_cor$x, d_cor$observed_score)
```

One more detail is handled explicitly, and is worth seeing in code rather
than just prose, because it was a real bug caught during development:

```{r}
#| label: show-nan-guard
#| eval: false
z <- rnorm(n)
if (r_theta_x == 0) {
  theta_z <- z
} else {
  xz <- as.numeric(scale(x))
  if (anyNA(xz)) {
    warning("x has zero variance (e.g. all identical values); r_with_x cannot be induced. Ignoring r_with_x.")
    theta_z <- z
  } else {
    theta_z <- r_theta_x * xz + sqrt(1 - r_theta_x^2) * z
  }
}
```

::: {.callout-note title="A real bug this guard fixes"}
`scale(x)` is undefined (`NaN`) when `n < 2` or `x` has zero variance, and
`0 * NaN` is `NaN` in R — so multiplying it in unconditionally, even at
the default `r_with_x = 0`, would silently turn every `theta` (and hence
every score) into `NA`. This exact case is `n = 1`, i.e. the simplest
possible call. The fix is to skip the standardization path entirely
whenever `r_theta_x == 0`, rather than relying on the multiplication to
zero it out.
:::

### Repeated studies: `simulate_studies()` and `simulate_parallel_forms()`

```{r}
#| label: demo-simulate-studies
studies <- simulate_studies(n_studies = 3, n_per_study = 15, seed = 4)
tapply(studies$observed_score, studies$study, mean)
```

`simulate_studies()` just calls `simulate_vocab_data()` once per study
(without resetting the seed between calls, so each study's sample differs
but the whole sequence stays reproducible from one top-level `seed`),
row-binding the results with a `study` label — this is what feeds the
`_grid()` plotting functions below.

`simulate_parallel_forms()` is a different use case: it draws **one**
`theta_i` per person but **two** independent test administrations
(`form1`, `form2`) from that same `theta_i` — i.e. the same person tested
twice (or on two equivalent forms). This is how the script's own
reliability calibration can be checked empirically:

```{r}
#| label: demo-parallel-forms
pf <- simulate_parallel_forms(n = 20000, mu = 30, reliability = 0.80, seed = 5)
cor(pf$form1, pf$form2)
```

### Figures: `plot_dots()`, `plot_dotpile()`, `plot_histogram()`, `plot_scatter()`, and their `_grid()` variants

These are comparatively simple `ggplot2` wrappers; the interesting parts
are the choices, not the mechanics:

- `plot_dots()` — one or two scores as points on a number line, with `y`
  nudged just above 0 so points sit on the axis line rather than being
  bisected by it.
- `plot_dotpile()` — a sample's distribution as discrete points via
  `ggdist::stat_dots()`, by design **not** a smoothed density curve or a
  binned histogram, so the figure reads as "these are the actual
  individual scores", not an idealized shape.
- `plot_dotpile_grid()` — the faceted version; `binwidth` is fixed
  explicitly rather than left to `ggdist`'s per-panel default, because
  that default sizes each panel's dot diameter to that panel's own
  tallest stack — which would make dots visibly different sizes across
  panels, undermining a "same scale, different samples" comparison.
- `plot_histogram()` / `plot_histogram_grid()` — standard binned
  histograms, for the "distribution looks approximately normal" step.
- `plot_scatter()` — score against a covariate, with an optional
  `geom_smooth(method = "lm")` fitted line.

```{r}
#| label: demo-figures
#| fig-width: 7
#| fig-height: 2.2
plot_dots(d$observed_score[1])
```

```{r}
#| label: demo-dotpile
#| fig-width: 7
#| fig-height: 2.5
plot_dotpile(simulate_vocab_data(n = 30, seed = 6)$observed_score)
```

```{r}
#| label: demo-histogram-grid
#| fig-width: 8
#| fig-height: 4
plot_histogram_grid(simulate_studies(6, 40, seed = 7))
```

```{r}
#| label: demo-scatter
#| fig-width: 6
#| fig-height: 4
plot_scatter(simulate_vocab_data(n = 80, r_with_x = 0.5, seed = 8))
```

## Quarto usage / integration

Setup-chunk lines:

```r
library(ggdist)
source("sampling-distributions.R")
```

Real usage from `doing-being-science.qmd` — paired with
`testing-sequence.R`'s icon, sharing one `simulate_vocab_data()` draw so
the icon's score label and the plotted point match:

````{verbatim}
```{r}
#| label: fig-one-score-dot
#| fig-cap: "Their score, as a point on the number line"
#| fig-width: 5
#| fig-height: 3.2
#| echo: false

plot_dots(d_one$observed_score)
```
````

The repeated-studies pairing (icon grid + matching dotpile grid):

````{verbatim}
```{r}
#| label: fig-many-studies-dotpiles
#| fig-width: 8
#| fig-height: 5

plot_dotpile_grid(studies_sample, ncol = 3)
```
````

### Parameter quick reference (`simulate_vocab_data()`)

| Parameter | Default | Meaning |
|---|---|---|
| `n` | — | Sample size. |
| `mu` | `30` | Target mean observed score (0–40 scale). |
| `reliability` | `0.80` | Target parallel-forms reliability; ignored if `sd_true` given. |
| `sd_true` | `NULL` | Alternate calibration target: between-person "true" score SD. |
| `r_with_x` | `0` | Target correlation between the covariate `x` and the **observed** score. |
| `x`, `x_mean`, `x_sd` | `NULL, 0, 1` | Pre-specified covariate, or generated from `N(x_mean, x_sd)`. |
| `n_items` | `40` | Number of test items. |
| `seed` | `NULL` | Reproducibility. |

### Common errors

| Message | Likely cause |
|---|---|
| `reliability = ... is not achievable at mu = ...` | Requested `reliability` is too close to 1 for the given `mu`/`n_items` (implausibly large ability spread). |
| `sd_true = ... is not achievable at mu = ...` | Requested `sd_true` exceeds the analytic ceiling `sqrt(mu * (n_items - mu))`. |
| Warning: `r_with_x = ... is not achievable at reliability = ...` | Requested correlation exceeds the `sqrt(reliability)` ceiling; value was capped. |
| Warning: `x has zero variance...` | `x` is constant (or `n < 2`); `r_with_x` can't be induced and was ignored. |

## Exercises

1. Call `calibrate_ability(mu = 30, reliability = 0.95)` and then
   `calibrate_ability(mu = 30, reliability = 0.99)`. Compare the two
   `$moments$sd_true` values. What happens if you push `reliability`
   toward `0.999`, and why does the script's error message call that
   "implausibly large"?
2. Simulate two samples with the same `mu` and `n` but different
   `reliability` (e.g. `0.6` vs `0.95`), and plot both with
   `plot_dotpile()`. Which one's spread looks bigger, and does that
   match what `implied_moments()` predicts?
3. Request `r_with_x = 0.9` at `reliability = 0.5` via
   `simulate_vocab_data()`. Confirm it warns and is capped, and compute
   `sqrt(0.5)` by hand to check the cap matches.
4. Use `sd_true` instead of `reliability` to calibrate two samples with
   the *same* spread but different means (e.g. `mu = 20` vs `mu = 35`,
   both `sd_true = 5`). Plot both as histograms side by side — do they
   look like plausible vocabulary-test distributions at both means?
5. Generate 6 studies with `simulate_studies(6, 15, seed = 1)` (a small
   `n` per study) and compare `plot_histogram_grid()` against
   `plot_dotpile_grid()` for the same data. Which representation makes
   the sample-to-sample variability at small `n` easier to see, and why?

## See also

- [testing-sequence-walkthrough.qmd](testing-sequence-walkthrough.qmd) —
  the icon-based half of the same sampling/measurement-error narrative;
  several slides pair one figure from each script using a shared
  simulated dataset.
