---
title: "Tutorial: the Aims 1-2 synthetic-data simulation pipeline"
subtitle: "How the data behind the power / error-rate / replication figures is generated, walked through piece by piece"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: cosmo
    embed-resources: true
    code-fold: false
execute:
  warning: false
  message: false
---

::: {.callout-note}
## What this document is (and isn't)

This is a **teaching walk-through** of the simulation/analysis pipeline
behind `synthetic-data-analyses-power-error-replication-aims-1-2-draft.qmd`
(Aims 1-2 of the Phase 2 parametric simulation program). It extracts every
piece of code and commentary needed to understand **how the input data for
that document's tables and figures is produced** -- the data-generating
function, the model-fitting logic, the sample-size/effect-size grids, and
the per-cell orchestration that ties them together -- stopping exactly where
that data becomes ready to plot.

It deliberately does **not** cover: the actual visualization code (`ggplot`
calls, plotting helpers), the substantive findings/discussion, or the
small-scale smoke test used to validate the pipeline before the full run
(that smoke test *is* a nice minimal worked example if you want to run this
code yourself interactively -- see the closing note).

Code below is reproduced verbatim from the source document (function names,
defaults, and comments unchanged) so it can be cross-checked directly
against it.
:::

## The big picture

Aims 1-2 answer: *if we knew the true effect sizes and variance components
exactly, how would our ability to detect and correctly characterize three
specific effects (`zHLVA`, `zPCREFp`, their interaction) change with
participant sample size, text sample size, and the size of the effect
itself?* Answering this requires, for many different combinations of
"true" parameter values and sample sizes:

1. **Simulate** a synthetic dataset with those exact, known-true properties.
2. **Fit** the same statistical model a real analysis would use.
3. **Classify** the fit's outcome against the known truth (right sign? CI
   excludes zero? how far off is the estimate?).
4. **Repeat** many times (replicates) per combination, and repeat across
   every combination in the design.

The rest of this document walks through the code that does exactly that, in
the order a reader would need to understand it.

## Step 1: packages

```{r}
#| label: setup

library(tidyverse)
library(lme4)
```

Only two packages are needed for the *generation* side of the pipeline:
`tidyverse` (for `dplyr`/`tidyr` data wrangling and `purrr::map`/`pmap` for
iterating over grids) and `lme4` (`glmer`, for fitting the mixed-effects
logistic regression models). Everything downstream of this point works with
plain R and these two packages -- no modeling-specific niceties are needed
until the visualization stage, which is out of scope here.

## Step 2: the data-generating function

The core building block is a function that, given a set of "true" parameter
values and sample sizes, simulates one complete synthetic dataset that
looks like the real study's data structure: participants, each answering
multiple-choice questions about several texts, with a known true effect of
participant vocabulary (`zHLVA`) and text cohesion (`zPCREFp`), plus their
interaction, on accuracy.

```{r}
#| label: generate-data-p2

generate_data_p2 <- function(
    n_participants = 200,
    n_texts        = 20,
    k              = min(10, n_texts),
    questions_per_text = 4,
    b_intercept    = 2.093,
    b_hlva         = 0.363,
    b_pcref        = 0.095,
    b_interaction  = 0.054,
    sd_participant = 0.69,
    sd_text        = 0.41,
    sd_pcref_slope = 0,
    sd_hlva_slope  = 0,
    cor_part       = 0,
    cor_text       = 0,
    cohesion_design = c("natural", "widened"),
    reliability_hlva = 1,
    reliability_pcref = 1,
    seed           = NULL
) {
  cohesion_design <- match.arg(cohesion_design)
  if (!is.null(seed)) set.seed(seed)
  k <- min(k, n_texts)

  # Participant-level random effects: intercept + zPCREFp slope (correlated) -- Aim 5
  Sigma_part <- matrix(
    c(sd_participant^2, cor_part * sd_participant * sd_pcref_slope,
      cor_part * sd_participant * sd_pcref_slope, sd_pcref_slope^2),
    nrow = 2
  )
  re_part <- if (sd_pcref_slope > 0) {
    MASS::mvrnorm(n_participants, mu = c(0, 0), Sigma = Sigma_part)
  } else {
    cbind(rnorm(n_participants, 0, sd_participant), 0)
  }

  # zHLVA measurement reliability -- Aim 6 strategy E; reliability_hlva = 1
  # (default) reproduces the exact prior behavior used in Aims 1-5.
  zHLVA_true <- rnorm(n_participants, 0, 1)
  zHLVA_obs  <- sqrt(reliability_hlva) * zHLVA_true +
    sqrt(1 - reliability_hlva) * rnorm(n_participants, 0, 1)

  participants <- tibble(
    participant_id = 1:n_participants,
    zHLVA          = zHLVA_true,
    zHLVA_obs      = zHLVA_obs,
    u_participant  = re_part[, 1],
    s_pcref        = re_part[, 2]
  )

  # Text-level random effects: intercept + zHLVA slope (correlated) -- Aim 5
  Sigma_text <- matrix(
    c(sd_text^2, cor_text * sd_text * sd_hlva_slope,
      cor_text * sd_text * sd_hlva_slope, sd_hlva_slope^2),
    nrow = 2
  )
  re_text <- if (sd_hlva_slope > 0) {
    MASS::mvrnorm(n_texts, mu = c(0, 0), Sigma = Sigma_text)
  } else {
    cbind(rnorm(n_texts, 0, sd_text), 0)
  }

  # zPCREFp sampling design -- Aim 6 strategy D (widened cohesion range) --
  # and measurement reliability -- Aim 6 strategy E.
  zPCREFp_true <- if (cohesion_design == "widened") {
    rep(c(-1.5, 1.5), length.out = n_texts) + rnorm(n_texts, 0, 0.3)
  } else {
    rnorm(n_texts, 0, 1)
  }
  zPCREFp_obs <- sqrt(reliability_pcref) * zPCREFp_true +
    sqrt(1 - reliability_pcref) * rnorm(n_texts, 0, 1)

  texts <- tibble(
    text_id  = 1:n_texts,
    zPCREFp  = zPCREFp_true,
    zPCREFp_obs = zPCREFp_obs,
    u_text   = re_text[, 1],
    s_hlva   = re_text[, 2]
  )

  assignments <- tibble(participant_id = 1:n_participants) |>
    dplyr::mutate(assigned = purrr::map(participant_id,
                          ~ sample(1:n_texts, size = k, replace = FALSE))) |>
    tidyr::unnest(assigned) |>
    dplyr::rename(text_id = assigned)

  assignments |>
    dplyr::left_join(participants, by = "participant_id") |>
    dplyr::left_join(texts,        by = "text_id") |>
    dplyr::slice(rep(1:dplyr::n(), each = questions_per_text)) |>
    dplyr::mutate(
      question_id = rep(1:questions_per_text, times = dplyr::n() / questions_per_text),
      eta = b_intercept +
        (b_hlva  + s_hlva)  * zHLVA +
        (b_pcref + s_pcref) * zPCREFp +
        b_interaction * zHLVA * zPCREFp +
        u_participant + u_text,
      prob     = plogis(eta),
      response = rbinom(dplyr::n(), 1, prob)
    ) |>
    dplyr::mutate(zHLVA_fit = zHLVA_obs, zPCREFp_fit = zPCREFp_obs) |>
    dplyr::select(participant_id, text_id, question_id,
                  zHLVA = zHLVA_fit, zPCREFp = zPCREFp_fit, response)
}
```

Walking through this piece by piece:

**a. Arguments and their defaults (lines defining the signature).** Every
"true" quantity the simulation needs is a function argument with a default
equal to this project's real, empirically-estimated value (e.g.
`b_hlva = 0.363` is the real fitted `zHLVA` coefficient). This is
deliberate: calling `generate_data_p2()` with no arguments reproduces the
"realistic" scenario, and any Aim-specific sweep just overrides the one or
two arguments it cares about (e.g. Aim 1 overrides `b_hlva`/`b_pcref`/
`b_interaction`; Aim 6 overrides `cohesion_design`/`reliability_hlva`).
`sd_pcref_slope`, `sd_hlva_slope`, `cor_part`, `cor_text` default to zero --
they only matter for Aim 5's richer random-slopes design, so Aims 1-2 never
touch them and effectively get a simpler random-**intercepts**-only model.

**b. Participant-level random effects.** `Sigma_part` is a 2x2
variance-covariance matrix for a participant's random intercept and (Aim
5-only) random `zPCREFp` slope. When `sd_pcref_slope = 0` (Aims 1-2's case),
building this correlated bivariate draw would be wasted work and
numerically fragile (a rank-deficient covariance matrix), so the code
branches: if there's no slope variance, it just draws ordinary independent
normal random intercepts (`rnorm`) and hard-codes the slope column to zero.
`MASS::mvrnorm()` (correlated bivariate normal draws) is only reached when
Aim 5 actually asks for slope variation.

**c. `zHLVA` and measurement reliability.** Every participant gets a "true"
`zHLVA` value drawn from a standard normal. `zHLVA_obs` is a
reliability-attenuated version: `reliability_hlva = 1` (the Aims 1-2
default) makes `zHLVA_obs` identical to `zHLVA_true`; a value below 1 (used
only in Aim 6's measurement-reliability strategy) blends in independent
noise, simulating an imperfect vocabulary test.

**d. Assembling the `participants` tibble.** One row per participant,
carrying both the true and observed `zHLVA`, plus the random-intercept/slope
draws from (b).

**e-g. Mirror image for texts.** Steps (e) text-level random effects, (f)
`zPCREFp`'s true value (with an optional "widened" sampling design used only
by Aim 6, which spaces cohesion values at two extremes rather than sampling
them continuously) and its own reliability attenuation, and (g) assembling
the `texts` tibble, all repeat the same logic as (b)-(d) with participant
and text swapped.

**h. Crossing participants with texts.** Real participants don't read every
text -- each is assigned a random subset of `k` texts (default: up to 10,
clamped to however many texts actually exist in this run, since a run with
`n_texts = 2` can't give every participant 10 distinct texts). `purrr::map`
draws one random sample of text IDs per participant, and `tidyr::unnest()`
turns that list-column into one row per (participant, assigned text) pair.

**i. Expanding to the question level.** The real outcome (`response`) is
recorded per-question, not per-text (each text has multiple comprehension
questions). `dplyr::slice(rep(..., each = questions_per_text))` duplicates
every participant-text row `questions_per_text` (default 4) times, and a
repeating `1:questions_per_text` sequence labels which question each
duplicate represents.

**j. The actual data-generating step.** `eta` is the model's linear
predictor on the logit scale: an intercept, the `zHLVA` and `zPCREFp` main
effects (each augmented by that participant's or text's own random slope,
which is zero for Aims 1-2), the fixed interaction term, and the
participant's and text's random intercepts. `plogis(eta)` converts that to
a probability, and `rbinom(..., 1, prob)` draws the actual 0/1 `response`
-- this is the only genuinely random step tied to the *outcome* itself
(everything upstream generates covariates and random effects; this line is
where "did this participant get this question right" is actually decided).

**k. Final selection: true vs. observed covariates.** The very last
`mutate()`/`select()` swaps in the (possibly reliability-attenuated)
*observed* `zHLVA`/`zPCREFp` values as what the fitted model will see, while
`eta` was computed from the *true* values. When `reliability_* = 1` (Aims
1-2), these are identical and this distinction is invisible; the separation
exists so Aim 6's measurement-reliability strategy can reuse this same
function unmodified.

::: {.callout-tip}
**A small but important implementation note:** `library(MASS)` is
deliberately never attached in the source document (only `MASS::mvrnorm()`
is called, via `::`). This is because `MASS` also defines a `select()`
function that would silently shadow `dplyr::select()` if attached, breaking
the pipeline above the moment it's loaded after `tidyverse`. This is a
common `MASS`/`dplyr` gotcha worth knowing if you reuse this pattern
elsewhere.
:::

## Step 3: fitting the model and classifying the result

Once a dataset exists, it needs to be fit with the same kind of model a
real analysis would use, and the fit's outcome needs to be captured in a
structured way (including failure modes) so later steps can build
diagnostics from it.

```{r}
#| label: fit-model-p2

extract_terms_of_interest_p2 <- function(fit) {
  terms <- c("zHLVA", "zPCREFp", "zHLVA:zPCREFp")
  co <- tryCatch(summary(fit)$coefficients, error = function(e) NULL)
  if (is.null(co)) return(NULL)
  avail <- intersect(terms, rownames(co))
  tibble(term = avail, estimate = co[avail, "Estimate"], std.error = co[avail, "Std. Error"])
}

fit_model_p2 <- function(data, re_spec = c("correct", "misspecified", "full_slopes")) {
  re_spec <- match.arg(re_spec)
  form <- switch(re_spec,
    correct      = response ~ zHLVA * zPCREFp + (1 | participant_id) + (1 | text_id),
    misspecified = response ~ zHLVA * zPCREFp + (1 | participant_id),
    full_slopes  = response ~ zHLVA * zPCREFp + (1 + zPCREFp | participant_id) + (1 + zHLVA | text_id)
  )
  warn_msgs <- character(0)
  fit <- tryCatch(
    withCallingHandlers(
      glmer(form, data = data, family = binomial, control = glmerControl(optimizer = "bobyqa")),
      warning = function(w) { warn_msgs <<- c(warn_msgs, conditionMessage(w)); invokeRestart("muffleWarning") }
    ),
    error = function(e) e
  )
  if (inherits(fit, "error")) return(list(status = "error", coefs = NULL))
  status <- if (length(warn_msgs) > 0 &&
                any(grepl("failed to converge|Model failed|large eigenvalue|unable to evaluate",
                          warn_msgs, ignore.case = TRUE))) {
    "optimizer_warning"
  } else if (isSingular(fit)) {
    "converged_singular"
  } else {
    "converged_clean"
  }
  list(status = status, coefs = extract_terms_of_interest_p2(fit))
}

fit_model_p2_guarded <- function(data, re_spec, timeout_secs = 15) {
  setTimeLimit(elapsed = timeout_secs, transient = TRUE)
  res <- tryCatch(
    fit_model_p2(data, re_spec),
    error = function(e) list(status = "error", coefs = NULL)
  )
  setTimeLimit(elapsed = Inf, transient = FALSE)
  res
}
```

Three functions, each doing one job:

**`extract_terms_of_interest_p2()`** pulls just the three coefficients this
whole program cares about (`zHLVA`, `zPCREFp`, their interaction) out of a
fitted model's coefficient table, as a tidy `term`/`estimate`/`std.error`
tibble. `intersect(terms, rownames(co))` guards against a term being
missing entirely (which can happen on a badly degenerate fit) rather than
erroring.

**`fit_model_p2()`** does the actual fitting. Three things worth noting:
- The formula is chosen via `re_spec`, a three-way switch between `"correct"`
  (random intercepts for both participant and text -- the specification
  Aims 1-2 always use), `"misspecified"` (participant intercept only, used
  by Aim 3 to quantify the cost of omitting a grouping factor), and
  `"full_slopes"` (adds correlated random slopes, used only by Aim 5).
- `withCallingHandlers()` around the `glmer()` call **captures warnings
  without stopping execution** -- `invokeRestart("muffleWarning")` lets the
  fit continue and the warning's text is stashed in `warn_msgs` for
  inspection afterward, rather than letting it print to the console
  thousands of times over a large simulation.
- The `status` classification turns those captured warnings (plus `lme4`'s
  own `isSingular()` check) into one of three buckets: `optimizer_warning`
  (the optimizer itself flagged non-convergence), `converged_singular` (a
  random-effect variance pinned at its zero boundary -- expected and common
  at small `n_t`, as later diagnostics show), or `converged_clean`. This
  three-way classification is what later analysis uses to report *how
  often* fits broke down, not just what they estimated when they didn't.

**`fit_model_p2_guarded()`** wraps `fit_model_p2()` in a per-fit wall-clock
timeout (`setTimeLimit()`, default 15 seconds) so that one pathological fit
in a 60,000+-fit simulation run can't hang the entire background job
indefinitely; a fit that blows through the timeout is treated the same as
any other error.

## Step 4: the sample-size and effect-size grids

```{r}
#| label: sample-size-grid

n_p_grid <- c(25, 100, 375, 725, 1300, 1497)
n_t_grid <- c(2, 10, 35, 85, 136)
```

```{r}
#| label: aim1-sweep-grids

b_hlva_grid        <- c(0, 0.18, 0.363, 0.7, 1.0)
b_pcref_grid       <- c(0, 0.05, 0.095, 0.2, 0.35)
b_interaction_grid <- c(0, 0.025, 0.054, 0.10, 0.20)
```

`n_p_grid` (participant counts) and `n_t_grid` (text counts) together
define 6 x 5 = 30 `(n_p, n_t)` "cells" -- every combination of participant
and text sample size the simulation will visit. These particular values
were carried over unchanged from an earlier, real-data resampling analysis
(Phase 1), so that this parametric-simulation program's results can be laid
directly alongside that earlier analysis's real-data findings.

The three effect-size grids implement Aim 1's **one-at-a-time sweep**
design: each grid has 5 values centered on that term's real,
empirically-estimated effect size (the middle value in every grid),
flanked by two smaller and two larger hypothetical true effects. Sweeping
one term while holding the other two fixed at their real values -- rather
than crossing all three simultaneously -- keeps the design a manageable
"3 sweeps" rather than a combinatorial explosion, at the cost of not being
able to see three-way interactions between the swept effects (not a
concern here, since the real fitted model has no such three-way term among
these predictors).

## Step 5: generating a discovery/replication pair (for Aim 2)

Aim 2 asks a different question from Aim 1: not "how well does one fit
recover the truth," but "if I split my sample into two independent halves,
does a significant finding in one replicate in the other?" That needs a
function that generates *two* related datasets at once, sharing (or not)
part of the design.

```{r}
#| label: generate-data-p2-pair

generate_data_p2_pair <- function(
    n_p_total, n_t_total, k = min(10, n_t_total), questions_per_text = 4,
    b_intercept = 2.093, b_hlva = 0.363, b_pcref = 0.095, b_interaction = 0.054,
    sd_participant = 0.69, sd_text = 0.41,
    sd_pcref_slope = 0, sd_hlva_slope = 0, cor_part = 0, cor_text = 0,
    split_margin = c("participants_only", "both"),
    seed
) {
  split_margin <- match.arg(split_margin)
  set.seed(seed)

  n_p_half <- max(1, floor(n_p_total / 2))
  n_t_half <- if (split_margin == "both") max(1, floor(n_t_total / 2)) else n_t_total

  make_texts <- function(n_t) {
    Sigma_text <- matrix(c(sd_text^2, cor_text * sd_text * sd_hlva_slope,
                            cor_text * sd_text * sd_hlva_slope, sd_hlva_slope^2), nrow = 2)
    re_text <- if (sd_hlva_slope > 0) MASS::mvrnorm(n_t, c(0, 0), Sigma_text) else cbind(rnorm(n_t, 0, sd_text), 0)
    tibble(text_id = 1:n_t, zPCREFp = rnorm(n_t, 0, 1), u_text = re_text[, 1], s_hlva = re_text[, 2])
  }
  make_participants <- function(n_p) {
    Sigma_part <- matrix(c(sd_participant^2, cor_part * sd_participant * sd_pcref_slope,
                            cor_part * sd_participant * sd_pcref_slope, sd_pcref_slope^2), nrow = 2)
    re_part <- if (sd_pcref_slope > 0) MASS::mvrnorm(n_p, c(0, 0), Sigma_part) else cbind(rnorm(n_p, 0, sd_participant), 0)
    tibble(participant_id = 1:n_p, zHLVA = rnorm(n_p, 0, 1), u_participant = re_part[, 1], s_pcref = re_part[, 2])
  }
  build_dataset <- function(participants, texts) {
    kk <- min(k, nrow(texts))
    assignments <- tibble(participant_id = participants$participant_id) |>
      dplyr::mutate(assigned = purrr::map(participant_id, ~ sample(texts$text_id, size = kk, replace = FALSE))) |>
      tidyr::unnest(assigned) |> dplyr::rename(text_id = assigned)
    assignments |>
      dplyr::left_join(participants, by = "participant_id") |>
      dplyr::left_join(texts, by = "text_id") |>
      dplyr::slice(rep(1:dplyr::n(), each = questions_per_text)) |>
      dplyr::mutate(
        question_id = rep(1:questions_per_text, times = dplyr::n() / questions_per_text),
        eta = b_intercept + (b_hlva + s_hlva) * zHLVA + (b_pcref + s_pcref) * zPCREFp +
          b_interaction * zHLVA * zPCREFp + u_participant + u_text,
        prob = plogis(eta), response = rbinom(dplyr::n(), 1, prob)
      ) |>
      dplyr::select(participant_id, text_id, question_id, zHLVA, zPCREFp, response)
  }

  texts_shared      <- make_texts(n_t_half)
  participants_disc <- make_participants(n_p_half)
  participants_repl <- make_participants(n_p_half)
  texts_repl        <- if (split_margin == "both") make_texts(n_t_half) else texts_shared

  list(
    discovery   = build_dataset(participants_disc, texts_shared),
    replication = build_dataset(participants_repl, texts_repl)
  )
}
```

This is deliberately a self-contained re-implementation of
`generate_data_p2()`'s logic (three internal helpers `make_texts()`,
`make_participants()`, `build_dataset()`, each doing the same job as the
corresponding piece of Step 2) rather than a call to it, because it needs
to build *two* datasets that share some pieces and not others, controlled
by `split_margin`:

- **`n_p_half`/`n_t_half`**: given a cell's *total* sample size, halve the
  participant count always; halve the text count only if `split_margin =
  "both"`. With `split_margin = "participants_only"` (what Aims 1-2 use),
  `n_t_half` is just the full `n_t_total` -- both halves see the *same*
  texts.
- **Why avoid splitting texts at small `n_t`**: the comment in the source
  document notes this choice specifically avoids degenerate 1-text halves
  at the smallest grid point (`n_t = 2`) -- splitting 2 texts into two
  halves of 1 would make the text-level random intercept trivially
  unidentifiable in a different, less interesting way than the phenomenon
  Aim 1/2 are trying to study.
- **Building the pair**: `texts_shared` is generated once; `discovery` and
  `replication` each get their *own*, independently-sampled participants
  (`participants_disc`/`participants_repl`), but share `texts_shared`
  unless `split_margin = "both"` asks for independent `texts_repl` too.
  Each half is then expanded to the full participant x text x question
  dataset via the shared `build_dataset()` helper -- functionally identical
  to steps (h)-(j) of `generate_data_p2()`.

## Step 6: diagnostics and the per-grid-cell orchestrator

Everything above produces raw data and raw model fits. This step turns a
raw fit's coefficients into the specific pass/fail diagnostics the report's
tables and figures need, and then wires the whole pipeline together so a
single function call handles one complete "cell" of the design (both aims,
at a given sweep/effect value/sample size/replicate).

```{r}
#| label: aim1-aim2-diagnostics-and-runner

diagnose_coefs_p2 <- function(coefs, true_vals) {
  coefs |>
    dplyr::mutate(
      true_value      = true_vals[term],
      ci_lower        = estimate - 1.96 * std.error,
      ci_upper        = estimate + 1.96 * std.error,
      ci_width        = ci_upper - ci_lower,
      excludes_zero   = ci_lower > 0 | ci_upper < 0,
      correct_sign    = sign(estimate) == sign(true_value),
      sign_error      = excludes_zero & !correct_sign,
      inflation_ratio = ifelse(true_value != 0, abs(estimate) / abs(true_value), NA_real_),
      bias            = estimate - true_value,
      covers_truth    = ci_lower <= true_value & true_value <= ci_upper
    )
}

combine_replication_p2 <- function(res_disc, res_repl, true_vals) {
  if (is.null(res_disc$coefs) || is.null(res_repl$coefs)) {
    return(tibble(term = names(true_vals), replicated = NA, discovery_significant = NA))
  }
  disc <- diagnose_coefs_p2(res_disc$coefs, true_vals) |>
    dplyr::select(term, estimate, excludes_zero)
  repl <- diagnose_coefs_p2(res_repl$coefs, true_vals) |>
    dplyr::select(term, estimate_repl = estimate, excludes_zero_repl = excludes_zero)
  dplyr::full_join(disc, repl, by = "term") |>
    dplyr::mutate(
      discovery_significant = excludes_zero,
      replicated = excludes_zero & excludes_zero_repl & sign(estimate) == sign(estimate_repl)
    )
}

run_aim1_aim2_cell <- function(sweep, effect_value, n_p, n_t, rep, split_margin = "participants_only") {
  b_hlva        <- if (sweep == "hlva")        effect_value else 0.363
  b_pcref       <- if (sweep == "pcref")       effect_value else 0.095
  b_interaction <- if (sweep == "interaction") effect_value else 0.054
  true_vals <- c(zHLVA = b_hlva, zPCREFp = b_pcref, `zHLVA:zPCREFp` = b_interaction)

  sweep_idx <- match(sweep, c("hlva", "pcref", "interaction"))
  seed_base <- rep * 1e7 + n_p * 1e3 + n_t + sweep_idx * 1e5 + round(effect_value * 1000)

  ## Aim 1: single full-sample fit
  df_full  <- generate_data_p2(n_participants = n_p, n_texts = n_t,
                                b_hlva = b_hlva, b_pcref = b_pcref, b_interaction = b_interaction,
                                seed = seed_base)
  res_full <- fit_model_p2_guarded(df_full, "correct")

  aim1_rows <- if (!is.null(res_full$coefs)) {
    diagnose_coefs_p2(res_full$coefs, true_vals) |> dplyr::mutate(status = res_full$status)
  } else {
    tibble(term = names(true_vals), status = res_full$status)
  }
  aim1_rows <- aim1_rows |>
    dplyr::mutate(sweep = sweep, effect_value = effect_value, n_p = n_p, n_t = n_t, rep = rep)

  ## Aim 2: discovery/replication split (same grid cell = total N)
  pair <- generate_data_p2_pair(n_p_total = n_p, n_t_total = n_t,
                                 b_hlva = b_hlva, b_pcref = b_pcref, b_interaction = b_interaction,
                                 split_margin = split_margin, seed = seed_base + 1)
  res_disc <- fit_model_p2_guarded(pair$discovery, "correct")
  res_repl <- fit_model_p2_guarded(pair$replication, "correct")

  aim2_rows <- combine_replication_p2(res_disc, res_repl, true_vals) |>
    dplyr::mutate(
      sweep = sweep, effect_value = effect_value, n_p = n_p, n_t = n_t, rep = rep,
      split_margin = split_margin,
      status_discovery = res_disc$status, status_replication = res_repl$status
    )

  list(aim1 = aim1_rows, aim2 = aim2_rows)
}
```

**`diagnose_coefs_p2()`** is where "estimate + standard error" turns into
the actual quantities the report's tables/figures show, given the true
value being simulated:
- `ci_lower`/`ci_upper`/`ci_width`: a 95% **Wald** interval
  (`estimate +/- 1.96 x std.error`) -- chosen for speed over a
  profile-likelihood interval, since this needs to run tens of thousands of
  times.
- `excludes_zero`: the conventional "statistically significant at p < .05"
  criterion, expressed directly as a CI check.
- `correct_sign`/`sign_error`: whether the point estimate's sign matches
  the true value's sign, and specifically whether a *significant* result
  has the *wrong* sign (Marek et al.'s "Type S" error).
- `inflation_ratio`: `|estimate| / |true value|`, undefined (`NA`) when the
  true value is exactly zero, since "how many times too large" isn't
  meaningful for a null effect (Marek et al.'s "Type M" error, computed
  later only among significant fits).
- `bias` and `covers_truth`: the raw signed estimation error, and whether
  the CI actually contains the true value (nominal 95% coverage would mean
  this is `TRUE` 95% of the time under correct model specification).

**`combine_replication_p2()`** takes two already-diagnosed fits (from two
independent halves of one cell) and defines **replication**: a term counts
as replicated only if it was significant in *both* halves *and* the sign
agreed between them. `discovery_significant` records whether it was even
significant in the first half at all -- needed later to compute the
*conditional* replication rate (replication rate among only the
discoveries), a distinct question from the *unconditional* rate.

**`run_aim1_aim2_cell()`** is the orchestrator: given one sweep name, one
swept effect value, one `(n_p, n_t)` cell, and one replicate index, it:
1. Works out the three true coefficient values for this call -- the swept
   term takes `effect_value`, the other two stay at their real,
   empirically-observed defaults.
2. Builds a **deterministic seed** from `(rep, n_p, n_t, sweep, effect_value)`
   via a simple hashing formula, so every individual simulated dataset is
   reproducible on its own, without needing to save the full random state
   of a multi-hour parallel run.
3. Runs the **Aim 1** side: one dataset at the full cell size, fit once,
   diagnosed once.
4. Runs the **Aim 2** side: one discovery/replication *pair* at the same
   total cell size (so a cell's total sample size means the same thing in
   both aims), each half fit and diagnosed, then combined into a
   replication verdict.
5. Returns both aims' results together as a two-element list, so a single
   function call produces everything one "cell" contributes to both aims'
   final datasets.

## Step 7: running the full design (outside this document)

Every piece above is a function -- nothing has actually been executed at
scale yet. Producing the real input data for the report meant calling
`run_aim1_aim2_cell()` once for **every** combination of:

- 3 sweeps (`hlva`, `pcref`, `interaction`)
- x 5 effect values each
- x 30 `(n_p, n_t)` cells
- x 50 replicates per cell

-- 22,500 calls to `run_aim1_aim2_cell()`, each of which performs 3
`glmer` fits internally (Aim 1's one fit, plus Aim 2's discovery and
replication fits), for **67,500 total model fits**. This was executed as a
**detached background R process** (`run_aim1_aim2_full.R`, not part of
this document), one sweep at a time, using `furrr::plan(multisession,
workers = 4)` to fit cells in parallel across 4 worker processes. All three
sweeps together took roughly 5 hours of wall-clock time. Results were saved
to six `.rds` files (`results_phase2/aim1_hlva.rds`,
`results_phase2/aim2_hlva.rds`, and the `pcref`/`interaction` equivalents).

This step is worth calling out explicitly in a tutorial, even though there
is no code chunk to show for it here: **a simulation of this size should
not be run inside the document that reports on it.** Rendering a report
should be fast and repeatable; a multi-hour parallel batch job belongs in
its own script, run once, with its output saved and loaded (as the next
step does).

## Step 8: turning saved results into the tibbles the visualizations use

This is the last step on the *generation* side -- the point at which raw,
already-computed simulation output becomes exactly the data structures the
report's tables and figures consume directly.

```{r}
#| label: load-aim1-aim2-results

aim1_files <- c("results_phase2/aim1_hlva.rds", "results_phase2/aim1_pcref.rds",
                 "results_phase2/aim1_interaction.rds")
aim2_files <- c("results_phase2/aim2_hlva.rds", "results_phase2/aim2_pcref.rds",
                 "results_phase2/aim2_interaction.rds")
missing_files <- c(aim1_files, aim2_files)[!file.exists(c(aim1_files, aim2_files))]
if (length(missing_files) > 0) {
  stop(
    "Missing Aim 1/2 result file(s): ", paste(missing_files, collapse = ", "),
    ". These are produced by the detached background job run_aim1_aim2_full.R ",
    "(~5h on 4 workers) -- this document does not regenerate them inline. ",
    "Run that script first if the files are genuinely absent."
  )
}

aim1_hlva        <- readRDS("results_phase2/aim1_hlva.rds")
aim1_pcref       <- readRDS("results_phase2/aim1_pcref.rds")
aim1_interaction <- readRDS("results_phase2/aim1_interaction.rds")
aim2_hlva        <- readRDS("results_phase2/aim2_hlva.rds")
aim2_pcref       <- readRDS("results_phase2/aim2_pcref.rds")
aim2_interaction <- readRDS("results_phase2/aim2_interaction.rds")

own_effect <- c(zHLVA = 0.363, zPCREFp = 0.095, `zHLVA:zPCREFp` = 0.054)
term_levels <- c("zHLVA", "zPCREFp", "zHLVA:zPCREFp")

# Each sweep's own term, evaluated only at its empirically-observed effect
# size -- the slice used by most tables/figures below.
aim1_all <- dplyr::bind_rows(
  aim1_hlva        |> dplyr::filter(term == "zHLVA",        effect_value == own_effect["zHLVA"]),
  aim1_pcref       |> dplyr::filter(term == "zPCREFp",       effect_value == own_effect["zPCREFp"]),
  aim1_interaction |> dplyr::filter(term == "zHLVA:zPCREFp", effect_value == own_effect["zHLVA:zPCREFp"])
) |> dplyr::mutate(term = factor(term, levels = term_levels))

aim2_all <- dplyr::bind_rows(
  aim2_hlva        |> dplyr::filter(term == "zHLVA",        effect_value == own_effect["zHLVA"]),
  aim2_pcref       |> dplyr::filter(term == "zPCREFp",       effect_value == own_effect["zPCREFp"]),
  aim2_interaction |> dplyr::filter(term == "zHLVA:zPCREFp", effect_value == own_effect["zHLVA:zPCREFp"])
) |> dplyr::mutate(term = factor(term, levels = term_levels))

# Full effect-size sweep, all points (not just the own-effect-size slice) --
# used for the Aim 1 power-curves figure below.
aim1_sweep_combined <- dplyr::bind_rows(
  aim1_hlva        |> dplyr::filter(term == "zHLVA"),
  aim1_pcref       |> dplyr::filter(term == "zPCREFp"),
  aim1_interaction |> dplyr::filter(term == "zHLVA:zPCREFp")
) |> dplyr::mutate(term = factor(term, levels = term_levels))
```

Piece by piece:

- **The `file.exists()` guard.** Before touching anything else, the code
  checks that all six expected `.rds` files exist and fails loudly with a
  specific, actionable error message if any are missing, rather than
  letting a later step fail confusingly (or silently proceed with stale
  data). This is a good habit any time a document depends on
  externally-produced files it doesn't (and shouldn't) regenerate itself.
- **Loading the six files.** Each sweep's Aim 1 and Aim 2 results were
  saved separately by the background job, so there are six objects to
  read in: `aim1_hlva`, `aim1_pcref`, `aim1_interaction` and their Aim 2
  counterparts. Each is a long tibble with one row per (term, effect
  value, `n_p`, `n_t`, replicate) combination.
- **`own_effect` and `term_levels`.** `own_effect` records each term's
  real, empirically-observed value -- the same numbers used as defaults
  back in Step 2/6 -- so that later filtering can pick out "the row where
  the swept effect equals its real value" without hard-coding that number
  three separate times. `term_levels` fixes a consistent factor ordering
  (`zHLVA`, `zPCREFp`, interaction) used throughout so tables and plots
  always present the three terms in the same order.
- **`aim1_all` / `aim2_all`: the "own effect size" slice.** Each sweep's
  saved results contain data for *all five* swept effect-size values (0,
  small, real, large, larger), but most tables/figures only want *each
  term's own real value* -- e.g. `zHLVA`'s row from the `hlva` sweep at
  `effect_value == 0.363`, not from the `pcref` or `interaction` sweeps
  (where `zHLVA` was just held fixed as a covariate, not the focus). The
  three `filter()` + `bind_rows()` calls pick exactly one term from each of
  the three sweeps' output and stack them into one tidy tibble.
- **`aim1_sweep_combined`: the full sweep, kept separate.** One figure (the
  power-curves figure, showing how power changes as the *true effect size
  itself* varies) genuinely needs every effect-size value, not just the
  real one -- so this object keeps all five values per term rather than
  filtering down to one.

At this point, `aim1_all`, `aim2_all`, and `aim1_sweep_combined` are exactly
the data structures every downstream table and figure consumes. Nothing
past this point generates new data -- everything from here on is
summarizing (`group_by()`/`summarise()`) and plotting what these three
tibbles already contain.

::: {.callout-tip}
## If you want to actually run this yourself

The full 67,500-fit design is a multi-hour job and shouldn't be your first
attempt at reproducing this. The source document
(`synthetic-data-analyses-power-error-replication-aims-1-2-draft.qmd`)
contains a small, self-contained **smoke test** in its collapsed
"Development history" section that calls exactly the functions walked
through above (`run_aim1_aim2_cell()`, and transitively everything else) on
a tiny 24-combination grid, completing in about 6 seconds -- a good
starting point for experimenting with this code interactively before
attempting anything at full scale.
:::
