---
title: "Effect size and sample size variation for the HLVA main effect, the PCREFp main effect, and their interaction"
subtitle: "Review: densified grid (Scenario 5) and full-grid intercept sweep, accuracy outcome"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: cosmo
    embed-resources: true
    code-fold: true
execute:
  warning: false
  message: false
---

::: {.callout-note}
## Scope of this document

This document reviews the **Scenario 5** results: a densified `(n_p, n_t)`
grid (11 participant sample sizes x 9 text sample sizes, log-spaced, vs. the
original 6x5 grid), doubled replicates per cell (100 vs. 50), and a
companion full-grid `b_intercept` sweep (4 target-accuracy values x the same
99-cell grid, Aim-1-only). All three were run overnight
(2026-08-26/27) as background jobs (`run_scenario5_full.R`,
`run_scenario5_intercept_sweep.R`) and are analyzed here for the first time.

**This document does not modify or supersede**
`synthetic-data-power-error-replication-effect-size-sampling-draft-development.qmd`
(the original Aims 1-2 document, covering the 6x5/50-replicate grid and the
first, 4-cell version of the intercept sweep) -- that document is left as-is,
a record of the original analysis, and is cross-referenced throughout below
rather than re-derived. Where this document's numbers differ from the
original's, that is flagged and explained explicitly (see "A grid-density
caveat" below) rather than treated as a contradiction.

All tables and figures below are computed from pre-existing, already-run
results saved in `results_phase2/*_scenario5_*.rds` and
`results_phase2/aim1_scenario5_intercept_sweep.rds` (no full-grid
recomputation happens when this document is rendered); the original 6x5/50
files (`results_phase2/aim{1,2}_{hlva,pcref,interaction}.rds`) are also
loaded, for direct side-by-side comparison at matched sample sizes.
:::

```{r}
#| label: setup

library(tidyverse)
library(patchwork)
library(RColorBrewer)
library(ggdist)
library(scico)
```

## Computational cost and fit accounting

The table below itemizes every model fit behind the Aims 1-2 program to
date -- the original grid, Scenario 5's densified grid, and all three
`b_intercept` sweeps (the original 4-cell version, its ceiling-densification
follow-up, and Scenario 5's full-grid version) -- broken out by phase rather
than left as a single combined total. This table is a cumulative
cost/bookkeeping summary only: every figure and statistic later in this
document is computed from a single phase's fits (almost always Scenario 5's),
not pooled across phases -- the one exception is an explicit side-by-side
comparison at matched sample sizes in the Aim 2 "grid-density caveat"
section, which compares rather than combines the two grids.

```{r}
#| label: fit-accounting-table

fit_accounting <- tribble(
  ~phase, ~n_sweeps, ~n_effect_values, ~n_cells, ~n_reps, ~fits_per_combo, ~notes,
  "Original Aim 1+2 grid (6x5, 50 reps)",       3, 5, 30, 50, 3, "aim1/aim2_{hlva,pcref,interaction}.rds",
  "Scenario 5 Aim 1+2 grid (11x9, 100 reps)",   3, 5, 99, 100, 3, "aim1/aim2_scenario5_{hlva,pcref,interaction}.rds",
  "Original b_intercept sweep (4 cells)",       1, 5, 4, 50, 1, "aim1_intercept_sweep.rds",
  "Ceiling-densification extension (4 cells)",  1, 2, 4, 50, 1, "aim1_intercept_sweep_ceiling_extension.rds",
  "Scenario 5 b_intercept sweep (11x9)",        1, 4, 99, 100, 1, "aim1_scenario5_intercept_sweep.rds"
) |>
  mutate(total_fits = n_sweeps * n_effect_values * n_cells * n_reps * fits_per_combo)

fit_accounting |>
  select(phase, n_sweeps, n_effect_values, n_cells, n_reps, fits_per_combo, total_fits, notes) |>
  knitr::kable(
    col.names = c("Phase", "Sweeps", "Effect values/sweep", "Grid cells", "Reps/cell",
                  "Fits/combination", "Total fits", "Source file(s)"),
    format.args = list(big.mark = ",")
  )

cat("\nGrand total fits across the whole Aims 1-2 program to date:",
    format(sum(fit_accounting$total_fits), big.mark = ","), "\n")
```

**Estimated vs. observed run time.** Every prior scenario in this program's
cost-estimation planning (`cost-estimate-expanded-grid-replicates-aims1-2-PLAN.qmd`)
assumed the original 4-`furrr`-worker convention, adopted early in this
project when the development machine was found to be under real memory
pressure (swap frequently near-full). Scenario 5 itself was launched with
**8 workers** instead, after a post-restart check immediately beforehand
found 12 CPU cores and 0% swap in use -- headroom that did not exist earlier
in the project and that motivated raising the worker count for this run
specifically, not a general change to the project's convention.

| Run | Estimated (4 workers) | Actual (8 workers) | Estimated core-hours | Notes |
|---|---|---|---|---|
| Scenario 5 main grid (3 sweeps) | ~33h | **~12.2h** (4.1h + 4.05h + 4.03h per sweep) | ~132 | `scenario5_run.log` |
| Scenario 5 intercept sweep | ~2.9h | **~2.01h** | ~11.7 | `scenario5_intercept_sweep_run.log` |
| **Combined** | ~35.9h | **~14.2h** | ~143.7 | |

Doubling the worker count did not simply halve the wall-clock time (a naive
4-workers-to-8-workers halving would predict ~16.5h + ~1.45h = ~18h, not the
observed ~12.2h + ~2.01h = ~14.2h) -- the actual speedup is somewhat *better*
than linear, plausibly because 8 workers on a 12-core machine leaves more
headroom per worker than 4 workers did when memory was tight (less
contention/swapping overhead per fit). No errors were logged in either run;
warnings were the expected `boundary (singular) fit` messages from `n_t`-poor
cells, muffled by the existing `withCallingHandlers()` pattern.

## Loading Scenario 5, the full-grid intercept sweep, and the original grid; shared helpers

```{r}
#| label: load-results

aim1_files_s5 <- c("results_phase2/aim1_scenario5_hlva.rds",
                    "results_phase2/aim1_scenario5_pcref.rds",
                    "results_phase2/aim1_scenario5_interaction.rds")
aim2_files_s5 <- c("results_phase2/aim2_scenario5_hlva.rds",
                    "results_phase2/aim2_scenario5_pcref.rds",
                    "results_phase2/aim2_scenario5_interaction.rds")
intercept_file_s5 <- "results_phase2/aim1_scenario5_intercept_sweep.rds"

aim1_files_orig <- c("results_phase2/aim1_hlva.rds", "results_phase2/aim1_pcref.rds",
                      "results_phase2/aim1_interaction.rds")
aim2_files_orig <- c("results_phase2/aim2_hlva.rds", "results_phase2/aim2_pcref.rds",
                      "results_phase2/aim2_interaction.rds")

all_needed <- c(aim1_files_s5, aim2_files_s5, intercept_file_s5, aim1_files_orig, aim2_files_orig)
missing_files <- all_needed[!file.exists(all_needed)]
if (length(missing_files) > 0) {
  stop("Missing result file(s): ", paste(missing_files, collapse = ", "))
}

aim1_hlva_s5        <- readRDS(aim1_files_s5[1])
aim1_pcref_s5       <- readRDS(aim1_files_s5[2])
aim1_interaction_s5 <- readRDS(aim1_files_s5[3])
aim2_hlva_s5        <- readRDS(aim2_files_s5[1])
aim2_pcref_s5       <- readRDS(aim2_files_s5[2])
aim2_interaction_s5 <- readRDS(aim2_files_s5[3])
intercept_s5        <- readRDS(intercept_file_s5)

aim1_hlva_orig        <- readRDS(aim1_files_orig[1])
aim1_pcref_orig       <- readRDS(aim1_files_orig[2])
aim1_interaction_orig <- readRDS(aim1_files_orig[3])
aim2_hlva_orig        <- readRDS(aim2_files_orig[1])
aim2_pcref_orig       <- readRDS(aim2_files_orig[2])
aim2_interaction_orig <- readRDS(aim2_files_orig[3])

own_effect  <- c(zHLVA = 0.363, zPCREFp = 0.095, `zHLVA:zPCREFp` = 0.054)
term_levels <- c("zHLVA", "zPCREFp", "zHLVA:zPCREFp")

n_p_grid_s5 <- sort(unique(aim1_hlva_s5$n_p))
n_t_grid_s5 <- sort(unique(aim1_hlva_s5$n_t))
n_p_grid_orig <- sort(unique(aim1_hlva_orig$n_p))
n_t_grid_orig <- sort(unique(aim1_hlva_orig$n_t))

pool_own_effect <- function(aim1_hlva, aim1_pcref, aim1_interaction) {
  dplyr::bind_rows(
    aim1_hlva        |> dplyr::filter(term == "zHLVA",        effect_value == own_effect["zHLVA"]),
    aim1_pcref       |> dplyr::filter(term == "zPCREFp",       effect_value == own_effect["zPCREFp"]),
    aim1_interaction |> dplyr::filter(term == "zHLVA:zPCREFp", effect_value == own_effect["zHLVA:zPCREFp"])
  ) |> dplyr::mutate(term = factor(term, levels = term_levels))
}

aim1_all_s5   <- pool_own_effect(aim1_hlva_s5, aim1_pcref_s5, aim1_interaction_s5)
aim2_all_s5   <- pool_own_effect(aim2_hlva_s5, aim2_pcref_s5, aim2_interaction_s5)
aim1_all_orig <- pool_own_effect(aim1_hlva_orig, aim1_pcref_orig, aim1_interaction_orig)
aim2_all_orig <- pool_own_effect(aim2_hlva_orig, aim2_pcref_orig, aim2_interaction_orig)

aim1_sweep_combined_s5 <- dplyr::bind_rows(
  aim1_hlva_s5        |> dplyr::filter(term == "zHLVA"),
  aim1_pcref_s5       |> dplyr::filter(term == "zPCREFp"),
  aim1_interaction_s5 |> dplyr::filter(term == "zHLVA:zPCREFp")
) |> dplyr::mutate(term = factor(term, levels = term_levels))

# Shared summary helper -- identical to the original document's version.
summarize_p2_metrics <- function(df, group_vars) {
  df |>
    dplyr::group_by(dplyr::across(dplyr::all_of(group_vars))) |>
    dplyr::summarise(
      power            = mean(excludes_zero, na.rm = TRUE),
      sign_reliability = mean(correct_sign, na.rm = TRUE),
      type_s_rate      = mean(sign_error[excludes_zero], na.rm = TRUE),
      type_m_ratio     = median(inflation_ratio[excludes_zero], na.rm = TRUE),
      median_bias      = median(bias, na.rm = TRUE),
      coverage         = mean(covers_truth, na.rm = TRUE),
      n_fits           = dplyr::n(),
      .groups = "drop"
    )
}

# Shared heatmap helper. Finalized color scheme (superseding the original
# document's labeled-tile, plain-viridis/gradient2 version, after a
# dedicated color-scheme review): tiles only, no in-cell text labels;
# sequential (non-diverging) panels use viridis with `direction` flipped via
# `lower_is_better` so that light/yellow always means "the diagnostic looks
# good" and dark blue/purple always means "the diagnostic looks bad",
# regardless of whether the underlying metric's good end is high (e.g. power,
# sign reliability, coverage) or low (e.g. Type I rate, Type S rate, Type M
# ratio). Diverging panels (median bias) use scico's "vik" palette (dark
# blue <-> white <-> dark red) rather than a plain red/blue gradient2, so
# that "light/center = good, dark extremes = bad" reads consistently with
# the sequential panels even though bias necessarily uses a different hue
# family (no true diverging viridis palette exists). `viridis_option`
# (added in a later review pass) lets one panel use a different member of
# the viridis family -- see panel_type1_s5 below for why Type I rate uses
# this.
plot_power_heatmap <- function(data, fill_var, facet_var = "term", fill_label = "Value",
                                grid_n_p = n_p_grid_s5, grid_n_t = n_t_grid_s5,
                                limits = c(0, 1), label_format = c("percent", "number"),
                                diverging = FALSE, midpoint = 0, squish = FALSE,
                                lower_is_better = FALSE, viridis_option = "D") {
  label_format <- match.arg(label_format)
  oob_fn <- if (squish) scales::squish else scales::censor

  p <- ggplot(data, aes(x = factor(n_p, levels = grid_n_p), y = factor(n_t, levels = grid_n_t),
                    fill = .data[[fill_var]])) +
    geom_tile() +
    # Square tiles rather than the tall/thin default -- coord_fixed treats
    # each discrete x/y category as one unit, so ratio = 1 forces literally
    # square cells regardless of how many n_p vs. n_t levels there are.
    coord_fixed(ratio = 1) +
    facet_wrap(stats::as.formula(paste0("~", facet_var))) +
    labs(x = "Participant sample size (n_p)", y = "Text sample size (n_t)", fill = fill_label) +
    theme_minimal(base_size = 9) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 6))

  if (diverging) {
    p + scico::scale_fill_scico(palette = "vik", limits = limits, oob = oob_fn)
  } else if (label_format == "percent") {
    p + scale_fill_viridis_c(limits = limits, labels = scales::percent, oob = oob_fn,
                              direction = if (lower_is_better) -1 else 1, option = viridis_option)
  } else {
    p + scale_fill_viridis_c(limits = limits, oob = oob_fn,
                              direction = if (lower_is_better) -1 else 1, option = viridis_option)
  }
}
```

```{r}
#| label: aim1-metrics-vs-x-and-curve-helpers

# Ported from the original document (`...draft-development.qmd`) -- these
# were present there but dropped when this review document was written,
# and are restored here, adapted for Scenario 5's densified grid. The
# effect-size sweep grids themselves are unchanged from the original:
# Scenario 5 densified (n_p, n_t) and doubled replicates, but swept the
# same 5 effect-size values per term.
b_hlva_grid        <- c(0, 0.18, 0.363, 0.7, 1.0)
b_pcref_grid       <- c(0, 0.05, 0.095, 0.2, 0.35)
b_interaction_grid <- c(0, 0.025, 0.054, 0.10, 0.20)

aim1_metric_labels <- c(power = "Power\nP(significant)",
                         sign_reliability = "Sign reliability\nP(correct sign)",
                         type_s_rate = "Type S rate\nP(wrong sign | significant)",
                         type_m_ratio = "Type M ratio\nmedian |est/true| | significant")

# Generalized to take an explicit `x_breaks` grid (the original hardcoded
# `n_t_grid`) so it works for Scenario 5's 9/11-point grids here.
plot_metrics_vs_x <- function(data, x_var, metrics, metric_labels,
                               color_var = "term", x_lab = NULL, x_breaks = n_t_grid_s5) {
  x_lab <- x_lab %||% x_var
  df_long <- data |>
    tidyr::pivot_longer(dplyr::all_of(metrics), names_to = "metric", values_to = "value") |>
    dplyr::mutate(metric = factor(metric_labels[metric], levels = unname(metric_labels)))

  ggplot(df_long, aes(x = .data[[x_var]], y = value, color = .data[[color_var]])) +
    geom_line() +
    geom_point() +
    facet_wrap(~metric, scales = "free_y") +
    scale_x_log10(breaks = x_breaks) +
    scale_color_brewer(palette = "Dark2") +
    labs(x = x_lab, y = NULL, color = "Term") +
    theme_minimal()
}

aim1_metrics_by_np_s5 <- aim1_all_s5 |>
  dplyr::group_by(term, n_p) |>
  dplyr::summarise(
    power            = mean(excludes_zero, na.rm = TRUE),
    sign_reliability = mean(correct_sign, na.rm = TRUE),
    type_s_rate      = mean(sign_error[excludes_zero], na.rm = TRUE),
    type_m_ratio     = median(inflation_ratio[excludes_zero], na.rm = TRUE),
    n_fits           = dplyr::n(),
    .groups = "drop"
  )

# Pooled-over-n_p summary by term x n_t -- feeds the metrics-vs-n_t figure
# below directly (this document keeps visualizations primary and does not
# print this as a standalone table).
aim1_type1_by_nt_s5 <- aim1_sweep_combined_s5 |>
  dplyr::filter(effect_value == 0) |>
  dplyr::group_by(term, n_t) |>
  dplyr::summarise(type_1_rate = mean(excludes_zero, na.rm = TRUE), .groups = "drop")

aim1_metrics_by_nt_s5 <- summarize_p2_metrics(aim1_all_s5, c("term", "n_t")) |>
  dplyr::left_join(aim1_type1_by_nt_s5, by = c("term", "n_t")) |>
  dplyr::relocate(type_1_rate, .after = power)

# Full effect-size sweep at every (n_p, n_t) cell (not just each term's own
# observed effect size) -- needed for the power/error-rate curve figures
# below, mirroring the original document's `aim1_metrics_by_effect`.
aim1_metrics_by_effect_s5 <- aim1_sweep_combined_s5 |>
  dplyr::group_by(term, effect_value, n_p, n_t) |>
  dplyr::summarise(
    power            = mean(excludes_zero, na.rm = TRUE),
    sign_reliability = mean(correct_sign, na.rm = TRUE),
    coverage         = mean(covers_truth, na.rm = TRUE),
    type_s_rate      = mean(sign_error[excludes_zero], na.rm = TRUE),
    type_m_ratio     = median(inflation_ratio[excludes_zero], na.rm = TRUE),
    .groups = "drop"
  )

# Ported from the original document's `make_aim1_curve_panel()`. The n_p
# grid is now an explicit argument (`x_grid`, defaulting to Scenario 5's
# 11-point grid) rather than hardcoded, since the densified grid has more
# than twice as many n_p values as the original 6-point grid.
make_aim1_curve_panel_s5 <- function(term_name, effect_grid, palette_name, metric_var, y_lab,
                                      axis_title = NULL, legend_title = NULL,
                                      include_zero = TRUE, zero_label = "no true effect",
                                      highlight_zero_orange = TRUE, ref_line = NULL, log_y = FALSE,
                                      y_limits = NULL, x_grid = n_p_grid_s5) {
  own_value <- own_effect[[term_name]]
  eff_grid <- if (include_zero) effect_grid else effect_grid[effect_grid != 0]

  # Skip the palette's palest 2 shades before interpolating -- the raw
  # Brewer palette's lightest color is too close to white to read as a line.
  pal <- grDevices::colorRampPalette(RColorBrewer::brewer.pal(9, palette_name)[3:9])(length(eff_grid))
  if (include_zero && highlight_zero_orange && 0 %in% eff_grid) {
    pal[eff_grid == 0] <- "#E69F00"  # Okabe-Ito orange, colorblind-friendly
  }
  effect_labels <- dplyr::case_when(
    eff_grid == 0 & include_zero ~ paste0(eff_grid, " (", zero_label, ")"),
    eff_grid == own_value        ~ paste0(eff_grid, " (observed)"),
    TRUE                         ~ as.character(eff_grid)
  )

  df <- aim1_metrics_by_effect_s5 |>
    dplyr::filter(term == term_name, effect_value %in% eff_grid) |>
    dplyr::mutate(
      effect_f = factor(effect_value, levels = eff_grid, labels = effect_labels),
      is_own   = effect_value == own_value
    )

  p <- ggplot(df, aes(x = n_p, y = .data[[metric_var]], color = effect_f, group = effect_f)) +
    geom_line(aes(linewidth = is_own)) +
    geom_point(aes(size = is_own)) +
    facet_wrap(~n_t, nrow = 1, labeller = label_both) +
    scale_x_log10(breaks = x_grid, labels = scales::label_comma()) +
    scale_color_manual(values = pal, name = legend_title %||% term_name) +
    scale_linewidth_manual(values = c(`FALSE` = 0.5, `TRUE` = 1.3), guide = "none") +
    scale_size_manual(values = c(`FALSE` = 1, `TRUE` = 2.2), guide = "none") +
    labs(x = axis_title, y = y_lab) +
    theme_minimal(base_size = 9) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 6),
      legend.position = "right"
    )

  if (!is.null(ref_line)) {
    p <- p + geom_hline(yintercept = ref_line, linetype = "dotted", color = "grey40", linewidth = 0.4)
  }
  if (log_y) {
    p <- p + scale_y_log10()
  }
  if (!is.null(y_limits)) {
    p <- p + coord_cartesian(ylim = y_limits)
  }
  p
}
```

## What are the implications for a range of design and analysis diagnostics (power, type-S/M errors, sign reliability, coverage, bias, RMSE) of potential variation in effect sizes and sample sizes?

Once all three effects (`zHLVA`, `zPCREFp`, and their interaction) are run
through the same effect-size x sample-size grid with the same diagnostic
taxonomy, the results answer questions that different
readers will care about differently:

- A reader planning a **new study** mainly wants **power** and **discovery
  rate**: how large a sample, on which margin, is needed to reliably detect
  a given effect?
- A reader trying to **interpret an existing significant finding** (their
  own or someone else's) mainly wants **Type S (sign) error rate** and
  **Type M (magnitude) ratio**: conditional on a result being "significant,"
  how likely is it to have the wrong sign, or to be substantially inflated?
- A reader thinking about **meta-scientific reproducibility** mainly wants
  the **discovery/replication rates** from Aim 2: if this study were run
  twice independently, how often would both runs agree?

The table below defines each diagnostic used throughout this document and
states exactly how it is computed here. All are derived from
`diagnose_coefs_p2()`'s per-replicate-fit columns (`estimate`, `std.error`,
`true_value`, and the **Wald** 95% CI `estimate +/- 1.96 * std.error`), then
aggregated by `summarize_p2_metrics()` over whichever grouping is in play
(typically `term` x `n_t`, pooled over `n_p` and the 100 replicates/cell).

| Diagnostic | Standard definition | Computed here as |
|---|---|---|
| `power` | P(reject H0 \| true effect is non-zero) -- probability of correctly detecting a real effect. | `mean(excludes_zero)`, restricted to rows at the term's non-zero (typically empirically-observed) effect size. `excludes_zero` = the 95% Wald CI does not contain 0. |
| `type_1_rate` | P(reject H0 \| true effect is exactly zero) -- the false-positive rate; nominally ~0.05 for a correctly calibrated 95% CI. | `mean(excludes_zero)`, restricted to rows at `effect_value == 0` on the same sweep. |
| `sign_reliability` | P(estimated sign matches true sign), unconditional on significance. | `mean(correct_sign)` over **all** fits (significant or not), where `correct_sign = sign(estimate) == sign(true_value)`. |
| `type_s_rate` | Type S (sign) error rate: P(estimated sign is wrong \| the result is significant) -- Gelman & Carlin's (2014)/Marek et al.'s (2022) usage. | `mean(sign_error[excludes_zero])`, i.e. computed only among rows where the CI excludes zero; `sign_error = excludes_zero & !correct_sign`. |
| `type_m_ratio` | Type M (magnitude/inflation) ratio: how much larger a *significant* estimate typically is than the true effect. | `median(inflation_ratio[excludes_zero])`, where `inflation_ratio = abs(estimate) / abs(true_value)` (undefined/`NA` when `true_value == 0`, since inflation relative to zero is not meaningful); the median (not mean) is used to resist the heavy right skew of this ratio at small `n_t`. |
| `median_bias` | Bias: the typical signed deviation of the estimate from the true value. | `median(bias)` over all fits, where `bias = estimate - true_value`. Median rather than mean, again for robustness to the occasional extreme estimate at small `n_t`. |
| `coverage` | Nominal 95% CI coverage: the proportion of CIs that actually contain the true value (should be ~0.95 if well-calibrated). | `mean(covers_truth)`, where `covers_truth = ci_lower <= true_value & true_value <= ci_upper`. |
| `rmse_raw` | Root-mean-squared error of the estimate around the true value. | `sqrt(mean(bias^2))` over all fits. Dominated by rare pathological fits at `n_t = 2`. |
| `rmse_trimmed` | RMSE, robustified against extreme outliers. | `sqrt(mean(pmin(bias^2, quantile(bias^2, 0.95))))` -- squared bias is winsorized at its own 95th percentile (within each group) before averaging, so a handful of extreme singular-fit estimates cannot dominate the summary. Preferred over `rmse_raw` for cross-cell comparison. |

Two things to note about this taxonomy as implemented: first, `type_s_rate`
and `type_m_ratio` are both computed *conditional on significance*
(`excludes_zero`), matching Marek et al.'s (2022) framing of "what does a
significant result look like at this sample size" -- they are not defined,
and are `NA`, for the (usually large) share of non-significant fits.
Second, all coverage/CI-based diagnostics here use **Wald** intervals
(`estimate +/- 1.96 * std.error`) rather than profile-likelihood or
bootstrap CIs, for computational speed across tens of thousands of `glmer`
fits -- this is a possible source of miscalibration in its own right,
distinct from the random-effects-identifiability mechanism discussed later
in this document, and is not separately investigated here.

```{r}
#| label: effect-intercept-grid-labels

# Build markdown-bolded, comma-separated strings of each swept grid, marking
# whichever grid point equals the empirically-observed value (the one value
# each term is held at in the power-surface/metrics-vs-x figures elsewhere
# in this document), so the table below shows the actual 5-point sweep
# design directly instead of collapsing it to a single number.
format_swept_grid <- function(grid, observed) {
  paste(ifelse(grid == observed, paste0("**", grid, "**"), as.character(grid)), collapse = ", ")
}

hlva_grid_label        <- format_swept_grid(b_hlva_grid, own_effect[["zHLVA"]])
pcref_grid_label        <- format_swept_grid(b_pcref_grid, own_effect[["zPCREFp"]])
interaction_grid_label  <- format_swept_grid(b_interaction_grid, own_effect[["zHLVA:zPCREFp"]])

intercept_grid_label <- intercept_s5 |>
  dplyr::distinct(target_accuracy, b_intercept) |>
  dplyr::arrange(target_accuracy) |>
  dplyr::mutate(
    lab = paste0(round(b_intercept, 3), " (", scales::percent(target_accuracy, accuracy = 1), " accuracy)"),
    lab = ifelse(target_accuracy == 0.89, paste0("**", lab, ", observed**"), lab)
  ) |>
  dplyr::pull(lab) |>
  paste(collapse = ", ")
```

| Input | Values used in this grid (**bold** = the empirically-observed value) | Source of the empirically-observed value |
|---|---|---|
| Sample-size grid (participants), `n_p` | `r paste(n_p_grid_s5, collapse = ", ")` | n/a -- this row is the sample-size axis itself, not an effect estimate; Scenario 5's densified, log-spaced grid |
| Sample-size grid (texts), `n_t` | `r paste(n_t_grid_s5, collapse = ", ")` | n/a -- Scenario 5's densified, log-spaced grid |
| Effect size, `zHLVA` | `r hlva_grid_label` | Real accuracy-outcome data, via the Phase 1 empirical-resampling analysis's (`sampling-variability-corridor-of-stability.qmd`) full-data reduced-model fit (Step 0) |
| Effect size, `zPCREFp` | `r pcref_grid_label` | Same (Phase 1 Step 0/Step 5); `glmer` and `brms` agree to within 0.094-0.098, of which 0.095 is used as the sweep's center |
| Effect size, `zHLVA:zPCREFp` | `r interaction_grid_label` | Same (Phase 1 Step 0/Step 5); `glmer` and `brms` agree on 0.054 |
| Baseline log-odds intercept, `b_intercept` | `r intercept_grid_label` | Real accuracy-outcome data (project's actual mean accuracy is ~89%); swept separately from the three effect sizes above, Aim-1-only, in the "`b_intercept` sweep" section further below -- not crossed with the effect-size sweeps |
| Variance component, `sd(ResponseId)` | held fixed at 0.69 (not swept in this document's Aim 1/2 grids -- see Phase 2's Aim 4 for a dedicated variance-component sweep) | Full core model (`dev.all.studies.noratings.ptqs.normal.0.1.normal.0.1.lkj.2.study.ri.rds`), extracted in `phase2-effect-variance-misspecification.qmd`'s `extract-aim5-reference` chunk -- differs from Phase 1's simpler reduced-model estimate (~0.61-0.62) because the full model's 26 fixed effects absorb some of that between-participant variance |
| Variance component, `sd(text.id)` | held fixed at 0.41 (not swept here -- see Aim 4) | Same source as above -- differs from Phase 1's reduced-model estimate (~0.70-0.72), same reason |
| RE-misspecification contrast | a "correct" `(1\|ResponseId)+(1\|text.id)` model fit alongside a "misspecified" `(1\|ResponseId)`-only model, at every cell | Design choice carried over from Phase 1's real-data resampling analysis |

"Phase 1" throughout this table refers to this project's accuracy-outcome
empirical **resampling** analysis
(`sampling-variability-corridor-of-stability.qmd`, condensed in
`accuracy-resampling-based-analyses.qmd`) -- the real-data study that this
document's parametric simulations (part of the separate "Phase 2" program)
are designed to generalize beyond. Phase 1's full-data model fit is where
every empirically-observed value above comes from.

The three effect sizes are **not** held at a single value anywhere in this
document's simulations -- as the table now shows explicitly, Aim 1 sweeps
each term across its own 5-point grid (0 through roughly 2-3x the
observed value), with the empirically-observed value included as one point
among five, precisely so that "what if the real effect were smaller/larger
than what this study happened to find" can be answered directly rather than
only at the one value seen in the real data. What Scenario 5 changed,
relative to the design that produced the original 6x5 grid, is only the
**sample-size** grid's density (11 x 9 vs. 6 x 5 values, log-spaced) and its
replicate count (100 vs. 50 per cell) -- the same three 5-point effect-size
grids and the same two fixed variance components are reused unchanged. The
baseline intercept was, by contrast, held fixed (at the observed ~89%
accuracy value) throughout Aim 1/2's main effect-size sweeps above, and is
only varied in a separate, Aim-1-only sweep (added to the table here for
completeness, detailed in its own section below). Reusing the same
effect-size grids and variance-component values means this document's
results remain directly comparable to Phase 1's real-data panels at any
shared `(n_p, n_t, effect_value)` combination -- demonstrated concretely in
the grid-density-comparability check later in this document.

::: {.callout-important}
## Note: the real core model's random-effects structure is richer than this DGP

The real fitted core model's random-effects terms let participants vary in
their *slopes* for all five text-level predictors (including `zPCREFp`), and
let texts vary in their *slopes* for the participant-level predictors
(including `zHLVA`), each correlated with its own grouping factor's
intercept. This grid uses a simpler, random-**intercept**-only "correct"
data-generating process -- it answers a coarser question (omitting a
grouping factor's clustering altogether) and should not be read as testing
the real model's richer slope/covariance structure. That richer structure
is the subject of a separate future document.
:::

::: {.callout-note}
## Forewarning: the analyses hold baseline accuracy fixed, at first

The heat maps and curve figures below holds baseline accuracy at this project's real observed value (`b_intercept = 2.093`, ~89% mean accuracy). But logistic regression's  non-collapsibility means baseline
accuracy is not a neutral modeling choice, and it will vary across tasks,
samples, and studies. A dedicated `b_intercept` sweep, later in this
document, tests this directly across the full grid at four target
accuracies (50%, 75%, 89% -- observed, 95%).
:::

### Aim 1 -- power surface and error surfaces

The figures below show 2-D surfaces, one per diagnostic, for each of the
three effects, giving each diagnostic's value as it varies over the full
`n_p` x `n_t` sample-size grid while holding each effect's size fixed at
its own empirically-observed value from the real study.

```{r}
#| label: fig-aim1-power-surface-s5
#| fig-width: 11
#| fig-height: 4.3
#| fig-cap: "Scenario 5: power surface across the full densified (n_p, n_t) grid, at each term's own empirically-observed effect size."

aim1_surface_s5 <- aim1_all_s5 |>
  dplyr::group_by(term, n_p, n_t) |>
  dplyr::summarise(power = mean(excludes_zero, na.rm = TRUE), .groups = "drop")

plot_power_heatmap(aim1_surface_s5, fill_var = "power", fill_label = "Power")
```

`zPCREFp`'s power at `n_t=2` meets or exceeds its power at every larger
`n_t` at the same `n_p`, in **9 of 11** `n_p` cells (only the two smallest,
`n_p=25` and `n_p=38`, break the pattern) -- a reversal of the usual
"more text data, more power" expectation, and a robust property of this
design rather than an artifact of any particular sample point.

```{r}
#| label: fig-aim1-error-surfaces-s5
#| fig-width: 15
#| fig-height: 6.5
#| fig-cap: "Scenario 5: the remaining diagnostics from the fuller metrics table, each as its own (n_p, n_t) surface. Color scheme is direction-consistent viridis for the sequential panels (light/yellow = the diagnostic looks good, dark blue/purple = it looks bad), with `direction` flipped for Type I rate, Type S rate, and Type M ratio so that their low-is-good polarity reads the same way as sign reliability's and coverage's high-is-good polarity; Type I rate, sign reliability, Type S rate, and coverage share the [0, 100%] color scale, Type M ratio uses a squished scale capped at 10. Type I rate additionally uses cividis rather than plain viridis, distinguishing it as the one panel computed at the true-null (effect = 0) slice rather than each term's own observed effect size. Median bias uses scico's diverging \"vik\" palette (dark blue/red = bad, white = good), centered at zero."

aim1_surfaces_extra_s5 <- aim1_all_s5 |>
  dplyr::group_by(term, n_p, n_t) |>
  dplyr::summarise(
    sign_reliability = mean(correct_sign, na.rm = TRUE),
    type_s_rate      = mean(sign_error[excludes_zero], na.rm = TRUE),
    type_m_ratio     = median(inflation_ratio[excludes_zero], na.rm = TRUE),
    median_bias      = median(bias, na.rm = TRUE),
    coverage         = mean(covers_truth, na.rm = TRUE),
    .groups = "drop"
  )

aim1_type1_surface_s5 <- aim1_sweep_combined_s5 |>
  dplyr::filter(effect_value == 0) |>
  dplyr::group_by(term, n_p, n_t) |>
  dplyr::summarise(type_1_rate = mean(excludes_zero, na.rm = TRUE), .groups = "drop")

aim1_surfaces_extra_s5 <- aim1_surfaces_extra_s5 |>
  dplyr::left_join(aim1_type1_surface_s5, by = c("term", "n_p", "n_t"))

bias_limit_s5 <- max(abs(aim1_surfaces_extra_s5$median_bias), na.rm = TRUE)

# Type I rate is the one panel here computed at effect_value == 0 (the
# true-null slice of the sweep) -- every other panel (sign reliability,
# Type S rate, coverage, Type M ratio, median bias) is computed at each
# term's own non-zero, empirically-observed effect size. Plain viridis for
# all six implied they're all "the same kind of thing," which understates
# the difference: Type I rate answers "how often does this design falsely
# detect an effect that isn't there," not "how well does it recover an
# effect that is." `cividis` (still part of the viridis palette family --
# same perceptually-uniform, colorblind-safe design lineage, so still
# "analogous") swaps viridis's purple-green-yellow hue range for a
# distinct blue-yellow one with no green/purple, flagging this one panel
# as a different kind of diagnostic without abandoning the viridis
# convention entirely. Matches the same treatment applied to the
# equivalent panel in
# synthetic-data-observed-effect-varying-sample-impacts-diagnostics-heatmaps-funnels-figure.qmd.
panel_type1_s5 <- plot_power_heatmap(aim1_surfaces_extra_s5, fill_var = "type_1_rate", fill_label = "Type I\nrate",
                                       lower_is_better = TRUE, viridis_option = "cividis") +
  labs(title = "Type I (false-positive) rate")
panel_signrel_s5 <- plot_power_heatmap(aim1_surfaces_extra_s5, fill_var = "sign_reliability", fill_label = "Sign\nreliability") +
  labs(title = "Sign reliability")
panel_types_s5 <- plot_power_heatmap(aim1_surfaces_extra_s5, fill_var = "type_s_rate", fill_label = "Type S\nrate",
                                       lower_is_better = TRUE) +
  labs(title = "Type S (sign) error rate")
panel_coverage_s5 <- plot_power_heatmap(aim1_surfaces_extra_s5, fill_var = "coverage", fill_label = "Coverage") +
  labs(title = "95% CI coverage")
panel_typem_s5 <- plot_power_heatmap(aim1_surfaces_extra_s5, fill_var = "type_m_ratio", fill_label = "Type M\nratio",
                                       limits = c(0, 10), label_format = "number", squish = TRUE, lower_is_better = TRUE) +
  labs(title = "Type M (magnitude) ratio")
panel_bias_s5 <- plot_power_heatmap(aim1_surfaces_extra_s5, fill_var = "median_bias", fill_label = "Median\nbias",
                                      limits = c(-bias_limit_s5, bias_limit_s5), label_format = "number", diverging = TRUE) +
  labs(title = "Median bias")

patchwork::wrap_plots(panel_type1_s5, panel_signrel_s5, panel_types_s5, panel_coverage_s5,
                       panel_typem_s5, panel_bias_s5, ncol = 2)
```

Three findings stand out from this figure:

1. **Coverage collapse for `zPCREFp` at `n_t=2`**: nominal 95% coverage
   falls toward single digits at the largest `n_p` values (994, 1497) --
   more participant data makes an already-miscalibrated text-margin
   estimate more confidently wrong, not less.
2. **`zHLVA` remains comparatively well-calibrated everywhere**: its Type I
   rate stays in the 2-11% range and its sign reliability saturates near
   100% by `n_t >= 6` -- a faster and more complete stabilization than
   either text-margin term reaches anywhere on this grid.
3. **Sign reliability, Type S rate, and Type M ratio all decline/rise
   monotonically with `n_t`** in this view -- visibly smoother, gentler
   gradients across the grid than the sharp reversal visible in the power
   surface above (and confirmed directly in the metrics-vs-`n_t` figure
   below). The paradox is concentrated in the raw significance-rate
   diagnostics, not the full error-taxonomy bundle.

::: callout-note
`zPCREFp`'s and the interaction's non-monotonic **power** pattern at small
`n_t` -- the "underpowered paradox" -- is examined further below. Coverage
shows a related but **monotonic** collapse, not the same non-monotonic
dip-and-recover signature as power (see finding 3 above).
:::

### Aim 1 -- metrics vs. participant sample size and vs. text sample size

This section shows two one-axis-at-a-time views of power, sign reliability, Type S rate, and Type M ratio: first pooled over `n_t` (shown vs. `n_p`), and second pooled over `n_p` (shown vs. `n_t`).
Diagnostic values are computed by pooling across replicate studies varying on the sampling dimension not shown.
In each plot, while diagnostic values are shown varying in relation to participant or text sample sizes, each term's effect size is held at the empirically-observed effect size.

These plots are followed by a full set of effect-size-sweep
curve figures, analysing diagnostic value variation for each effect size, participant sample size, and text sample size.

We show first the participant-margin variation over key diagnostics, pooled over all 9 `n_t` values instead of
all 11 `n_p` values. 

```{r}
#| label: fig-aim1-metrics-vs-np-s5
#| fig-cap: "Scenario 5: power, sign reliability, Type S error rate, and Type M exaggeration ratio vs. participant sample size (n_p), at each term's own empirically-observed effect size, pooled across all 9 text sample sizes and 100 replicates/cell."

plot_metrics_vs_x(
  aim1_metrics_by_np_s5, x_var = "n_p",
  metrics = c("power", "sign_reliability", "type_s_rate", "type_m_ratio"),
  metric_labels = aim1_metric_labels,
  x_lab = "Participant sample size (n_p, log scale)",
  x_breaks = n_p_grid_s5
)
```

`zHLVA`'s curves are the best-behaved of the three,
approaching ceiling/floor values (power and sign reliability near 1, Type
S/M near 0) by moderate `n_p` -- consistent with it being the
participant-margin-limited effect. `zPCREFp` and the interaction are
pooled here across text sample sizes spanning the pathological `n_t = 2`
through the well-behaved `n_t = 136`, so the implications of smaller text samples are diluted rather than isolated in this view: the `n_t`-axis figure below, and the full `(n_p, n_t)` surfaces or curves plots are more informative.

Next we show the text-margin variation over key diagnostics, pooled over all 11 `n_p` values instead of 9 `n_t` values. 

```{r}
#| label: fig-aim1-metrics-vs-nt-s5
#| fig-cap: "Scenario 5: power, sign reliability, Type S error rate, and Type M exaggeration ratio vs. text sample size (n_t), at each term's own empirically-observed effect size, pooled across all 11 participant sample sizes and 100 replicates/cell."

plot_metrics_vs_x(
  aim1_metrics_by_nt_s5, x_var = "n_t",
  metrics = c("power", "sign_reliability", "type_s_rate", "type_m_ratio"),
  metric_labels = aim1_metric_labels,
  x_lab = "Text sample size (n_t, log scale)",
  x_breaks = n_t_grid_s5
)
```

This is the one-axis (pooled-over-`n_p`) counterpart of the power surface and error surfaces above, and puts Marek et al.'s (2022) "underpowered paradox" and its error-taxonomy consequences on a single set of panels for the first time in
this document. `zPCREFp`'s sign reliability, Type S rate, and Type M ratio are all worst at `n_t = 2` and improve monotonically from there, while its **power** dips non-monotonically: power falls from 0.567 at `n_t=2` to a minimum of **0.186 at `n_t=16`**, then rises to 0.447 by `n_t=136` -- a
genuine U-shape, directly visible in the power panel above rather than inferred from a handful of grid points. The same panel makes clear that this non-monotonicity is specific to the *raw significance rate* (power, and by extension the Type I rate and, in Aim 2 below, the replication rate): the other three diagnostics in this same figure -- sign reliability,
Type S rate, and Type M ratio -- simply improve, if slowly, as `n_t` grows.

#### Aim 1 -- the underpowered-paradox mechanism: individual-estimate funnel plots at finer text-sample resolution

`zPCREFp`'s power shows the "underpowered paradox" signature directly:
power is *higher* at `n_t = 2` (0.567) than at `n_t = 10` (0.197) -- not
because small samples are better powered, but because the noisiest, most
degenerate cell produces wildly inflated estimates that cross the
significance threshold more often, while being wrong about the sign
roughly 40% of the time among those "significant" fits. A researcher who
only sees "significant, p < .05" from a 2-text study cannot distinguish
this from a well-powered result.

The summary diagnostics above (power, sign reliability, Type S/M rates) are
all aggregates over 100 replicate fits per cell. It can be easier to see
why the underpowered paradox happens by looking at the individual replicate
estimates themselves, rather than only their summary statistics. The three
panels below show every individual estimate from Aim 1's fits at
`n_p = 1497` -- the largest participant sample size in the grid, held
constant across every row and column so that the funnel shown reflects the
text margin (`n_t`) specifically, not participant-margin noise -- for
`zHLVA` (top), `zPCREFp` (middle), and the `zHLVA:zPCREFp` interaction
(bottom): one dot per replicate fit, one column per simulated true effect
size, one row per text sample size. `zHLVA` is included here for the same
reason the tables/heatmaps above already treat all three terms
symmetrically: since the whole point of this document is to compare all
three effects on the same footing, not just the interaction and its
text-level counterpart, the mechanism-level view does the same. This uses
only already-computed replicate estimates
(`results_phase2/aim1_scenario5_hlva.rds`) -- no new simulation.

```{r}
#| label: fig-aim1-funnel-s5
#| fig-width: 12
#| fig-height: 20
#| fig-cap: "Scenario 5: individual estimates (one dot per replicate fit, n_p held at its largest grid value, 1497) for zHLVA (top), zPCREFp (middle), and the interaction (bottom), across all 9 text sample sizes (rows) and each term's 5 simulated effect sizes (columns). Dashed line = true value; dark red (vik's diverging \"bad\" pole) = 95% CI excludes zero, matching the same red used to flag extremes in the bias heatmap above; dark purple (viridis's \"bad\"/low pole, used here as a muted color) = not significant."

illustrative_n_p_s5 <- max(n_p_grid_s5)

make_aim1_distribution_panel <- function(data, term_name, x_clip, illustrative_n_p, n_t_levels,
                                          show_x_axis = TRUE) {
  df <- data |>
    dplyr::filter(term == term_name, n_p == illustrative_n_p) |>
    dplyr::mutate(
      n_t_f = factor(n_t, levels = n_t_levels),
      significant = ifelse(excludes_zero, "Significant (95% CI excludes 0)", "Not significant"),
      estimate_clipped = pmax(pmin(estimate, x_clip), -x_clip)
    )

  ggplot(df, aes(x = estimate_clipped, y = n_t_f, color = significant, fill = significant)) +
    ggdist::stat_dots(side = "both", scale = 0.85, alpha = 0.55,
                       binwidth = ggplot2::unit(c(1.3, 2.2), "mm"), overflow = "compress") +
    geom_vline(aes(xintercept = effect_value), linetype = "dashed", color = "black", linewidth = 0.5) +
    facet_wrap(~effect_value, nrow = 1, labeller = as_labeller(function(x) paste0("true = ", x))) +
    scale_x_continuous(breaks = c(-x_clip, 0, x_clip)) +
    # Colors drawn from the same finalized palette family as the heatmaps
    # above, for document-wide consistency: vik's dark-red pole (the
    # diverging scale's "bad" extreme) flags significant estimates, and
    # viridis's dark pole (the sequential scale's "bad"/low end, used here
    # simply as a muted, receding color) marks non-significant ones.
    scale_color_manual(values = c("Significant (95% CI excludes 0)" = "#590007", "Not significant" = "#440154")) +
    scale_fill_manual(values = c("Significant (95% CI excludes 0)" = "#590007", "Not significant" = "#440154")) +
    labs(x = if (show_x_axis) paste0(term_name, " estimate (clipped to [-", x_clip, ", ", x_clip, "])") else NULL,
         y = "n_t", color = NULL, fill = NULL) +
    theme_bw(base_size = 10) +
    theme(panel.border = element_rect(color = "grey35", fill = NA, linewidth = 0.7),
          panel.spacing = unit(1.1, "lines"), strip.background = element_rect(fill = "grey85", color = "grey35"),
          strip.text = element_text(face = "bold", size = 8.5), panel.grid.minor = element_blank(),
          panel.grid.major.y = element_blank(), legend.position = "bottom")
}

panel_dist_hlva_s5 <- make_aim1_distribution_panel(
  aim1_hlva_s5, "zHLVA", x_clip = 2, illustrative_n_p = illustrative_n_p_s5,
  n_t_levels = n_t_grid_s5, show_x_axis = FALSE
) + labs(title = "zHLVA (participant-level main effect)")

panel_dist_pcref_s5 <- make_aim1_distribution_panel(
  aim1_pcref_s5, "zPCREFp", x_clip = 2, illustrative_n_p = illustrative_n_p_s5,
  n_t_levels = n_t_grid_s5, show_x_axis = FALSE
) + labs(title = "zPCREFp (text-level main effect)")

panel_dist_interaction_s5 <- make_aim1_distribution_panel(
  aim1_interaction_s5, "zHLVA:zPCREFp", x_clip = 1.5, illustrative_n_p = illustrative_n_p_s5,
  n_t_levels = n_t_grid_s5, show_x_axis = TRUE
) + labs(title = "zHLVA:zPCREFp (interaction)")

(panel_dist_hlva_s5 / panel_dist_pcref_s5 / panel_dist_interaction_s5) +
  patchwork::plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

Five things are visible here that the summary tables/plots only imply.

First, the funnel shape itself, present in **all three rows**: each cloud
of dots is wide at the bottom (`n_t = 2`) and collapses to a tighter
cluster around the dashed true-value line moving up the grid -- the
textbook "wide at small *N*, narrow at large *N*" funnel, for every true
effect size simulated, for every term (participant-level or text-level).

Second, and this is the paradox in its most direct form: in `zPCREFp`'s
leftmost column (`true = 0`, i.e. **no true effect at all**), a substantial
share of the `n_t = 2` dots are still colored dark red -- individually
indistinguishable, in a real analysis, from a genuine discovery. Those red
dots at `n_t = 2` are false positives purely from noise, and they are just
as "significant" (95% CI excludes zero) as the true positives further
right. A significance threshold cannot tell these apart; only a larger
text sample -- more rows lower down turning into fewer, tighter, more-
often-correctly-colored rows higher up -- can.

Third, `zHLVA`'s top row makes the *contrast* with the text-level terms
directly visible: at `n_p = 1497, n_t = 2` its power is
`r scales::percent(aim1_surface_s5 |> dplyr::filter(term == "zHLVA", n_p == max(n_p_grid_s5), n_t == 2) |> dplyr::pull(power), accuracy = 1)`,
its funnel is comparatively narrow already, and its Type I rate stays much
closer to nominal at every `n_t` (see the error-surfaces figure above). The
instability visible in the `zPCREFp` and interaction rows is not a generic
property of "estimating anything at `n_t = 2`" -- it is specific to terms
whose identification depends on the text margin.

Fourth, comparing the `zPCREFp` and interaction rows directly answers
whether the interaction's funnel looks like `zPCREFp`'s: **not really**. At
`n_p = 1497, n_t = 2`, `zPCREFp`'s power is
`r scales::percent(aim1_surface_s5 |> dplyr::filter(term == "zPCREFp", n_p == max(n_p_grid_s5), n_t == 2) |> dplyr::pull(power), accuracy = 1)`
while the interaction's is only
`r scales::percent(aim1_surface_s5 |> dplyr::filter(term == "zHLVA:zPCREFp", n_p == max(n_p_grid_s5), n_t == 2) |> dplyr::pull(power), accuracy = 1)`
-- visible directly in the interaction row as mostly grey (not red) dots at
this cell, across every column including its own empirically-observed
effect size. Where `zPCREFp`'s problem at `n_t = 2` is "too many
confidently-wrong answers," the interaction's is closer to "usually no
answer at all, and on the rare occasions there is one, it is not obviously
more trustworthy than `zPCREFp`'s" -- the same underlying text-margin
scarcity, but manifesting as a power problem more than a false-discovery
problem for this particular term.

Fifth, two things are visible only because of this document's finer,
9-row grid: (a) **`zHLVA` visibly stabilizes by `n_t=6`**, several rows
earlier than either text-margin term; (b) **`zPCREFp`'s miscalibration
decays gradually, not abruptly** -- its `true=0` column stays visibly noisy
through `n_t=3` and `n_t=6` and only approaches nominal by `n_t=28`-`47`.
As throughout, exact red/grey proportions are unreliable to read directly
off a heavily-overplotted static image at this density; the quantitative
figures above are the trustworthy source, and this plot is best read for
the qualitative funnel-narrowing pattern and the relative-speed contrast
between terms.

##### Why can an estimate be "significant" when the true effect is exactly zero?

This is the question the `true = 0` column above raises most sharply, and
it is worth answering mechanistically rather than leaving it as a
statistical curiosity. Two different things are going on, and it is
important to distinguish them.

**It is not simply the ordinary 5% false-positive rate.** A correctly
calibrated 95% CI is *expected* to exclude a true value of zero 5% of the
time, by construction -- that is what "95%" means. But the observed rate at
`n_t = 2` is far above that, as the table below shows.

```{r}
#| label: fig-aim1-singular-fit-diagnostic-s5

null_nt2_diagnostic_s5 <- aim1_pcref_s5 |>
  dplyr::filter(term == "zPCREFp", n_t == 2, effect_value == 0)

cat("Fit status, all", nrow(null_nt2_diagnostic_s5),
    "replicates (n_t = 2, true zPCREFp effect = 0):\n")
table(null_nt2_diagnostic_s5$status)

null_nt2_diagnostic_s5 |>
  dplyr::group_by(excludes_zero) |>
  dplyr::summarise(
    n = dplyr::n(),
    `median |estimate|` = round(median(abs(estimate), na.rm = TRUE), 3),
    `median std. error` = round(median(std.error, na.rm = TRUE), 3),
    .groups = "drop"
  )
```

Across these 1,100 replicates (11 `n_p` values x 100 replicates), 56.1% are
"significant" under a true null -- roughly 11x the nominal 5% rate.
Something is actively miscalibrated here, not merely unlucky.

**The mechanism is a random-effects identifiability breakdown, not
sampling bad luck on the covariate itself.** With only `n_t = 2` texts,
there is exactly one degree of freedom available to describe how the two
sampled texts differ from each other -- two points determine a line
perfectly, with nothing left over to check whether that line is
trustworthy. The "correct" model asks `glmer` to split any such
between-text difference into two pieces: a text-level random intercept
(`(1 | text_id)`, capturing idiosyncratic, unexplained differences between
texts) and the fixed `zPCREFp` slope (capturing the cohesion effect
specifically). Estimating a random-intercept *variance* from only 2 groups
is close to impossible -- there is essentially no information to
distinguish "these two texts differ a lot by chance" from "these two texts
differ a lot systematically." In practice, `lme4` resolves this by pinning
the text-level variance at (or extremely near) its zero boundary -- a
**singular fit**: 1,099 of the 1,100 replicates above converge to
`converged_singular` (the remaining one to `error`) -- the text-level
random-intercept variance collapses to zero almost every time `n_t = 2`,
regardless of how many participants are simulated.

Once that variance is pinned at zero, the model has nowhere else to put
the two sampled texts' chance difference except into the only fixed-effect
predictor that differs between them -- `zPCREFp`. This inflates the
fixed-effect point estimate (median |estimate| = 0.701 among the
"significant" fits, versus a true value of exactly 0) and simultaneously
produces an overconfident, too-narrow Wald standard error (median SE =
0.149 among the "significant" fits, versus 0.287 among the fits that
correctly stayed non-significant) -- because the SE formula implicitly
assumes the random-effects variance structure has been reliably estimated,
an assumption that fails precisely at this boundary. Inflated estimate +
deflated SE together push a large share of purely-noise-driven fits across
the 95%-CI-excludes-zero line.

In short: at `n_t = 2`, the model is not really testing "does cohesion
predict comprehension" -- it is testing "do these two specific,
randomly-sampled texts differ," and then attributing 100% of whatever
difference it finds to cohesion, because there is no other place left for
that difference to go. This is a more extreme version of the same family
of problem documented elsewhere in this project as random-effects
misspecification (Aim 3's omitted-`text.id` bias, and Phase 1's finding
that pseudoreplication introduces genuine bias, not just imprecision) --
here the grouping factor is nominally *in* the model formula, but with only
2 groups it is statistically unidentifiable in practice, which behaves much
like omitting it.

### Individual effect-size sweep: power and error-rate curves by (n_p, n_t, effect size)

The two figures above pool across five simulated effect-size values per
term (using only each term's own empirically-observed value). The curve
figures below instead show **all five effect-size values per term
separately**, so the full effect-size x `n_p` x `n_t` dependence is
visible rather than collapsed to one slice. Strictly, **power** is the
probability of correctly detecting an effect that truly exists in the
population -- it is only a meaningful concept when the true effect is
non-zero. Rather than color the `effect size = 0` case as if it answered
the same question as its non-zero neighbors, it is drawn in a distinct,
colorblind-friendly bright orange, held constant across all three term
rows, in every figure below where it is shown at all.

```{r}
#| label: fig-aim1-power-curves-s5
#| fig-width: 18
#| fig-height: 9.5
#| fig-cap: "Scenario 5: classic power curves -- P(significant) vs. participant sample size (n_p, log scale), one row per term, faceted by text sample size (n_t, 9 columns) across the densified grid. Each row uses its own sequential color palette (darker = larger effect size); the thicker line/larger points mark each term's empirically-observed effect size; the bright orange line/points are the Type I (false-positive) rate, i.e. the effect size = 0 case."

panel_hlva_pc_s5 <- make_aim1_curve_panel_s5(
  "zHLVA", b_hlva_grid, "Greens", metric_var = "power", y_lab = "Power",
  legend_title = "zHLVA effect size", zero_label = "no true effect -- Type I rate",
  y_limits = c(0, 1)
)
panel_pcref_pc_s5 <- make_aim1_curve_panel_s5(
  "zPCREFp", b_pcref_grid, "Blues", metric_var = "power", y_lab = "Power",
  legend_title = "zPCREFp effect size", zero_label = "no true effect -- Type I rate",
  y_limits = c(0, 1)
)
panel_interaction_pc_s5 <- make_aim1_curve_panel_s5(
  "zHLVA:zPCREFp", b_interaction_grid, "Purples", metric_var = "power", y_lab = "Power",
  axis_title = "Participant sample size (n_p, log scale)", legend_title = "Interaction effect size",
  zero_label = "no true effect -- Type I rate", y_limits = c(0, 1)
)

panel_hlva_pc_s5 / panel_pcref_pc_s5 / panel_interaction_pc_s5
```

At the densified resolution, the `zPCREFp` row's `n_t = 2`/`n_t = 3`
columns show the bright orange Type I (false-positive) rate curve staying
non-trivial and non-monotonic across the entire `n_p` range, exactly as
the power surface and metrics-vs-`n_t` figure above already established,
but here the transition across the 7 intermediate `n_t` columns (6 through 80) is
visible directly rather than inferred from two endpoints -- the orange
curve's flattening toward zero is gradual, not a step change at any single
`n_t`. `zHLVA`'s row again saturates fastest and most completely across
every `n_t` column; the interaction's row sits between the two, needing
growth on both margins simultaneously to approach ceiling power.

```{r}
#| label: fig-aim1-sign-reliability-curves-s5
#| fig-width: 18
#| fig-height: 9.5
#| fig-cap: "Scenario 5: sign reliability -- P(estimated sign matches true sign) -- vs. participant sample size (n_p, log scale), one row per term, faceted by text sample size (n_t). Layout and color convention match the power curves above, except the true-null (effect = 0) point is omitted: sign reliability has no meaningful value there, since sign(true_value) = 0 makes `correct_sign` essentially never TRUE by construction, not because of instability."

panel_hlva_src_s5 <- make_aim1_curve_panel_s5(
  "zHLVA", b_hlva_grid, "Greens", metric_var = "sign_reliability", y_lab = "Sign reliability",
  legend_title = "zHLVA effect size", include_zero = FALSE, y_limits = c(0, 1)
)
panel_pcref_src_s5 <- make_aim1_curve_panel_s5(
  "zPCREFp", b_pcref_grid, "Blues", metric_var = "sign_reliability", y_lab = "Sign reliability",
  legend_title = "zPCREFp effect size", include_zero = FALSE, y_limits = c(0, 1)
)
panel_interaction_src_s5 <- make_aim1_curve_panel_s5(
  "zHLVA:zPCREFp", b_interaction_grid, "Purples", metric_var = "sign_reliability", y_lab = "Sign reliability",
  axis_title = "Participant sample size (n_p, log scale)", legend_title = "Interaction effect size",
  include_zero = FALSE, y_limits = c(0, 1)
)

panel_hlva_src_s5 / panel_pcref_src_s5 / panel_interaction_src_s5
```

`zPCREFp`'s row shows the same "flat, coin-flip-adjacent" pattern at
`n_t = 2`-`3` documented in the metrics-vs-`n_t` figure above, now visible
across the whole effect-size sweep rather than just its own observed value: even
the largest simulated effect (0.35) stays well short of the near-ceiling
sign reliability it reaches once `n_t >= 10`. `zHLVA` again saturates
earliest and most completely of the three terms, independent of `n_t`.

```{r}
#| label: fig-aim1-coverage-curves-s5
#| fig-width: 18
#| fig-height: 9.5
#| fig-cap: "Scenario 5: 95% CI coverage -- P(the Wald CI contains the true value) -- vs. participant sample size (n_p, log scale), one row per term, faceted by text sample size (n_t). Dotted horizontal line marks nominal 95% coverage. Coverage is well-defined at effect = 0 (a CI can cover a true value of exactly zero), so the true-null line is not a qualitatively different question the way power's Type I rate is -- but it is still drawn in the same bright orange as the power curves' Type I line, purely so it is visually distinguishable from the palest of the four non-zero effect-size lines, which would otherwise be nearly indistinguishable shades of the same sequential palette."

panel_hlva_covc_s5 <- make_aim1_curve_panel_s5(
  "zHLVA", b_hlva_grid, "Greens", metric_var = "coverage", y_lab = "95% CI coverage",
  legend_title = "zHLVA effect size", highlight_zero_orange = TRUE, zero_label = "no true effect",
  ref_line = 0.95, y_limits = c(0, 1)
)
panel_pcref_covc_s5 <- make_aim1_curve_panel_s5(
  "zPCREFp", b_pcref_grid, "Blues", metric_var = "coverage", y_lab = "95% CI coverage",
  legend_title = "zPCREFp effect size", highlight_zero_orange = TRUE, zero_label = "no true effect",
  ref_line = 0.95, y_limits = c(0, 1)
)
panel_interaction_covc_s5 <- make_aim1_curve_panel_s5(
  "zHLVA:zPCREFp", b_interaction_grid, "Purples", metric_var = "coverage", y_lab = "95% CI coverage",
  axis_title = "Participant sample size (n_p, log scale)", legend_title = "Interaction effect size",
  highlight_zero_orange = TRUE, zero_label = "no true effect", ref_line = 0.95, y_limits = c(0, 1)
)

panel_hlva_covc_s5 / panel_pcref_covc_s5 / panel_interaction_covc_s5
```

`zPCREFp`'s coverage curves reproduce the coverage-collapse pattern already
identified in the error-surfaces figure earlier -- most severe at
`n_t = 2`-`3`, growing *worse* as `n_p` increases rather than better. The
now-visible orange true-null line tracks closely with the other four
effect-size lines at every `n_t`, confirming the collapse is not specific
to the one non-zero effect size shown in the earlier surface figure -- it
is a general property of the small-`n_t` random-effects-identifiability
breakdown, present whether or not there is a true effect to detect.
`zHLVA` and, to a lesser extent, the interaction stay closer to the
nominal 95% line throughout.

```{r}
#| label: fig-aim1-type-s-curves-s5
#| fig-width: 18
#| fig-height: 9.5
#| fig-cap: "Scenario 5: Type S (sign) error rate -- P(wrong sign | significant) -- vs. participant sample size (n_p, log scale), one row per term, faceted by text sample size (n_t). The true-null point is omitted, as for sign reliability above: conditioning on significance at a true effect of exactly zero makes Type S trivially ~100% by construction, not informative of miscalibration."

panel_hlva_tsc_s5 <- make_aim1_curve_panel_s5(
  "zHLVA", b_hlva_grid, "Greens", metric_var = "type_s_rate", y_lab = "Type S rate",
  legend_title = "zHLVA effect size", include_zero = FALSE, y_limits = c(0, 1)
)
panel_pcref_tsc_s5 <- make_aim1_curve_panel_s5(
  "zPCREFp", b_pcref_grid, "Blues", metric_var = "type_s_rate", y_lab = "Type S rate",
  legend_title = "zPCREFp effect size", include_zero = FALSE, y_limits = c(0, 1)
)
panel_interaction_tsc_s5 <- make_aim1_curve_panel_s5(
  "zHLVA:zPCREFp", b_interaction_grid, "Purples", metric_var = "type_s_rate", y_lab = "Type S rate",
  axis_title = "Participant sample size (n_p, log scale)", legend_title = "Interaction effect size",
  include_zero = FALSE, y_limits = c(0, 1)
)

panel_hlva_tsc_s5 / panel_pcref_tsc_s5 / panel_interaction_tsc_s5
```

`zPCREFp` again shows the familiar flat, margin-dependent pattern at small
`n_t`, mirroring its own sign-reliability curves (Type S rate is,
definitionally, close to `1 - sign_reliability` among significant fits).
Both main effects' Type S rates fall toward zero by moderate `n_t`; `zHLVA`
remains the most consistently well-behaved row.

```{r}
#| label: fig-aim1-type-m-curves-s5
#| fig-width: 18
#| fig-height: 9.5
#| fig-cap: "Scenario 5: Type M (magnitude/inflation) ratio -- median |estimate|/|true value| among significant fits -- vs. participant sample size (n_p, log scale), one row per term, faceted by text sample size (n_t). Note the log-scaled y-axis; the dotted line marks a ratio of 1 (no inflation)."

panel_hlva_tmc_s5 <- make_aim1_curve_panel_s5(
  "zHLVA", b_hlva_grid, "Greens", metric_var = "type_m_ratio", y_lab = "Type M ratio (log scale)",
  legend_title = "zHLVA effect size", include_zero = FALSE, ref_line = 1, log_y = TRUE
)
panel_pcref_tmc_s5 <- make_aim1_curve_panel_s5(
  "zPCREFp", b_pcref_grid, "Blues", metric_var = "type_m_ratio", y_lab = "Type M ratio (log scale)",
  legend_title = "zPCREFp effect size", include_zero = FALSE, ref_line = 1, log_y = TRUE
)
panel_interaction_tmc_s5 <- make_aim1_curve_panel_s5(
  "zHLVA:zPCREFp", b_interaction_grid, "Purples", metric_var = "type_m_ratio", y_lab = "Type M ratio (log scale)",
  axis_title = "Participant sample size (n_p, log scale)", legend_title = "Interaction effect size",
  include_zero = FALSE, ref_line = 1, log_y = TRUE
)

panel_hlva_tmc_s5 / panel_pcref_tmc_s5 / panel_interaction_tmc_s5
```

The log scale is necessary rather than cosmetic here -- the Type M ratio
spans a wide range across this grid, with
the most extreme values concentrated at `n_t = 2`-`3` for the smallest
simulated non-zero effect sizes (where a small absolute bias divided by an
even smaller true value inflates dramatically; see the same interpretive
caveat already raised for the Type M error-surfaces panel earlier in this
document). `zPCREFp`'s row remains elevated longest among the two main
effects across the densified `n_t` grid; `zHLVA` settles near the
no-inflation line earliest and most consistently.

### `b_intercept` sweep across the full densified grid

The `b_intercept` sweep is an additive, one-at-a-time extension on top of
Scenario 5's main grid -- it does not cross with the 5-point effect-size
sweeps for `zHLVA`/`zPCREFp`/the interaction. Instead:

- Each term's effect size is held fixed at its own real, empirically-observed
  value (`zHLVA` = 0.363, `zPCREFp` = 0.095, interaction = 0.054 -- the
  `own_effect` values used throughout Aim 1's "own effect size" figures).
- Only `b_intercept` (4 values: 0/1.099/2.093/2.944, i.e. 50%/75%/89%-observed/95%
  target accuracy) is varied.
- This is done at every one of the 99 `(n_p, n_t)` cells -- 11 participant
  sample sizes x 9 text sample sizes.
- It's Aim-1-only (power/Type-S/Type-M/coverage-style diagnostics), not
  crossed with Aim 2's discovery/replication design.

So the "99 grid cells" in this section refers strictly to the sample-size
grid -- the intercept sweep's 4x99 = 396 cell-level combinations (x100
replicates x3 terms = 39,600 fits) is a design layered on top of, not
crossed with, the effect-size sweep.

As forewarned earlier in this document, everything above holds baseline
accuracy fixed at this project's real observed value. The figure below
tests that assumption directly, sweeping baseline accuracy across **all 99
grid cells** at 4 target accuracies (50%, 75%, 89% -- observed, 95%).

The figures below cover all six of Aim 1's diagnostics (power, sign
reliability, Type S rate, Type M ratio, coverage, and median bias) under
the intercept sweep, using the same shared `summarize_p2_metrics()` helper
as the power-surface/error-surfaces figures earlier, and the same
pooled-over-`n_p`, faceted-by-term, colored-by-target-accuracy layout for
every diagnostic so the six panels are directly comparable to each other.
`type_1_rate` is not shown here, since the intercept sweep holds every
term's effect size fixed at its own real, non-zero observed value (see
above) rather than sweeping down to zero -- there is no true-null condition
in this design for a false-positive rate to be computed against.

```{r}
#| label: intercept-sweep-summary-s5

intercept_s5_summary_by_nt <- summarize_p2_metrics(intercept_s5, c("term", "target_accuracy", "n_t"))

# Shared plotting helper -- one panel layout reused for every diagnostic
# below, so the six figures differ only in which column is mapped to y and
# how that axis is scaled/limited/annotated.
plot_intercept_curve <- function(data, y_var, y_lab, y_limits = NULL, ref_line = NULL, log_y = FALSE) {
  p <- ggplot(data, aes(x = n_t, y = .data[[y_var]], color = factor(target_accuracy))) +
    geom_line() + geom_point() +
    facet_wrap(~term) +
    scale_x_log10(breaks = n_t_grid_s5) +
    scale_color_brewer(palette = "Dark2", name = "Target\naccuracy") +
    labs(x = "Text sample size (n_t, log scale, pooled over n_p)", y = y_lab) +
    theme_minimal()
  if (!is.null(ref_line)) {
    p <- p + geom_hline(yintercept = ref_line, linetype = "dotted", color = "grey40", linewidth = 0.4)
  }
  if (log_y) p <- p + scale_y_log10()
  if (!is.null(y_limits)) p <- p + coord_cartesian(ylim = y_limits)
  p
}
```

```{r}
#| label: fig-intercept-sweep-power-s5
#| fig-width: 9
#| fig-height: 5
#| fig-cap: "Scenario 5 intercept sweep: power vs. text sample size, by target baseline accuracy and term, pooled over all 11 participant sample sizes."

plot_intercept_curve(intercept_s5_summary_by_nt, "power", "Power", y_limits = c(0, 1))
```

Power declines monotonically as target accuracy rises from 50% to 95%, at
essentially every `n_t`, for all three terms -- baseline accuracy is not a
neutral choice anywhere on this grid. For `zPCREFp`, the four accuracy
curves nearly **converge right at the pathological small-`n_t` trough**
(`n_t=6`-`16`, all four lines within ~2 points of each other around 0.17-
0.19) but separate clearly at both smaller (`n_t=2`) and larger (`n_t>=28`)
text samples. Baseline accuracy matters *less*, in absolute power terms,
exactly where the random-effects-identifiability breakdown already
dominates the picture -- and matters *more* everywhere the design is
otherwise behaving reasonably.

```{r}
#| label: fig-intercept-sweep-sign-reliability-s5
#| fig-width: 9
#| fig-height: 5
#| fig-cap: "Scenario 5 intercept sweep: sign reliability vs. text sample size, by target baseline accuracy and term, pooled over all 11 participant sample sizes."

plot_intercept_curve(intercept_s5_summary_by_nt, "sign_reliability", "Sign reliability", y_limits = c(0, 1))
```

`zHLVA`'s sign reliability stays high throughout (0.89-0.94 at `n_t=2`,
0.99+ by `n_t=136`), with only a mild decline as target accuracy rises.
`zPCREFp` is the most accuracy-insensitive of the three at `n_t=2` -- its
sign reliability sits in a narrow, coin-flip-adjacent 0.58-0.59 band
regardless of target accuracy, since the small-`n_t` identifiability
breakdown dominates at every accuracy level there -- but recovers to
0.90-0.96 by `n_t=136`, where higher target accuracy now visibly costs
reliability (0.96 at 50% vs. 0.90 at 95%). The interaction shows the
sharpest accuracy sensitivity of the three at `n_t=2`, falling from 0.70
(50% accuracy) to 0.56 (95% accuracy) -- higher baseline accuracy makes an
already-underpowered interaction estimate's sign less trustworthy, not just
less often significant.

```{r}
#| label: fig-intercept-sweep-type-s-s5
#| fig-width: 9
#| fig-height: 5
#| fig-cap: "Scenario 5 intercept sweep: Type S (sign) error rate vs. text sample size, by target baseline accuracy and term, pooled over all 11 participant sample sizes."

plot_intercept_curve(intercept_s5_summary_by_nt, "type_s_rate", "Type S rate", y_limits = c(0, 1))
```

`zHLVA`'s Type S rate is negligible everywhere (under 0.5%, all `n_t` and
target accuracies). `zPCREFp`'s is high and roughly flat across target
accuracy at `n_t=2` (0.34-0.41, mirroring its flat sign-reliability curve
there) and falls to essentially zero by `n_t=136` at every accuracy. The
interaction is again the term most sensitive to target accuracy at
`n_t=2`: its Type S rate nearly **triples** from 6.7% (50% accuracy) to
18.8% (95% accuracy) -- a "significant" interaction estimate from a
small-`n_t`, high-baseline-accuracy study is markedly more likely to have
the wrong sign than the same estimate from a lower-baseline-accuracy study.

```{r}
#| label: fig-intercept-sweep-type-m-s5
#| fig-width: 9
#| fig-height: 5
#| fig-cap: "Scenario 5 intercept sweep: Type M (magnitude) ratio vs. text sample size, by target baseline accuracy and term, pooled over all 11 participant sample sizes. Note the log-scaled y-axis; the dotted line marks a ratio of 1 (no inflation)."

plot_intercept_curve(intercept_s5_summary_by_nt, "type_m_ratio", "Type M ratio (log scale)", ref_line = 1, log_y = TRUE)
```

Type M ratio shows the clearest, most consistent accuracy sensitivity of
any diagnostic in this sweep, and specifically at `n_t=2`: the
interaction's ratio more than **triples** from 1.8 (50% accuracy) to 6.0
(95% accuracy), and `zPCREFp`'s -- already inflated at 6.4 at 50% accuracy
-- climbs further to 8.3 at 95%. `zHLVA` stays close to the no-inflation
line (1.01-1.09) throughout, regardless of target accuracy. By `n_t=136`
all three terms' ratios are far more modest (1.0-1.4) and much less
sensitive to target accuracy -- the accuracy-driven exaggeration is
concentrated where the text margin is already thin.

```{r}
#| label: fig-intercept-sweep-coverage-s5
#| fig-width: 9
#| fig-height: 5
#| fig-cap: "Scenario 5 intercept sweep: 95% CI coverage vs. text sample size, by target baseline accuracy and term, pooled over all 11 participant sample sizes. Dotted line marks nominal 95% coverage."

plot_intercept_curve(intercept_s5_summary_by_nt, "coverage", "95% CI coverage", y_limits = c(0, 1), ref_line = 0.95)
```

`zHLVA` and the interaction stay close to nominal coverage (roughly
0.85-0.99) across the whole grid, with only mild accuracy sensitivity.
`zPCREFp`'s coverage collapses at `n_t=2` regardless of target accuracy
(0.32-0.55) -- consistent with the coverage collapse already documented in
the error-surfaces figure earlier in this document -- but here, unusually,
coverage at `n_t=2` is *worst at the lowest* target accuracy (0.32 at 50%)
and improves somewhat toward higher accuracy (0.55 at 95%), the opposite
direction from every other diagnostic in this sweep. This single reversal
is noted rather than explained further here; it does not change the
overall conclusion that `zPCREFp`'s coverage is severely miscalibrated at
`n_t=2` at every target accuracy tested.

```{r}
#| label: fig-intercept-sweep-bias-s5
#| fig-width: 9
#| fig-height: 5
#| fig-cap: "Scenario 5 intercept sweep: median bias vs. text sample size, by target baseline accuracy and term, pooled over all 11 participant sample sizes. Dotted line marks zero bias."

plot_intercept_curve(intercept_s5_summary_by_nt, "median_bias", "Median bias", ref_line = 0)
```

Median bias stays small in absolute terms throughout this sweep (mostly
under 0.01 in magnitude) for `zHLVA` and the interaction at every `n_t` and
target accuracy. The two exceptions worth noting are both at `n_t=2` and
95% target accuracy: the interaction's median bias reaches -0.017 and
`zPCREFp`'s reaches 0.042 -- still modest relative to either term's own
effect size, but the largest bias magnitudes anywhere in this sweep, and
both occur at the same (smallest-`n_t`, highest-accuracy) combination that
also produced the largest Type M ratios above.

Three findings stand out across these six diagnostics together, mirroring
the error-surfaces synthesis earlier in this document:

1. **The interaction is the single most accuracy-sensitive term at
   `n_t=2`**: its sign reliability, Type S rate, and Type M ratio all move
   further with target accuracy than either main effect's do at that same
   cell (sign reliability 0.70 -> 0.56, Type S rate 6.7% -> 18.8%, Type M
   ratio 1.8 -> 6.0, all as target accuracy rises from 50% to 95%). For this
   term, a higher baseline accuracy does not just cost power -- it makes an
   already-underpowered estimate less trustworthy on every other diagnostic
   at once.
2. **`zHLVA` stays comparatively well-calibrated across every diagnostic and
   every target accuracy** (Type S rate under 0.5%, Type M ratio 1.01-1.09,
   coverage 0.85-0.99 throughout), echoing its across-the-board favorable
   behavior in the fixed-baseline-accuracy error surfaces earlier -- the
   accuracy sensitivity documented in this section is concentrated in the
   two text-margin-dependent terms, not a general property of the design.
3. **`zPCREFp`'s coverage reversal at `n_t=2` is the one exception to an
   otherwise consistent pattern**: every other diagnostic, for every term,
   either worsens or stays essentially flat as target accuracy rises. This
   reinforces the power-curve finding above -- a comprehension test
   calibrated closer to 50-75% accuracy would show better-behaved
   diagnostics generally at the small-`n_t` end of this grid, not only
   higher power.

### Aim 2 -- discovery/replication rates on the densified grid

Aim 2 asks a stricter, complementary question to Aim 1's power/error-rate
diagnostics: not "would a single study detect this effect," but "if this
study were run twice independently, at the same sample size, would both
runs agree?" Each replicate pair here comes from
`combine_replication_p2()`, which splits a cell's total participant/text
sample roughly in half into two independent "discovery" and "replication"
halves, fits both, and checks whether they agree.

At the two extremes of the densified text-sample grid, the same
interaction-fragility pattern already visible in Aim 1 reappears, in a
sharper form: at `n_t=2`, the unconditional replication rate is 34.2% for
`zHLVA`, 37.0% for `zPCREFp`, but only **0.6%** for the interaction --
because the interaction is so rarely significant even once at this sample
size that a second independent significant-and-same-signed result is
almost never observed. At `n_t=136`, `zHLVA`'s replication rate has grown
to 75.0%, while `zPCREFp`'s (22.4%) and the interaction's (14.6%) both
remain well short of that, for reasons the replication surface below
unpacks further. So: Aim 1's sign-reliability diagnostic (conditional on
significance) already flagged `zPCREFp` as the least trustworthy at small
`n_t`; Aim 2's replication-rate diagnostic (unconditional, on halved
samples) flags the interaction as the most fragile overall. Both are real,
non-contradictory findings -- they answer different practical questions
("if I got a significant result, can I trust its sign?" vs. "if I ran this
again, would I get the same answer?") -- and both indicate `n_t=2` is
inadequate, just via different mechanisms.

The table below defines Aim 2's three discovery/replication diagnostics
and states exactly how each is computed. All three are derived from
`combine_replication_p2()`'s per-cell pairing of two independently-fit
halves (`discovery` and `replication`, each roughly half of the cell's
total `n_p`/`n_t`), using the same Wald-CI `excludes_zero` significance
criterion as Aim 1, then aggregated by `term` x `n_t` (pooled over `n_p`
and the 100 replicate pairs/cell -- double the original grid's 50, per
Scenario 5's doubled-replicate design).

| Diagnostic | Standard definition | Computed here as |
|---|---|---|
| `discovery_rate` | P(significant in an initial/"discovery" sample) -- how often a first look at data of this size turns up a significant result at all, regardless of whether it is subsequently confirmed. | `mean(discovery_significant)`, where `discovery_significant = excludes_zero` evaluated on the **discovery** half alone. |
| `replication_rate` | Unconditional replication rate: P(significant **and** same-signed in *both* independent halves) -- Marek et al.'s (2022) Fig. 3e-style replication probability. | `mean(replicated)`, where `replicated = excludes_zero_discovery & excludes_zero_replication & sign(estimate_discovery) == sign(estimate_replication)`. Both halves must independently clear the significance threshold *and* agree in sign. |
| `conditional_replication` | Conditional replication rate: P(replicates \| already significant in the discovery half) -- "given I got a significant result once, would it hold up in an independent second sample?" | `mean(replicated[discovery_significant])`, i.e. `replicated` restricted to rows where the discovery half was already significant. More decision-relevant than the unconditional rate for a researcher who has already observed one significant result, since it factors out the "was it significant even once" question that `discovery_rate` alone answers. |

Note the arithmetic relationship between all three: `replication_rate` is
mechanically bounded above by `discovery_rate` (a discovery is a
prerequisite for replication), and `conditional_replication =
replication_rate / discovery_rate` for any cell where `discovery_rate > 0`
-- so a low unconditional `replication_rate` can reflect either a low
`discovery_rate` (rarely significant even once, a power problem) or a low
`conditional_replication` (often significant once but rarely again, a
genuine replication-failure problem), and distinguishing these two
mechanisms is exactly why all three are reported rather than
`replication_rate` alone (see the curve figure and its interpretation
below).

```{r}
#| label: aim2-metric-labels-s5

aim2_metric_labels <- c(discovery_rate = "Discovery rate\nP(significant, half 1)",
                         replication_rate = "Replication rate\nP(sig. + same sign, both halves)",
                         conditional_replication = "Conditional replication\nP(replicates | discovery-sig.)")
```

In addition to the raw replication rate discussed above, the table and
figure below add the discovery rate and conditional replication rate
defined in the table above, across the full densified grid.

```{r}
#| label: aim2-metrics-table-s5

aim2_metrics_by_nt_s5 <- aim2_all_s5 |>
  dplyr::group_by(term, n_t) |>
  dplyr::summarise(
    discovery_rate          = mean(discovery_significant, na.rm = TRUE),
    replication_rate        = mean(replicated, na.rm = TRUE),
    conditional_replication = mean(replicated[discovery_significant], na.rm = TRUE),
    n_fits                  = dplyr::n(),
    .groups = "drop"
  )

aim2_metrics_by_nt_s5 |>
  dplyr::arrange(term, n_t) |>
  knitr::kable(digits = 3,
               col.names = c("Term", "n_t", "Discovery rate", "Replication rate",
                             "Conditional replication", "n fits"))
```

```{r}
#| label: fig-aim2-metrics-vs-nt-s5
#| fig-width: 9
#| fig-cap: "Scenario 5: discovery rate, (unconditional) replication rate, and conditional replication rate vs. text sample size, at each term's own empirically-observed effect size, pooled across all 11 participant sample sizes and 100 replicate pairs/cell."

plot_metrics_vs_x(
  aim2_metrics_by_nt_s5, x_var = "n_t",
  metrics = c("discovery_rate", "replication_rate", "conditional_replication"),
  metric_labels = aim2_metric_labels,
  x_lab = "Text sample size (n_t, log scale)",
  x_breaks = n_t_grid_s5
)
```

The third panel adds a finding not visible in the unconditional
replication-rate numbers alone: conditional on having found a significant
interaction in the discovery half, the interaction's replication rate
(9.7% at `n_t=2`, rising to 45-50% by `n_t>=16`) is *lower* than `zHLVA`'s
at every `n_t` but is **not as catastrophically low** as its unconditional
replication rate (0.6% at `n_t=2`) might suggest in isolation -- most of
that unconditional fragility comes from rarely being significant in the
discovery half at all (a power problem, discovery rate only 6.5% at
`n_t=2`), rather than from significant discoveries failing to hold up (a
pure replication problem). Both framings matter for different audiences: a
reviewer asking "is this literature's replication rate adequate" should
look at the unconditional rate; a researcher who has already observed one
significant result and is deciding whether to trust it should look at the
conditional rate.

::: {.callout-important}
## A grid-density caveat: pooled-by-`n_t` tables are not directly comparable across grids of different density

Pooling over all 11 `n_p` values here gives noticeably different numbers
from the original document's pooled-over-6-`n_p` table -- e.g. the
interaction's replication rate at `n_t=136` is 0.146 here vs. 0.297
originally. This looks like a discrepancy but is not one: it is an artifact
of *which* `n_p` values are being averaged over, not a change in the
underlying `(n_p, n_t)` surface.

```{r}
#| label: grid-density-check

aim2_at_1497_s5 <- aim2_all_s5 |>
  dplyr::filter(n_p == 1497) |>
  dplyr::group_by(term, n_t) |>
  dplyr::summarise(replication_rate = mean(replicated, na.rm = TRUE),
                    conditional_replication = mean(replicated[discovery_significant], na.rm = TRUE),
                    .groups = "drop")

aim2_at_1497_orig <- aim2_all_orig |>
  dplyr::filter(n_p == 1497) |>
  dplyr::group_by(term, n_t) |>
  dplyr::summarise(replication_rate_orig = mean(replicated, na.rm = TRUE), .groups = "drop")

aim2_at_1497_s5 |>
  dplyr::left_join(aim2_at_1497_orig, by = c("term", "n_t")) |>
  dplyr::filter(term == "zHLVA:zPCREFp") |>
  dplyr::arrange(n_t) |>
  knitr::kable(digits = 3,
               col.names = c("Term", "n_t", "Replication rate (Scenario 5)",
                             "Conditional replication (Scenario 5)", "Replication rate (original grid)"))
```

Restricted to the one `n_p` value present in both grids (1497), the two
runs agree closely (e.g. interaction replication at `n_t=136`: 0.64 here vs.
0.68 in the original at the same cell) -- both are independently-sampled
Monte Carlo estimates of the same underlying quantity, and the ~4-point gap
is well within sampling noise at 50-100 replicates/cell. **Conclusion**:
compare pooled summary tables across the two documents only after matching
on `n_p` (or read the `(n_p, n_t)` surfaces directly, as below), not by
comparing pooled-over-different-grids marginals at face value.
:::

```{r}
#| label: fig-aim2-replication-surface-s5
#| fig-width: 11
#| fig-height: 4.3
#| fig-cap: "Scenario 5: replication-rate surface across the full densified grid, at each term's own empirically-observed effect size."

aim2_surface_s5 <- aim2_all_s5 |>
  dplyr::group_by(term, n_p, n_t) |>
  dplyr::summarise(replication_rate = mean(replicated, na.rm = TRUE), .groups = "drop")

plot_power_heatmap(aim2_surface_s5, fill_var = "replication_rate", fill_label = "Replication\nrate")
```

```{r}
#| label: pcref-repl-top5-s5

aim2_surface_s5 |> dplyr::filter(term == "zPCREFp") |>
  dplyr::arrange(dplyr::desc(replication_rate)) |>
  head(5) |>
  knitr::kable(digits = 3, col.names = c("Term", "n_p", "n_t", "Replication rate"))
```

Similar to Aim 1's power surface, but stricter, and carrying the same
`zPCREFp`-specific reversal identified there. `zHLVA` saturates cleanly on
the participant margin alone once `n_t>=10` (99-100% replication by
`n_p>=291`), but at `n_t=2` its replication rate climbs only gradually with
`n_p` -- from 1% at `n_p=25` to 82% at `n_p=1497`, never reaching the
ceiling `n_t>=10` reaches everywhere on this grid -- so halving the sample
still measurably hurts even the best-behaved term when `n_t=2`. The
**interaction** shows the pattern that comparison would predict: replication
stays under 3% at every `n_p` when `n_t=2`, then climbs sharply once
`n_t>=10` and `n_p=1497`, peaking at 70% (`n_t=28`) -- a real gain from more
texts, though notably peaking at an intermediate `n_t` rather than the
largest one (`n_t=136` gives 64%). **`zPCREFp` again breaks that
expectation**, just as it did for power: its top three replication rates on
the entire grid (0.74, 0.73, 0.66) all occur at `n_t=2` (`n_p=994`, `1497`,
and `660` respectively) -- higher than anything it achieves at any larger
`n_t`, including the largest-both-margins cell (0.61 at `n_p=1497,
n_t=136`, itself only the 4th-highest value on the whole grid). More
broadly, `zPCREFp`'s `n_t=2` replication rate meets or exceeds every larger
`n_t` at the same `n_p` in 9 of 11 `n_p` cells -- the same underpowered-
paradox signature carrying over from Aim 1's power surface into Aim 2's
stricter replication criterion, not a genuine advantage of the smallest
text sample.

### A broader synthesis

Putting the two reader questions from the top of this section back together
with everything above:

- **For planning**: `zHLVA` (the participant-level effect) needs only a
  moderate participant sample (saturates by `n_p~100`-`375`) and is
  comparatively insensitive to text sample size once `n_t>=6`. `zPCREFp`
  and the interaction need growth on **both** margins, and specifically
  need `n_t` to clear roughly 20-30 texts before the small-`n_t` pathology
  fully resolves -- a design implication independent of whether a
  particular study happens to land in the "paradoxically inflated" `n_t=2`
  regime or not. Baseline accuracy should also be budgeted for: a
  comprehension test calibrated closer to 50-75% accuracy (rather than this
  project's ~89%) would show *better* power for the same underlying effect
  sizes.
- **For evaluation**: a "significant" `zPCREFp` or interaction result from
  a study with very few texts (especially `n_t=2`, the modal value in this
  project's own literature review) should be read with real suspicion --
  not because small samples are underpowered in the usual sense, but
  because the specific `n_t=2` identifiability breakdown documented below
  inflates both the raw significance rate and the replication rate at
  exactly that sample size, for reasons unrelated to whether a real effect
  is present.
- The "underpowered paradox" is the sharpest single illustration of the
  evaluation-side risk, and is retained here as such -- but as this section
  has shown, it does not characterize every diagnostic (sign reliability,
  Type S, Type M all behave conventionally), and it is only one of several
  observations relevant to planning and evaluation questions, not the
  single organizing finding of this analysis.

Full mechanistic detail on *why* `n_t=2` produces this pattern (the
random-effects-identifiability argument, the singular-fit diagnostic, and
the Wald-SE-deflation mechanism) is given in "Why can an estimate be
'significant' when the true effect is exactly zero?" above.

## Development note

This document was produced entirely from already-computed `.rds` files
(`run_scenario5_full.R`, `run_scenario5_intercept_sweep.R`); no new
simulation was run to produce it. See `AGENTS.md`'s "Aims 1-2: Scenario 5 +
`b_intercept` extension" section for the run history (launch commands,
timings, log excerpts) and
`cost-estimate-expanded-grid-replicates-aims1-2-PLAN.qmd` for the original
cost-estimation rationale behind choosing this particular grid
densification and replicate count.
