---
title: "Effect size variation for the HLVA main effect, the PCREFp main effect, and their interaction"
subtitle: "Aims 1-2: power, error rates, and replication rates (accuracy outcome)"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: cosmo
    embed-resources: true
    code-fold: true
execute:
  warning: false
  message: false
---

::: {.callout-note}
## Scope of this document

This document covers **Aim 1** (effect size x sample size, with a fuller
statistical-error taxonomy) and **Aim 2** (discovery/replication split) of
the Phase 2 parametric simulation program, for the `response` (comprehension
accuracy) outcome only. It is a restructured development version of
`synthetic-data-analyses-power-error-replication-aims-1-2-draft.qmd` (which
is left unchanged as the reference/working copy), reframed around this
project's actual, broader purpose: **mapping how power, the full Type
I/II/S/M error taxonomy, and discovery/replication rates vary with effect
size and sample size, for all three effects this project cares about**
(the `zHLVA` participant-level main effect, the `zPCREFp` text-level main
effect, and their interaction) -- not only the specific sign-instability
question that originally motivated the design (see below). Aims 3-6 and the
normative-F1-ANOVA-malpractice extension will get their own documents
later, built the same way.

All tables and figures below are computed from pre-existing, already-run
results saved in `results_phase2/aim1_*.rds` and `results_phase2/aim2_*.rds`
(no full-grid recomputation happens when this document is rendered). Full
design rationale, the complete six-aim plan, and every aim's development
history live untouched in `phase2-effect-variance-misspecification-PLAN.qmd`
and `phase2-effect-variance-misspecification.qmd` -- this document condenses
and refines the Aim 1/2 material from those sources rather than re-deriving
it. The DGP/fitting code below is kept live (not just for reference) because
it may be revisited for informative modifications in a follow-up pass; see
the closing note.
:::

::: {.callout-important}
## Terminology check: Type I vs. Type II

Throughout this document, **Type I error** = a false positive (rejecting a
true null -- concluding an effect exists when it does not), and **Type II
error** = a false negative (failing to reject a false null -- missing a
real effect). This is the standard convention, and matches how the rest of
this document already uses "false-positive rate" (e.g. the `n_t = 2`
`zPCREFp` discussion below). If a different convention was intended when
this restructuring was requested, flag it and the language below can be
adjusted.
:::

```{r}
#| label: setup

library(tidyverse)
library(lme4)
library(patchwork)
library(RColorBrewer)
library(ggdist)
```

## Relationship to Phase 1 and motivating framework

Phase 1 (`sampling-variability-corridor-of-stability.qmd`) was an
**empirical resampling** study of this project's one real dataset,
answering "how unstable is our *actual* estimate at smaller *N*?" Phase 2 is
a **parametric simulation** study, answering the complementary question:
"how unstable would an estimate be across a *range* of plausible true
effects, variances, and model specifications -- are Phase 1's findings a
generic property of this kind of crossed design, or specific to this
project's particular dataset?"

Aims 1-2 port Marek et al.'s (2022) brain-wide-association-study
reproducibility framework onto this setting. Marek et al. documented a
statistical-error taxonomy -- false negatives, false positives,
**magnitude/inflation error (Type M)**, and **sign error (Type S)** -- each
as a function of sample size, and showed that smaller samples' significant
effects are disproportionately the most inflated and sign-flipped ones (the
"underpowered paradox": stricter significance thresholds select for, rather
than screen out, the most inflated estimates). They also used an explicit
**discovery/replication split** design to quantify out-of-sample replication
probability directly. Aims 1-2 were originally motivated by testing whether
this project's own documented sign-instability (the "reverse cohesion
effect" finding -- see project memory) is a domain-specific instance of the
same generic phenomenon; the design covers **three effects** rather than
the interaction alone (the `zHLVA` main effect, the `zPCREFp` main effect,
and their `zHLVA:zPCREFp` interaction) precisely so that question could be
answered by comparison.

That motivating question, however, is narrower than what this document is
actually useful for. Once all three effects are run through the same
effect-size x sample-size grid with the same diagnostic taxonomy, the
results answer a more general question that different readers will care
about differently:

- A reader planning a **new study** mainly wants **power** and **discovery
  rate**: how large a sample, on which margin, is needed to reliably detect
  a given effect?
- A reader trying to **interpret an existing significant finding** (their
  own or someone else's) mainly wants **Type S (sign) error rate** and
  **Type M (magnitude) ratio**: conditional on a result being "significant,"
  how likely is it to have the wrong sign, or to be substantially inflated?
- A reader thinking about **meta-scientific reproducibility** mainly wants
  the **discovery/replication rates** from Aim 2: if this study were run
  twice independently, how often would both runs agree?

All three questions are answered for all three effects below; the original
sign-instability question is one specific instance of the first two,
revisited explicitly in "Aim 1: is small-text-*N* instability a
text-margin property or an interaction-specific property?" below.

The table below defines each diagnostic used throughout this document and
states exactly how it is computed here. All are derived from
`diagnose_coefs_p2()`'s per-replicate-fit columns (`estimate`, `std.error`,
`true_value`, and the **Wald** 95% CI `estimate +/- 1.96 * std.error`), then
aggregated by `summarize_p2_metrics()` / the `aim1-full-metrics-table`
chunk over whichever grouping is in play (typically `term` x `n_t`, pooled
over `n_p` and the 50 replicates/cell).

| Diagnostic | Standard definition | Computed here as |
|---|---|---|
| `power` | P(reject H0 \| true effect is non-zero) -- probability of correctly detecting a real effect. | `mean(excludes_zero)`, restricted to rows at the term's non-zero (typically empirically-observed) effect size. `excludes_zero` = the 95% Wald CI does not contain 0. |
| `type_1_rate` | P(reject H0 \| true effect is exactly zero) -- the false-positive rate; nominally ~0.05 for a correctly calibrated 95% CI. | `mean(excludes_zero)`, restricted to rows at `effect_value == 0` on the same sweep. |
| `sign_reliability` | P(estimated sign matches true sign), unconditional on significance. | `mean(correct_sign)` over **all** fits (significant or not), where `correct_sign = sign(estimate) == sign(true_value)`. |
| `type_s_rate` | Type S (sign) error rate: P(estimated sign is wrong \| the result is significant) -- Gelman & Carlin's (2014)/Marek et al.'s (2022) usage. | `mean(sign_error[excludes_zero])`, i.e. computed only among rows where the CI excludes zero; `sign_error = excludes_zero & !correct_sign`. |
| `type_m_ratio` | Type M (magnitude/inflation) ratio: how much larger a *significant* estimate typically is than the true effect. | `median(inflation_ratio[excludes_zero])`, where `inflation_ratio = abs(estimate) / abs(true_value)` (undefined/`NA` when `true_value == 0`, since inflation relative to zero is not meaningful); the median (not mean) is used to resist the heavy right skew of this ratio at small `n_t`. |
| `median_bias` | Bias: the typical signed deviation of the estimate from the true value. | `median(bias)` over all fits, where `bias = estimate - true_value`. Median rather than mean, again for robustness to the occasional extreme estimate at small `n_t`. |
| `coverage` | Nominal 95% CI coverage: the proportion of CIs that actually contain the true value (should be ~0.95 if well-calibrated). | `mean(covers_truth)`, where `covers_truth = ci_lower <= true_value & true_value <= ci_upper`. |
| `rmse_raw` | Root-mean-squared error of the estimate around the true value. | `sqrt(mean(bias^2))` over all fits. Included for completeness but dominated by rare pathological fits at `n_t = 2` (see prose below the fuller metrics table). |
| `rmse_trimmed` | RMSE, robustified against extreme outliers. | `sqrt(mean(pmin(bias^2, quantile(bias^2, 0.95))))` -- squared bias is winsorized at its own 95th percentile (within each group) before averaging, so a handful of extreme singular-fit estimates cannot dominate the summary. Preferred over `rmse_raw` for cross-cell comparison. |

Two things to note about this taxonomy as implemented: first, `type_s_rate`
and `type_m_ratio` are both computed *conditional on significance*
(`excludes_zero`), matching Marek et al.'s (2022) framing of "what does a
significant result look like at this sample size" -- they are not defined,
and are `NA`, for the (usually large) share of non-significant fits.
Second, all coverage/CI-based diagnostics here use **Wald** intervals
(`estimate +/- 1.96 * std.error`) rather than profile-likelihood or
bootstrap CIs, for computational speed across tens of thousands of `glmer`
fits -- this is a possible source of miscalibration in its own right,
distinct from the random-effects-identifiability mechanism discussed later
in this document, and is not separately investigated here.

| Input | Value(s) | Source |
|---|---|---|
| Sample-size grid (participants) | 25, 100, 375, 725, 1300, 1497 | Phase 1 Step 3's joint grid |
| Sample-size grid (texts) | 2, 10, 35, 85, 136 | Phase 1 Step 3's joint grid |
| Effect size, `zHLVA` | 0.363 | Phase 1 Step 0 full-data fit |
| Effect size, `zPCREFp` | 0.094-0.098 | Phase 1 Step 0 / Step 5 |
| Effect size, `zHLVA:zPCREFp` | 0.054 | Phase 1 Step 0 / Step 5 (glmer and brms agree) |
| Variance component, `sd(ResponseId)` | 0.69 | Full core model (`dev.all.studies.noratings.ptqs.normal.0.1.normal.0.1.lkj.2.study.ri.rds`), extracted in `phase2-effect-variance-misspecification.qmd`'s `extract-aim5-reference` chunk -- differs from Phase 1's simpler reduced-model estimate (~0.61-0.62) because the full model's 26 fixed effects absorb some of that between-participant variance |
| Variance component, `sd(text.id)` | 0.41 | Same source as above -- differs from Phase 1's reduced-model estimate (~0.70-0.72), same reason |
| RE-misspecification contrast | "correct" `(1\|ResponseId)+(1\|text.id)` vs. "misspecified" `(1\|ResponseId)` only | Phase 1, all steps |

Reusing the same sample-size grid means these results can be laid
side-by-side against Phase 1's real-data panels directly.

::: {.callout-important}
## Note: the real core model's random-effects structure is richer than this DGP

The real fitted core model's random-effects terms let participants vary in
their *slopes* for all five text-level predictors (including `zPCREFp`), and
let texts vary in their *slopes* for the participant-level predictors
(including `zHLVA`), each correlated with its own grouping factor's
intercept. Aims 1-2 (like Aim 3-4 and 6) use a simpler, random-**intercept**-
only "correct"/"misspecified" data-generating process, following Phase 1's
reduced model -- they answer a coarser question (omitting a grouping
factor's clustering altogether) and should not be read as testing the real
model's richer slope/covariance structure. That richer structure is the
subject of Aim 5, covered in a separate future document.
:::

## Design: Aim 1 (effect size x sample size) and Aim 2 (discovery/replication split)

Both aims share one data-generating run: RE specification fixed at
"correct" (random intercepts only, for `ResponseId`/`participant_id` and
`text.id`/`text_id`); variance components fixed at their empirical values;
the full joint sample-size grid (30 `(n_p, n_t)` cells); 50 replicates/cell.
A grid cell's `(n_p, n_t)` is the **total** sample -- Aim 2 splits each cell
into two roughly-equal-sized discovery/replication halves.

**Aim 1** runs three **one-at-a-time** effect-size sweeps -- each holding
the other two effects fixed at their empirical values while varying the
swept effect:

| Sweep | Effect | Grid (centered on empirical estimate) |
|---|---|---|
| (i) | `zHLVA` (main effect, participant-level) | 0, 0.18, 0.363 (observed), 0.7, 1.0 |
| (ii) | `zPCREFp` (main effect, text-level) | 0, 0.05, ~0.095 (observed), 0.2, 0.35 |
| (iii) | `zHLVA:zPCREFp` (interaction) | 0, 0.025, 0.054 (observed), 0.10, 0.20 |

Diagnostics extend the project's existing four (bias, sign reliability,
coverage, RMSE) with a Marek-style decomposition, directly computable here
because the true value is known exactly: **Type S (sign) error rate** (among
CI-excludes-0 draws, proportion with the wrong sign), **Type M
(magnitude/inflation) ratio** (`|estimate|/|true value|` among
CI-excludes-0 draws), and **power** on its own scale.

**The original motivating question Aim 1 answers** (one specific instance
of its general purpose): is the sign-instability pattern found in Phase 1
(near-coin-flip sign consistency for the interaction at *n* = 2 texts) (a) a
property of the *specific* small effect size observed, (b) specific to the
interaction term, or (c) a general property of being estimated from the
*text* margin (in which case `zPCREFp`, also text-level, should show a
similar pattern)? Comparing all three effects on the same diagnostics
distinguishes these explanations directly. But the same grid and diagnostics
also answer the broader questions above (power/discovery for study
planning, Type S/M for interpreting an existing result) for **any** of the
three effects, independent of that original motivating question -- readers
who don't care about the reverse-cohesion-effect history can use the tables
and figures below directly for whichever effect and diagnostic interests
them.

**Aim 2** ports Marek et al.'s replication design (their Fig. 3e): for each
grid cell (all three effects), split the simulated sample into two
independent, equal-sized discovery/replication halves, fit both, and compute
the proportion of replicates in which a discovery-significant effect also
replicates (same sign, both significant) out-of-sample. The split axis is
configurable (`split_margin`): `"participants_only"` (same text pool in both
halves, new participants only) vs. `"both"` (independent participants *and*
texts). Both variants can be compared once built; the run below uses
`"participants_only"`.

The significance-threshold/multiple-comparisons sweep (Marek's
P<0.05-to-Bonferroni comparison) is explicitly **deferred** here -- see
"Flagged future work" below.

## Shared data-generating function

Generalizes `simulate_interaction_data()` (from
`power-simulation-vocabulary-cohesion.qmd`) and `generate_data_s3()` (from
`simulation_study_complete.qmd`): crossed participant x text design with
`k`-texts-per-participant subsetting (clamped to `n_texts`), independently
settable effect sizes for all three effects, independently settable
intercept variances for both margins, and optional by-margin slope
variance/correlation (zero by default here, so Aims 1-2 reduce to the
intercept-only DGP; non-zero values are used only by Aim 5, covered
elsewhere).

```{r}
#| label: generate-data-p2

generate_data_p2 <- function(
    n_participants = 200,
    n_texts        = 20,
    k              = min(10, n_texts),
    questions_per_text = 4,
    b_intercept    = 2.093,
    b_hlva         = 0.363,
    b_pcref        = 0.095,
    b_interaction  = 0.054,
    sd_participant = 0.69,
    sd_text        = 0.41,
    sd_pcref_slope = 0,
    sd_hlva_slope  = 0,
    cor_part       = 0,
    cor_text       = 0,
    cohesion_design = c("natural", "widened"),
    reliability_hlva = 1,
    reliability_pcref = 1,
    seed           = NULL
) {
  cohesion_design <- match.arg(cohesion_design)
  if (!is.null(seed)) set.seed(seed)
  k <- min(k, n_texts)

  # Participant-level random effects: intercept + zPCREFp slope (correlated) -- Aim 5
  Sigma_part <- matrix(
    c(sd_participant^2, cor_part * sd_participant * sd_pcref_slope,
      cor_part * sd_participant * sd_pcref_slope, sd_pcref_slope^2),
    nrow = 2
  )
  re_part <- if (sd_pcref_slope > 0) {
    MASS::mvrnorm(n_participants, mu = c(0, 0), Sigma = Sigma_part)
  } else {
    cbind(rnorm(n_participants, 0, sd_participant), 0)
  }

  # zHLVA measurement reliability -- Aim 6 strategy E; reliability_hlva = 1
  # (default) reproduces the exact prior behavior used in Aims 1-5.
  zHLVA_true <- rnorm(n_participants, 0, 1)
  zHLVA_obs  <- sqrt(reliability_hlva) * zHLVA_true +
    sqrt(1 - reliability_hlva) * rnorm(n_participants, 0, 1)

  participants <- tibble(
    participant_id = 1:n_participants,
    zHLVA          = zHLVA_true,
    zHLVA_obs      = zHLVA_obs,
    u_participant  = re_part[, 1],
    s_pcref        = re_part[, 2]
  )

  # Text-level random effects: intercept + zHLVA slope (correlated) -- Aim 5
  Sigma_text <- matrix(
    c(sd_text^2, cor_text * sd_text * sd_hlva_slope,
      cor_text * sd_text * sd_hlva_slope, sd_hlva_slope^2),
    nrow = 2
  )
  re_text <- if (sd_hlva_slope > 0) {
    MASS::mvrnorm(n_texts, mu = c(0, 0), Sigma = Sigma_text)
  } else {
    cbind(rnorm(n_texts, 0, sd_text), 0)
  }

  # zPCREFp sampling design -- Aim 6 strategy D (widened cohesion range) --
  # and measurement reliability -- Aim 6 strategy E.
  zPCREFp_true <- if (cohesion_design == "widened") {
    rep(c(-1.5, 1.5), length.out = n_texts) + rnorm(n_texts, 0, 0.3)
  } else {
    rnorm(n_texts, 0, 1)
  }
  zPCREFp_obs <- sqrt(reliability_pcref) * zPCREFp_true +
    sqrt(1 - reliability_pcref) * rnorm(n_texts, 0, 1)

  texts <- tibble(
    text_id  = 1:n_texts,
    zPCREFp  = zPCREFp_true,
    zPCREFp_obs = zPCREFp_obs,
    u_text   = re_text[, 1],
    s_hlva   = re_text[, 2]
  )

  assignments <- tibble(participant_id = 1:n_participants) |>
    dplyr::mutate(assigned = purrr::map(participant_id,
                          ~ sample(1:n_texts, size = k, replace = FALSE))) |>
    tidyr::unnest(assigned) |>
    dplyr::rename(text_id = assigned)

  assignments |>
    dplyr::left_join(participants, by = "participant_id") |>
    dplyr::left_join(texts,        by = "text_id") |>
    dplyr::slice(rep(1:dplyr::n(), each = questions_per_text)) |>
    dplyr::mutate(
      question_id = rep(1:questions_per_text, times = dplyr::n() / questions_per_text),
      eta = b_intercept +
        (b_hlva  + s_hlva)  * zHLVA +
        (b_pcref + s_pcref) * zPCREFp +
        b_interaction * zHLVA * zPCREFp +
        u_participant + u_text,
      prob     = plogis(eta),
      response = rbinom(dplyr::n(), 1, prob)
    ) |>
    dplyr::mutate(zHLVA_fit = zHLVA_obs, zPCREFp_fit = zPCREFp_obs) |>
    dplyr::select(participant_id, text_id, question_id,
                  zHLVA = zHLVA_fit, zPCREFp = zPCREFp_fit, response)
}
```

Note: `library(MASS)` is deliberately **not** attached (only `MASS::mvrnorm`
is called via `::`), because `MASS::select()` masks `dplyr::select()` and
breaks the pipeline above if attached. The fitted model always sees
`zHLVA`/`zPCREFp` as **observed** (possibly measurement-error-attenuated)
values, while the response is generated from the **true** latent values --
with the default `reliability_hlva = reliability_pcref = 1` used throughout
Aims 1-2, observed and true are identical.

## Shared fitting and outcome-classification helpers

Reused from `sampling-variability-corridor-of-stability.qmd`'s
`fit_reduced_model()` / `extract_terms_of_interest()` /
`fit_reduced_model_guarded()` pattern, extended with a third RE
specification (`"full_slopes"`, used only by Aim 5) alongside the
`"correct"`/`"misspecified"` pair used here and by Aim 3.

```{r}
#| label: fit-model-p2

extract_terms_of_interest_p2 <- function(fit) {
  terms <- c("zHLVA", "zPCREFp", "zHLVA:zPCREFp")
  co <- tryCatch(summary(fit)$coefficients, error = function(e) NULL)
  if (is.null(co)) return(NULL)
  avail <- intersect(terms, rownames(co))
  tibble(term = avail, estimate = co[avail, "Estimate"], std.error = co[avail, "Std. Error"])
}

fit_model_p2 <- function(data, re_spec = c("correct", "misspecified", "full_slopes")) {
  re_spec <- match.arg(re_spec)
  form <- switch(re_spec,
    correct      = response ~ zHLVA * zPCREFp + (1 | participant_id) + (1 | text_id),
    misspecified = response ~ zHLVA * zPCREFp + (1 | participant_id),
    full_slopes  = response ~ zHLVA * zPCREFp + (1 + zPCREFp | participant_id) + (1 + zHLVA | text_id)
  )
  warn_msgs <- character(0)
  fit <- tryCatch(
    withCallingHandlers(
      glmer(form, data = data, family = binomial, control = glmerControl(optimizer = "bobyqa")),
      warning = function(w) { warn_msgs <<- c(warn_msgs, conditionMessage(w)); invokeRestart("muffleWarning") }
    ),
    error = function(e) e
  )
  if (inherits(fit, "error")) return(list(status = "error", coefs = NULL))
  status <- if (length(warn_msgs) > 0 &&
                any(grepl("failed to converge|Model failed|large eigenvalue|unable to evaluate",
                          warn_msgs, ignore.case = TRUE))) {
    "optimizer_warning"
  } else if (isSingular(fit)) {
    "converged_singular"
  } else {
    "converged_clean"
  }
  list(status = status, coefs = extract_terms_of_interest_p2(fit))
}

fit_model_p2_guarded <- function(data, re_spec, timeout_secs = 15) {
  setTimeLimit(elapsed = timeout_secs, transient = TRUE)
  res <- tryCatch(
    fit_model_p2(data, re_spec),
    error = function(e) list(status = "error", coefs = NULL)
  )
  setTimeLimit(elapsed = Inf, transient = FALSE)
  res
}
```

## Sample-size grid and Aim 1 sweep grids

```{r}
#| label: sample-size-grid

n_p_grid <- c(25, 100, 375, 725, 1300, 1497)
n_t_grid <- c(2, 10, 35, 85, 136)
joint_grid <- tidyr::expand_grid(n_p = n_p_grid, n_t = n_t_grid)
nrow(joint_grid)
```

```{r}
#| label: aim1-sweep-grids

b_hlva_grid        <- c(0, 0.18, 0.363, 0.7, 1.0)
b_pcref_grid       <- c(0, 0.05, 0.095, 0.2, 0.35)
b_interaction_grid <- c(0, 0.025, 0.054, 0.10, 0.20)
```

## Discovery/replication data-generating pair (Aim 2)

For `split_margin = "participants_only"`, both halves share the same
`n_t_total` texts (avoiding degenerate 1-text halves at the smallest grid
point, `n_t = 2`); only the participant pool is split. For `split_margin =
"both"`, both margins split independently.

```{r}
#| label: generate-data-p2-pair

generate_data_p2_pair <- function(
    n_p_total, n_t_total, k = min(10, n_t_total), questions_per_text = 4,
    b_intercept = 2.093, b_hlva = 0.363, b_pcref = 0.095, b_interaction = 0.054,
    sd_participant = 0.69, sd_text = 0.41,
    sd_pcref_slope = 0, sd_hlva_slope = 0, cor_part = 0, cor_text = 0,
    split_margin = c("participants_only", "both"),
    seed
) {
  split_margin <- match.arg(split_margin)
  set.seed(seed)

  n_p_half <- max(1, floor(n_p_total / 2))
  n_t_half <- if (split_margin == "both") max(1, floor(n_t_total / 2)) else n_t_total

  make_texts <- function(n_t) {
    Sigma_text <- matrix(c(sd_text^2, cor_text * sd_text * sd_hlva_slope,
                            cor_text * sd_text * sd_hlva_slope, sd_hlva_slope^2), nrow = 2)
    re_text <- if (sd_hlva_slope > 0) MASS::mvrnorm(n_t, c(0, 0), Sigma_text) else cbind(rnorm(n_t, 0, sd_text), 0)
    tibble(text_id = 1:n_t, zPCREFp = rnorm(n_t, 0, 1), u_text = re_text[, 1], s_hlva = re_text[, 2])
  }
  make_participants <- function(n_p) {
    Sigma_part <- matrix(c(sd_participant^2, cor_part * sd_participant * sd_pcref_slope,
                            cor_part * sd_participant * sd_pcref_slope, sd_pcref_slope^2), nrow = 2)
    re_part <- if (sd_pcref_slope > 0) MASS::mvrnorm(n_p, c(0, 0), Sigma_part) else cbind(rnorm(n_p, 0, sd_participant), 0)
    tibble(participant_id = 1:n_p, zHLVA = rnorm(n_p, 0, 1), u_participant = re_part[, 1], s_pcref = re_part[, 2])
  }
  build_dataset <- function(participants, texts) {
    kk <- min(k, nrow(texts))
    assignments <- tibble(participant_id = participants$participant_id) |>
      dplyr::mutate(assigned = purrr::map(participant_id, ~ sample(texts$text_id, size = kk, replace = FALSE))) |>
      tidyr::unnest(assigned) |> dplyr::rename(text_id = assigned)
    assignments |>
      dplyr::left_join(participants, by = "participant_id") |>
      dplyr::left_join(texts, by = "text_id") |>
      dplyr::slice(rep(1:dplyr::n(), each = questions_per_text)) |>
      dplyr::mutate(
        question_id = rep(1:questions_per_text, times = dplyr::n() / questions_per_text),
        eta = b_intercept + (b_hlva + s_hlva) * zHLVA + (b_pcref + s_pcref) * zPCREFp +
          b_interaction * zHLVA * zPCREFp + u_participant + u_text,
        prob = plogis(eta), response = rbinom(dplyr::n(), 1, prob)
      ) |>
      dplyr::select(participant_id, text_id, question_id, zHLVA, zPCREFp, response)
  }

  texts_shared      <- make_texts(n_t_half)
  participants_disc <- make_participants(n_p_half)
  participants_repl <- make_participants(n_p_half)
  texts_repl        <- if (split_margin == "both") make_texts(n_t_half) else texts_shared

  list(
    discovery   = build_dataset(participants_disc, texts_shared),
    replication = build_dataset(participants_repl, texts_repl)
  )
}
```

## Diagnostics and combined per-cell runner

Coverage/CI-width/Type S/Type M diagnostics use **Wald confidence intervals**
(`estimate +/- 1.96 * std.error`) rather than profile-likelihood CIs, for
speed -- consistent with the project's `glmer`-as-fast-workhorse convention.

```{r}
#| label: aim1-aim2-diagnostics-and-runner

diagnose_coefs_p2 <- function(coefs, true_vals) {
  coefs |>
    dplyr::mutate(
      true_value      = true_vals[term],
      ci_lower        = estimate - 1.96 * std.error,
      ci_upper        = estimate + 1.96 * std.error,
      ci_width        = ci_upper - ci_lower,
      excludes_zero   = ci_lower > 0 | ci_upper < 0,
      correct_sign    = sign(estimate) == sign(true_value),
      sign_error      = excludes_zero & !correct_sign,
      inflation_ratio = ifelse(true_value != 0, abs(estimate) / abs(true_value), NA_real_),
      bias            = estimate - true_value,
      covers_truth    = ci_lower <= true_value & true_value <= ci_upper
    )
}

combine_replication_p2 <- function(res_disc, res_repl, true_vals) {
  if (is.null(res_disc$coefs) || is.null(res_repl$coefs)) {
    return(tibble(term = names(true_vals), replicated = NA, discovery_significant = NA))
  }
  disc <- diagnose_coefs_p2(res_disc$coefs, true_vals) |>
    dplyr::select(term, estimate, excludes_zero)
  repl <- diagnose_coefs_p2(res_repl$coefs, true_vals) |>
    dplyr::select(term, estimate_repl = estimate, excludes_zero_repl = excludes_zero)
  dplyr::full_join(disc, repl, by = "term") |>
    dplyr::mutate(
      discovery_significant = excludes_zero,
      replicated = excludes_zero & excludes_zero_repl & sign(estimate) == sign(estimate_repl)
    )
}

run_aim1_aim2_cell <- function(sweep, effect_value, n_p, n_t, rep, split_margin = "participants_only") {
  b_hlva        <- if (sweep == "hlva")        effect_value else 0.363
  b_pcref       <- if (sweep == "pcref")       effect_value else 0.095
  b_interaction <- if (sweep == "interaction") effect_value else 0.054
  true_vals <- c(zHLVA = b_hlva, zPCREFp = b_pcref, `zHLVA:zPCREFp` = b_interaction)

  sweep_idx <- match(sweep, c("hlva", "pcref", "interaction"))
  seed_base <- rep * 1e7 + n_p * 1e3 + n_t + sweep_idx * 1e5 + round(effect_value * 1000)

  ## Aim 1: single full-sample fit
  df_full  <- generate_data_p2(n_participants = n_p, n_texts = n_t,
                                b_hlva = b_hlva, b_pcref = b_pcref, b_interaction = b_interaction,
                                seed = seed_base)
  res_full <- fit_model_p2_guarded(df_full, "correct")

  aim1_rows <- if (!is.null(res_full$coefs)) {
    diagnose_coefs_p2(res_full$coefs, true_vals) |> dplyr::mutate(status = res_full$status)
  } else {
    tibble(term = names(true_vals), status = res_full$status)
  }
  aim1_rows <- aim1_rows |>
    dplyr::mutate(sweep = sweep, effect_value = effect_value, n_p = n_p, n_t = n_t, rep = rep)

  ## Aim 2: discovery/replication split (same grid cell = total N)
  pair <- generate_data_p2_pair(n_p_total = n_p, n_t_total = n_t,
                                 b_hlva = b_hlva, b_pcref = b_pcref, b_interaction = b_interaction,
                                 split_margin = split_margin, seed = seed_base + 1)
  res_disc <- fit_model_p2_guarded(pair$discovery, "correct")
  res_repl <- fit_model_p2_guarded(pair$replication, "correct")

  aim2_rows <- combine_replication_p2(res_disc, res_repl, true_vals) |>
    dplyr::mutate(
      sweep = sweep, effect_value = effect_value, n_p = n_p, n_t = n_t, rep = rep,
      split_margin = split_margin,
      status_discovery = res_disc$status, status_replication = res_repl$status
    )

  list(aim1 = aim1_rows, aim2 = aim2_rows)
}
```

The full grid (3 sweeps x 5 effect values x 30 cells x 50 replicates x 3
fits/combination = 67,500 total `glmer` fits) was run once, as a detached
background process, well before this document was written -- see the
"Development history" appendix for the validation/timing/cost detail. It is
**not** re-run here; the sections below load its saved output directly.

## Loading Aim 1/2 results and shared helpers

```{r}
#| label: load-aim1-aim2-results

aim1_files <- c("results_phase2/aim1_hlva.rds", "results_phase2/aim1_pcref.rds",
                 "results_phase2/aim1_interaction.rds")
aim2_files <- c("results_phase2/aim2_hlva.rds", "results_phase2/aim2_pcref.rds",
                 "results_phase2/aim2_interaction.rds")
missing_files <- c(aim1_files, aim2_files)[!file.exists(c(aim1_files, aim2_files))]
if (length(missing_files) > 0) {
  stop(
    "Missing Aim 1/2 result file(s): ", paste(missing_files, collapse = ", "),
    ". These are produced by the detached background job run_aim1_aim2_full.R ",
    "(~5h on 4 workers) -- this document does not regenerate them inline. ",
    "Run that script first if the files are genuinely absent."
  )
}

aim1_hlva        <- readRDS("results_phase2/aim1_hlva.rds")
aim1_pcref       <- readRDS("results_phase2/aim1_pcref.rds")
aim1_interaction <- readRDS("results_phase2/aim1_interaction.rds")
aim2_hlva        <- readRDS("results_phase2/aim2_hlva.rds")
aim2_pcref       <- readRDS("results_phase2/aim2_pcref.rds")
aim2_interaction <- readRDS("results_phase2/aim2_interaction.rds")

own_effect <- c(zHLVA = 0.363, zPCREFp = 0.095, `zHLVA:zPCREFp` = 0.054)
term_levels <- c("zHLVA", "zPCREFp", "zHLVA:zPCREFp")

# Each sweep's own term, evaluated only at its empirically-observed effect
# size -- the slice used by most tables/figures below.
aim1_all <- dplyr::bind_rows(
  aim1_hlva        |> dplyr::filter(term == "zHLVA",        effect_value == own_effect["zHLVA"]),
  aim1_pcref       |> dplyr::filter(term == "zPCREFp",       effect_value == own_effect["zPCREFp"]),
  aim1_interaction |> dplyr::filter(term == "zHLVA:zPCREFp", effect_value == own_effect["zHLVA:zPCREFp"])
) |> dplyr::mutate(term = factor(term, levels = term_levels))

aim2_all <- dplyr::bind_rows(
  aim2_hlva        |> dplyr::filter(term == "zHLVA",        effect_value == own_effect["zHLVA"]),
  aim2_pcref       |> dplyr::filter(term == "zPCREFp",       effect_value == own_effect["zPCREFp"]),
  aim2_interaction |> dplyr::filter(term == "zHLVA:zPCREFp", effect_value == own_effect["zHLVA:zPCREFp"])
) |> dplyr::mutate(term = factor(term, levels = term_levels))

# Full effect-size sweep, all points (not just the own-effect-size slice) --
# used for the Aim 1 power-curves figure below.
aim1_sweep_combined <- dplyr::bind_rows(
  aim1_hlva        |> dplyr::filter(term == "zHLVA"),
  aim1_pcref       |> dplyr::filter(term == "zPCREFp"),
  aim1_interaction |> dplyr::filter(term == "zHLVA:zPCREFp")
) |> dplyr::mutate(term = factor(term, levels = term_levels))

# Shared summary helper: power, sign reliability, Type S error rate,
# Type M ratio, median bias, and coverage, grouped by whatever columns are
# passed in `group_vars`.
summarize_p2_metrics <- function(df, group_vars) {
  df |>
    dplyr::group_by(dplyr::across(dplyr::all_of(group_vars))) |>
    dplyr::summarise(
      power            = mean(excludes_zero, na.rm = TRUE),
      sign_reliability = mean(correct_sign, na.rm = TRUE),
      type_s_rate      = mean(sign_error[excludes_zero], na.rm = TRUE),
      type_m_ratio     = median(inflation_ratio[excludes_zero], na.rm = TRUE),
      median_bias      = median(bias, na.rm = TRUE),
      coverage         = mean(covers_truth, na.rm = TRUE),
      n_fits           = dplyr::n(),
      .groups = "drop"
    )
}

aim1_metric_labels <- c(power = "Power\nP(significant)",
                         sign_reliability = "Sign reliability\nP(correct sign)",
                         type_s_rate = "Type S rate\nP(wrong sign | significant)",
                         type_m_ratio = "Type M ratio\nmedian |est/true| | significant")

aim2_metric_labels <- c(discovery_rate = "Discovery rate\nP(significant, half 1)",
                         replication_rate = "Replication rate\nP(sig. + same sign, both halves)",
                         conditional_replication = "Conditional replication\nP(replicates | discovery-sig.)")
```

## Shared plotting helpers

Aim 1 and Aim 2 each need a metrics-vs-`n_t` line/facet plot and an
`n_p`x`n_t` heatmap that differ only in which columns/labels are used. Both
are written once here and reused for each aim below (and are intended to be
reused again for the Aims 3-6 documents still to come).

```{r}
#| label: plot-helpers

plot_metrics_vs_x <- function(data, x_var, metrics, metric_labels,
                               color_var = "term", x_lab = NULL, x_breaks = n_t_grid) {
  x_lab <- x_lab %||% x_var
  df_long <- data |>
    tidyr::pivot_longer(dplyr::all_of(metrics), names_to = "metric", values_to = "value") |>
    dplyr::mutate(metric = factor(metric_labels[metric], levels = unname(metric_labels)))

  ggplot(df_long, aes(x = .data[[x_var]], y = value, color = .data[[color_var]])) +
    geom_line() +
    geom_point() +
    facet_wrap(~metric, scales = "free_y") +
    scale_x_log10(breaks = x_breaks) +
    scale_color_brewer(palette = "Dark2") +
    labs(x = x_lab, y = NULL, color = "Term") +
    theme_minimal()
}

plot_power_heatmap <- function(data, fill_var, facet_var = "term", fill_label = "Value",
                                grid_n_p = n_p_grid, grid_n_t = n_t_grid,
                                limits = c(0, 1), label_format = c("percent", "number"),
                                digits = 2, diverging = FALSE, midpoint = 0, squish = FALSE) {
  # Generalized from a power-only heatmap to any (n_p, n_t) diagnostic
  # surface. Defaults (limits = c(0, 1), percent labels, no squishing)
  # exactly reproduce the original power/replication-rate behavior, so
  # existing calls below are unaffected. `label_format = "number"` and/or
  # `diverging = TRUE` support diagnostics that aren't bounded proportions
  # (Type M ratio, median bias, RMSE); `squish = TRUE` lets a handful of
  # extreme outlier cells (common at n_t = 2) recolor to the scale's edge
  # color rather than stretching the whole gradient to accommodate them --
  # the text label still shows the real, uncapped value.
  label_format <- match.arg(label_format)
  label_fn <- if (label_format == "percent") {
    function(x) scales::percent(x, accuracy = 1)
  } else {
    function(x) sprintf(paste0("%.", digits, "f"), x)
  }
  oob_fn <- if (squish) scales::squish else scales::censor

  p <- ggplot(data, aes(x = factor(n_p, levels = grid_n_p), y = factor(n_t, levels = grid_n_t),
                    fill = .data[[fill_var]])) +
    geom_tile() +
    geom_text(aes(label = label_fn(.data[[fill_var]])), size = 2.8,
              color = if (diverging) "black" else "white") +
    facet_wrap(stats::as.formula(paste0("~", facet_var))) +
    labs(x = "Participant sample size (n_p)", y = "Text sample size (n_t)", fill = fill_label) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

  if (diverging) {
    p + scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B",
                              midpoint = midpoint, limits = limits, oob = oob_fn)
  } else if (label_format == "percent") {
    p + scale_fill_viridis_c(limits = limits, labels = scales::percent, oob = oob_fn)
  } else {
    p + scale_fill_viridis_c(limits = limits, oob = oob_fn)
  }
}
```

## Aim 1: is small-text-*N* instability a text-margin property or an interaction-specific property?

At each effect's own empirically-observed value, sign reliability and power
by `n_t`:

| `n_t` | zHLVA sign reliability | zPCREFp sign reliability | interaction sign reliability | zHLVA power | zPCREFp power | interaction power |
|---|---|---|---|---|---|---|
| 2 | 0.940 | **0.553** | 0.717 | 0.650 | 0.627 | 0.127 |
| 10 | 0.997 | 0.730 | 0.920 | 0.897 | 0.177 | 0.603 |
| 35 | 0.997 | 0.860 | 0.930 | 0.917 | 0.300 | 0.577 |
| 85 | 0.993 | 0.930 | 0.897 | 0.920 | 0.420 | 0.570 |
| 136 | 0.993 | 0.937 | 0.913 | 0.910 | 0.510 | 0.620 |

**This resolves the question the design set out to answer, and the answer is
not what the "reverse cohesion effect" framing might suggest.** At `n_t =
2`, `zPCREFp`'s sign reliability (0.553) is *worse* than the interaction's
(0.717) -- barely better than a coin flip -- while `zHLVA` (the
participant-level effect) is comparatively well-behaved (0.940). This is a
**text-margin property, not an interaction-specific property**: the
text-level main effect is at least as unstable as the interaction at the
smallest realistic text sample size, and by this particular diagnostic, more
so.

`zPCREFp`'s power also shows the "underpowered paradox" signature directly:
power is *higher* at `n_t = 2` (0.627) than at `n_t = 10` (0.177) -- not
because small samples are better powered, but because the noisiest, most
degenerate cell produces wildly inflated estimates that cross the
significance threshold more often *while being wrong about the sign* half
the time. A researcher who only sees "significant, p < .05" from a 2-text
study cannot distinguish this from a well-powered result.

### Aim 1 -- fuller metrics table

The table above reports sign reliability and power only. For completeness
against Marek et al.'s (2022) full diagnostic set, this expands to
**Type I (false-positive) rate**, Type S error rate, Type M ratio, median
bias, coverage, and RMSE. All columns except the new Type I column are
computed at each term's own empirically-observed effect size, pooled across
the six participant sample sizes; the Type I column instead comes from the
`effect_value = 0` point on the same sweep (the true-null case), since a
false-positive rate is only defined when there is nothing true to detect.
Surfacing it as an explicit table column (rather than leaving it visible
only as the orange curve in Figure 4, further below) makes it directly
comparable, row by row, against power and the other error rates.

```{r}
#| label: aim1-full-metrics-table

aim1_metrics_by_nt <- aim1_all |>
  dplyr::group_by(term, n_t) |>
  dplyr::summarise(
    power            = mean(excludes_zero, na.rm = TRUE),
    sign_reliability = mean(correct_sign, na.rm = TRUE),
    type_s_rate      = mean(sign_error[excludes_zero], na.rm = TRUE),
    type_m_ratio     = median(inflation_ratio[excludes_zero], na.rm = TRUE),
    median_bias      = median(bias, na.rm = TRUE),
    coverage         = mean(covers_truth, na.rm = TRUE),
    rmse_raw         = sqrt(mean(bias^2, na.rm = TRUE)),
    rmse_trimmed     = sqrt(mean(pmin(bias^2, quantile(bias^2, 0.95, na.rm = TRUE)), na.rm = TRUE)),
    n_fits           = dplyr::n(),
    .groups = "drop"
  )

aim1_type1_by_nt <- aim1_sweep_combined |>
  dplyr::filter(effect_value == 0) |>
  dplyr::group_by(term, n_t) |>
  dplyr::summarise(type_1_rate = mean(excludes_zero, na.rm = TRUE), .groups = "drop")

aim1_metrics_by_nt <- aim1_metrics_by_nt |>
  dplyr::left_join(aim1_type1_by_nt, by = c("term", "n_t")) |>
  dplyr::relocate(type_1_rate, .after = power)

aim1_metrics_by_nt |>
  dplyr::arrange(term, n_t) |>
  knitr::kable(
    digits = 3,
    col.names = c("Term", "n_t", "Power", "Type I rate", "Sign reliability",
                  "Type S rate", "Type M ratio", "Median bias", "Coverage",
                  "RMSE (raw)", "RMSE (trimmed)", "n fits")
  )
```

`rmse_raw` is included for completeness but is itself dominated by rare
pathological fits at small `n_t` (e.g. an outlying -20.3 estimate at
`n_t = 2`); `rmse_trimmed` (winsorizing squared bias at its 95th percentile
before averaging) is the more trustworthy summary of typical-case error
magnitude and should be preferred when comparing across cells. Note that
`type_1_rate` is nominally expected to be ~0.05 under correct calibration --
values well above that (as for `zPCREFp` at small `n_t`, previewed in the
summary table above and detailed mechanistically below) indicate
miscalibration, not merely bad luck.

### Aim 1 -- visualizations

#### Visualizations of power and error surfaces: showing power or error rate variation over variation in text sample size x participant sample size, for observed effects

```{r}
#| label: fig-aim1-power-surface
#| fig-width: 8
#| fig-height: 4
#| fig-cap: "Aim 1: power surface across the full joint (n_p, n_t) grid, at each term's own empirically-observed effect size. This is the direct 2-D analog of a Marek et al. (2022)-style Figure 3 sample-size surface: both axes are the two sample-size margins, and tile fill is the probability a 95% Wald CI excludes zero."

aim1_surface <- aim1_all |>
  dplyr::group_by(term, n_p, n_t) |>
  dplyr::summarise(power = mean(excludes_zero, na.rm = TRUE), .groups = "drop")

plot_power_heatmap(aim1_surface, fill_var = "power", fill_label = "Power")
```

The surface makes visible what the by-`n_t` table (pooled over `n_p`)
cannot -- and the accurate picture is more nuanced than a coarse read
suggests (the numbers below were checked directly against
`aim1_surface`). `zHLVA` is the best-behaved term, but its saturation is
**not** independent of `n_t`: at `n_t >= 10`, power already reaches
92-100% by `n_p = 100`; at `n_t = 2`, the same `n_p = 100` cell reaches
only 46%, and power doesn't approach that ceiling (92%) until `n_p =
1300`. `n_t = 2` still measurably slows `zHLVA` down -- just far less than
it does the two text-margin terms.

`zPCREFp` and the interaction, once the full surface is examined, turn out
to diverge sharply from each other, not just from `zHLVA`. The
**interaction** genuinely needs growth on both margins: at `n_t = 2` its
power never exceeds 24% anywhere on the `n_p` grid (2-24%), while at
`n_t >= 10` it reaches 90-96% by `n_p = 1300`+ -- a real, large power gain
from a bigger text sample. **`zPCREFp` does not show this pattern.** At
matched `n_p`, its power at `n_t = 2` (20-88% across the `n_p` grid) meets
or exceeds its power at *every* larger `n_t` in the grid (10, 35, 85, and
136) in 5 of 6 `n_p` cells -- e.g. at `n_p = 1300`, `n_t = 2` reaches 88%
while the best any larger `n_t` achieves there is 72% (`n_t = 136`). This
is not `zPCREFp` being well-powered at `n_t = 2`; it is the underpowered
paradox previewed above, visible directly on the surface -- the apparent
"power" at `n_t = 2` is inflated by the same singular-fit false-positive
mechanism documented later in this section, not a sign that more
participants are working as they should for this term.

::: {.callout-important}
## A second, independent illustration of the underpowered paradox

The `n_t = 2` reversal above is worth flagging explicitly as its **own**
piece of evidence for the underpowered paradox, distinct from the sign-
reliability/Type S evidence discussed earlier (`zPCREFp`'s 0.553 sign
reliability at `n_t = 2`, worse than the interaction's) and later
(the Type S/Type M curves). That evidence all requires first conditioning
on which fits were "significant" and then asking whether their *sign* was
right. The power-surface reversal needs no such conditioning: it shows up
in the raw significance rate itself, at every single `n_p` on the grid
(not just pooled, as in the earlier by-`n_t` table), without ever
inspecting sign or magnitude. Two structurally different diagnostics --
one about whether significant results have the correct sign, one about
how often results are significant at all -- point to the same underlying
`n_t = 2` identifiability breakdown, which is stronger corroboration than
either diagnostic could provide alone.
:::

The same `(n_p, n_t)` surface can be drawn for every other diagnostic in
the fuller metrics table above, not just power -- doing so shows directly
whether the text-margin concentration visible in the power surface is a
general property of the smallest cells or specific to power itself. These
seven surfaces are computed by term x `n_p` x `n_t` (i.e. one level finer
than the by-`n_t`-only fuller metrics table above, which pools over `n_p`).

```{r}
#| label: fig-aim1-error-surfaces
#| fig-width: 13
#| fig-height: 16
#| fig-cap: "Aim 1: the remaining seven diagnostics from the fuller metrics table, each drawn as its own (n_p, n_t) surface (the same layout as the power surface above). Type I rate, sign reliability, Type S rate, and coverage share the fixed [0, 100%] color scale used for power, for direct comparability; Type M ratio and RMSE (trimmed) use a free color scale, capped (squished, not truncated -- the printed number is always the true value) at a fixed upper limit so that rare extreme outlier cells at n_t = 2 don't wash out the rest of the scale; median bias uses a diverging scale centered at zero."

aim1_surfaces_extra <- aim1_all |>
  dplyr::group_by(term, n_p, n_t) |>
  dplyr::summarise(
    sign_reliability = mean(correct_sign, na.rm = TRUE),
    type_s_rate      = mean(sign_error[excludes_zero], na.rm = TRUE),
    type_m_ratio     = median(inflation_ratio[excludes_zero], na.rm = TRUE),
    median_bias      = median(bias, na.rm = TRUE),
    coverage         = mean(covers_truth, na.rm = TRUE),
    rmse_trimmed     = sqrt(mean(pmin(bias^2, quantile(bias^2, 0.95, na.rm = TRUE)), na.rm = TRUE)),
    .groups = "drop"
  )

aim1_type1_surface <- aim1_sweep_combined |>
  dplyr::filter(effect_value == 0) |>
  dplyr::group_by(term, n_p, n_t) |>
  dplyr::summarise(type_1_rate = mean(excludes_zero, na.rm = TRUE), .groups = "drop")

aim1_surfaces_extra <- aim1_surfaces_extra |>
  dplyr::left_join(aim1_type1_surface, by = c("term", "n_p", "n_t"))

bias_limit <- max(abs(aim1_surfaces_extra$median_bias), na.rm = TRUE)

panel_surface_type1 <- plot_power_heatmap(
  aim1_surfaces_extra, fill_var = "type_1_rate", fill_label = "Type I\nrate"
) + labs(title = "Type I (false-positive) rate")

panel_surface_signrel <- plot_power_heatmap(
  aim1_surfaces_extra, fill_var = "sign_reliability", fill_label = "Sign\nreliability"
) + labs(title = "Sign reliability")

panel_surface_types <- plot_power_heatmap(
  aim1_surfaces_extra, fill_var = "type_s_rate", fill_label = "Type S\nrate"
) + labs(title = "Type S (sign) error rate")

panel_surface_coverage <- plot_power_heatmap(
  aim1_surfaces_extra, fill_var = "coverage", fill_label = "Coverage"
) + labs(title = "95% CI coverage")

panel_surface_typem <- plot_power_heatmap(
  aim1_surfaces_extra, fill_var = "type_m_ratio", fill_label = "Type M\nratio",
  limits = c(0, 10), label_format = "number", digits = 1, squish = TRUE
) + labs(title = "Type M (magnitude) ratio")

panel_surface_bias <- plot_power_heatmap(
  aim1_surfaces_extra, fill_var = "median_bias", fill_label = "Median\nbias",
  limits = c(-bias_limit, bias_limit), label_format = "number", digits = 3, diverging = TRUE
) + labs(title = "Median bias")

panel_surface_rmse <- plot_power_heatmap(
  aim1_surfaces_extra, fill_var = "rmse_trimmed", fill_label = "RMSE\n(trimmed)",
  limits = c(0, 1.5), label_format = "number", digits = 2, squish = TRUE
) + labs(title = "RMSE (trimmed)")

patchwork::wrap_plots(
  panel_surface_type1, panel_surface_signrel, panel_surface_types, panel_surface_coverage,
  panel_surface_typem, panel_surface_bias, panel_surface_rmse,
  ncol = 2
)
```

Three findings stand out. First, **Type I rate, sign reliability, Type S
rate, and coverage all show the same qualitative pattern as power**: the
worst cells are concentrated at `n_t = 2`, and (for `zPCREFp` and the
interaction) growing `n_p` alone at `n_t = 2` does not fix them --
confirming that the text-margin concentration already established for
power is a general property of this design's smallest cells, not specific
to the power diagnostic. Coverage is the starkest illustration: nominal
coverage should be ~95% everywhere, but `zPCREFp`'s coverage at `n_t = 2`
falls to as low as 10-12% at the *largest* participant sample sizes (`n_p`
= 1300-1497) -- more data on the wrong margin makes the miscalibration
*more* confidently wrong, not less, exactly the "overconfident, too-narrow
Wald SE" mechanism documented later in this section.

Second, **the Type M ratio panel needs a specific interpretive caveat**:
its most extreme cell (the interaction, `n_p = 25`, `n_t = 2`, median ratio
~85x, squished to the scale's dark edge here) is not evidence of an
enormous absolute error -- the interaction's own effect size is only
0.054, so even a modest absolute bias becomes a huge *ratio* when divided
by a true value that small. The median bias panel alongside it shows this
directly: median bias stays within +/-0.12 in every cell on this entire
grid, including the ones with triple-digit Type M ratios. Ratio-based
inflation measures like Type M are most informative for effects of
moderate-to-large size and can be misleadingly dramatic-looking for a
near-zero true effect; this document's own `zHLVA:zPCREFp` sweep is a
concrete illustration of that general statistical caveat, not a sign of a
uniquely severe problem with the interaction term.

(A related, more mundane artifact: one cell -- the interaction at
`n_p = 100, n_t = 85` -- shows as blank/grey in the Type S and Type M
panels. Both diagnostics are conditional on significance, and this
particular cell happened to have zero significant fits among its 50
replicates; dividing by zero significant fits leaves the diagnostic
undefined rather than zero, which `ggplot2` renders as a missing tile
rather than a colored one. This is a small-sample artifact of this
specific `(n_p, n_t)` cell, not a data error.)

Third, the **RMSE (trimmed) surface differs from the by-`n_t`-only version
in the fuller metrics table above** because it is computed per `(term,
n_p, n_t)` cell here (50 replicates) rather than pooled over all six `n_p`
values (300 replicates) -- the finer grouping means the within-cell 95th-
percentile winsorization threshold is itself estimated from fewer points,
so it is less robust to a single extreme replicate (e.g. `zHLVA` at `n_p =
25, n_t = 2`, where one pathological fit still pulls the trimmed RMSE up
to 2.86 despite `zHLVA` otherwise behaving well at that cell). Treat the
by-`n_t`-only table's RMSE column as the more stable summary, and this
finer surface as showing *where* that instability is concentrated rather
than as a precise per-cell estimate in its own right.

#### Visualizations of power or error rate (sign reliability, type S, type M) variation over variation in text sample size and over participant sample size, for observed effects

```{r}
#| label: fig-aim1-metrics-vs-np
#| fig-cap: "Aim 1: power, sign reliability, Type S error rate, and Type M exaggeration ratio vs. participant sample size (n_p), at each term's own empirically-observed effect size, pooled across all five text sample sizes and 50 replicates/cell."

aim1_metrics_by_np <- aim1_all |>
  dplyr::group_by(term, n_p) |>
  dplyr::summarise(
    power            = mean(excludes_zero, na.rm = TRUE),
    sign_reliability = mean(correct_sign, na.rm = TRUE),
    type_s_rate      = mean(sign_error[excludes_zero], na.rm = TRUE),
    type_m_ratio     = median(inflation_ratio[excludes_zero], na.rm = TRUE),
    n_fits           = dplyr::n(),
    .groups = "drop"
  )

plot_metrics_vs_x(
  aim1_metrics_by_np, x_var = "n_p",
  metrics = c("power", "sign_reliability", "type_s_rate", "type_m_ratio"),
  metric_labels = aim1_metric_labels,
  x_lab = "Participant sample size (n_p, log scale)",
  x_breaks = n_p_grid
)
```

This is the participant-margin counterpart of the text-margin figure
below: the same four diagnostics, pooled over all five `n_t` values
instead of all six `n_p` values, so the two figures together examine each
diagnostic over both margins. `zHLVA`'s curves here should look like
conventional, well-behaved diagnostics improving monotonically with `n_p`
(power and sign reliability rising, Type S/M falling), consistent with it
being the participant-margin-limited effect; `zPCREFp` and the interaction
are pooled here across text sample sizes that range from the pathological
`n_t = 2` up to the well-behaved `n_t = 136`, so any distinctive `n_t = 2`
signature (the underpowered paradox) is diluted rather than isolated in
this view -- the `n_t`-axis figure below, and the full `(n_p, n_t)`
surfaces above, remain the more direct way to see that pathology.

```{r}
#| label: fig-aim1-metrics-vs-nt
#| fig-cap: "Aim 1: power, sign reliability, Type S error rate, and Type M exaggeration ratio vs. text sample size (n_t), at each term's own empirically-observed effect size, pooled across all six participant sample sizes and 50 replicates/cell."

plot_metrics_vs_x(
  aim1_metrics_by_nt, x_var = "n_t",
  metrics = c("power", "sign_reliability", "type_s_rate", "type_m_ratio"),
  metric_labels = aim1_metric_labels,
  x_lab = "Text sample size (n_t, log scale)"
)
```

This is the one-axis (pooled-over-`n_p`) counterpart of the surface above,
and the visual counterpart of Aim 1's prose finding: `zPCREFp` has the
*worst* sign reliability and the *highest* Type S/Type M values at
`n_t = 2`, worse even than the interaction -- while also having power
*higher* at `n_t = 2` than at `n_t = 10`, the "underpowered paradox"
signature (a significant result here is more likely to be wrong, not more
trustworthy).

##### Mechanism: what the underpowered paradox looks like at the individual-estimate level

The summary diagnostics above (power, sign reliability, Type S/M rates) are
all *aggregates* over 50 replicate fits per cell. It can be easier to see
why the paradox happens by looking at the individual replicate estimates
themselves, rather than only their summary statistics. The three panels
below show every individual estimate from Aim 1's fits at `n_p = 1497` --
the **largest** participant sample size in the grid, held constant across
every row and column so that the funnel shown reflects the text margin
(`n_t`) specifically, not participant-margin noise -- for `zHLVA` (top),
`zPCREFp` (middle), and the `zHLVA:zPCREFp` interaction (bottom): one dot
per replicate fit, one column per simulated true effect size, one row per
text sample size. `zHLVA` is included here (extending the original
two-panel version of this figure) for the same reason the tables/heatmaps
above already treat all three terms symmetrically: since the whole point of
this document is now to compare all three effects on the same footing, not
just the interaction and its text-level counterpart, the mechanism-level
view should do the same. This uses only already-computed replicate
estimates (`results_phase2/aim1_hlva.rds`) -- no new simulation.

```{r}
#| label: fig-aim1-distribution-explainer
#| fig-width: 12
#| fig-height: 12.5
#| fig-cap: "Aim 1: individual estimates (one dot per replicate fit, n_p held constant at its largest grid value, 1497) for zHLVA (top), zPCREFp (middle), and the zHLVA:zPCREFp interaction (bottom), by text sample size (rows, narrower at top) and true effect size (columns, labeled \"true = ...\"). Dashed vertical line = the true value being simulated in that column. Red = a 95% CI that excludes zero (\"significant\"); grey = not significant. A handful of pathological estimates beyond the plotted range (all at n_t = 2) are clipped to the axis edge for readability."

illustrative_n_p <- max(n_p_grid)

make_aim1_distribution_panel <- function(data, term_name, x_clip, illustrative_n_p,
                                          show_x_axis = TRUE) {
  df <- data |>
    dplyr::filter(term == term_name, n_p == illustrative_n_p) |>
    dplyr::mutate(
      n_t_f = factor(n_t, levels = n_t_grid),
      significant = ifelse(excludes_zero, "Significant (95% CI excludes 0)", "Not significant"),
      estimate_clipped = pmax(pmin(estimate, x_clip), -x_clip)
    )

  ggplot(df, aes(x = estimate_clipped, y = n_t_f, color = significant, fill = significant)) +
    ggdist::stat_dots(
      side = "both", scale = 0.85, alpha = 0.55,
      binwidth = ggplot2::unit(c(1.3, 2.2), "mm"),
      overflow = "compress"
    ) +
    geom_vline(aes(xintercept = effect_value), linetype = "dashed", color = "black", linewidth = 0.5) +
    facet_wrap(~effect_value, nrow = 1,
               labeller = as_labeller(function(x) paste0("true = ", x))) +
    scale_x_continuous(breaks = c(-x_clip, 0, x_clip)) +
    scale_color_manual(values = c("Significant (95% CI excludes 0)" = "#b2182b",
                                   "Not significant" = "grey45")) +
    scale_fill_manual(values = c("Significant (95% CI excludes 0)" = "#b2182b",
                                  "Not significant" = "grey45")) +
    labs(
      x = if (show_x_axis) paste0(term_name, " estimate (clipped to [-", x_clip, ", ", x_clip, "])") else NULL,
      y = "n_t", color = NULL, fill = NULL
    ) +
    theme_bw(base_size = 10) +
    theme(
      panel.border = element_rect(color = "grey35", fill = NA, linewidth = 0.7),
      panel.spacing = unit(1.1, "lines"),
      strip.background = element_rect(fill = "grey85", color = "grey35"),
      strip.text = element_text(face = "bold", size = 8.5),
      panel.grid.minor = element_blank(),
      panel.grid.major.y = element_blank(),
      legend.position = "bottom"
    )
}

panel_dist_hlva <- make_aim1_distribution_panel(
  aim1_hlva, "zHLVA", x_clip = 2, illustrative_n_p = illustrative_n_p, show_x_axis = FALSE
) + labs(title = "zHLVA (participant-level main effect)")

panel_dist_pcref <- make_aim1_distribution_panel(
  aim1_pcref, "zPCREFp", x_clip = 2, illustrative_n_p = illustrative_n_p, show_x_axis = FALSE
) + labs(title = "zPCREFp (text-level main effect)")

panel_dist_interaction <- make_aim1_distribution_panel(
  aim1_interaction, "zHLVA:zPCREFp", x_clip = 1.5, illustrative_n_p = illustrative_n_p, show_x_axis = TRUE
) + labs(title = "zHLVA:zPCREFp (interaction)")

(panel_dist_hlva / panel_dist_pcref / panel_dist_interaction) +
  patchwork::plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

Four things are visible here that the summary tables/plots only imply.
First, the funnel shape itself, present in **all three** rows: each cloud
of dots is wide at the bottom (`n_t = 2`) and collapses to a tight cluster
around the dashed true-value line by the top (`n_t = 136`) -- the textbook
"wide at small *N*, narrow at large *N*" funnel, for every true effect size
simulated, for every term (participant-level or text-level). Second, and
this is the paradox in its most direct form: in `zPCREFp`'s leftmost
column (`true = 0`, i.e. **no true effect at all**), a substantial share of
the `n_t = 2` dots are still colored red -- individually indistinguishable,
in a real analysis, from a genuine discovery. Those red dots at `n_t = 2`
are false positives purely from noise, and they are just as red (just as
"significant", p < .05) as the true positives further right. A
significance threshold cannot tell these apart; only a larger text sample
-- more rows lower down turning into fewer, tighter, more-often-
correctly-colored rows higher up -- can.

Third, `zHLVA`'s top row makes the *contrast* with the text-level terms
directly visible for the first time in this document: its `n_t = 2` funnel
is comparatively narrow already (few dots reach the clip edge), and its
`true = 0` column has only a modest excess of red dots relative to the
nominal 5% -- consistent with its Type I rate staying much closer to
nominal at every `n_t` (Table 2 above). The instability visible in the
`zPCREFp` and interaction rows is not a generic property of "estimating
anything at `n_t = 2`" -- it is specific to terms whose identification
depends on the text margin.

Fourth, comparing the `zPCREFp` and interaction rows directly answers
whether the interaction's funnel looks like `zPCREFp`'s: **not really**. At
`n_t = 2`, the interaction's dots are mostly grey, not red, across every
column including its own empirically-observed effect size (`true = 0.054`)
-- consistent with its much lower power at that cell (0.127, vs.
`zPCREFp`'s 0.627 from the table above). Where `zPCREFp`'s problem at
`n_t = 2` is "too many confidently-wrong answers," the interaction's is
closer to "usually no answer at all, and on the rare occasions there is
one, it is not obviously more trustworthy than `zPCREFp`'s" -- the same
underlying text-margin scarcity, but manifesting as a power problem more
than a false-discovery problem for this particular term.

###### Why can an estimate be "significant" when the true effect is exactly zero?

This is the question the `true = 0` column above raises most sharply, and
it is worth answering mechanistically rather than leaving it as a
statistical curiosity. Two different things are going on, and it is
important to distinguish them.

**It is not simply the ordinary 5% false-positive rate.** A correctly
calibrated 95% CI is *expected* to exclude a true value of zero 5% of the
time, by construction -- that is what "95%" means. But the observed rate at
`n_t = 2` is far above that: the table below shows that 66.3% of the
`zPCREFp` fits at `n_t = 2` under a true null are "significant," roughly
13x the nominal rate. Something is actively miscalibrated here, not merely
unlucky.

**The mechanism is a random-effects identifiability breakdown, not sampling
bad luck on the covariate itself.** With only `n_t = 2` texts, there is
exactly one degree of freedom available to describe how the two sampled
texts differ from each other -- two points determine a line perfectly, with
nothing left over to check whether that line is trustworthy. The "correct"
model asks `glmer` to split any such between-text difference into two
pieces: a text-level random intercept (`(1 | text_id)`, capturing
idiosyncratic, unexplained differences between texts) and the fixed
`zPCREFp` slope (capturing the cohesion effect specifically). Estimating a
random-intercept *variance* from only 2 groups is close to impossible --
there is essentially no information to distinguish "these two texts differ
a lot by chance" from "these two texts differ a lot systematically." In
practice, `lme4` resolves this by pinning the text-level variance at (or
extremely near) its zero boundary -- a **singular fit**:

```{r}
#| label: fig-aim1-singular-fit-diagnostic

null_nt2_diagnostic <- aim1_pcref |>
  dplyr::filter(term == "zPCREFp", n_t == 2, effect_value == 0)

cat("Fit status, all 300 replicates (n_t = 2, true zPCREFp effect = 0):\n")
table(null_nt2_diagnostic$status)

null_nt2_diagnostic |>
  dplyr::group_by(excludes_zero) |>
  dplyr::summarise(
    n = dplyr::n(),
    `median |estimate|` = round(median(abs(estimate), na.rm = TRUE), 3),
    `median std. error` = round(median(std.error, na.rm = TRUE), 3),
    .groups = "drop"
  )
```

Every single one of these 300 replicates converges to a singular fit --
the text-level random-intercept variance collapses to zero every time
`n_t = 2`, regardless of how many participants are simulated. Once that
variance is pinned at zero, the model has nowhere else to put the two
sampled texts' chance difference except into the only fixed-effect
predictor that differs between them -- `zPCREFp`. This inflates the fixed-
effect point estimate (median |estimate| = 0.734 among the "significant"
fits, versus a true value of exactly 0) and simultaneously produces an
overconfident, too-narrow Wald standard error (median SE = 0.106 among the
"significant" fits, versus 0.294 among the fits that correctly stayed
non-significant) -- because the SE formula implicitly assumes the
random-effects variance structure has been reliably estimated, an
assumption that fails precisely at this boundary. Inflated estimate +
deflated SE together push a large share of purely-noise-driven fits across
the 95%-CI-excludes-zero line.

In short: at `n_t = 2`, the model is not really testing "does cohesion
predict comprehension" -- it is testing "do these two specific,
randomly-sampled texts differ," and then attributing 100% of whatever
difference it finds to cohesion, because there is no other place left for
that difference to go. This is a more extreme version of the same
family of problem documented elsewhere in this project as random-effects
misspecification (Aim 3's omitted-`text.id` bias, and Phase 1's finding
that pseudoreplication introduces genuine bias, not just imprecision) --
here the grouping factor is nominally *in* the model formula, but with only
2 groups it is statistically unidentifiable in practice, which behaves much
like omitting it.

#### Visualizations of power and error surfaces: showing power or error rate variation over variation in text sample size x participant sample size, for varying effects

Strictly, **power** is the probability of correctly detecting an effect
that truly exists in the population, under conventional null-hypothesis
significance testing -- it is only a meaningful concept when the true
effect is non-zero. The power surface and the metrics-vs-`n_t` plot above
both hold the true effect size fixed at each term's own observed value;
the figure below instead sweeps the true effect size down to and through
zero for all three terms, so that the same plot shows genuine power curves
(non-zero effect) alongside the **Type I (false-positive) rate** (the
`effect size = 0` case, where there is nothing to detect). Rather than
color that `effect size = 0` line as if it were simply the lightest point
on the same power-curve palette -- which would imply it answers the same
question as its non-zero neighbors -- it is drawn in a distinct,
colorblind-friendly bright orange, held constant across all three rows.

```{r}
#| label: fig-aim1-power-curves
#| fig-width: 10
#| fig-height: 9.5
#| fig-cap: "Aim 1: classic power curves -- P(significant) vs. participant sample size (n_p, log scale), one row per term, faceted by text sample size (n_t) across columns. Each row uses its own sequential color palette (darker = larger effect size); the thicker line/larger points mark each term's empirically-observed effect size; the bright orange line/points are the Type I (false-positive) rate, i.e. the effect size = 0 case."

aim1_metrics_by_effect <- aim1_sweep_combined |>
  dplyr::group_by(term, effect_value, n_p, n_t) |>
  dplyr::summarise(
    power            = mean(excludes_zero, na.rm = TRUE),
    sign_reliability = mean(correct_sign, na.rm = TRUE),
    coverage         = mean(covers_truth, na.rm = TRUE),
    type_s_rate      = mean(sign_error[excludes_zero], na.rm = TRUE),
    type_m_ratio     = median(inflation_ratio[excludes_zero], na.rm = TRUE),
    .groups = "drop"
  )

# Each term sweeps a different set of effect-size values (see the sweep
# grids above), so a single shared color scale across all three rows would
# either compress each term's own 5-point range into indistinguishable
# shades or mix three unrelated numeric scales under one legend. Instead,
# each term gets its own sequential ColorBrewer palette (darker = larger
# effect), and the curve at that term's empirically-observed effect size
# (the middle grid point in every sweep) is drawn thicker and with larger
# points so it is identifiable at a glance. Generalized from the original
# power-only panel builder so the same machinery produces sign-reliability
# and coverage curves below: `metric_var`/`y_lab` select which column to
# plot, `include_zero` drops the true-null point entirely when a metric
# has no meaningful value there (sign reliability -- see prose below), and
# `highlight_zero_orange` controls whether a still-included zero point gets
# the "different question" orange treatment used for power's Type I rate.
make_aim1_curve_panel <- function(term_name, effect_grid, palette_name, metric_var, y_lab,
                                   axis_title = NULL, legend_title = NULL,
                                   include_zero = TRUE, zero_label = "no true effect",
                                   highlight_zero_orange = TRUE, ref_line = NULL, log_y = FALSE,
                                   y_limits = NULL) {
  own_value <- own_effect[[term_name]]
  eff_grid <- if (include_zero) effect_grid else effect_grid[effect_grid != 0]

  # Skip the palette's palest 2 shades (indices 1-2 of the 9-class Brewer
  # palette) before interpolating -- the raw 5-class palette's lightest
  # color is too close to white/background to read easily as a line.
  pal <- grDevices::colorRampPalette(RColorBrewer::brewer.pal(9, palette_name)[3:9])(length(eff_grid))
  if (include_zero && highlight_zero_orange && 0 %in% eff_grid) {
    pal[eff_grid == 0] <- "#E69F00"  # Okabe-Ito orange, colorblind-friendly
  }
  effect_labels <- dplyr::case_when(
    eff_grid == 0 & include_zero ~ paste0(eff_grid, " (", zero_label, ")"),
    eff_grid == own_value        ~ paste0(eff_grid, " (observed)"),
    TRUE                         ~ as.character(eff_grid)
  )

  df <- aim1_metrics_by_effect |>
    dplyr::filter(term == term_name, effect_value %in% eff_grid) |>
    dplyr::mutate(
      effect_f = factor(effect_value, levels = eff_grid, labels = effect_labels),
      is_own   = effect_value == own_value
    )

  p <- ggplot(df, aes(x = n_p, y = .data[[metric_var]], color = effect_f, group = effect_f)) +
    geom_line(aes(linewidth = is_own)) +
    geom_point(aes(size = is_own)) +
    facet_wrap(~n_t, nrow = 1, labeller = label_both) +
    scale_x_log10(breaks = n_p_grid, labels = scales::label_comma()) +
    scale_color_manual(values = pal, name = legend_title %||% term_name) +
    scale_linewidth_manual(values = c(`FALSE` = 0.5, `TRUE` = 1.3), guide = "none") +
    scale_size_manual(values = c(`FALSE` = 1, `TRUE` = 2.2), guide = "none") +
    labs(x = axis_title, y = y_lab) +
    theme_minimal(base_size = 9) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 6.5),
      legend.position = "right"
    )

  if (!is.null(ref_line)) {
    p <- p + geom_hline(yintercept = ref_line, linetype = "dotted", color = "grey40", linewidth = 0.4)
  }
  if (log_y) {
    p <- p + scale_y_log10()
  }
  # coord_cartesian (not scale limits) so a fixed [0, 1] y-axis is shared
  # across power/sign-reliability/coverage/Type S -- directly comparable
  # panel-to-panel -- without silently dropping any data point that might
  # fall (even marginally) outside that range.
  if (!is.null(y_limits)) {
    p <- p + coord_cartesian(ylim = y_limits)
  }
  p
}

panel_hlva <- make_aim1_curve_panel("zHLVA", b_hlva_grid, "Greens",
                                     metric_var = "power", y_lab = "Power",
                                     legend_title = "zHLVA effect size",
                                     zero_label = "no true effect -- Type I rate",
                                     y_limits = c(0, 1))
panel_pcref <- make_aim1_curve_panel("zPCREFp", b_pcref_grid, "Blues",
                                      metric_var = "power", y_lab = "Power",
                                      legend_title = "zPCREFp effect size",
                                      zero_label = "no true effect -- Type I rate",
                                      y_limits = c(0, 1))
panel_interaction <- make_aim1_curve_panel("zHLVA:zPCREFp", b_interaction_grid, "Purples",
                                            metric_var = "power", y_lab = "Power",
                                            axis_title = "Participant sample size (n_p, log scale)",
                                            legend_title = "Interaction effect size",
                                            zero_label = "no true effect -- Type I rate",
                                            y_limits = c(0, 1))

panel_hlva / panel_pcref / panel_interaction
```

The `zPCREFp` row is the most diagnostic: at `n_t = 2`, the bright orange
Type I (false-positive) rate curve (true effect = 0) stays **non-monotonic
and never near zero across the entire `n_p` range** (~20-60%) -- i.e., a
sizeable share of fits at this text sample size are "significant" purely
from noise, regardless of how many participants are added. At larger
`n_t`, that same orange curve flattens out near zero as expected (a
well-calibrated Type I rate), and the colored (genuine power) curves
separate out into a conventional monotonically-rising power curve as `n_p`
grows -- confirming the pathology is specific to the `n_t = 2` regime, not
a general property of `zPCREFp`. This is the clearest single illustration
in this document of the underpowered paradox's mechanism: at the modal
literature text sample size, growing the participant sample does not
rescue the false-positive problem -- it is a text-margin problem, not a
participant-margin one. The highlighted "(observed)" curve in each row --
the effect size actually estimated from this project's real data -- sits in
the middle of its own sweep by construction, letting readers see directly
how much power/error-rate behavior would differ if the true effect were
smaller or larger than what was actually observed.

The same effect-size x `n_p` x `n_t` curve layout can be drawn for **sign
reliability** and **95% CI coverage** as well, using the same
`make_aim1_curve_panel()` machinery as the power curves above -- this
shows whether the same margin-dependence and effect-size sensitivity
apply to those two diagnostics, not just power.

```{r}
#| label: fig-aim1-sign-reliability-curves
#| fig-width: 10
#| fig-height: 9.5
#| fig-cap: "Aim 1: sign reliability -- P(estimated sign matches true sign) -- vs. participant sample size (n_p, log scale), one row per term, faceted by text sample size (n_t). Layout and color convention match the power curves above, except the true-null (effect = 0) point is omitted (see prose below)."

# Sign reliability has no meaningful value at effect_value = 0: with no
# true effect, sign(true_value) = 0, so `correct_sign` (sign(estimate) ==
# sign(true_value)) is essentially never TRUE by construction, not because
# of any instability -- there is no "correct sign" to recover when there
# is nothing there. The zero point is therefore dropped entirely
# (`include_zero = FALSE`) rather than plotted as a degenerate near-0%
# curve or given the power curves' orange "different question" treatment.
panel_hlva_sr <- make_aim1_curve_panel("zHLVA", b_hlva_grid, "Greens",
                                        metric_var = "sign_reliability", y_lab = "Sign reliability",
                                        legend_title = "zHLVA effect size", include_zero = FALSE,
                                        y_limits = c(0, 1))
panel_pcref_sr <- make_aim1_curve_panel("zPCREFp", b_pcref_grid, "Blues",
                                         metric_var = "sign_reliability", y_lab = "Sign reliability",
                                         legend_title = "zPCREFp effect size", include_zero = FALSE,
                                         y_limits = c(0, 1))
panel_interaction_sr <- make_aim1_curve_panel("zHLVA:zPCREFp", b_interaction_grid, "Purples",
                                               metric_var = "sign_reliability", y_lab = "Sign reliability",
                                               axis_title = "Participant sample size (n_p, log scale)",
                                               legend_title = "Interaction effect size", include_zero = FALSE,
                                               y_limits = c(0, 1))

panel_hlva_sr / panel_pcref_sr / panel_interaction_sr
```

`zPCREFp`'s row again shows the underpowered paradox's flatness signature,
this time on sign reliability rather than power: at `n_t = 2`, sign
reliability sits in the 0.4-0.8 range **across the entire effect-size
sweep**, including its largest simulated effect size (0.35, comparable in
magnitude to `zHLVA`'s own effect) -- and it stays there regardless of
`n_p`, from 25 participants up to the full 1497. Growing the participant
sample at `n_t = 2` does not resolve the coin-flip-adjacent sign
uncertainty even for a comparatively large text-level effect. By contrast,
the same effect size (0.35) reaches ~0.92-1.0 sign reliability once
`n_t = 10`, and a perfect 1.0 by `n_t = 136` at every `n_p`. `zHLVA`'s row
again saturates near 1.0 already at modest `n_p`, independent of `n_t` --
the same participant-margin/text-margin split documented throughout this
section holds for sign reliability just as it does for power.

```{r}
#| label: fig-aim1-coverage-curves
#| fig-width: 10
#| fig-height: 9.5
#| fig-cap: "Aim 1: 95% CI coverage -- P(the Wald CI contains the true value) -- vs. participant sample size (n_p, log scale), one row per term, faceted by text sample size (n_t). Dotted horizontal line marks nominal 95% coverage. Unlike sign reliability, coverage is well-defined at effect = 0 (a CI can cover a true value of exactly zero) and is included without the orange re-color, since it is not answering a qualitatively different question there -- 95% coverage is the target at every effect size, including zero."

panel_hlva_cov <- make_aim1_curve_panel("zHLVA", b_hlva_grid, "Greens",
                                         metric_var = "coverage", y_lab = "95% CI coverage",
                                         legend_title = "zHLVA effect size",
                                         highlight_zero_orange = FALSE, zero_label = "no true effect",
                                         ref_line = 0.95, y_limits = c(0, 1))
panel_pcref_cov <- make_aim1_curve_panel("zPCREFp", b_pcref_grid, "Blues",
                                          metric_var = "coverage", y_lab = "95% CI coverage",
                                          legend_title = "zPCREFp effect size",
                                          highlight_zero_orange = FALSE, zero_label = "no true effect",
                                          ref_line = 0.95, y_limits = c(0, 1))
panel_interaction_cov <- make_aim1_curve_panel("zHLVA:zPCREFp", b_interaction_grid, "Purples",
                                                metric_var = "coverage", y_lab = "95% CI coverage",
                                                axis_title = "Participant sample size (n_p, log scale)",
                                                legend_title = "Interaction effect size",
                                                highlight_zero_orange = FALSE, zero_label = "no true effect",
                                                ref_line = 0.95, y_limits = c(0, 1))

panel_hlva_cov / panel_pcref_cov / panel_interaction_cov
```

`zPCREFp`'s coverage curves show the clearest and most alarming pattern of
the three metrics covered in this section: at `n_t = 2`, coverage falls
from ~80-86% at `n_p = 25` down to as low as **10-24%** at `n_p` = 1300 or
1497 -- and this collapse happens **at every effect size on the sweep,
including the true-null case** (coverage 0.80 -> 0.16 across the same
`n_p` range when the true effect is exactly zero). That "more data makes
it worse" direction is the opposite of what more participants should do,
and confirms that the overconfident-Wald-SE mechanism documented earlier
in this section (the text-level random-intercept variance collapsing to
zero at `n_t = 2`, in "Why can an estimate be 'significant' when the true
effect is exactly zero?" above) is not a quirk of the one effect size
shown in the earlier error-surfaces figure -- it operates identically
whether or not there is a true effect to detect. `zHLVA` and, to a lesser extent, the interaction
stay much closer to the nominal 95% line (dotted) throughout, again
consistent with their comparative insulation from this project's
chronic text-sample scarcity.

The same treatment extends to the two remaining diagnostics from the
metrics-vs-`n_t` four-panel figure earlier in this section -- **Type S
(sign) error rate** and **Type M (magnitude) ratio** -- both conditional
on significance, so both use the same `include_zero = FALSE` convention
as sign reliability above (a "wrong sign" or "inflation ratio" needs a
true, non-zero sign/magnitude to be wrong or inflated relative to; see the
prose below the figures for the mechanistic detail).

```{r}
#| label: fig-aim1-type-s-curves
#| fig-width: 10
#| fig-height: 9.5
#| fig-cap: "Aim 1: Type S (sign) error rate -- P(wrong sign | significant) -- vs. participant sample size (n_p, log scale), one row per term, faceted by text sample size (n_t). The true-null (effect = 0) point is omitted, as for sign reliability above: conditioning on significance at a true effect of exactly zero makes Type S trivially ~100% by construction, not informative of miscalibration."

panel_hlva_ts <- make_aim1_curve_panel("zHLVA", b_hlva_grid, "Greens",
                                        metric_var = "type_s_rate", y_lab = "Type S rate",
                                        legend_title = "zHLVA effect size", include_zero = FALSE,
                                        y_limits = c(0, 1))
panel_pcref_ts <- make_aim1_curve_panel("zPCREFp", b_pcref_grid, "Blues",
                                         metric_var = "type_s_rate", y_lab = "Type S rate",
                                         legend_title = "zPCREFp effect size", include_zero = FALSE,
                                         y_limits = c(0, 1))
panel_interaction_ts <- make_aim1_curve_panel("zHLVA:zPCREFp", b_interaction_grid, "Purples",
                                               metric_var = "type_s_rate", y_lab = "Type S rate",
                                               axis_title = "Participant sample size (n_p, log scale)",
                                               legend_title = "Interaction effect size", include_zero = FALSE,
                                               y_limits = c(0, 1))

panel_hlva_ts / panel_pcref_ts / panel_interaction_ts
```

`zPCREFp` again shows the by-now-familiar flat, margin-dependent pattern,
mirroring its own sign-reliability curves (Type S rate is, definitionally,
close to `1 - sign_reliability` among significant fits): at `n_t = 2`,
Type S rates of 30-80% recur across the effect-size sweep regardless of
`n_p` -- a "significant" `zPCREFp` result at this text sample size has a
real chance of pointing the wrong direction even for its largest simulated
effect size. Both main effects' Type S rates fall toward, and mostly reach,
zero by `n_t = 10`-`35`; `zHLVA` is again the most consistently
well-behaved row.

```{r}
#| label: fig-aim1-type-m-curves
#| fig-width: 10
#| fig-height: 9.5
#| fig-cap: "Aim 1: Type M (magnitude/inflation) ratio -- median |estimate|/|true value| among significant fits -- vs. participant sample size (n_p, log scale), one row per term, faceted by text sample size (n_t). Note the log-scaled y-axis (the ratio spans nearly three orders of magnitude across this grid); the dotted line marks a ratio of 1 (no inflation)."

panel_hlva_tm <- make_aim1_curve_panel("zHLVA", b_hlva_grid, "Greens",
                                        metric_var = "type_m_ratio", y_lab = "Type M ratio (log scale)",
                                        legend_title = "zHLVA effect size", include_zero = FALSE,
                                        ref_line = 1, log_y = TRUE)
panel_pcref_tm <- make_aim1_curve_panel("zPCREFp", b_pcref_grid, "Blues",
                                         metric_var = "type_m_ratio", y_lab = "Type M ratio (log scale)",
                                         legend_title = "zPCREFp effect size", include_zero = FALSE,
                                         ref_line = 1, log_y = TRUE)
panel_interaction_tm <- make_aim1_curve_panel("zHLVA:zPCREFp", b_interaction_grid, "Purples",
                                               metric_var = "type_m_ratio", y_lab = "Type M ratio (log scale)",
                                               axis_title = "Participant sample size (n_p, log scale)",
                                               legend_title = "Interaction effect size", include_zero = FALSE,
                                               ref_line = 1, log_y = TRUE)

panel_hlva_tm / panel_pcref_tm / panel_interaction_tm
```

The log scale is necessary here, not cosmetic: across this grid, the
median Type M ratio among significant fits ranges from ~0.92 (a slight
*under*-estimate, the well-behaved end) up to ~163 at its most extreme
cell -- a span a linear axis could not show without flattening every other
line to the bottom of the plot. As with the error-surfaces figure earlier,
the most extreme ratios cluster at `n_t = 2` for the smallest simulated
non-zero effect sizes (where a small absolute bias divided by an even
smaller true value inflates dramatically) -- and the same interpretive
caveat applies: a large Type M ratio at a small true effect size reflects
that effect size's small denominator as much as it reflects the estimate's
absolute error. `zPCREFp`'s row still stands out among the two main-effect
terms, with ratios remaining elevated (typically 1.5-10x, worse at smaller
`n_p`) at `n_t = 2` even for its larger simulated effect sizes (0.2, 0.35).
`zHLVA` settles close to the dotted no-inflation line by `n_t = 35` at
every `n_p` and every non-zero effect size; the interaction does too for
its own empirically-observed effect size (0.054) and larger, though its
*smallest* simulated effect size (0.025) still shows non-trivial inflation
at `n_t = 35` when `n_p` is also small (ratio ~4-8x at `n_p` = 25-100) --
consistent with the general point above that Type M ratios are most
exaggerated, and least representative of a "typical" absolute error, when
the true effect being divided by is itself very small.

## Aim 2: discovery/replication rates

| `n_t` | zHLVA replication | zPCREFp replication | interaction replication |
|---|---|---|---|
| 2 | 0.478 | 0.473 | **0.013** |
| 10 | 0.803 | 0.157 | 0.250 |
| 35 | 0.797 | 0.163 | 0.267 |
| 85 | 0.820 | 0.237 | 0.273 |
| 136 | 0.797 | 0.340 | 0.297 |

A different, complementary picture emerges here: under the stricter
"significant in both independent halves, same sign" criterion, the
**interaction is the most fragile of the three** at `n_t = 2` (1.3%
replication vs. ~47-48% for both main effects) -- because it is so rarely
even significant in the first place with half the already-small sample. So:
Aim 1's sign-reliability diagnostic (conditional on significance) points to
`zPCREFp` as the least trustworthy; Aim 2's replication-rate diagnostic
(unconditional, halved samples) points to the interaction as the most
fragile overall. Both are real, non-contradictory findings -- they answer
different practical questions ("if I got a significant result, can I trust
its sign?" vs. "if I ran this again, would I get the same answer?") -- and
both indicate `n_t = 2` is inadequate, just via different mechanisms.

The table below defines Aim 2's three discovery/replication diagnostics and
states exactly how each is computed. All three are derived from
`combine_replication_p2()`'s per-cell pairing of two independently-fit
halves (`discovery` and `replication`, each `n_p_total`/`n_t_total` split
roughly in half by `generate_data_p2_pair()`), using the same Wald-CI
`excludes_zero` significance criterion as Aim 1, then aggregated by
`term` x `n_t` (pooled over `n_p` and the 50 replicate pairs/cell).

| Diagnostic | Standard definition | Computed here as |
|---|---|---|
| `discovery_rate` | P(significant in an initial/"discovery" sample) -- how often a first look at data of this size turns up a significant result at all, regardless of whether it is subsequently confirmed. | `mean(discovery_significant)`, where `discovery_significant = excludes_zero` evaluated on the **discovery** half alone. |
| `replication_rate` | Unconditional replication rate: P(significant **and** same-signed in *both* independent halves) -- Marek et al.'s (2022) Fig. 3e-style replication probability. | `mean(replicated)`, where `replicated = excludes_zero_discovery & excludes_zero_replication & sign(estimate_discovery) == sign(estimate_replication)`. Both halves must independently clear the significance threshold *and* agree in sign. |
| `conditional_replication` | Conditional replication rate: P(replicates \| already significant in the discovery half) -- "given I got a significant result once, would it hold up in an independent second sample?" | `mean(replicated[discovery_significant])`, i.e. `replicated` restricted to rows where the discovery half was already significant. More decision-relevant than the unconditional rate for a researcher who has already observed one significant result, since it factors out the "was it significant even once" question that `discovery_rate` alone answers. |

Note the arithmetic relationship between all three: `replication_rate` is
mechanically bounded above by `discovery_rate` (a discovery is a
prerequisite for replication), and `conditional_replication =
replication_rate / discovery_rate` for any cell where `discovery_rate > 0`
-- so a low unconditional `replication_rate` can reflect either a low
`discovery_rate` (rarely significant even once, a power problem) or a low
`conditional_replication` (often significant once but rarely again, a
genuine replication-failure problem), and distinguishing these two
mechanisms is exactly why all three are reported rather than
`replication_rate` alone (see the discussion of the interaction's `n_t = 2`
numbers below).

### Aim 2 -- fuller metrics table and visualizations

In addition to the raw replication rate above, the table and figure below
add the discovery rate and conditional replication rate defined above.

```{r}
#| label: aim2-full-metrics-table

aim2_metrics_by_nt <- aim2_all |>
  dplyr::group_by(term, n_t) |>
  dplyr::summarise(
    discovery_rate           = mean(discovery_significant, na.rm = TRUE),
    replication_rate         = mean(replicated, na.rm = TRUE),
    conditional_replication  = mean(replicated[discovery_significant], na.rm = TRUE),
    n_fits                   = dplyr::n(),
    .groups = "drop"
  )

aim2_metrics_by_nt |>
  dplyr::arrange(term, n_t) |>
  knitr::kable(
    digits = 3,
    col.names = c("Term", "n_t", "Discovery rate", "Replication rate",
                  "Conditional replication", "n fits")
  )
```

```{r}
#| label: fig-aim2-metrics-vs-nt
#| fig-width: 9
#| fig-cap: "Aim 2: discovery rate, (unconditional) replication rate, and conditional replication rate vs. text sample size, at each term's own empirically-observed effect size."

plot_metrics_vs_x(
  aim2_metrics_by_nt, x_var = "n_t",
  metrics = c("discovery_rate", "replication_rate", "conditional_replication"),
  metric_labels = aim2_metric_labels,
  x_lab = "Text sample size (n_t, log scale)"
)
```

The third panel adds a finding not visible in the unconditional
replication-rate table alone: conditional on having found a significant
interaction in the discovery half, the interaction's replication rate (~22%
at `n_t = 2`, rising to ~70-92% at larger `n_t`) is *lower* than `zHLVA`'s at
every `n_t` but is **not as catastrophically low** as its unconditional
replication rate (1.3% at `n_t = 2`) might suggest in isolation -- most of
that unconditional fragility comes from rarely being significant in the
discovery half at all (a power problem), rather than from significant
discoveries failing to hold up (a pure replication problem). Both framings
matter for different audiences: a reviewer asking "is this literature's
replication rate adequate" should look at the unconditional rate; a
researcher who has already observed one significant result and is deciding
whether to trust it should look at the conditional rate.

```{r}
#| label: fig-aim2-replication-surface
#| fig-width: 8
#| fig-height: 4
#| fig-cap: "Aim 2: unconditional replication-rate surface across the full joint (n_p, n_t) grid, at each term's own empirically-observed effect size -- the discovery/replication analog of the Aim 1 power surface above."

aim2_surface <- aim2_all |>
  dplyr::group_by(term, n_p, n_t) |>
  dplyr::summarise(replication_rate = mean(replicated, na.rm = TRUE), .groups = "drop")

plot_power_heatmap(aim2_surface, fill_var = "replication_rate", fill_label = "Replication\nrate")
```

Similar to Aim 1's power surface, but stricter, and carrying the same
`zPCREFp`-specific caveat identified there (numbers checked directly
against `aim2_surface`). `zHLVA` saturates cleanly on the participant
margin alone once `n_t >= 10` (100% replication by `n_p = 375`), but at
`n_t = 2` its replication rate lags well behind -- 54-80% across `n_p =
375`-1497, never reaching 100% anywhere on this grid -- so halving the
sample still measurably hurts even the best-behaved term when `n_t = 2`.
The **interaction** shows the pattern the paragraph above expected:
replication stays under 5% at every `n_p` when `n_t = 2`, then climbs to
48-68% once `n_t >= 10` and `n_p >= 1300`, topping out at 68% (`n_p =
1497`, `n_t = 85` or `136`) -- a real gain from more texts. **`zPCREFp`
again breaks that expectation**, just as it did for power: its two
highest replication rates on the entire grid (0.78 and 0.80) both occur
at `n_t = 2` (`n_p = 1300` and `1497` respectively) -- higher than
anything it achieves at any larger `n_t`, including the largest-both-
margins cell (0.64 at `n_p = 1497, n_t = 136`). This is the same
underpowered-paradox signature carrying over from Aim 1's power surface
into Aim 2's stricter replication criterion, not a genuine advantage of
the smallest text sample.

## Aim 1 & Aim 2: discussion

Comparing the two aims' diagnostics side by side clarifies what each is, and
isn't, telling us:

- **Aim 1** (conditional on significance) says `zPCREFp` is the *least*
  trustworthy term at `n_t = 2` -- its significant results are the most
  likely to have the wrong sign.
- **Aim 2** (unconditional, halved samples) says the **interaction** is the
  most fragile overall -- not because its significant results are
  especially untrustworthy, but because it is rarely significant at all
  once the sample is halved.
- Both agree that `n_t = 2` -- the modal text sample size in this project's
  literature review (Phase 1's overlay finding: ~95% of reviewed studies
  fall outside the trustworthy region) -- is inadequate for any of the
  three effects, via different mechanisms depending on which question is
  asked.
- `zHLVA`, the sole participant-level effect among the three, is
  comparatively well-behaved on both diagnostics and saturates on the
  participant margin alone, independent of text sample size -- consistent
  with Phase 1's finding that participant-margin effects are far less
  sensitive to this project's chronic text-sample scarcity than text-margin
  effects are. This is now visible on every diagnostic in this document
  (the fuller metrics table's explicit Type I column, the reordered power
  surface/curves, and the three-panel funnel plot), not just the original
  sign-reliability/power comparison -- the general pattern (text-margin
  terms behave qualitatively differently from the participant-margin term,
  regardless of which specific error-taxonomy diagnostic is examined) holds
  up consistently, which is itself the more general finding this
  restructured document is built to surface.

A full cross-aim synthesis, tying these findings together with Aims 3-6
(random-effects misspecification, variance-component sensitivity, slope/
covariance recovery, and design remedies), will appear once those aims have
their own reporting-ready documents.

## Flagged future work

- **The significance-threshold/multiple-comparisons sweep** (Marek et al.'s
  P<0.05-to-Bonferroni comparison) was explicitly deferred at the design
  stage, given the real core model's 26-term fixed-effects formula. Aim 1/2
  above use a single 95%-CI-based threshold throughout; whether Aim 1's
  error-rate findings hold up under a stricter multiple-comparisons
  correction remains untested.
- **Extension decisions** (recorded after discussion with the user; none
  implemented in this pass -- see `cost-estimate-expanded-grid-replicates-aims1-2-PLAN.qmd`
  for the full cost-estimation method and all six candidate scenarios; the
  ones referenced by the decisions below are summarized here so this
  document is self-contained). All scenarios assume the same 4-worker
  `furrr` setup and 3 sweeps x 5 effect values x 3 fits/replicate as the
  current run; scale factor and wall-clock are relative to/estimated from
  the achieved 67,500-fit, ~5h baseline (Scenario 0):

All seven scenarios from the source document are listed below, for
completeness. Scenarios 1-3 (greyed out) are not referenced by name in
the decisions that follow -- they sit between Scenario 0 and Scenario 4 in
both cost and scope, and were superseded by the bundled Scenario 4+5
decision below rather than chosen individually.

  | Scenario | Change | Total fits | Scale factor | Est. wall-clock (4 workers) |
  |---|---|---|---|---|
  | 0 (baseline, achieved) | Current 6x5 `n_p`/`n_t` grid, 50 replicates/cell | 67,500 | 1.0x | ~5h |
  | <span style="color:grey">1</span> | <span style="color:grey">Replicates 50->100 only</span> | <span style="color:grey">135,000</span> | <span style="color:grey">2.0x</span> | <span style="color:grey">~10h</span> |
  | <span style="color:grey">2</span> | <span style="color:grey">Densify `n_p` only: 6->10 values (add log-spaced midpoints), `n_t` unchanged</span> | <span style="color:grey">112,500</span> | <span style="color:grey">1.67x</span> | <span style="color:grey">~8.3h</span> |
  | <span style="color:grey">3</span> | <span style="color:grey">Densify `n_t` only: 5->9 values, `n_p` unchanged</span> | <span style="color:grey">121,500</span> | <span style="color:grey">1.8x</span> | <span style="color:grey">~9h</span> |
  | 4 | Densify both margins (~doubling each): `n_p` 6->11 values, `n_t` 5->9 values | 222,750 | 3.3x | ~16.5h |
  | 5 | Scenario 4 + replicates 50->100 in the same run | 445,500 | 6.6x | ~33h |
  | 6 | Scenario 4, plus extend the range upward beyond this project's real N (e.g. add `n_p` = 3000/5000, `n_t` = 200/300) | ~290,000-335,000 | ~4.3-5x | **now reliably extrapolable (benchmarked below)**: ~0.06 sec/1,000 rows, no superlinear penalty detected |

  1. **Densify both margins of the `n_p`/`n_t` grid** (Scenario 4, ~3.3x
     cost) **and increase replicates from 50 to 100 in the same run**
     (bundled with densification per the user's preference, rather than as
     a separate job) -- together, this is Scenario 5 in the cost-estimate
     document: ~445,500 fits, ~33h wall-clock on 4 workers, ~132
     core-hours. This becomes the next full-grid run superseding the
     current 30-cell/50-replicate results once launched.
  2. **Extending the sample-size range beyond this project's real N**
     (Scenario 6 above; e.g. `n_p` = 3000/5000, `n_t` = 200/300) is not
     ruled out, and the benchmark the user asked for (a handful of single
     fits at candidate large `n_p`/`n_t` values) has now been run --
     summarized below -- resolving the "not reliably extrapolable"
     uncertainty. Whether to commit to a full Scenario 6 run remains a
     separate, still-open decision; this benchmark only removes the cost
     uncertainty, not the decision itself.

::: {.callout-note}
## Scenario 6 benchmark (completed)

6 candidate cells at the proposed extremes (`n_p` up to 5000, `n_t` up to
300), 2 replicates each, `"correct"` RE spec, timed with the same
`generate_data_p2()`/`fit_model_p2()` helpers used throughout this
document (full detail and the underlying findings are in
`cost-estimate-expanded-grid-replicates-aims1-2-PLAN.qmd`'s "Scenario 6
benchmark" section):

| Cell | `n_p` | `n_t` | Rows | Mean fit time | Status |
|---|---|---|---|---|---|
| Text-margin only | 1497 | 200 | 59,880 | 2.93s | clean, both reps |
| Text-margin only | 1497 | 300 | 59,880 | 3.50s | clean, both reps |
| Participant-margin only | 3000 | 136 | 120,000 | 7.18s | clean, both reps |
| Participant-margin only | 5000 | 136 | 200,000 | 12.18s | clean, both reps |
| Both margins | 3000 | 300 | 120,000 | 7.65s | clean, both reps |
| Both margins (max) | 5000 | 300 | 200,000 | 11.68s | clean, both reps |

All 12 fits converged cleanly (no singular fits, optimizer warnings, or
timeouts). Fit time normalizes to **0.049-0.064 sec per 1,000 rows**
(mean 0.058) across every cell -- matching the existing baseline's implied
rate (0.059 sec/1,000 rows) almost exactly, with no evidence that fitting
gets disproportionately harder at very large `n_p`/`n_t`. A second,
practically useful finding: because `generate_data_p2()` caps each
participant at `k = min(10, n_texts)` assigned texts, extending `n_t`
*alone* past 10 adds no rows and costs essentially nothing extra (the
`n_t = 200`/`300` cells cost the same as the existing `n_t = 136`
baseline at matched `n_p`) -- only extending `n_p`, or adding more
texts-per-participant, increases cost.

As one concrete illustration: a dedicated sweep restricted to just these 6
cells, at the existing Aims 1-2 structure (3 sweeps x 5 effect values x 50
replicates/cell), would cost ~9.4h serial / ~2.4h on 4 workers (Aim 1
only, 4,500 fits) or ~28.2h serial / ~7.0h on 4 workers (full Aim 1 + Aim
2 structure, 13,500 fits, upper-bound estimate). A full densified-and-
extended Scenario 6 grid (the original "~130-150 cells" estimate) can now
be costed with the same confidence as Scenarios 1-4 via `rows x 0.06 /
1000` seconds per fit, once a specific grid is chosen.
:::
  3. **A dedicated `b_intercept` sweep** was judged worth scoping as its
     own small follow-up study to check whether logistic regression's
     non-collapsibility makes the text-margin instability findings above
     sensitive to this project's unusually high (~89%) baseline accuracy.
     It has since been designed and costed in full (see
     `cost-estimate-b-intercept-sweep-PLAN.qmd`), summarized here:

     ::: {.callout-note}
     ## `b_intercept` sweep design and cost (run)

     A 5-point one-at-a-time sweep over target baseline accuracies (0.50,
     0.65, 0.80, **0.89 -- observed**, 0.95, i.e. `b_intercept` = 0, 0.619,
     1.386, 2.093, 2.944), holding the three focal effect sizes at their
     empirically-observed values, at **4 representative cells** rather
     than the full 30-cell grid: the existing small (25, 2), moderate
     (375, 35), and large (1497, 136) benchmark cells, plus a new
     `n_t = 2` pathology cell at large `n_p` (1497, 2) -- benchmarked
     specifically for this plan at 0.78s/fit (mean of 2 reps, both
     `converged_singular`) -- chosen to isolate whether the text-margin
     identifiability breakdown itself is intercept-sensitive. Aim 1
     diagnostics only (no discovery/replication split), 50 replicates/
     cell: **1,000 total fits, ~22 minutes serial, ~5.6 minutes
     wall-clock on 4 workers** -- cheap enough to run interactively, no
     background job needed. A design check run as part of this planning
     pass found the simulated data's actual mean response rate runs
     1-2 percentage points below the naive `plogis(b_intercept)` target
     at every grid point (expected, from the same non-collapsibility/
     Jensen's-inequality logic this sweep investigates) -- the achieved,
     not nominal, baseline should be reported in any write-up.
     :::

     The sweep was subsequently run in full; see "`b_intercept` sweep:
     results" below (its own top-level section, placed after this list,
     since its code chunk did not nest cleanly inside this indented
     callout/list structure).
- **The shareable self-contained workflow** (a standalone `.R` process
  script + bash runner, updated generation/visualization tutorials sourcing
  from it, and a final trimmed reporting document) is planned as a
  follow-on piece of work once the extension decisions above are settled,
  so that any new grid only needs to be run once against the final
  pipeline.

Aims 3-6's own flagged-future-work items (Aim 3's `brms` spot-checks, Aim
4's flat interaction-CI-width finding, Aim 5's deferred joint
correlation x variance sweep, Aim 6's remedy-specific follow-ups, and the
`furrr` load-imbalance engineering note) are unchanged and documented in
`phase2-effect-variance-misspecification.qmd`; they will surface again in
their own future documents rather than being duplicated here.

## `b_intercept` sweep: results

The sweep designed and costed above (design summary: "`b_intercept`
sweep design and cost" callout in "Flagged future work") was run (script:
`run_b_intercept_sweep.R`) -- **sequentially rather than with `furrr`'s 4
workers**, given the real memory pressure this machine was found to be
under when this sweep was designed (swap ~87% full at the time); actual
runtime was **20.7 minutes** (vs. the ~5.6-minute parallel estimate),
still cheap enough to run interactively. Results are saved to
`results_phase2/aim1_intercept_sweep.rds` (1,000 fits x 3 terms = 3,000
rows) and loaded directly below, not recomputed.

See `cost-estimate-b-intercept-sweep-PLAN.qmd` for the fuller design,
execution, and results write-up (including the same findings below in
more detail, a projected-vs-achieved cost comparison, and the follow-on
"Scenario 5 + intercept variation" work this sweep's findings motivated).

```{r}
#| label: fig-b-intercept-sweep-results
#| fig-width: 9
#| fig-height: 6
#| fig-cap: "b_intercept sweep: power vs. target baseline accuracy, by representative cell and term. Vertical dotted line marks this project's real, observed baseline (~89%)."

intercept_sweep_results <- readRDS("results_phase2/aim1_intercept_sweep.rds")

intercept_sweep_summary <- intercept_sweep_results |>
  dplyr::group_by(label, n_p, n_t, target_accuracy, term) |>
  dplyr::summarise(
    power            = mean(excludes_zero, na.rm = TRUE),
    sign_reliability = mean(correct_sign, na.rm = TRUE),
    type_s_rate      = mean(sign_error[excludes_zero], na.rm = TRUE),
    type_m_ratio     = median(inflation_ratio[excludes_zero], na.rm = TRUE),
    coverage         = mean(covers_truth, na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::mutate(cell_label = paste0(label, " (n_p=", n_p, ", n_t=", n_t, ")"))

ggplot(intercept_sweep_summary, aes(x = target_accuracy, y = power, color = term)) +
  geom_vline(xintercept = 0.890, linetype = "dotted", color = "grey40") +
  geom_line() +
  geom_point() +
  facet_wrap(~cell_label, scales = "free_y") +
  scale_color_brewer(palette = "Dark2") +
  labs(x = "Target baseline accuracy (plogis(b_intercept))", y = "Power", color = "Term") +
  theme_minimal()
```

**Headline finding: power for all three effects generally declines as
baseline accuracy rises toward this project's real ~89% value**, most
clearly for the interaction and `zPCREFp` -- confirming the concern that
motivated this sweep. At the "moderate" cell (n_p=375, n_t=35), the
interaction's power declines close to monotonically from 84% (50%
accuracy) to 28% (95% accuracy); at the "small" cell (n_p=25, n_t=2),
`zHLVA`'s power declines from 32% to 6% over the same range. The `n_t =
2` pathology cell's Type S rate for `zPCREFp` also *worsens* toward
higher accuracy (36% -> 55% wrong-sign-given-significant), as does its
Type M ratio (4.2x -> 7.5x) -- so the text-margin identifiability
breakdown documented throughout the main part of this document is not
merely co-occurring with this project's high baseline accuracy, it is
measurably *exacerbated* by it.

**A specific mechanistic finding, investigated directly**: at the
"large" cell (n_p=1497, n_t=136), the interaction's power looks like it
drops suddenly -- a flat 100% from 50-89% accuracy, then a step down to
78% only at the most extreme point (95%) -- which could look like a
discontinuity or an artifact. Inspecting the underlying per-replicate
`estimate/std.error` ratios shows this is not a sudden change in the
data-generating mechanism: the median ratio declines smoothly and
continuously across the *entire* sweep (5.70 -> 5.58 -> 4.92 -> 3.73 ->
2.86), and so does the minimum (3.26 -> 3.25 -> 2.53 -> 2.05 -> 1.17).
What changes abruptly is only whether that smooth decline crosses the
fixed 95%-significance threshold (a ratio of 1.96): because the starting
margin at 50% accuracy is generous (median ratio ~5.7, nearly 3x the
threshold), it takes until the most extreme grid point for enough
replicates' ratios to dip below 1.96 (11 of 50 do, at 95% accuracy; zero
do at every lower level, including 89%, where the *minimum* ratio was
already down to 2.05 -- a near miss). This is a textbook case of a
smoothly-varying quantity producing an apparently sudden change in a
*thresholded* summary (power) -- compounded by a genuine, accelerating
increase in standard error as the baseline approaches the ceiling: mean
SE rises from 0.0095 to 0.0174 across the sweep, and the *rate* of
increase itself accelerates (successive ratios 1.02x, 1.14x, 1.22x,
1.30x), loosely tracking the Fisher-information-style prediction that
variance in a Bernoulli outcome scales with `1/(p(1-p))` (a naive
`1/sqrt(p(1-p))` calculation predicts a 2.29x relative SE increase from
50% to 95% accuracy; the observed increase, 1.83x, is smaller, plausibly
buffered by the participant/text random-effects variance components, but
points the same direction). This prediction -- that a denser grid near
the top of the accuracy range would reveal a smooth decline rather than
a true discontinuity at exactly 95% -- was tested directly (see
"Follow-up: ceiling-densification check" immediately below) and
confirmed.

### Follow-up: ceiling-densification check (confirmed)

Two more intercept values, 92% and 97% target accuracy, were added at
the same 4 cells and 50 replicates/cell (script:
`run_b_intercept_sweep_ceiling_extension.R`; 400 additional fits, 8.7
minutes serial), specifically to test the smooth-decline prediction
above. Results are saved separately
(`results_phase2/aim1_intercept_sweep_ceiling_extension.rds`) and
combined with the original sweep below.

```{r}
#| label: fig-b-intercept-ceiling-extension
#| fig-width: 9
#| fig-height: 6
#| fig-cap: "b_intercept sweep with ceiling densification: power vs. target baseline accuracy, by representative cell and term, now at 7 accuracy points (adding 92% and 97% to the original 5). Vertical dotted line marks this project's real, observed baseline (~89%)."

intercept_sweep_ceiling_ext <- readRDS("results_phase2/aim1_intercept_sweep_ceiling_extension.rds")

intercept_sweep_combined <- dplyr::bind_rows(intercept_sweep_results, intercept_sweep_ceiling_ext)

intercept_sweep_combined_summary <- intercept_sweep_combined |>
  dplyr::group_by(label, n_p, n_t, target_accuracy, term) |>
  dplyr::summarise(power = mean(excludes_zero, na.rm = TRUE), .groups = "drop") |>
  dplyr::mutate(cell_label = paste0(label, " (n_p=", n_p, ", n_t=", n_t, ")"))

ggplot(intercept_sweep_combined_summary, aes(x = target_accuracy, y = power, color = term)) +
  geom_vline(xintercept = 0.890, linetype = "dotted", color = "grey40") +
  geom_line() +
  geom_point() +
  facet_wrap(~cell_label, scales = "free_y") +
  scale_color_brewer(palette = "Dark2") +
  labs(x = "Target baseline accuracy (plogis(b_intercept))", y = "Power", color = "Term") +
  theme_minimal()
```

**Confirmed**: at the "large" cell, the interaction's power across all 7
points is 100%, 100%, 100%, 100%, **92%, 78%, 68%** (50%/65%/80%/89%/
**92%/95%/97%** accuracy) -- a smooth, monotonic decline once the two
new points fill in the gap, not a discontinuity at exactly 95%. The
underlying `estimate/std.error` ratios decline just as smoothly across
all 7 points (median: 5.70, 5.58, 4.92, 3.73, **3.31, 2.86, 2.32**;
minimum: 3.26, 3.25, 2.53, 2.05, **1.31, 1.17, 0.297**), and mean SE
keeps rising with the same accelerating pattern (0.0095, 0.0097, 0.0110,
0.0134, **0.0146, 0.0174, 0.0221**). The same pattern generalizes beyond
the one cell originally flagged: the "moderate" cell's interaction power
also continues declining smoothly past the original grid's endpoint,
from 48% (89%) down to 36%, 28%, and 22% at 92%/95%/97% accuracy. The
apparent "cliff" in the original 5-point grid was, as predicted, a grid-
sparsity artifact rather than a true discontinuity -- confirming that
this project's underpowered-paradox-adjacent findings do not depend on a
sudden qualitative change happening to fall near the observed baseline;
the underlying erosion is continuous and only *looks* abrupt because of
where the significance threshold happens to sit relative to a sparse
grid.

**Interpretation**: this project's own real baseline accuracy (~89%)
sits close enough to the region where these effects begin measurably
eroding that the main Aims 1-2 document's findings should be read as
somewhat conservative, not baseline-independent -- a comprehension test
with a more moderate pass rate would likely show *better* power and
*less* severe Type S/M error for the same underlying effect sizes, not
worse.

::: {.callout-note collapse="true"}
## Development history (collapsed -- process detail, not reportable findings)

This section preserves the process/decision narrative that accompanied the
original build of Aims 1-2, condensed from `phase2-effect-variance-misspecification-PLAN.qmd`
and `phase2-effect-variance-misspecification.qmd`. Nothing in this section
is required to understand the results above; it documents how they were
produced.

### Relationship to existing project simulation infrastructure

This work explicitly reused conventions already established in
`simulation_study_complete.qmd` (Simulations 1, 2, 3, 4, and 6) and
`power-simulation-vocabulary-cohesion.qmd`, rather than inventing new ones:

- **One-at-a-time sweeps, not a full factorial.** Existing simulations vary
  one dimension (effect size, a single variance component, or a single
  intercept-slope correlation) at a time while holding others fixed at a
  realistic value, to avoid combinatorial explosion. Phase 2 follows the
  same pattern (six aims, altogether) rather than crossing effect size x
  variance x covariance x misspecification x design strategy x
  two-dimensional sample size all at once.
- **Fitting**: `glmer` (fast) as the sweep workhorse, with `brms`
  spot-checks at a handful of cells using the project's precompile-once,
  `update()`-per-replicate pattern to avoid Stan recompilation cost across
  replicates (not run for Aims 1-2 specifically; see flagged future work).
- **Parallelization**: `furrr::future_pmap_dfr()` with `multisession`,
  `furrr_options(seed = TRUE)`, hierarchical seeds, 4 workers with a
  `setTimeLimit()` guard, following Phase 1's memory-safety lessons.
- **Outcome-quality classification**: converged-clean / converged-singular
  / optimizer-warning / error, reused verbatim from Phase 1's helper
  functions.

### Confirmed decisions (made before building Aims 1-2)

1. **Scope**: all six aims cover **three effects** -- `zHLVA` (main),
   `zPCREFp` (main), and `zHLVA:zPCREFp` (interaction) -- rather than the
   interaction alone.
2. **Aim 2 split axis**: left configurable (`split_margin =
   "participants_only"` vs. `"both"`) rather than fixed; both variants can
   be run and compared. The results above use `"participants_only"`.
3. **Significance-threshold sweep**: deferred (see "Flagged future work").
4. **Pacing**: build and review Aim 1 (together with Aim 2) first, as the
   more traditional power-analysis-style starting point extended with the
   replication-split question.
5. **Replicate count**: 50 resamples/cell, matching Phase 1 Step 3's
   memory-safe, coarsened configuration.

### Timing benchmark

A small number of representative cells (small/moderate/large) were timed
once each for the `"correct"` and `"misspecified"` RE specifications (the
two used by Aims 1-3), before committing to the full grid:

| Cell | n_p | n_t | `correct` | `misspecified` |
|---|---|---|---|---|
| small | 25 | 2 | 0.34s (singular) | 0.03s (clean) |
| moderate | 375 | 35 | 0.71s (clean) | 0.30s (clean) |
| large | 1497 | 136 | 3.53s (clean) | 1.43s (clean) |

Mean per-fit time across these three cells (a rough, order-of-magnitude
estimate, not a proper average over the full grid's cell-size
distribution): `correct` ~1.53s, `misspecified` ~0.59s. (The original
benchmark also timed a `"full_slopes"` arm for Aim 5's covariance sweep,
using reference slope-variance/correlation values extracted from the fitted
core model; that arm is out of scope for this Aims-1/2-only document and is
reported instead in the Aim 5 document.)

Extrapolated cost: Aim 1 (22,500 fits, `correct` spec) ~9.6h serial / ~2.4h
on 4 workers; Aim 2 (~22,500 fits, `correct` spec) ~9.6h serial / ~2.4h
parallel. Combined, each `(sweep, effect_value, n_p, n_t, rep)` combination
triggers **three** `glmer` fits (one Aim 1 full-sample fit, plus Aim 2's
discovery and replication half-sample fits): 3 sweeps x 5 effect values x 30
cells x 50 reps = 22,500 combinations x 3 fits = **67,500 total fits**,
estimated at ~20h serial / ~5h on 4 parallel workers.

### Validation: small-scale smoke test

Before committing to the full grid, the pipeline was validated on a 24-cell
subset (2 effect values x 2 participant Ns x 2 text Ns x 3 reps, interaction
sweep only, at small sample sizes). Kept live here (cheap, ~6 seconds) as a
lightweight regression check that `generate_data_p2()`/`fit_model_p2()`
above still behave as expected.

```{r}
#| label: aim1-aim2-smoke-test
#| cache: true

smoke_grid <- tidyr::crossing(
  sweep = "interaction",
  effect_value = c(0, 0.054),
  n_p = c(25, 100),
  n_t = c(2, 10),
  rep = 1:3
)

smoke_results <- purrr::pmap(smoke_grid, run_aim1_aim2_cell)
aim1_smoke <- purrr::map_dfr(smoke_results, "aim1")
aim2_smoke <- purrr::map_dfr(smoke_results, "aim2")
```

```{r}
#| label: aim1-aim2-smoke-test-check

aim1_smoke |> dplyr::filter(term == "zHLVA:zPCREFp") |>
  dplyr::select(n_p, n_t, effect_value, rep, estimate, ci_lower, ci_upper,
                excludes_zero, correct_sign, status)

aim2_smoke |> dplyr::filter(term == "zHLVA:zPCREFp") |>
  dplyr::select(n_p, n_t, effect_value, rep, discovery_significant, replicated,
                status_discovery, status_replication)
```

The pipeline ran the 24-cell subset in ~5.6 seconds and produced sensible,
correctly-structured output for both Aim 1 (per-term bias/CI/sign
diagnostics) and Aim 2 (discovery/replication agreement) -- singular fits
were common at these smallest cells (`n_p = 25`, `n_t = 2`), as expected.

### Cost note and launch record

The full grid was not run interactively. It was launched as a detached
background process (`run_aim1_aim2_full.R`), one sweep at a time (`hlva`,
`pcref`, `interaction`), using `furrr::plan(multisession, workers = 4)`. All
three sweeps finished in ~99-101 minutes each (~5h total on 4 parallel
workers, matching the benchmark-derived estimate closely). Results were
saved to `results_phase2/aim{1,2}_{sweep}.rds` and are what this document
loads above.
:::

## Follow-up note

The DGP and fitting code above (`generate_data_p2()`, `fit_model_p2()`,
`generate_data_p2_pair()`, `run_aim1_aim2_cell()`) is kept live in this
document, unmodified from the version used to produce the saved results, so
it can be inspected and -- in a future, separate turn -- modified if
specific changes turn out to be informative. No new simulation runs have
been made in this pass; any modification to this code that would justify a
new full-grid run (a multi-hour, background-job undertaking) should be
discussed and scoped before being launched.
