---
title: "Effect size variation for the HLVA main effect, the PCREFp main effect, and their interaction"
subtitle: "Aims 1-2: power, error rates, and replication rates (accuracy outcome)"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: cosmo
    embed-resources: true
    code-fold: true
execute:
  warning: false
  message: false
---

::: {.callout-note}
## Scope of this document

This document covers **Aim 1** (effect size x sample size, with a fuller
statistical-error taxonomy) and **Aim 2** (discovery/replication split) of
the Phase 2 parametric simulation program, for the `response` (comprehension
accuracy) outcome only. It is the first of what will be several smaller
documents replacing a single originally-planned six-aim report; Aims 3-6 and
the normative-F1-ANOVA-malpractice extension will get their own documents
later, built the same way.

All tables and figures below are computed from pre-existing, already-run
results saved in `results_phase2/aim1_*.rds` and `results_phase2/aim2_*.rds`
(no full-grid recomputation happens when this document is rendered). Full
design rationale, the complete six-aim plan, and every aim's development
history live untouched in `phase2-effect-variance-misspecification-PLAN.qmd`
and `phase2-effect-variance-misspecification.qmd` -- this document condenses
and refines the Aim 1/2 material from those sources rather than re-deriving
it. The DGP/fitting code below is kept live (not just for reference) because
it may be revisited for informative modifications in a follow-up pass; see
the closing note.
:::

```{r}
#| label: setup

library(tidyverse)
library(lme4)
library(patchwork)
library(RColorBrewer)
library(ggdist)
```

## Relationship to Phase 1 and motivating framework

Phase 1 (`sampling-variability-corridor-of-stability.qmd`) was an
**empirical resampling** study of this project's one real dataset,
answering "how unstable is our *actual* estimate at smaller *N*?" Phase 2 is
a **parametric simulation** study, answering the complementary question:
"how unstable would an estimate be across a *range* of plausible true
effects, variances, and model specifications -- are Phase 1's findings a
generic property of this kind of crossed design, or specific to this
project's particular dataset?"

Aims 1-2 port Marek et al.'s (2022) brain-wide-association-study
reproducibility framework onto this setting. Marek et al. documented a
statistical-error taxonomy -- false negatives, false positives,
**magnitude/inflation error (Type M)**, and **sign error (Type S)** -- each
as a function of sample size, and showed that smaller samples' significant
effects are disproportionately the most inflated and sign-flipped ones (the
"underpowered paradox": stricter significance thresholds select for, rather
than screen out, the most inflated estimates). They also used an explicit
**discovery/replication split** design to quantify out-of-sample replication
probability directly. Aims 1-2 test whether this project's own documented
sign-instability (the "reverse cohesion effect" finding -- see project
memory) is a domain-specific instance of the same generic phenomenon, now
covering **three effects** rather than the interaction alone: the `zHLVA`
main effect, the `zPCREFp` main effect, and their `zHLVA:zPCREFp`
interaction.

| Input | Value(s) | Source |
|---|---|---|
| Sample-size grid (participants) | 25, 100, 375, 725, 1300, 1497 | Phase 1 Step 3's joint grid |
| Sample-size grid (texts) | 2, 10, 35, 85, 136 | Phase 1 Step 3's joint grid |
| Effect size, `zHLVA` | 0.363 | Phase 1 Step 0 full-data fit |
| Effect size, `zPCREFp` | 0.094-0.098 | Phase 1 Step 0 / Step 5 |
| Effect size, `zHLVA:zPCREFp` | 0.054 | Phase 1 Step 0 / Step 5 (glmer and brms agree) |
| Variance component, `sd(ResponseId)` | ~0.61-0.62 | Full-data reduced model |
| Variance component, `sd(text.id)` | ~0.70-0.72 | Full-data reduced model |
| RE-misspecification contrast | "correct" `(1\|ResponseId)+(1\|text.id)` vs. "misspecified" `(1\|ResponseId)` only | Phase 1, all steps |

Reusing the same sample-size grid means these results can be laid
side-by-side against Phase 1's real-data panels directly.

::: {.callout-important}
## Note: the real core model's random-effects structure is richer than this DGP

The real fitted core model's random-effects terms let participants vary in
their *slopes* for all five text-level predictors (including `zPCREFp`), and
let texts vary in their *slopes* for the participant-level predictors
(including `zHLVA`), each correlated with its own grouping factor's
intercept. Aims 1-2 (like Aim 3-4 and 6) use a simpler, random-**intercept**-
only "correct"/"misspecified" data-generating process, following Phase 1's
reduced model -- they answer a coarser question (omitting a grouping
factor's clustering altogether) and should not be read as testing the real
model's richer slope/covariance structure. That richer structure is the
subject of Aim 5, covered in a separate future document.
:::

## Design: Aim 1 (effect size x sample size) and Aim 2 (discovery/replication split)

Both aims share one data-generating run: RE specification fixed at
"correct" (random intercepts only, for `ResponseId`/`participant_id` and
`text.id`/`text_id`); variance components fixed at their empirical values;
the full joint sample-size grid (30 `(n_p, n_t)` cells); 50 replicates/cell.
A grid cell's `(n_p, n_t)` is the **total** sample -- Aim 2 splits each cell
into two roughly-equal-sized discovery/replication halves.

**Aim 1** runs three **one-at-a-time** effect-size sweeps -- each holding
the other two effects fixed at their empirical values while varying the
swept effect:

| Sweep | Effect | Grid (centered on empirical estimate) |
|---|---|---|
| (i) | `zHLVA` (main effect, participant-level) | 0, 0.18, 0.363 (observed), 0.7, 1.0 |
| (ii) | `zPCREFp` (main effect, text-level) | 0, 0.05, ~0.095 (observed), 0.2, 0.35 |
| (iii) | `zHLVA:zPCREFp` (interaction) | 0, 0.025, 0.054 (observed), 0.10, 0.20 |

Diagnostics extend the project's existing four (bias, sign reliability,
coverage, RMSE) with a Marek-style decomposition, directly computable here
because the true value is known exactly: **Type S (sign) error rate** (among
CI-excludes-0 draws, proportion with the wrong sign), **Type M
(magnitude/inflation) ratio** (`|estimate|/|true value|` among
CI-excludes-0 draws), and **power** on its own scale.

**Question Aim 1 answers:** is the sign-instability pattern found in Phase 1
(near-coin-flip sign consistency for the interaction at *n* = 2 texts) (a) a
property of the *specific* small effect size observed, (b) specific to the
interaction term, or (c) a general property of being estimated from the
*text* margin (in which case `zPCREFp`, also text-level, should show a
similar pattern)? Comparing all three effects on the same diagnostics
distinguishes these explanations directly.

**Aim 2** ports Marek et al.'s replication design (their Fig. 3e): for each
grid cell (all three effects), split the simulated sample into two
independent, equal-sized discovery/replication halves, fit both, and compute
the proportion of replicates in which a discovery-significant effect also
replicates (same sign, both significant) out-of-sample. The split axis is
configurable (`split_margin`): `"participants_only"` (same text pool in both
halves, new participants only) vs. `"both"` (independent participants *and*
texts). Both variants can be compared once built; the run below uses
`"participants_only"`.

The significance-threshold/multiple-comparisons sweep (Marek's
P<0.05-to-Bonferroni comparison) is explicitly **deferred** here -- see
"Flagged future work" below.

## Shared data-generating function

Generalizes `simulate_interaction_data()` (from
`power-simulation-vocabulary-cohesion.qmd`) and `generate_data_s3()` (from
`simulation_study_complete.qmd`): crossed participant x text design with
`k`-texts-per-participant subsetting (clamped to `n_texts`), independently
settable effect sizes for all three effects, independently settable
intercept variances for both margins, and optional by-margin slope
variance/correlation (zero by default here, so Aims 1-2 reduce to the
intercept-only DGP; non-zero values are used only by Aim 5, covered
elsewhere).

```{r}
#| label: generate-data-p2

generate_data_p2 <- function(
    n_participants = 200,
    n_texts        = 20,
    k              = min(10, n_texts),
    questions_per_text = 4,
    b_intercept    = 2.093,
    b_hlva         = 0.363,
    b_pcref        = 0.095,
    b_interaction  = 0.054,
    sd_participant = 0.69,
    sd_text        = 0.41,
    sd_pcref_slope = 0,
    sd_hlva_slope  = 0,
    cor_part       = 0,
    cor_text       = 0,
    cohesion_design = c("natural", "widened"),
    reliability_hlva = 1,
    reliability_pcref = 1,
    seed           = NULL
) {
  cohesion_design <- match.arg(cohesion_design)
  if (!is.null(seed)) set.seed(seed)
  k <- min(k, n_texts)

  # Participant-level random effects: intercept + zPCREFp slope (correlated) -- Aim 5
  Sigma_part <- matrix(
    c(sd_participant^2, cor_part * sd_participant * sd_pcref_slope,
      cor_part * sd_participant * sd_pcref_slope, sd_pcref_slope^2),
    nrow = 2
  )
  re_part <- if (sd_pcref_slope > 0) {
    MASS::mvrnorm(n_participants, mu = c(0, 0), Sigma = Sigma_part)
  } else {
    cbind(rnorm(n_participants, 0, sd_participant), 0)
  }

  # zHLVA measurement reliability -- Aim 6 strategy E; reliability_hlva = 1
  # (default) reproduces the exact prior behavior used in Aims 1-5.
  zHLVA_true <- rnorm(n_participants, 0, 1)
  zHLVA_obs  <- sqrt(reliability_hlva) * zHLVA_true +
    sqrt(1 - reliability_hlva) * rnorm(n_participants, 0, 1)

  participants <- tibble(
    participant_id = 1:n_participants,
    zHLVA          = zHLVA_true,
    zHLVA_obs      = zHLVA_obs,
    u_participant  = re_part[, 1],
    s_pcref        = re_part[, 2]
  )

  # Text-level random effects: intercept + zHLVA slope (correlated) -- Aim 5
  Sigma_text <- matrix(
    c(sd_text^2, cor_text * sd_text * sd_hlva_slope,
      cor_text * sd_text * sd_hlva_slope, sd_hlva_slope^2),
    nrow = 2
  )
  re_text <- if (sd_hlva_slope > 0) {
    MASS::mvrnorm(n_texts, mu = c(0, 0), Sigma = Sigma_text)
  } else {
    cbind(rnorm(n_texts, 0, sd_text), 0)
  }

  # zPCREFp sampling design -- Aim 6 strategy D (widened cohesion range) --
  # and measurement reliability -- Aim 6 strategy E.
  zPCREFp_true <- if (cohesion_design == "widened") {
    rep(c(-1.5, 1.5), length.out = n_texts) + rnorm(n_texts, 0, 0.3)
  } else {
    rnorm(n_texts, 0, 1)
  }
  zPCREFp_obs <- sqrt(reliability_pcref) * zPCREFp_true +
    sqrt(1 - reliability_pcref) * rnorm(n_texts, 0, 1)

  texts <- tibble(
    text_id  = 1:n_texts,
    zPCREFp  = zPCREFp_true,
    zPCREFp_obs = zPCREFp_obs,
    u_text   = re_text[, 1],
    s_hlva   = re_text[, 2]
  )

  assignments <- tibble(participant_id = 1:n_participants) |>
    dplyr::mutate(assigned = purrr::map(participant_id,
                          ~ sample(1:n_texts, size = k, replace = FALSE))) |>
    tidyr::unnest(assigned) |>
    dplyr::rename(text_id = assigned)

  assignments |>
    dplyr::left_join(participants, by = "participant_id") |>
    dplyr::left_join(texts,        by = "text_id") |>
    dplyr::slice(rep(1:dplyr::n(), each = questions_per_text)) |>
    dplyr::mutate(
      question_id = rep(1:questions_per_text, times = dplyr::n() / questions_per_text),
      eta = b_intercept +
        (b_hlva  + s_hlva)  * zHLVA +
        (b_pcref + s_pcref) * zPCREFp +
        b_interaction * zHLVA * zPCREFp +
        u_participant + u_text,
      prob     = plogis(eta),
      response = rbinom(dplyr::n(), 1, prob)
    ) |>
    dplyr::mutate(zHLVA_fit = zHLVA_obs, zPCREFp_fit = zPCREFp_obs) |>
    dplyr::select(participant_id, text_id, question_id,
                  zHLVA = zHLVA_fit, zPCREFp = zPCREFp_fit, response)
}
```

Note: `library(MASS)` is deliberately **not** attached (only `MASS::mvrnorm`
is called via `::`), because `MASS::select()` masks `dplyr::select()` and
breaks the pipeline above if attached. The fitted model always sees
`zHLVA`/`zPCREFp` as **observed** (possibly measurement-error-attenuated)
values, while the response is generated from the **true** latent values --
with the default `reliability_hlva = reliability_pcref = 1` used throughout
Aims 1-2, observed and true are identical.

## Shared fitting and outcome-classification helpers

Reused from `sampling-variability-corridor-of-stability.qmd`'s
`fit_reduced_model()` / `extract_terms_of_interest()` /
`fit_reduced_model_guarded()` pattern, extended with a third RE
specification (`"full_slopes"`, used only by Aim 5) alongside the
`"correct"`/`"misspecified"` pair used here and by Aim 3.

```{r}
#| label: fit-model-p2

extract_terms_of_interest_p2 <- function(fit) {
  terms <- c("zHLVA", "zPCREFp", "zHLVA:zPCREFp")
  co <- tryCatch(summary(fit)$coefficients, error = function(e) NULL)
  if (is.null(co)) return(NULL)
  avail <- intersect(terms, rownames(co))
  tibble(term = avail, estimate = co[avail, "Estimate"], std.error = co[avail, "Std. Error"])
}

fit_model_p2 <- function(data, re_spec = c("correct", "misspecified", "full_slopes")) {
  re_spec <- match.arg(re_spec)
  form <- switch(re_spec,
    correct      = response ~ zHLVA * zPCREFp + (1 | participant_id) + (1 | text_id),
    misspecified = response ~ zHLVA * zPCREFp + (1 | participant_id),
    full_slopes  = response ~ zHLVA * zPCREFp + (1 + zPCREFp | participant_id) + (1 + zHLVA | text_id)
  )
  warn_msgs <- character(0)
  fit <- tryCatch(
    withCallingHandlers(
      glmer(form, data = data, family = binomial, control = glmerControl(optimizer = "bobyqa")),
      warning = function(w) { warn_msgs <<- c(warn_msgs, conditionMessage(w)); invokeRestart("muffleWarning") }
    ),
    error = function(e) e
  )
  if (inherits(fit, "error")) return(list(status = "error", coefs = NULL))
  status <- if (length(warn_msgs) > 0 &&
                any(grepl("failed to converge|Model failed|large eigenvalue|unable to evaluate",
                          warn_msgs, ignore.case = TRUE))) {
    "optimizer_warning"
  } else if (isSingular(fit)) {
    "converged_singular"
  } else {
    "converged_clean"
  }
  list(status = status, coefs = extract_terms_of_interest_p2(fit))
}

fit_model_p2_guarded <- function(data, re_spec, timeout_secs = 15) {
  setTimeLimit(elapsed = timeout_secs, transient = TRUE)
  res <- tryCatch(
    fit_model_p2(data, re_spec),
    error = function(e) list(status = "error", coefs = NULL)
  )
  setTimeLimit(elapsed = Inf, transient = FALSE)
  res
}
```

## Sample-size grid and Aim 1 sweep grids

```{r}
#| label: sample-size-grid

n_p_grid <- c(25, 100, 375, 725, 1300, 1497)
n_t_grid <- c(2, 10, 35, 85, 136)
joint_grid <- tidyr::expand_grid(n_p = n_p_grid, n_t = n_t_grid)
nrow(joint_grid)
```

```{r}
#| label: aim1-sweep-grids

b_hlva_grid        <- c(0, 0.18, 0.363, 0.7, 1.0)
b_pcref_grid       <- c(0, 0.05, 0.095, 0.2, 0.35)
b_interaction_grid <- c(0, 0.025, 0.054, 0.10, 0.20)
```

## Discovery/replication data-generating pair (Aim 2)

For `split_margin = "participants_only"`, both halves share the same
`n_t_total` texts (avoiding degenerate 1-text halves at the smallest grid
point, `n_t = 2`); only the participant pool is split. For `split_margin =
"both"`, both margins split independently.

```{r}
#| label: generate-data-p2-pair

generate_data_p2_pair <- function(
    n_p_total, n_t_total, k = min(10, n_t_total), questions_per_text = 4,
    b_intercept = 2.093, b_hlva = 0.363, b_pcref = 0.095, b_interaction = 0.054,
    sd_participant = 0.69, sd_text = 0.41,
    sd_pcref_slope = 0, sd_hlva_slope = 0, cor_part = 0, cor_text = 0,
    split_margin = c("participants_only", "both"),
    seed
) {
  split_margin <- match.arg(split_margin)
  set.seed(seed)

  n_p_half <- max(1, floor(n_p_total / 2))
  n_t_half <- if (split_margin == "both") max(1, floor(n_t_total / 2)) else n_t_total

  make_texts <- function(n_t) {
    Sigma_text <- matrix(c(sd_text^2, cor_text * sd_text * sd_hlva_slope,
                            cor_text * sd_text * sd_hlva_slope, sd_hlva_slope^2), nrow = 2)
    re_text <- if (sd_hlva_slope > 0) MASS::mvrnorm(n_t, c(0, 0), Sigma_text) else cbind(rnorm(n_t, 0, sd_text), 0)
    tibble(text_id = 1:n_t, zPCREFp = rnorm(n_t, 0, 1), u_text = re_text[, 1], s_hlva = re_text[, 2])
  }
  make_participants <- function(n_p) {
    Sigma_part <- matrix(c(sd_participant^2, cor_part * sd_participant * sd_pcref_slope,
                            cor_part * sd_participant * sd_pcref_slope, sd_pcref_slope^2), nrow = 2)
    re_part <- if (sd_pcref_slope > 0) MASS::mvrnorm(n_p, c(0, 0), Sigma_part) else cbind(rnorm(n_p, 0, sd_participant), 0)
    tibble(participant_id = 1:n_p, zHLVA = rnorm(n_p, 0, 1), u_participant = re_part[, 1], s_pcref = re_part[, 2])
  }
  build_dataset <- function(participants, texts) {
    kk <- min(k, nrow(texts))
    assignments <- tibble(participant_id = participants$participant_id) |>
      dplyr::mutate(assigned = purrr::map(participant_id, ~ sample(texts$text_id, size = kk, replace = FALSE))) |>
      tidyr::unnest(assigned) |> dplyr::rename(text_id = assigned)
    assignments |>
      dplyr::left_join(participants, by = "participant_id") |>
      dplyr::left_join(texts, by = "text_id") |>
      dplyr::slice(rep(1:dplyr::n(), each = questions_per_text)) |>
      dplyr::mutate(
        question_id = rep(1:questions_per_text, times = dplyr::n() / questions_per_text),
        eta = b_intercept + (b_hlva + s_hlva) * zHLVA + (b_pcref + s_pcref) * zPCREFp +
          b_interaction * zHLVA * zPCREFp + u_participant + u_text,
        prob = plogis(eta), response = rbinom(dplyr::n(), 1, prob)
      ) |>
      dplyr::select(participant_id, text_id, question_id, zHLVA, zPCREFp, response)
  }

  texts_shared      <- make_texts(n_t_half)
  participants_disc <- make_participants(n_p_half)
  participants_repl <- make_participants(n_p_half)
  texts_repl        <- if (split_margin == "both") make_texts(n_t_half) else texts_shared

  list(
    discovery   = build_dataset(participants_disc, texts_shared),
    replication = build_dataset(participants_repl, texts_repl)
  )
}
```

## Diagnostics and combined per-cell runner

Coverage/CI-width/Type S/Type M diagnostics use **Wald confidence intervals**
(`estimate +/- 1.96 * std.error`) rather than profile-likelihood CIs, for
speed -- consistent with the project's `glmer`-as-fast-workhorse convention.

```{r}
#| label: aim1-aim2-diagnostics-and-runner

diagnose_coefs_p2 <- function(coefs, true_vals) {
  coefs |>
    dplyr::mutate(
      true_value      = true_vals[term],
      ci_lower        = estimate - 1.96 * std.error,
      ci_upper        = estimate + 1.96 * std.error,
      ci_width        = ci_upper - ci_lower,
      excludes_zero   = ci_lower > 0 | ci_upper < 0,
      correct_sign    = sign(estimate) == sign(true_value),
      sign_error      = excludes_zero & !correct_sign,
      inflation_ratio = ifelse(true_value != 0, abs(estimate) / abs(true_value), NA_real_),
      bias            = estimate - true_value,
      covers_truth    = ci_lower <= true_value & true_value <= ci_upper
    )
}

combine_replication_p2 <- function(res_disc, res_repl, true_vals) {
  if (is.null(res_disc$coefs) || is.null(res_repl$coefs)) {
    return(tibble(term = names(true_vals), replicated = NA, discovery_significant = NA))
  }
  disc <- diagnose_coefs_p2(res_disc$coefs, true_vals) |>
    dplyr::select(term, estimate, excludes_zero)
  repl <- diagnose_coefs_p2(res_repl$coefs, true_vals) |>
    dplyr::select(term, estimate_repl = estimate, excludes_zero_repl = excludes_zero)
  dplyr::full_join(disc, repl, by = "term") |>
    dplyr::mutate(
      discovery_significant = excludes_zero,
      replicated = excludes_zero & excludes_zero_repl & sign(estimate) == sign(estimate_repl)
    )
}

run_aim1_aim2_cell <- function(sweep, effect_value, n_p, n_t, rep, split_margin = "participants_only") {
  b_hlva        <- if (sweep == "hlva")        effect_value else 0.363
  b_pcref       <- if (sweep == "pcref")       effect_value else 0.095
  b_interaction <- if (sweep == "interaction") effect_value else 0.054
  true_vals <- c(zHLVA = b_hlva, zPCREFp = b_pcref, `zHLVA:zPCREFp` = b_interaction)

  sweep_idx <- match(sweep, c("hlva", "pcref", "interaction"))
  seed_base <- rep * 1e7 + n_p * 1e3 + n_t + sweep_idx * 1e5 + round(effect_value * 1000)

  ## Aim 1: single full-sample fit
  df_full  <- generate_data_p2(n_participants = n_p, n_texts = n_t,
                                b_hlva = b_hlva, b_pcref = b_pcref, b_interaction = b_interaction,
                                seed = seed_base)
  res_full <- fit_model_p2_guarded(df_full, "correct")

  aim1_rows <- if (!is.null(res_full$coefs)) {
    diagnose_coefs_p2(res_full$coefs, true_vals) |> dplyr::mutate(status = res_full$status)
  } else {
    tibble(term = names(true_vals), status = res_full$status)
  }
  aim1_rows <- aim1_rows |>
    dplyr::mutate(sweep = sweep, effect_value = effect_value, n_p = n_p, n_t = n_t, rep = rep)

  ## Aim 2: discovery/replication split (same grid cell = total N)
  pair <- generate_data_p2_pair(n_p_total = n_p, n_t_total = n_t,
                                 b_hlva = b_hlva, b_pcref = b_pcref, b_interaction = b_interaction,
                                 split_margin = split_margin, seed = seed_base + 1)
  res_disc <- fit_model_p2_guarded(pair$discovery, "correct")
  res_repl <- fit_model_p2_guarded(pair$replication, "correct")

  aim2_rows <- combine_replication_p2(res_disc, res_repl, true_vals) |>
    dplyr::mutate(
      sweep = sweep, effect_value = effect_value, n_p = n_p, n_t = n_t, rep = rep,
      split_margin = split_margin,
      status_discovery = res_disc$status, status_replication = res_repl$status
    )

  list(aim1 = aim1_rows, aim2 = aim2_rows)
}
```

The full grid (3 sweeps x 5 effect values x 30 cells x 50 replicates x 3
fits/combination = 67,500 total `glmer` fits) was run once, as a detached
background process, well before this document was written -- see the
"Development history" appendix for the validation/timing/cost detail. It is
**not** re-run here; the sections below load its saved output directly.

## Loading Aim 1/2 results and shared helpers

```{r}
#| label: load-aim1-aim2-results

aim1_files <- c("results_phase2/aim1_hlva.rds", "results_phase2/aim1_pcref.rds",
                 "results_phase2/aim1_interaction.rds")
aim2_files <- c("results_phase2/aim2_hlva.rds", "results_phase2/aim2_pcref.rds",
                 "results_phase2/aim2_interaction.rds")
missing_files <- c(aim1_files, aim2_files)[!file.exists(c(aim1_files, aim2_files))]
if (length(missing_files) > 0) {
  stop(
    "Missing Aim 1/2 result file(s): ", paste(missing_files, collapse = ", "),
    ". These are produced by the detached background job run_aim1_aim2_full.R ",
    "(~5h on 4 workers) -- this document does not regenerate them inline. ",
    "Run that script first if the files are genuinely absent."
  )
}

aim1_hlva        <- readRDS("results_phase2/aim1_hlva.rds")
aim1_pcref       <- readRDS("results_phase2/aim1_pcref.rds")
aim1_interaction <- readRDS("results_phase2/aim1_interaction.rds")
aim2_hlva        <- readRDS("results_phase2/aim2_hlva.rds")
aim2_pcref       <- readRDS("results_phase2/aim2_pcref.rds")
aim2_interaction <- readRDS("results_phase2/aim2_interaction.rds")

own_effect <- c(zHLVA = 0.363, zPCREFp = 0.095, `zHLVA:zPCREFp` = 0.054)
term_levels <- c("zHLVA", "zPCREFp", "zHLVA:zPCREFp")

# Each sweep's own term, evaluated only at its empirically-observed effect
# size -- the slice used by most tables/figures below.
aim1_all <- dplyr::bind_rows(
  aim1_hlva        |> dplyr::filter(term == "zHLVA",        effect_value == own_effect["zHLVA"]),
  aim1_pcref       |> dplyr::filter(term == "zPCREFp",       effect_value == own_effect["zPCREFp"]),
  aim1_interaction |> dplyr::filter(term == "zHLVA:zPCREFp", effect_value == own_effect["zHLVA:zPCREFp"])
) |> dplyr::mutate(term = factor(term, levels = term_levels))

aim2_all <- dplyr::bind_rows(
  aim2_hlva        |> dplyr::filter(term == "zHLVA",        effect_value == own_effect["zHLVA"]),
  aim2_pcref       |> dplyr::filter(term == "zPCREFp",       effect_value == own_effect["zPCREFp"]),
  aim2_interaction |> dplyr::filter(term == "zHLVA:zPCREFp", effect_value == own_effect["zHLVA:zPCREFp"])
) |> dplyr::mutate(term = factor(term, levels = term_levels))

# Full effect-size sweep, all points (not just the own-effect-size slice) --
# used for the Aim 1 power-curves figure below.
aim1_sweep_combined <- dplyr::bind_rows(
  aim1_hlva        |> dplyr::filter(term == "zHLVA"),
  aim1_pcref       |> dplyr::filter(term == "zPCREFp"),
  aim1_interaction |> dplyr::filter(term == "zHLVA:zPCREFp")
) |> dplyr::mutate(term = factor(term, levels = term_levels))

# Shared summary helper: power, sign reliability, Type S error rate,
# Type M ratio, median bias, and coverage, grouped by whatever columns are
# passed in `group_vars`.
summarize_p2_metrics <- function(df, group_vars) {
  df |>
    dplyr::group_by(dplyr::across(dplyr::all_of(group_vars))) |>
    dplyr::summarise(
      power            = mean(excludes_zero, na.rm = TRUE),
      sign_reliability = mean(correct_sign, na.rm = TRUE),
      type_s_rate      = mean(sign_error[excludes_zero], na.rm = TRUE),
      type_m_ratio     = median(inflation_ratio[excludes_zero], na.rm = TRUE),
      median_bias      = median(bias, na.rm = TRUE),
      coverage         = mean(covers_truth, na.rm = TRUE),
      n_fits           = dplyr::n(),
      .groups = "drop"
    )
}

aim1_metric_labels <- c(power = "Power\nP(significant)",
                         sign_reliability = "Sign reliability\nP(correct sign)",
                         type_s_rate = "Type S rate\nP(wrong sign | significant)",
                         type_m_ratio = "Type M ratio\nmedian |est/true| | significant")

aim2_metric_labels <- c(discovery_rate = "Discovery rate\nP(significant, half 1)",
                         replication_rate = "Replication rate\nP(sig. + same sign, both halves)",
                         conditional_replication = "Conditional replication\nP(replicates | discovery-sig.)")
```

## Shared plotting helpers

Aim 1 and Aim 2 each need a metrics-vs-`n_t` line/facet plot and an
`n_p`x`n_t` heatmap that differ only in which columns/labels are used. Both
are written once here and reused for each aim below (and are intended to be
reused again for the Aims 3-6 documents still to come).

```{r}
#| label: plot-helpers

plot_metrics_vs_x <- function(data, x_var, metrics, metric_labels,
                               color_var = "term", x_lab = NULL, x_breaks = n_t_grid) {
  x_lab <- x_lab %||% x_var
  df_long <- data |>
    tidyr::pivot_longer(dplyr::all_of(metrics), names_to = "metric", values_to = "value") |>
    dplyr::mutate(metric = factor(metric_labels[metric], levels = unname(metric_labels)))

  ggplot(df_long, aes(x = .data[[x_var]], y = value, color = .data[[color_var]])) +
    geom_line() +
    geom_point() +
    facet_wrap(~metric, scales = "free_y") +
    scale_x_log10(breaks = x_breaks) +
    scale_color_brewer(palette = "Dark2") +
    labs(x = x_lab, y = NULL, color = "Term") +
    theme_minimal()
}

plot_power_heatmap <- function(data, fill_var, facet_var = "term", fill_label = "Value",
                                grid_n_p = n_p_grid, grid_n_t = n_t_grid) {
  ggplot(data, aes(x = factor(n_p, levels = grid_n_p), y = factor(n_t, levels = grid_n_t),
                    fill = .data[[fill_var]])) +
    geom_tile() +
    geom_text(aes(label = scales::percent(.data[[fill_var]], accuracy = 1)), size = 2.8, color = "white") +
    facet_wrap(stats::as.formula(paste0("~", facet_var))) +
    scale_fill_viridis_c(limits = c(0, 1), labels = scales::percent) +
    labs(x = "Participant sample size (n_p)", y = "Text sample size (n_t)", fill = fill_label) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}
```

## Aim 1: is small-text-*N* instability a text-margin property or an interaction-specific property?

At each effect's own empirically-observed value, sign reliability and power
by `n_t`:

| `n_t` | zHLVA sign reliability | zPCREFp sign reliability | interaction sign reliability | zHLVA power | zPCREFp power | interaction power |
|---|---|---|---|---|---|---|
| 2 | 0.940 | **0.553** | 0.717 | 0.650 | 0.627 | 0.127 |
| 10 | 0.997 | 0.730 | 0.920 | 0.897 | 0.177 | 0.603 |
| 35 | 0.997 | 0.860 | 0.930 | 0.917 | 0.300 | 0.577 |
| 85 | 0.993 | 0.930 | 0.897 | 0.920 | 0.420 | 0.570 |
| 136 | 0.993 | 0.937 | 0.913 | 0.910 | 0.510 | 0.620 |

**This resolves the question the design set out to answer, and the answer is
not what the "reverse cohesion effect" framing might suggest.** At `n_t =
2`, `zPCREFp`'s sign reliability (0.553) is *worse* than the interaction's
(0.717) -- barely better than a coin flip -- while `zHLVA` (the
participant-level effect) is comparatively well-behaved (0.940). This is a
**text-margin property, not an interaction-specific property**: the
text-level main effect is at least as unstable as the interaction at the
smallest realistic text sample size, and by this particular diagnostic, more
so.

`zPCREFp`'s power also shows the "underpowered paradox" signature directly:
power is *higher* at `n_t = 2` (0.627) than at `n_t = 10` (0.177) -- not
because small samples are better powered, but because the noisiest, most
degenerate cell produces wildly inflated estimates that cross the
significance threshold more often *while being wrong about the sign* half
the time. A researcher who only sees "significant, p < .05" from a 2-text
study cannot distinguish this from a well-powered result.

### Aim 1 -- fuller metrics table

The table above reports sign reliability and power only. For completeness
against Marek et al.'s (2022) full diagnostic set, this expands to Type S
error rate, Type M ratio, median bias, coverage, and RMSE, all at each
term's own empirically-observed effect size, pooled across the six
participant sample sizes.

```{r}
#| label: aim1-full-metrics-table

aim1_metrics_by_nt <- aim1_all |>
  dplyr::group_by(term, n_t) |>
  dplyr::summarise(
    power            = mean(excludes_zero, na.rm = TRUE),
    sign_reliability = mean(correct_sign, na.rm = TRUE),
    type_s_rate      = mean(sign_error[excludes_zero], na.rm = TRUE),
    type_m_ratio     = median(inflation_ratio[excludes_zero], na.rm = TRUE),
    median_bias      = median(bias, na.rm = TRUE),
    coverage         = mean(covers_truth, na.rm = TRUE),
    rmse_raw         = sqrt(mean(bias^2, na.rm = TRUE)),
    rmse_trimmed     = sqrt(mean(pmin(bias^2, quantile(bias^2, 0.95, na.rm = TRUE)), na.rm = TRUE)),
    n_fits           = dplyr::n(),
    .groups = "drop"
  )

aim1_metrics_by_nt |>
  dplyr::mutate(dplyr::across(dplyr::where(is.numeric) & !c(n_t, n_fits), ~round(., 3))) |>
  dplyr::arrange(term, n_t)
```

`rmse_raw` is included for completeness but is itself dominated by rare
pathological fits at small `n_t` (e.g. an outlying -20.3 estimate at
`n_t = 2`); `rmse_trimmed` (winsorizing squared bias at its 95th percentile
before averaging) is the more trustworthy summary of typical-case error
magnitude and should be preferred when comparing across cells.

### Aim 1 -- visualizations

```{r}
#| label: fig-aim1-metrics-vs-nt
#| fig-cap: "Aim 1: power, sign reliability, Type S error rate, and Type M exaggeration ratio vs. text sample size (n_t), at each term's own empirically-observed effect size, pooled across all six participant sample sizes and 50 replicates/cell."

plot_metrics_vs_x(
  aim1_metrics_by_nt, x_var = "n_t",
  metrics = c("power", "sign_reliability", "type_s_rate", "type_m_ratio"),
  metric_labels = aim1_metric_labels,
  x_lab = "Text sample size (n_t, log scale)"
)
```

This is the visual counterpart of the table above and of Aim 1's prose
finding: `zPCREFp` has the *worst* sign reliability and the *highest* Type
S/Type M values at `n_t = 2`, worse even than the interaction -- while also
having power *higher* at `n_t = 2` than at `n_t = 10`, the "underpowered
paradox" signature (a significant result here is more likely to be wrong,
not more trustworthy).

### What the "underpowered paradox" looks like at the level of individual estimates

The summary diagnostics above (power, sign reliability, Type S/M rates) are
all *aggregates* over 50 replicate fits per cell. It can be easier to see
why the paradox happens by looking at the individual replicate estimates
themselves, rather than only their summary statistics. The two panels below
show every individual estimate from Aim 1's fits at `n_p = 1497` -- the
**largest** participant sample size in the grid, held constant across every
row and column so that the funnel shown reflects the text margin (`n_t`)
specifically, not participant-margin noise -- for `zPCREFp` (top) and the
`zHLVA:zPCREFp` interaction (bottom): one dot per replicate fit, one column
per simulated true effect size, one row per text sample size.

```{r}
#| label: fig-aim1-distribution-explainer
#| fig-width: 12
#| fig-height: 8.5
#| fig-cap: "Aim 1: individual estimates (one dot per replicate fit, n_p held constant at its largest grid value, 1497) for zPCREFp (top) and the zHLVA:zPCREFp interaction (bottom), by text sample size (rows, narrower at top) and true effect size (columns, labeled \"true = ...\"). Dashed vertical line = the true value being simulated in that column. Red = a 95% CI that excludes zero (\"significant\"); grey = not significant. A handful of pathological estimates beyond the plotted range (all at n_t = 2) are clipped to the axis edge for readability."

illustrative_n_p <- max(n_p_grid)

make_aim1_distribution_panel <- function(data, term_name, x_clip, illustrative_n_p,
                                          show_x_axis = TRUE) {
  df <- data |>
    dplyr::filter(term == term_name, n_p == illustrative_n_p) |>
    dplyr::mutate(
      n_t_f = factor(n_t, levels = n_t_grid),
      significant = ifelse(excludes_zero, "Significant (95% CI excludes 0)", "Not significant"),
      estimate_clipped = pmax(pmin(estimate, x_clip), -x_clip)
    )

  ggplot(df, aes(x = estimate_clipped, y = n_t_f, color = significant, fill = significant)) +
    ggdist::stat_dots(
      side = "both", scale = 0.85, alpha = 0.55,
      binwidth = ggplot2::unit(c(1.3, 2.2), "mm"),
      overflow = "compress"
    ) +
    geom_vline(aes(xintercept = effect_value), linetype = "dashed", color = "black", linewidth = 0.5) +
    facet_wrap(~effect_value, nrow = 1,
               labeller = as_labeller(function(x) paste0("true = ", x))) +
    scale_x_continuous(breaks = c(-x_clip, 0, x_clip)) +
    scale_color_manual(values = c("Significant (95% CI excludes 0)" = "#b2182b",
                                   "Not significant" = "grey45")) +
    scale_fill_manual(values = c("Significant (95% CI excludes 0)" = "#b2182b",
                                  "Not significant" = "grey45")) +
    labs(
      x = if (show_x_axis) paste0(term_name, " estimate (clipped to [-", x_clip, ", ", x_clip, "])") else NULL,
      y = "n_t", color = NULL, fill = NULL
    ) +
    theme_bw(base_size = 10) +
    theme(
      panel.border = element_rect(color = "grey35", fill = NA, linewidth = 0.7),
      panel.spacing = unit(1.1, "lines"),
      strip.background = element_rect(fill = "grey85", color = "grey35"),
      strip.text = element_text(face = "bold", size = 8.5),
      panel.grid.minor = element_blank(),
      panel.grid.major.y = element_blank(),
      legend.position = "bottom"
    )
}

panel_dist_pcref <- make_aim1_distribution_panel(
  aim1_pcref, "zPCREFp", x_clip = 2, illustrative_n_p = illustrative_n_p, show_x_axis = FALSE
) + labs(title = "zPCREFp (text-level main effect)")

panel_dist_interaction <- make_aim1_distribution_panel(
  aim1_interaction, "zHLVA:zPCREFp", x_clip = 1.5, illustrative_n_p = illustrative_n_p, show_x_axis = TRUE
) + labs(title = "zHLVA:zPCREFp (interaction)")

(panel_dist_pcref / panel_dist_interaction) +
  patchwork::plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

Three things are visible here that the summary tables/plots only imply.
First, the funnel shape itself, present in **both** rows: each cloud of
dots is wide at the bottom (`n_t = 2`) and collapses to a tight cluster
around the dashed true-value line by the top (`n_t = 136`) -- the textbook
"wide at small *N*, narrow at large *N*" funnel, for every true effect size
simulated. Second, and this is the paradox in its most direct form: in
`zPCREFp`'s leftmost column (`true = 0`, i.e. **no true effect at all**), a
substantial share of the `n_t = 2` dots are still colored red --
individually indistinguishable, in a real analysis, from a genuine
discovery. Those red dots at `n_t = 2` are false positives purely from
noise, and they are just as red (just as "significant", p < .05) as the
true positives further right. A significance threshold cannot tell these
apart; only a larger text sample -- more rows lower down turning into
fewer, tighter, more-often-correctly-colored rows higher up -- can.

Third, comparing the two rows directly answers whether the interaction's
funnel looks like `zPCREFp`'s: **not really**. At `n_t = 2`, the
interaction's dots are mostly grey, not red, across every column including
its own empirically-observed effect size (`true = 0.054`) -- consistent
with its much lower power at that cell (0.127, vs. `zPCREFp`'s 0.627 from
the table above). Where `zPCREFp`'s problem at `n_t = 2` is "too many
confidently-wrong answers," the interaction's is closer to "usually no
answer at all, and on the rare occasions there is one, it is not obviously
more trustworthy than `zPCREFp`'s" -- the same underlying text-margin
scarcity, but manifesting as a power problem more than a false-discovery
problem for this particular term.

### Why can an estimate be "significant" when the true effect is exactly zero?

This is the question the `true = 0` column above raises most sharply, and
it is worth answering mechanistically rather than leaving it as a
statistical curiosity. Two different things are going on, and it is
important to distinguish them.

**It is not simply the ordinary 5% false-positive rate.** A correctly
calibrated 95% CI is *expected* to exclude a true value of zero 5% of the
time, by construction -- that is what "95%" means. But the observed rate at
`n_t = 2` is far above that: the table below shows that 66.3% of the
`zPCREFp` fits at `n_t = 2` under a true null are "significant," roughly
13x the nominal rate. Something is actively miscalibrated here, not merely
unlucky.

**The mechanism is a random-effects identifiability breakdown, not sampling
bad luck on the covariate itself.** With only `n_t = 2` texts, there is
exactly one degree of freedom available to describe how the two sampled
texts differ from each other -- two points determine a line perfectly, with
nothing left over to check whether that line is trustworthy. The "correct"
model asks `glmer` to split any such between-text difference into two
pieces: a text-level random intercept (`(1 | text_id)`, capturing
idiosyncratic, unexplained differences between texts) and the fixed
`zPCREFp` slope (capturing the cohesion effect specifically). Estimating a
random-intercept *variance* from only 2 groups is close to impossible --
there is essentially no information to distinguish "these two texts differ
a lot by chance" from "these two texts differ a lot systematically." In
practice, `lme4` resolves this by pinning the text-level variance at (or
extremely near) its zero boundary -- a **singular fit**:

```{r}
#| label: fig-aim1-singular-fit-diagnostic

null_nt2_diagnostic <- aim1_pcref |>
  dplyr::filter(term == "zPCREFp", n_t == 2, effect_value == 0)

cat("Fit status, all 300 replicates (n_t = 2, true zPCREFp effect = 0):\n")
table(null_nt2_diagnostic$status)

null_nt2_diagnostic |>
  dplyr::group_by(excludes_zero) |>
  dplyr::summarise(
    n = dplyr::n(),
    `median |estimate|` = round(median(abs(estimate), na.rm = TRUE), 3),
    `median std. error` = round(median(std.error, na.rm = TRUE), 3),
    .groups = "drop"
  )
```

Every single one of these 300 replicates converges to a singular fit --
the text-level random-intercept variance collapses to zero every time
`n_t = 2`, regardless of how many participants are simulated. Once that
variance is pinned at zero, the model has nowhere else to put the two
sampled texts' chance difference except into the only fixed-effect
predictor that differs between them -- `zPCREFp`. This inflates the fixed-
effect point estimate (median |estimate| = 0.734 among the "significant"
fits, versus a true value of exactly 0) and simultaneously produces an
overconfident, too-narrow Wald standard error (median SE = 0.106 among the
"significant" fits, versus 0.294 among the fits that correctly stayed
non-significant) -- because the SE formula implicitly assumes the
random-effects variance structure has been reliably estimated, an
assumption that fails precisely at this boundary. Inflated estimate +
deflated SE together push a large share of purely-noise-driven fits across
the 95%-CI-excludes-zero line.

In short: at `n_t = 2`, the model is not really testing "does cohesion
predict comprehension" -- it is testing "do these two specific,
randomly-sampled texts differ," and then attributing 100% of whatever
difference it finds to cohesion, because there is no other place left for
that difference to go. This is a more extreme version of the same
family of problem documented elsewhere in this project as random-effects
misspecification (Aim 3's omitted-`text.id` bias, and Phase 1's finding
that pseudoreplication introduces genuine bias, not just imprecision) --
here the grouping factor is nominally *in* the model formula, but with only
2 groups it is statistically unidentifiable in practice, which behaves much
like omitting it.

### False negatives (power) and false positive rates

Strictly, **power** is the probability of correctly detecting an effect
that truly exists in the population, under conventional null-hypothesis
significance testing -- it is only a meaningful concept when the true
effect is non-zero. Figure 3 (the power surface below) and Figure 4 (the
power curves) both plot the same quantity, P(95% CI excludes zero), but
Figure 4 sweeps the true effect size down to and through zero. At that
point, P(significant) is no longer "power" in the strict sense: it is the
**false-positive rate**, since there is nothing there to detect. Rather
than color that `effect size = 0` line as if it were simply the lightest
point on the same power-curve palette -- which would imply it answers the
same question as its non-zero neighbors -- Figure 4 draws it in a distinct,
colorblind-friendly bright orange, held constant across all three rows. It
answers a different question ("how often does noise alone
cross the significance threshold?") from the colored curves next to it
("how often is a real effect of this size detected?"), and this is exactly
the distinction the "underpowered paradox" discussion above turns on.

```{r}
#| label: fig-aim1-power-surface
#| fig-width: 8
#| fig-height: 4
#| fig-cap: "Aim 1: power surface across the full joint (n_p, n_t) grid, at each term's own empirically-observed effect size. This is the direct 2-D analog of a Marek et al. (2022)-style Figure 3 sample-size surface: both axes are the two sample-size margins, and tile fill is the probability a 95% Wald CI excludes zero."

aim1_surface <- aim1_all |>
  dplyr::group_by(term, n_p, n_t) |>
  dplyr::summarise(power = mean(excludes_zero, na.rm = TRUE), .groups = "drop")

plot_power_heatmap(aim1_surface, fill_var = "power", fill_label = "Power")
```

The surface makes visible what the by-`n_t` table (pooled over `n_p`)
cannot: `zHLVA`'s power saturates near 100% once `n_p` reaches ~100,
*independent* of `n_t` -- it is a purely participant-margin-limited effect.
`zPCREFp` and the interaction, by contrast, need growth on **both** margins
simultaneously to reach comparable power; increasing `n_p` alone at `n_t =
2` caps out around 78-92% and 14-24% respectively, well short of what larger
`n_t` unlocks at the same `n_p`.

```{r}
#| label: fig-aim1-power-curves
#| fig-width: 10
#| fig-height: 9.5
#| fig-cap: "Aim 1: classic power curves -- P(significant) vs. participant sample size (n_p, log scale), one row per term, faceted by text sample size (n_t) across columns. Each row uses its own sequential color palette (darker = larger effect size); the thicker line/larger points mark each term's empirically-observed effect size."

aim1_power_by_effect <- aim1_sweep_combined |>
  dplyr::group_by(term, effect_value, n_p, n_t) |>
  dplyr::summarise(power = mean(excludes_zero, na.rm = TRUE), .groups = "drop")

# Each term sweeps a different set of effect-size values (see the sweep
# grids above), so a single shared color scale across all three rows would
# either compress each term's own 5-point range into indistinguishable
# shades or mix three unrelated numeric scales under one legend. Instead,
# each term gets its own sequential ColorBrewer palette (darker = larger
# effect), and the curve at that term's empirically-observed effect size
# (the middle grid point in every sweep) is drawn thicker and with larger
# points so it is identifiable at a glance. The effect-size-0 curve is a
# false-positive rate, not power (see the note above) -- it is recolored a
# single, colorblind-friendly bright orange (Okabe-Ito palette, held
# constant across all three terms/rows) rather than left as each palette's
# lightest shade, so it reads as answering a different question from its
# colored, non-zero neighbors.
make_aim1_power_panel <- function(term_name, effect_grid, palette_name,
                                   axis_title = NULL, legend_title = NULL) {
  own_value <- own_effect[[term_name]]
  # Skip the palette's palest 2 shades (indices 1-2 of the 9-class Brewer
  # palette) before interpolating -- the raw 5-class palette's lightest
  # color is too close to white/background to read easily as a line.
  pal <- grDevices::colorRampPalette(RColorBrewer::brewer.pal(9, palette_name)[3:9])(length(effect_grid))
  pal[effect_grid == 0] <- "#E69F00"  # Okabe-Ito orange, colorblind-friendly
  effect_labels <- dplyr::case_when(
    effect_grid == 0         ~ paste0(effect_grid, " (no true effect -- false-positive rate)"),
    effect_grid == own_value ~ paste0(effect_grid, " (observed)"),
    TRUE                     ~ as.character(effect_grid)
  )

  df <- aim1_power_by_effect |>
    dplyr::filter(term == term_name) |>
    dplyr::mutate(
      effect_f = factor(effect_value, levels = effect_grid, labels = effect_labels),
      is_own   = effect_value == own_value
    )

  ggplot(df, aes(x = n_p, y = power, color = effect_f, group = effect_f)) +
    geom_line(aes(linewidth = is_own)) +
    geom_point(aes(size = is_own)) +
    facet_wrap(~n_t, nrow = 1, labeller = label_both) +
    scale_x_log10(breaks = n_p_grid, labels = scales::label_comma()) +
    scale_color_manual(values = pal, name = legend_title %||% term_name) +
    scale_linewidth_manual(values = c(`FALSE` = 0.5, `TRUE` = 1.3), guide = "none") +
    scale_size_manual(values = c(`FALSE` = 1, `TRUE` = 2.2), guide = "none") +
    labs(x = axis_title, y = "Power") +
    theme_minimal(base_size = 9) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 6.5),
      legend.position = "right"
    )
}

panel_hlva <- make_aim1_power_panel("zHLVA", b_hlva_grid, "Greens",
                                     legend_title = "zHLVA effect size")
panel_pcref <- make_aim1_power_panel("zPCREFp", b_pcref_grid, "Blues",
                                      legend_title = "zPCREFp effect size")
panel_interaction <- make_aim1_power_panel("zHLVA:zPCREFp", b_interaction_grid, "Purples",
                                            axis_title = "Participant sample size (n_p, log scale)",
                                            legend_title = "Interaction effect size")

panel_hlva / panel_pcref / panel_interaction
```

The `zPCREFp` row is the most diagnostic: at `n_t = 2`, the bright orange
false-positive-rate curve (true effect = 0) stays **non-monotonic and never
near zero across the entire `n_p` range** (~20-60%) -- i.e., a sizeable
share of fits at this text sample size are "significant" purely from noise,
regardless of how many participants are added. At larger `n_t`, that same
orange curve flattens out near zero as expected (a well-calibrated
false-positive rate), and the colored (genuine power) curves separate out
into a conventional monotonically-rising power curve as `n_p` grows --
confirming the pathology is specific to the `n_t = 2` regime, not a general
property of `zPCREFp`. This is the clearest single illustration in this
document of the underpowered paradox's mechanism: at the modal literature
text sample size, growing the participant sample does not rescue the
false-positive problem -- it is a text-margin problem, not a
participant-margin one. The highlighted "(observed)" curve in each row --
the effect size actually estimated from this project's real data -- sits in
the middle of its own sweep by construction, letting readers see directly
how much power/error-rate behavior would differ if the true effect were
smaller or larger than what was actually observed.

## Aim 2: discovery/replication rates

| `n_t` | zHLVA replication | zPCREFp replication | interaction replication |
|---|---|---|---|
| 2 | 0.478 | 0.473 | **0.013** |
| 10 | 0.803 | 0.157 | 0.250 |
| 35 | 0.797 | 0.163 | 0.267 |
| 85 | 0.820 | 0.237 | 0.273 |
| 136 | 0.797 | 0.340 | 0.297 |

A different, complementary picture emerges here: under the stricter
"significant in both independent halves, same sign" criterion, the
**interaction is the most fragile of the three** at `n_t = 2` (1.3%
replication vs. ~47-48% for both main effects) -- because it is so rarely
even significant in the first place with half the already-small sample. So:
Aim 1's sign-reliability diagnostic (conditional on significance) points to
`zPCREFp` as the least trustworthy; Aim 2's replication-rate diagnostic
(unconditional, halved samples) points to the interaction as the most
fragile overall. Both are real, non-contradictory findings -- they answer
different practical questions ("if I got a significant result, can I trust
its sign?" vs. "if I ran this again, would I get the same answer?") -- and
both indicate `n_t = 2` is inadequate, just via different mechanisms.

### Aim 2 -- fuller metrics table and visualizations

In addition to the raw replication rate above, this adds discovery rate
(P[significant in the first, discovery half]) and conditional replication
(P[replicates | discovery-significant] -- "given I got a significant result
in one half, would it hold up in an independent second half?"), the more
directly decision-relevant quantity for a researcher who has already seen a
significant discovery-sample result.

```{r}
#| label: aim2-full-metrics-table

aim2_metrics_by_nt <- aim2_all |>
  dplyr::group_by(term, n_t) |>
  dplyr::summarise(
    discovery_rate           = mean(discovery_significant, na.rm = TRUE),
    replication_rate         = mean(replicated, na.rm = TRUE),
    conditional_replication  = mean(replicated[discovery_significant], na.rm = TRUE),
    n_fits                   = dplyr::n(),
    .groups = "drop"
  )

aim2_metrics_by_nt |>
  dplyr::mutate(dplyr::across(dplyr::where(is.numeric) & !c(n_t, n_fits), ~round(., 3))) |>
  dplyr::arrange(term, n_t)
```

```{r}
#| label: fig-aim2-metrics-vs-nt
#| fig-width: 9
#| fig-cap: "Aim 2: discovery rate, (unconditional) replication rate, and conditional replication rate vs. text sample size, at each term's own empirically-observed effect size."

plot_metrics_vs_x(
  aim2_metrics_by_nt, x_var = "n_t",
  metrics = c("discovery_rate", "replication_rate", "conditional_replication"),
  metric_labels = aim2_metric_labels,
  x_lab = "Text sample size (n_t, log scale)"
)
```

The third panel adds a finding not visible in the unconditional
replication-rate table alone: conditional on having found a significant
interaction in the discovery half, the interaction's replication rate (~22%
at `n_t = 2`, rising to ~70-92% at larger `n_t`) is *lower* than `zHLVA`'s at
every `n_t` but is **not as catastrophically low** as its unconditional
replication rate (1.3% at `n_t = 2`) might suggest in isolation -- most of
that unconditional fragility comes from rarely being significant in the
discovery half at all (a power problem), rather than from significant
discoveries failing to hold up (a pure replication problem). Both framings
matter for different audiences: a reviewer asking "is this literature's
replication rate adequate" should look at the unconditional rate; a
researcher who has already observed one significant result and is deciding
whether to trust it should look at the conditional rate.

```{r}
#| label: fig-aim2-replication-surface
#| fig-width: 8
#| fig-height: 4
#| fig-cap: "Aim 2: unconditional replication-rate surface across the full joint (n_p, n_t) grid, at each term's own empirically-observed effect size -- the discovery/replication analog of the Aim 1 power surface above."

aim2_surface <- aim2_all |>
  dplyr::group_by(term, n_p, n_t) |>
  dplyr::summarise(replication_rate = mean(replicated, na.rm = TRUE), .groups = "drop")

plot_power_heatmap(aim2_surface, fill_var = "replication_rate", fill_label = "Replication\nrate")
```

Same qualitative pattern as Aim 1's power surface, but stricter: `zHLVA`
again saturates on the participant margin alone (100% replication once
`n_p >= 375`, at any `n_t`), while `zPCREFp` and the interaction remain
below 70% replication everywhere on this grid except the single
largest-both-margins cell -- reinforcing that halving the sample (as Aim 2's
discovery/replication design necessarily does) makes the two
text-margin-dependent effects substantially harder to recover than Aim 1's
full-sample power numbers alone would suggest.

## Aim 1 & Aim 2: discussion

Comparing the two aims' diagnostics side by side clarifies what each is, and
isn't, telling us:

- **Aim 1** (conditional on significance) says `zPCREFp` is the *least*
  trustworthy term at `n_t = 2` -- its significant results are the most
  likely to have the wrong sign.
- **Aim 2** (unconditional, halved samples) says the **interaction** is the
  most fragile overall -- not because its significant results are
  especially untrustworthy, but because it is rarely significant at all
  once the sample is halved.
- Both agree that `n_t = 2` -- the modal text sample size in this project's
  literature review (Phase 1's overlay finding: ~95% of reviewed studies
  fall outside the trustworthy region) -- is inadequate for any of the
  three effects, via different mechanisms depending on which question is
  asked.
- `zHLVA`, the sole participant-level effect among the three, is
  comparatively well-behaved on both diagnostics and saturates on the
  participant margin alone, independent of text sample size -- consistent
  with Phase 1's finding that participant-margin effects are far less
  sensitive to this project's chronic text-sample scarcity than text-margin
  effects are.

A full cross-aim synthesis, tying these findings together with Aims 3-6
(random-effects misspecification, variance-component sensitivity, slope/
covariance recovery, and design remedies), will appear once those aims have
their own reporting-ready documents.

## Flagged future work

- **The significance-threshold/multiple-comparisons sweep** (Marek et al.'s
  P<0.05-to-Bonferroni comparison) was explicitly deferred at the design
  stage, given the real core model's 26-term fixed-effects formula. Aim 1/2
  above use a single 95%-CI-based threshold throughout; whether Aim 1's
  error-rate findings hold up under a stricter multiple-comparisons
  correction remains untested.

Aims 3-6's own flagged-future-work items (Aim 3's `brms` spot-checks, Aim
4's flat interaction-CI-width finding, Aim 5's deferred joint
correlation x variance sweep, Aim 6's remedy-specific follow-ups, and the
`furrr` load-imbalance engineering note) are unchanged and documented in
`phase2-effect-variance-misspecification.qmd`; they will surface again in
their own future documents rather than being duplicated here.

::: {.callout-note collapse="true"}
## Development history (collapsed -- process detail, not reportable findings)

This section preserves the process/decision narrative that accompanied the
original build of Aims 1-2, condensed from `phase2-effect-variance-misspecification-PLAN.qmd`
and `phase2-effect-variance-misspecification.qmd`. Nothing in this section
is required to understand the results above; it documents how they were
produced.

### Relationship to existing project simulation infrastructure

This work explicitly reused conventions already established in
`simulation_study_complete.qmd` (Simulations 1, 2, 3, 4, and 6) and
`power-simulation-vocabulary-cohesion.qmd`, rather than inventing new ones:

- **One-at-a-time sweeps, not a full factorial.** Existing simulations vary
  one dimension (effect size, a single variance component, or a single
  intercept-slope correlation) at a time while holding others fixed at a
  realistic value, to avoid combinatorial explosion. Phase 2 follows the
  same pattern (six aims, altogether) rather than crossing effect size x
  variance x covariance x misspecification x design strategy x
  two-dimensional sample size all at once.
- **Fitting**: `glmer` (fast) as the sweep workhorse, with `brms`
  spot-checks at a handful of cells using the project's precompile-once,
  `update()`-per-replicate pattern to avoid Stan recompilation cost across
  replicates (not run for Aims 1-2 specifically; see flagged future work).
- **Parallelization**: `furrr::future_pmap_dfr()` with `multisession`,
  `furrr_options(seed = TRUE)`, hierarchical seeds, 4 workers with a
  `setTimeLimit()` guard, following Phase 1's memory-safety lessons.
- **Outcome-quality classification**: converged-clean / converged-singular
  / optimizer-warning / error, reused verbatim from Phase 1's helper
  functions.

### Confirmed decisions (made before building Aims 1-2)

1. **Scope**: all six aims cover **three effects** -- `zHLVA` (main),
   `zPCREFp` (main), and `zHLVA:zPCREFp` (interaction) -- rather than the
   interaction alone.
2. **Aim 2 split axis**: left configurable (`split_margin =
   "participants_only"` vs. `"both"`) rather than fixed; both variants can
   be run and compared. The results above use `"participants_only"`.
3. **Significance-threshold sweep**: deferred (see "Flagged future work").
4. **Pacing**: build and review Aim 1 (together with Aim 2) first, as the
   more traditional power-analysis-style starting point extended with the
   replication-split question.
5. **Replicate count**: 50 resamples/cell, matching Phase 1 Step 3's
   memory-safe, coarsened configuration.

### Timing benchmark

A small number of representative cells (small/moderate/large) were timed
once each for the `"correct"` and `"misspecified"` RE specifications (the
two used by Aims 1-3), before committing to the full grid:

| Cell | n_p | n_t | `correct` | `misspecified` |
|---|---|---|---|---|
| small | 25 | 2 | 0.34s (singular) | 0.03s (clean) |
| moderate | 375 | 35 | 0.71s (clean) | 0.30s (clean) |
| large | 1497 | 136 | 3.53s (clean) | 1.43s (clean) |

Mean per-fit time across these three cells (a rough, order-of-magnitude
estimate, not a proper average over the full grid's cell-size
distribution): `correct` ~1.53s, `misspecified` ~0.59s. (The original
benchmark also timed a `"full_slopes"` arm for Aim 5's covariance sweep,
using reference slope-variance/correlation values extracted from the fitted
core model; that arm is out of scope for this Aims-1/2-only document and is
reported instead in the Aim 5 document.)

Extrapolated cost: Aim 1 (22,500 fits, `correct` spec) ~9.6h serial / ~2.4h
on 4 workers; Aim 2 (~22,500 fits, `correct` spec) ~9.6h serial / ~2.4h
parallel. Combined, each `(sweep, effect_value, n_p, n_t, rep)` combination
triggers **three** `glmer` fits (one Aim 1 full-sample fit, plus Aim 2's
discovery and replication half-sample fits): 3 sweeps x 5 effect values x 30
cells x 50 reps = 22,500 combinations x 3 fits = **67,500 total fits**,
estimated at ~20h serial / ~5h on 4 parallel workers.

### Validation: small-scale smoke test

Before committing to the full grid, the pipeline was validated on a 24-cell
subset (2 effect values x 2 participant Ns x 2 text Ns x 3 reps, interaction
sweep only, at small sample sizes). Kept live here (cheap, ~6 seconds) as a
lightweight regression check that `generate_data_p2()`/`fit_model_p2()`
above still behave as expected.

```{r}
#| label: aim1-aim2-smoke-test
#| cache: true

smoke_grid <- tidyr::crossing(
  sweep = "interaction",
  effect_value = c(0, 0.054),
  n_p = c(25, 100),
  n_t = c(2, 10),
  rep = 1:3
)

smoke_results <- purrr::pmap(smoke_grid, run_aim1_aim2_cell)
aim1_smoke <- purrr::map_dfr(smoke_results, "aim1")
aim2_smoke <- purrr::map_dfr(smoke_results, "aim2")
```

```{r}
#| label: aim1-aim2-smoke-test-check

aim1_smoke |> dplyr::filter(term == "zHLVA:zPCREFp") |>
  dplyr::select(n_p, n_t, effect_value, rep, estimate, ci_lower, ci_upper,
                excludes_zero, correct_sign, status)

aim2_smoke |> dplyr::filter(term == "zHLVA:zPCREFp") |>
  dplyr::select(n_p, n_t, effect_value, rep, discovery_significant, replicated,
                status_discovery, status_replication)
```

The pipeline ran the 24-cell subset in ~5.6 seconds and produced sensible,
correctly-structured output for both Aim 1 (per-term bias/CI/sign
diagnostics) and Aim 2 (discovery/replication agreement) -- singular fits
were common at these smallest cells (`n_p = 25`, `n_t = 2`), as expected.

### Cost note and launch record

The full grid was not run interactively. It was launched as a detached
background process (`run_aim1_aim2_full.R`), one sweep at a time (`hlva`,
`pcref`, `interaction`), using `furrr::plan(multisession, workers = 4)`. All
three sweeps finished in ~99-101 minutes each (~5h total on 4 parallel
workers, matching the benchmark-derived estimate closely). Results were
saved to `results_phase2/aim{1,2}_{sweep}.rds` and are what this document
loads above.
:::

## Follow-up note

The DGP and fitting code above (`generate_data_p2()`, `fit_model_p2()`,
`generate_data_p2_pair()`, `run_aim1_aim2_cell()`) is kept live in this
document, unmodified from the version used to produce the saved results, so
it can be inspected and -- in a future, separate turn -- modified if
specific changes turn out to be informative. No new simulation runs have
been made in this pass; any modification to this code that would justify a
new full-grid run (a multi-hour, background-job undertaking) should be
discussed and scoped before being launched.
