---
title: "`testing-sequence.R`: one person, two people, a sample — at a computer"
subtitle: "Walkthrough: how the script works, and how to use it in Quarto"
format:
  html:
    code-fold: false
    toc: true
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(fig.align = "center")
```

## Aims

`testing-sequence.R` extends the pictogram generator in `person-grid.R`
with a "person sitting at a computer" composite icon, and uses it to draw
the narrative sequence: **one** person doing a data-collection task → **two**
people doing it → a whole **sample** of people doing it (shown as a grid,
with no individual scores). It also provides a faceted "grid of grids"
version for showing several independent *studies*, each with its own
sample of people. Paired with `sampling-distributions.R`'s synthetic score
figures, this is the icon half of a two-track illustration of sampling and
measurement error.

After working through it, you should be able to:

- explain how `person_at_computer()` composes two independent polygon sets
  (person, monitor) into one icon, rather than drawing a new icon style;
- generate the one-person/two-person/sample sequence with your own scores;
- build a faceted grid of independent "studies" with `testing_icons_studies()`.

## Prerequisites / assumptions

- **Packages**: `ggplot2` only.
- **Sourcing**: `person-grid.R` **must** be sourced first — this script
  reuses `person_parts()`, `sample_counts()`, and `diverging_colors()`
  from it directly (not via a package namespace, so order matters).
- **Reader background**: read
  [person-grid-walkthrough.qmd](person-grid-walkthrough.qmd) first if the
  underlying icon mechanics (geometry, composition sampling, colour modes)
  aren't already familiar — this document assumes them.

```{r}
#| label: source-scripts
source("person-grid.R")
source("testing-sequence.R")
```

## How it works: the process

```{mermaid}
flowchart LR
    A["person_parts()\n(from person-grid.R)"] --> C["person_at_computer()\n(person + monitor, one role column)"]
    B["monitor_parts()\n(screen geometry,\npositioned relative to the head)"] --> C
    C --> D["testing_icons()\nloop over n icons,\nplace in a grid,\noptional score labels"]
    D --> E["testing_icons_studies()\nrepeat testing_icons() per study,\nfacet_wrap(~study)"]
```

The key idea is **composition over inheritance**: rather than writing a new
icon from scratch, `monitor_parts()` draws just the screen/stand, and
`person_at_computer()` glues one `person_parts()` output and one
`monitor_parts()` output into a single polygon set, tagged with a `role`
column (`"person"` vs `"monitor"`) so the two pieces can be filled
differently. Everything above that (grids, score labels, faceted studies)
is built the same way `person_grid()` was.

## Element-by-element walkthrough

### Screen geometry: `monitor_parts()`

```{r}
#| label: show-monitor-parts
#| eval: false
monitor_parts <- function(scale = 1, ox = 0, oy = 0,
                           center_y = 1.62, x_offset = 0.80) {
  bezel_hw <- 0.40; bezel_hh <- 0.34; inset <- 0.05
  neck_h <- 0.10; neck_hw <- 0.04; base_hw <- 0.22; base_h <- 0.05

  bezel_top <- center_y + bezel_hh
  bezel_bottom <- center_y - bezel_hh
  neck_top <- bezel_bottom
  neck_bottom <- neck_top - neck_h
  base_top <- neck_bottom
  base_bottom <- base_top - base_h

  parts <- list(
    bezel  = data.frame(x = x_offset + c(-bezel_hw, bezel_hw, bezel_hw, -bezel_hw),
                         y = c(bezel_top, bezel_top, bezel_bottom, bezel_bottom)),
    screen = data.frame(x = x_offset + c(-bezel_hw + inset, bezel_hw - inset,
                                          bezel_hw - inset, -bezel_hw + inset),
                         y = c(bezel_top - inset, bezel_top - inset,
                               bezel_bottom + inset, bezel_bottom + inset)),
    neck   = data.frame(x = x_offset + c(-neck_hw, neck_hw, neck_hw, -neck_hw),
                         y = c(neck_top, neck_top, neck_bottom, neck_bottom)),
    base   = data.frame(x = x_offset + c(-base_hw, base_hw, base_hw, -base_hw),
                         y = c(base_top, base_top, base_bottom, base_bottom))
  )
  # ... offset by (ox, oy) and scale, same as person_parts() ...
}
```

This draws four stacked polygons (bezel, inset screen, neck, base — an
inverted-T stand) in the **same local coordinate frame** `person_parts()`
uses (head centred at `x = 0, y = 1.62`), with every y-coordinate chained
off the previous one (`neck_top <- bezel_bottom`, etc.) so the stand
always sits flush under the screen regardless of bezel size.

::: {.callout-note title="Why center_y and x_offset, specifically?"}
`center_y = 1.62` vertically centres the monitor *on the head*, not on the
whole body, so the composite reads as "looking at a screen" rather than
"a person standing next to an unrelated box". `x_offset = 0.80` is a
fixed horizontal gap from the body's centre line — both defaults exist
specifically to make the pairing legible, not just to place a monitor
somewhere on the canvas.
:::

```{r}
#| label: demo-monitor-parts
#| fig-cap: "monitor_parts() alone"
ggplot(monitor_parts(), aes(x, y, group = part)) +
  geom_polygon(fill = "grey70") +
  coord_equal() +
  theme_void()
```

### The composite icon: `person_at_computer()`

```{r}
#| label: show-person-at-computer
#| eval: false
person_at_computer <- function(kind = "A", age = "adult", disabled = FALSE,
                                scale = 1, ox = 0, oy = 0,
                                person_color = "#3B6EA5",
                                screen_color = "#CFE8FF",
                                frame_color = "#555555") {
  p <- person_parts(kind, age, disabled, scale, ox, oy)
  p$role <- "person"
  p$fill <- person_color

  m <- monitor_parts(scale = scale, ox = ox, oy = oy)
  m$role <- "monitor"
  m$fill <- ifelse(m$part == "screen", screen_color, frame_color)

  rbind(p, m)
}
```

This simply calls `person_parts()` and `monitor_parts()` with the *same*
`scale`/`ox`/`oy`, stamps a `fill` colour on each (the monitor's bezel/neck/
base all get `frame_color`, only the inset `screen` part gets
`screen_color`), and row-binds the two polygon sets. Because both pieces
share the same offset, the composite moves as one unit when placed in a
grid later.

```{r}
#| label: demo-person-at-computer
#| fig-cap: "One composite icon"
one_icon <- person_at_computer(kind = "A")
ggplot(one_icon, aes(x, y, group = paste(part, role), fill = fill)) +
  geom_polygon() +
  scale_fill_identity() +
  coord_equal() +
  theme_void()
```

### The main sequence function: `testing_icons()`

This mirrors `person_grid()`'s structure closely (composition sampling,
colour modes, grid placement), with two differences specific to the
narrative sequence it's built for:

- an optional `scores` argument: when supplied (length must equal `n`),
  each icon gets an `"x = <score>"` label underneath it — used for the
  one-person and two-person steps. Left `NULL` for the "sample" step,
  where the narrative says scores exist but deliberately doesn't show
  them.
- it inherits `person_grid()`'s two colour modes (`"diversity"` /
  `"value"`) unchanged, so the same colour-by-effect-size trick is
  available here too, not just on plain person icons.

```{r}
#| label: demo-testing-icons-one
#| fig-cap: "n = 1, with a score label"
testing_icons(n = 1, scores = 27)$plot
```

```{r}
#| label: demo-testing-icons-two
#| fig-cap: "n = 2, with score labels"
testing_icons(n = 2, scores = c(31, 24))$plot
```

```{r}
#| label: demo-testing-icons-sample
#| fig-cap: "n = 20, no scores shown -- 'a sample of people'"
#| fig-width: 7
#| fig-height: 3.5
testing_icons(n = 20, seed = 7)$plot
```

### Many studies at once: `testing_icons_studies()`

```{r}
#| label: show-testing-icons-studies
#| eval: false
testing_icons_studies <- function(n_studies, n_per_study, ncol = NULL,
                                   seed = 1, ...) {
  panels <- vector("list", n_studies)
  for (s in seq_len(n_studies)) {
    g <- testing_icons(n = n_per_study, seed = seed + s, ...)
    g$data$study <- paste("Study", s)
    panels[[s]] <- g$data
  }
  df <- do.call(rbind, panels)
  df$poly_id <- paste(df$study, df$poly_id)
  # ... ggplot() + facet_wrap(~study) ...
}
```

This is the icon analogue of `sampling-distributions.R`'s
`plot_histogram_grid()`/`plot_dotpile_grid()`: it calls `testing_icons()`
once per "study" with an offset seed (`seed + s`), so each panel is an
**independently drawn** sample (not the same 20 icons repeated), tags each
panel's polygons with a `study` column, and facets.

::: {.callout-note title="Why re-prefix poly_id with study?"}
Without `df$poly_id <- paste(df$study, df$poly_id)`, icon 1 in Study 1 and
icon 1 in Study 2 would share the same `poly_id`, and `ggplot`'s `group`
aesthetic would try to connect/group their polygons together across
facet panels — a subtle bug that wouldn't throw an error, just quietly
distort the shapes.
:::

```{r}
#| label: demo-testing-icons-studies
#| fig-cap: "Six independent 20-person samples ('studies')"
#| fig-width: 8
#| fig-height: 5
testing_icons_studies(n_studies = 6, n_per_study = 20, ncol = 3, seed = 104)$plot
```

## Quarto usage / integration

Setup-chunk lines (order matters):

```r
source("person-grid.R")
source("testing-sequence.R")
```

Real usage from `doing-being-science.qmd` — one person, paired in a
two-column layout with a `sampling-distributions.R` dot plot of the same
simulated score (`d_one$observed_score`), so the icon's label and the
point on the number line visibly match:

````{verbatim}
```{r}
#| label: fig-one-person-icon
#| fig-cap: "You do the vocabulary test"
#| fig-width: 4
#| fig-height: 3.75
#| echo: false

d_one <- simulate_vocab_data(n = 1, seed = 101)
testing_icons(n = 1, scores = d_one$observed_score)$plot
```
````

The "many studies" step:

````{verbatim}
```{r}
#| label: fig-many-studies-icons
#| fig-width: 8
#| fig-height: 5

testing_icons_studies(n_studies = 6, n_per_study = 20, ncol = 3, seed = 104)$plot
```
````

::: {.callout-note title="seed controls cosmetics, not scores"}
The icon grid's `seed` only controls *cosmetic* sampling (silhouette,
age, disability, skin tone) — it has no bearing on simulated scores. If
you want an icon's label to match a score drawn elsewhere (as in the
one-person/two-person slides), pass that score in explicitly via
`scores =`, rather than relying on the seeds to line up.
:::

`#| echo: false` is used on the slide chunks so only the figure shows on
the slide, not the simulation code — drop it if you want to show the
code for teaching purposes.

### Parameter quick reference (`testing_icons()`)

| Parameter | Default | Meaning |
|---|---|---|
| `n` | — | Number of icons. |
| `scores` | `NULL` | Optional length-`n` vector; adds `"x = <score>"` labels. |
| `ncol` | `NULL` (auto) | Icons per row. |
| `comp_kind`, `comp_age`, `disabled_prop` | as `person_grid()` | Composition mix. |
| `mode` | `"diversity"` | `"diversity"` or `"value"` (see `person-grid.R`). |
| `skin_tones`, `value` | as `person_grid()` | Colour controls. |
| `seed` | `1` | Cosmetic-sampling reproducibility (not score data). |
| `dx`, `dy` | `1.9, 2.5` | Icon spacing. |

`testing_icons_studies()` adds `n_studies`, `n_per_study`, and forwards
everything else to `testing_icons()` via `...` (except `scores`, which
isn't supported in the faceted version).

### Common errors

| Message | Likely cause |
|---|---|
| `length(scores) == n is not TRUE` | `scores` was supplied but its length doesn't match `n`. |
| ``value` must be supplied and length n when mode = 'value'` | `mode = "value"` set without a matching-length `value` (inherited from `person-grid.R`). |
| ``props` must have non-negative values, with at least one positive` | A composition argument summed to zero (inherited from `person-grid.R`). |
| Icon and score label mismatch | Relying on matching `seed`s instead of passing the same simulated `scores` explicitly (see callout above). |

## Exercises

1. Call `testing_icons(n = 3, scores = c(20, 25, 30))` and check that the
   label positions don't overlap. Then try `scores` of very different
   lengths, e.g. `c(5, 38)` for `n = 2` — does label placement still
   read clearly?
2. In `monitor_parts()`, change `x_offset` from `0.80` to `0.40` and
   re-render `person_at_computer()`. At what point does the monitor start
   visually overlapping the body?
3. Build a 12-study, 10-person-each grid with
   `testing_icons_studies(n_studies = 12, n_per_study = 10, ncol = 4,
   seed = 1)`. Compare it to `n_studies = 12, n_per_study = 10, ncol = 2`
   — which layout reads better on a wide slide vs. a narrow one?
4. `testing_icons()`'s `seed` only affects cosmetic sampling, not scores.
   Demonstrate this: call it twice with the same `seed` but different
   `scores`, and confirm the icons' composition (silhouette/age/skin
   tone) is identical both times while only the labels differ.
5. Try passing `mode = "value"` with a numeric `value` vector to
   `testing_icons()` instead of the default diversity mode. What does an
   icon-at-a-computer grid coloured by effect size communicate that a
   plain `person_grid()` in value mode doesn't?

## See also

- [person-grid-walkthrough.qmd](person-grid-walkthrough.qmd) — the
  dependency this script builds on; read first.
- [sampling-distributions-walkthrough.qmd](sampling-distributions-walkthrough.qmd) —
  the synthetic-data figures this icon sequence is paired with in the
  slide deck (same narrative, data side instead of icon side).
- [peer-review-workflow-walkthrough.qmd](peer-review-workflow-walkthrough.qmd) —
  reuses `person_at_computer()` from this script.
