---
title: "Working with citations, references, and citation style guides"
---

## Working with citations

You will be working with reference information.

Th Quarto guide to working with citations is here:

<https://quarto.org/docs/authoring/citations.html>

As it says there:

Quarto will use Pandoc to automatically generate citations and a bibliography in a number of styles. To use this capability, you will need:

- A quarto document formatted with citations (see Citation Markdown).
- A bibliographic data source, for example a BibLaTeX (.bib) or BibTeX (.bibtex) file.
- Optionally, a CSL file which specifies the formatting to use when generating the citations and bibliography (when not using natbib or biblatex to generate the bibliography).

We will need to build a references list as we develop our writing, and we will want the citations in text and the reference list or lists that are produced to be presented following APA 7 style guidance.

### You will need a reference manager: Zotero is good.

I assume that you can work with a reference manager like Zotero:

<https://www.zotero.org>

#### Install and open Zotero

Have Zotero open and running on your machine while you write. 

RStudio's Visual Editor auto-detects a local Zotero installation and reads your library directly: no separate setup or API key needed for basic use.

#### Switch to Visual Editor mode

In RStudio, open your .qmd file and click the 'Visual' tab (top-left of the source pane, next to 'Source'). 

This switches from raw markdown editing to a WYSIWYG-style view that supports the citation picker. 

You can switch back to Source mode any time — the underlying markdown is identical either way e.g.

```{r}
#| eval: false
@knuth84
```

- If you write this key, then you get: @knuth84 -- where you do the citation.

#### Insert a citation

With cursor where you want a citation, go to Insert > Citation (or the keyboard shortcut Cmd/Ctrl+Shift+F8). 

Choose 'Your Zotero Library' as the source, search for the reference, and select it.

#### Quarto auto-adds it to references.bib

When you insert a Zotero citation this way, Quarto automatically adds the full reference entry to your references.bib file and inserts the citation key at your cursor — you don't need to manually export or copy anything.

#### (Optional) Install Better BibTeX for stable citation keys

By default Zotero generates citation keys like `davies2026` automatically. If you want more control over key format (e.g. author-year-word style, or matching a specific journal convention), install the Better BibTeX plugin for Zotero, then in RStudio go to Tools > Global Options > R Markdown > Citations and enable 'Use Better BibTeX for citation keys'. 
### Using a required style

We will need to present citations and the reference list using APA 7 style.

#### Download the APA CSL file

Go to <https://www.zotero.org/styles> and search 'APA'. 

Click on 'American Psychological Association 7th edition'.

On the style's page, right-click the download/link icon and choose 'Save Link As' to download the raw .csl file rather than letting your browser install it into Zotero.

#### Save it into your book's project folder

Rename the downloaded file to something like apa.csl if it isn't already, and place it in your project root — the same folder as _quarto.yml, alongside references.bib.

I am using `psychological-bulletin.csl` for now.

#### Point _quarto.yml at the CSL file

We edit the book yaml to add information on the .bib, the .csl, and how we want the citations to behave (in-text citations to hyperlink down to their entry in the reference list).

```{r}
#| eval: false
bibliography: references.bib
csl: psychological-bulletin.csl
link-citations: true
```

#### Render and check citations

Render or refresh your preview.

Your in-text citations and bibliography will now follow APA formatting — e.g. (Davies, 2026) in-text, and a properly formatted APA reference list where your #refs div sits, instead of Pandoc's default numbered/bracketed style.
